### 2016-09-28 Paper_v1.tex
First commit

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}
\usepackage{multirow}
\usepackage{amsmath}
\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
\DeclareMathOperator*{\argmin}{\arg\!\min}

\usepackage{algorithm2e}
\usepackage{algpseudocode}
%\usepackage[space]{cite}
\urlstyle{rm}

%\textwidth = 6.50 in
%\textheight = 9.5 in
%\oddsidemargin =  0.0 in
%\evensidemargin = 0.0 in
%\topmargin = -0.50 in
%\headheight = 0.0 in
%\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
\doublespace

%\bibliographystyle{chicago}
\bibliographystyle{plos2009}

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\large\bfseries}}
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}

% Change the number format in the ref list -
\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Fig.
\renewcommand{\figurename}{Fig.}

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Analysis of TX-TL Synthetic Circuits using Sequence Specific Constraints Based Modeling}}}
\vspace{0.05in}
{\par \centering \large{Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Constraints based models of synthetic circuits}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}

In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} TX-TL system.
A core \emph{E.~coli} metabolic model, consisting of XX metabolites and YY reactions, was developed from literature [REF].
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples
the transcription and translation of circuit components with the availability of metabolic resources.
Model parameters were largely taken from literature; our approach had very few adjustable parameters thereby allowing the a first principles prediction of circuit performance.
We tested this approach by first simulating $\sigma_{70}$-induced deGFP expression and then expanded these studies to more complex multicomponent circuits.
First principles predictions of circuit performance were consistent with measurements for a variety of cases.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.

\vspace{0.1in}
{\noindent \textbf{Keywords:}~Synthetic biology, Constraints based modeling, Biochemical modeling}

\pagebreak

\setcounter{page}{1}

\linenumbers

\section*{Introduction}

Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
Interestingly, many of the challenges confronting in-vivo genome-scale kinetic modeling can potentially be overcome in a cell-free system.
For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
Thus, cell-free operation holds several significant advantages for model development, identification and validation.
Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E. coli} or \textit{B. subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.

Stoichiometric reconstructions of microbial metabolism popularized by constraint based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \citep{2012_lewis_palsson_NatRevMicrobio}.
Since the first genome-scale stoichiometric model of \textit{E. coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E. coli} \citep{Feist:2007aa} or \textit{B. subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems using linear programming.
Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}. Thus, modern constraint-based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.

In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} TX-TL system.
A core \emph{E.~coli} cell free metabolic model, consisting of XX metabolites and YY reactions, was developed from literature [REF].
This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
Thus, sequence specific constraints based modeling explicitly couples
the transcription and translation of circuit components with the availability of metabolic resources.
Model parameters were largely taken from literature; our approach had very few adjustable parameters thereby allowing the a first principles prediction of circuit performance.
We tested this approach by first simulating $\sigma_{70}$-induced deGFP expression and then expanded these studies to more complex multicomponent circuits.
First principles predictions of circuit performance were consistent with measurements for a variety of cases.
Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance.
Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.

\clearpage

%Constraints based models are important tools to understand and ultimately to predict how cells utilize nutrients to produce products.
%Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and network decomposition approaches such as elementary modes (EMs) \cite{Schuster:2000aa}
%or extreme pathways (EPs) \cite{Schilling:2000aa}
%model intracellular metabolism using the biochemical stoichiometry and other constraints such as thermodynamical feasibility under pseudo-steady state conditions.
%FBA has been used to efficiently estimate the performance of metabolic networks of arbitrary complexity, including genome scale networks, using linear programming  \cite{Covert:2004aa}.
%On the other hand, EMs (or EPs) catalog all possible metabolic behaviors such that any flux distribution predicted by FBA is a convex combination
%of the EMs (or EPs) \cite{Wiback:2003aa}.
%However, the calculation of EMs (or EPs) is computationally expensive and currently infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.

\section*{Results}

\clearpage
\section*{Discussion}

\clearpage

\section*{Materials and Methods}

\subsection*{Formulation and solution of the model equations.}
The flux balance analysis problem was formulated as:
\begin{equation}\nonumber
 \begin{multlined}
	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
	\mathrm{Subject \; to:}
	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
 \end{multlined}
\end{equation}
where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
In this study, the objective $w_{obj}$ was to maximize the production of circuit output.

\subsection*{Transcription and translation template reactions.}

\subsection*{Global sensitivity analysis.}
We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble.
We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 10,000 and $d$ = 22, so the total sensitivity indices were computed from
460,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.

\clearpage

\section*{Acknowledgements}
This study was supported by an award from the Army Research Office (ARO \#59155-LS).
\clearpage

\bibliography{References_v1}

\clearpage

% Supplemental figures -
% Set the S-
\renewcommand\thefigure{S\arabic{figure}}
\renewcommand\thetable{T\arabic{table}}
\renewcommand\thepage{S-\arabic{page}}
\renewcommand\theequation{S\arabic{equation}}

% Reset the counters -
\setcounter{equation}{0}
\setcounter{table}{0}
\setcounter{figure}{0}
\setcounter{page}{1}

\end{document}

### 2016-09-30 Paper_v1.tex
update_MV

@@
 \section*{Results}
+The protein concentration levels of deGFP produced with TX-TL 2.0 at different plasmid concentrations was modeled using a sequence specific constraints based modeling approach in a \textit{E.
+coli} cell free network (Fig.
+\ref{fig:deGFP_Plasmid}).
+The metabolic network was formulated by removing all cell wall associated reactions and incorporating amino acid synthesis and transcription/translation associated reactions. 
+The network consisted of 281 reactions and 132 species. 
+The transcription and translation rates were constrained to experimental measurements \cite{2016_Garamella_etal_TXTL} for the RNA polymerase and ribosome elongation rates. 
+The production time for TX-TL 2.0 is approximately 8 hours \cite{2016_Garamella_etal_TXTL}. 
+The concentration of deGFP was determined at each plasmid concentration by multiplying the corresponding production flux by the production time. 
+The sequence specific constraints based model was able to capture the deGFP concentration up to a plasmid concentration of 5 nM and overestimates deGFP measurements above that.
+TX-TL 2.0 has a saturation on protein production mostly due to limited resources which our model begins to see saturating effects at ~9 nM plasmid concentration. 
+This may be due to oxidative phosphorylation activity which has been observed to take place in other cell free systems \cite{2008_Jewett_etal_OxPhos}.
+
+\begin{figure}[!t]\centering
+\includegraphics[width=0.45\textwidth]{./figures/deGFP_Plasmid.png}
+\caption{Model simulation (solid line) versus deGFP protein concentration (dots) at different plasmid concentrations produced in TX-TL 2.0. 
+}
+\label{fig:deGFP_Plasmid}
+\end{figure}
+
+\begin{figure}[!t]\centering
+\includegraphics[width=0.45\textwidth]{./figures/CAT_oxphos.pdf}
+\caption{
+}
+\label{fig:CAT_oxphos}
+\end{figure}
+
+Oxidative phosphorylation activity influences the protein production flux as well as acetate and lactate flux.    
+Our model suggests higher oxidative phosphorylation activity increases the protein production rate for CAT (Fig. \ref{fig:oxphos.pdf}) and other proteins. 
+Consistent with literature, limiting oxidative phosphorylation results in lower protein yields in cell free systems as seen in our model. 
+Oxidative phosphorylation activity was limited to have a maximum flux of 3.0 mmol/hr in it's reactions, determined from a tradeoff between acetate and lactate flux (Fig. \ref{fig:oxhos}). 
+Experimental data has shown lactate and acetate to have similar fluxes during glucose consumption during the production of CAT (data not shown).
+Low oxidative phosphorylation leads to NADH overflow which is re-oxidized with lactate dehydrogenase reducing pyruvate to lactate leading to high lactate formation.
+High oxidative phosphorylation relative to our cell free system leads to lower lactate and higher acetate formation.
+
+%\begin{figure}[!t]\centering
+%\includegraphics[width=0.45\textwidth]{./figures/YieldNumber.pdf}
+%\caption{Oxidative phosphorylation activity influences protein, acetate and lactate production.
+A) Increasing oxidative phosphorylation leads to higher protein (CAT) production %rates and begins to saturate.
+B) Lactate versus acetate flux at different oxidative phosphorylation activity levels.
+As oxidative phosphorylation increases, lactate formation %decreases and acetate formation increases.
+%}
+%\label{fig:oxphos}
+%\end{figure}
+
+Theoritical carbon yields were calculated for nine different proteins using a sequence specific constraints based modeling approach in a \textit{E. coli} cell free network. 
+Carbon yields were calculated for three different cases: unconstrained, limited oxidative phosphorylation and TX-TL constraints (Fig. \ref{fig:CarbonYield}). 
+In the unconstrained case, amino acids uptake and oxidative phosphorylation activity were unbounded. 
+Most of the proteins studied in this paper resulted in a carbon yield near forty percent for a cell free system.
+In the limited oxidative phosphorylation case, carbon yields were reduced for each of the proteins.
+
+In the TX-TL 2.0 case, oxidative phosphorylation activity was limited in addition to limited transcription and translation fluxes.
+Following these constraints, an optimal plasmid concentration was determined for each protein which was used to determine the carbon yield for TX-TL 2.0 at that plasmid concentration.
+Carbon yields are significantly reduced compared to the unconstrained and limited oxidative phosphorylation cases.
+The carbon yields are being lost mostly due to lactate and acetate formation.
+The constrained TX-TL case requires additional ATP to optimize for the protein of interest production leading to acetate formation.
+
+TX-TL 2.0 carbon yields were then compared to the carbon number of each protein of interest (Fig. \ref{fig:YieldNumber}). 
+There is a relative inverse relationship between the carbon yield and the carbon number of each protein with the exception of FGF21. 
+In comparing the amino acids of each protein, the proteins with a carbon yield below ten percent have a high abundance of proline which is a non polar amino acid and may have a negative impact on increasing carbon yield.
+
+\begin{figure*}[!t]\centering
+\includegraphics[width=0.90\textwidth]{./figures/CarbonYield.pdf}
+\caption{Theoriticcal carbon yield of nine different proteins for three different cases in a cell free system. 
+Unconstrained yields (white bar) produced the highest carbon yield out of the three cases.
+In the limited oxidative phosphorylation case (gray bar), carbon yields were slightly lower than for the unconstrained case.
+TXTL 2.0 (black bar) constraints had low carbon yields compared to unconstrained and limited oxidative phosphorylation cases.
+}
+\label{fig:CarbonYield}
+\end{figure*}
+
+\begin{figure}[!t]\centering
+\includegraphics[width=0.45\textwidth]{./figures/YieldNumber.pdf}
+\caption{Theoretical maximum carbon yield of nine proteins versus the corresponding carbon number of each protein.
+There is a relative inverse relationship of carbon yield to number of carbons in each protein except for the protein FGF21.
+}
+\label{fig:YieldNumber}
+\end{figure}
 
@@
 In this study, the objective $w_{obj}$ was to maximize the production of circuit output.
+The specific glucose uptake rate was constrained to allow a maximum flux of 10 mmol/hr \cite{2002_Mahadevan_BiophysJ}; the amino acids were also bound to allow a maximum flux of 10 mmol/hr, but did not reach this maximum flux.
 
 \subsection*{Transcription and translation template reactions.}
+The transcription and translation template reactions are based off sequence specific FBA \cite{2002_allen_palsson} involving transcription initiaion, transcription, mRNA degradation, translation initiation, translation, and tRNA charging.
+The mRNA and protein sequence of each protein was determined from literature. 
+The transcription rate was constrained by the following formulation:
+
+\begin{equation}\nonumber
+	w_{tx} = [RNAP]\frac{v_{RNAP}}{l_{mRNA}}\left(\frac{[Gene]}{km+[Gene]}\right)P
+\end{equation}
+
+where $[RNAP]$ is the concentration of RNA polymerase which was determined from literature values based on the number of copies per cell, $v_{RNAP}$ is the elongation rate (nucleotides/hr) of the RNA polymerase, $l_{mRNA}$ is the number of nucleotides in the mRNA for the protein of interest, $[Gene]$ is the gene concentration of the protein of interest, $km$ is the plasmid saturation coefficient, and $P$ is the promoter activity.
+The promoter activity was formulated following Moon et al. for synthetic circuits by the following:
+
+\begin{equation}\nonumber
+	P = \frac{K_{1}+K_{2}f_{p70}}{1 + K_{1}+K_{2}f_{p70}}
+\end{equation}
+
+where $K_{1}$ represents the state of RNA polymerase binding, $K_{2}$ is the state of sigma-70 binding along with RNA polymerase, and $f_{p70}$ is the fraction of the transcription factor, sigma-70, bound to the promoter following Hills kinetics.
+
+The translation rate was constrained by the following formulation:
+
+ \begin{equation}\nonumber
+	w_{tl} = [Ribo]K_{P}\frac{v_{Ribo}}{l_{protein}}[mRNA_{ss}]
+\end{equation}
+
+where $[Ribo]$ is the ribosome concentration determined from literature values based on the number of copies per cell, $K_{P}$ is the polysome amplification constant, $v_{Ribo}$ is the elongation rate (amino acids/hr) of the ribosome, $l_{protein}$ is the number of amino acids in the protein of interest, and $[mRNA_{ss}]$ is the mRNA concentration at steady state determined by the transcription rate divided by the degradation rate of mRNA.
+
+\subsection*{Theoritical carbon yield.}
+The theortical carbon yield of each protein was formulated as:
+
+\begin{equation}\nonumber
+	Yield = \frac{C_{POI}v_{POI}}{\sum_{i = 1}^{\mathcal{R}}C_{i}v_{i}}
+\end{equation}
+
+where $C_{POI}$ and $C_{i}$ denote the carbon number of the protein of interest (POI) and substrate $i$, respectively, $v_{POI}$ and $v_{i}$ denote the flux of the POI and substrate $i$, respectively,
+and $\mathcal{R}$ denotes the number of substrates consumed.
 

### 2016-12-08 Paper_v1.tex
figs

--- 2016-09-28 Paper_v1.tex
@@
-\documentclass[12pt]{article}
-% Load packages
+\documentclass[journal=asbcd6,manuscript=article]{achemso}
+
+\usepackage{natbib}
+\usepackage{setspace}
+\usepackage{xkeyval}
 \usepackage{url}  % Formatting web addresses
@@
 \usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
-\usepackage{setspace}
-%\usepackage{longtable}
-\usepackage{colortbl}
-%\usepackage{palatino,lettrine}
-%\usepackage{times}
-%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
-%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
-\usepackage[wide]{sidecap}
-%\usepackage[authoryear,round,comma,sort&compress]{natbib}
-\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
-%\usepackage[authoryear,round]{natbib}
-\usepackage{supertabular}
-\usepackage{simplemargins}
-\usepackage{fullpage}
-\usepackage{comment}
-\usepackage{lineno}
-%\usepackage{chicago}
-\usepackage{textcomp}
-\usepackage{multirow}
-\usepackage{amsmath}
-\usepackage[linesnumbered,lined,boxed,commentsnumbered]{algorithm2e}
-\DeclareMathOperator*{\argmin}{\arg\!\min}
+\usepackage[version=3]{mhchem} % Formula subscripts using \ce{}
+\usepackage[T1]{fontenc}       % Use modern font encodings
+%\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% If issues arise when submitting your manuscript, you may want to
+%% un-comment the next line.  This provides information on the
+%% version of every file you have used.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%\listfiles
 
-\usepackage{algorithm2e}
-\usepackage{algpseudocode}
-%\usepackage[space]{cite}
-\urlstyle{rm}
+\bibliographystyle{achemso}
+\newcommand*\mycommand[1]{\texttt{\emph{#1}}}
 
-%\textwidth = 6.50 in
-%\textheight = 9.5 in
-%\oddsidemargin =  0.0 in
-%\evensidemargin = 0.0 in
-%\topmargin = -0.50 in
-%\headheight = 0.0 in
-%\headsep = 0.25 in
-%\parskip = 0.15in
-%\linespread{1.75}
-\doublespace
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Meta-data block
+%% ---------------
+%% Each author should be given as a separate \author command.
+%%
+%% Corresponding authors should have an e-mail given after the author
+%% name as an \email command. Phone and fax numbers can be given
+%% using \phone and \fax, respectively; this information is optional.
+%%
+%% The affiliation of authors is given after the authors; each
+%% \affiliation command applies to all preceding authors not already
+%% assigned an affiliation.
+%%
+%% The affiliation takes an option argument for the short name.  This
+%% will typically be something like "University of Somewhere".
+%%
+%% The \altaffiliation macro should be used for new address, etc.
+%% On the other hand, \alsoaffiliation is used on a per author basis
+%% when authors are associated with multiple institutions.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\author{Michael Vilkhovoy}
+\author{Nicholas Horvath}
+\author{Jeffrey D. Varner}
+\email{jdv27@cornell.edu}
+\phone{+1 607 2554258}
+\fax{+1 607 2559166}
+\affiliation[Cornell University]
+{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\title{Analysis of Cell-Free Synthetic Circuits using Sequence Specific Constraints Based Modeling}
+%\abbreviations{MV,NH,JDV}
+\keywords{Synthetic biology, Constraints based modeling, Biochemical modeling}
 
-%\bibliographystyle{chicago}
-\bibliographystyle{plos2009}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The manuscript does not need to include \maketitle, which is
+%% executed automatically.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{document}
 
-\makeatletter
-\renewcommand\subsection{\@startsection
-	{subsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\bfseries}}
-\renewcommand\subsubsection{\@startsection
-	{subsubsection}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\renewcommand\section{\@startsection
-	{subsection}{2}{0mm}
-	{-0.2in}
-	{0.05\baselineskip}
-	{\normalfont\large\bfseries}}
-\renewcommand\paragraph{\@startsection
-	{paragraph}{2}{0mm}
-	{-0.05in}
-	{-0.5\baselineskip}
-	{\normalfont\normalsize\itshape}}
-\makeatother
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The "tocentry" environment can be used to create an entry for the
+%% graphical table of contents. It is given here as some journals
+%% require that it is printed as part of the abstract page. It will
+%% be automatically moved as appropriate.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{tocentry}
 
-%Review style settings
-%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}
+Some journals require a graphical entry for the Table of Contents.
+This should be laid out ``print ready'' so that the sizing of the
+text is correct.
 
-%Publication style settings
+Inside the \texttt{tocentry} environment, the font used is Helvetica
+8\,pt, as required by \emph{Journal of the American Chemical
+Society}.
 
-% Single space'd bib -
-\setlength\bibsep{0pt}
+The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
+permitted for  \emph{Journal of the American Chemical Society}
+graphical table of content entries. The box will not resize if the
+content is too big: instead it will overflow the edge of the box.
 
-\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}
-\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
+This box and the associated title will always be printed on a
+separate page at the end of the document.
 
-% Change the number format in the ref list -
-\renewcommand{\bibnumfmt}[1]{#1.}
+\end{tocentry}
 
-% Change Figure to Fig.
-\renewcommand{\figurename}{Fig.}
-
-% Begin ...
-\begin{document}
-\begin{titlepage}
-{\par\centering\textbf{\Large {Analysis of TX-TL Synthetic Circuits using Sequence Specific Constraints Based Modeling}}}
-\vspace{0.05in}
-{\par \centering \large{Jeffrey D. Varner$^{*}$}}
-\vspace{0.10in}
-{\par \centering {School of Chemical and Biomolecular Engineering}}
-{\par \centering {Cornell University, Ithaca NY 14853}}
-\vspace{0.1in}
-{\par \centering \textbf{Running Title:}~Constraints based models of synthetic circuits}
-\vspace{0.1in}
-{\par \centering \textbf{To be submitted:}~\emph{Scientific~Reports}}
-\vspace{0.5in}
-{\par \centering $^{*}$Corresponding author:}
-{\par \centering Jeffrey D. Varner,}
-{\par \centering Professor, School of Chemical and Biomolecular Engineering,}
-{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
-{\par \centering Email: jdv27@cornell.edu}
-{\par \centering Phone: (607) 255 - 4258}
-{\par \centering Fax: (607) 255 - 9166}
-\end{titlepage}
-\date{}
-\thispagestyle{empty}
-\pagebreak
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\section*{Abstract}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The abstract environment will automatically gobble the contents
+%% if an abstract is not used by the target journal.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{abstract}
 In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} TX-TL system.
@@
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
+\end{abstract}
 
-\vspace{0.1in}
-{\noindent \textbf{Keywords:}~Synthetic biology, Constraints based modeling, Biochemical modeling}
-
-\pagebreak
-
-\setcounter{page}{1}
-
-\linenumbers
-
-\section*{Introduction}
-
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% Start the main part of the manuscript here.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Introduction}
 Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
@@
 Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
-Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \citep{Jewett:2008aa}.
-However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \citep{MATTHAEI:1961aa,NIRENBERG:1961aa}.
-Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \citep{Lu:2014aa} to synthetic biology \citep{Hodgman:2012aa}.
+Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
+However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
+Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
 Interestingly, many of the challenges confronting in-vivo genome-scale kinetic modeling can potentially be overcome in a cell-free system.
@@
 
-Stoichiometric reconstructions of microbial metabolism popularized by constraint based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \citep{2012_lewis_palsson_NatRevMicrobio}.
+Stoichiometric reconstructions of microbial metabolism popularized by constraint based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \cite{2012_lewis_palsson_NatRevMicrobio}.
 Since the first genome-scale stoichiometric model of \textit{E.
-coli}, developed by Edwards and Palsson \citep{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.
-coli} \citep{Feist:2007aa} or \textit{B.
-subtilis} \citep{Oh:2007aa}, are now available \citep{2009_feist_palsson_NatRevMicrobio}.
+coli}, developed by Edwards and Palsson \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.
+coli} \cite{Feist:2007aa} or \textit{B.
+subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
 Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems using linear programming.
 Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
-Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \citep{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \citep{2007_schuetz_sauer_MolSysBio}.
-Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \citep{2013_hyduke_lewis_palsson_MolBioSys}.
-Thus, modern constraint-based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \citep{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \cite{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \cite{2007_schuetz_sauer_MolSysBio}.
+Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}.
+Thus, modern constraint-based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
 
@@
 
-\clearpage
+\section{Results and discussion}
 
-%Constraints based models are important tools to understand and ultimately to predict how cells utilize nutrients to produce products.
-%Constraints based methods such as flux balance analysis (FBA) \cite{2010_orth_NatBiotech} and network decomposition approaches such as elementary modes (EMs) \cite{Schuster:2000aa}
-%or extreme pathways (EPs) \cite{Schilling:2000aa}
-%model intracellular metabolism using the biochemical stoichiometry and other constraints such as thermodynamical feasibility under pseudo-steady state conditions.
-%FBA has been used to efficiently estimate the performance of metabolic networks of arbitrary complexity, including genome scale networks, using linear programming  \cite{Covert:2004aa}.
-%On the other hand, EMs (or EPs) catalog all possible metabolic behaviors such that any flux distribution predicted by FBA is a convex combination
-%of the EMs (or EPs) \cite{Wiback:2003aa}.
-%However, the calculation of EMs (or EPs) is computationally expensive and currently infeasible for genome scale networks \cite{2004_lee_varner_ko_ieee}.
+\subsection{Outline}
 
-\section*{Results}
+The document layout should follow the style of the journal concerned.
+Where appropriate, sections and subsections should be added in the
+normal way. If the class options are set correctly, warnings will be
+given if these should not be present.
 
-\clearpage
-\section*{Discussion}
+\subsection{References}
 
-\clearpage
+The class makes various changes to the way that references are
+handled.  The class loads \textsf{natbib}, and also the
+appropriate bibliography style.  References can be made using
+the normal method; the citation should be placed before any
+punctuation, as the class will move it if using a superscript
+citation style
+The use of \textsf{natbib} allows the use of the various citation
+commands of that package: \citeauthor{Abernethy2003} have shown
+something, in \citeyear{Cotton1999}, or as given by
+Ref.~\citenum{Mena2000}.  Long lists of authors will be
+automatically truncated in most article formats, but not in
+supplementary information or reviews. If you
+encounter problems with the citation macros, please check that
+your copy of \textsf{natbib} is up to date. The demonstration
+database file \texttt{achemso-demo.bib} shows how to complete
+entries correctly. Notice that ``\latin{et al.}'' is auto-formatted
+using the \texttt{\textbackslash latin} command.
 
-\section*{Materials and Methods}
+Multiple citations to be combined into a list can be given as
+a single citation.  This uses the \textsf{mcitelus} package
+\cite{Johnson1972,*Arduengo1992,*Eisenstein2005,*Arduengo1994}.
+Citations other than the first of the list should be indicated
+with a star. If the \textsf{mcitelus} package is not installed,
+the standard bibliography tools will still work but starred
+references will be ignored. Individual references can be referred
+to using \texttt{\textbackslash mciteSubRef}:
+``ref.~\mciteSubRef{Eisenstein2005}''.
 
-\subsection*{Formulation and solution of the model equations.}
-The flux balance analysis problem was formulated as:
-\begin{equation}\nonumber
- \begin{multlined}
-	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
-	\mathrm{Subject \; to:}
-	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
- \end{multlined}
+The class also handles notes to be added to the bibliography.  These
+should be given in place in the document \bibnote{This is a note.
+The text will be moved the the references section.  The title of the
+section will change to ``Notes and References''.}.  As with
+citations, the text should be placed before punctuation.  A note is
+also generated if a citation has an optional note.  This assumes that
+the whole work has already been cited: odd numbering will result if
+this is not the case .
+
+\subsection{Floats}
+
+New float types are automatically set up by the class file.  The
+means graphics are included as follows (Scheme~\ref{sch:example}).  As
+illustrated, the float is ``here'' if possible.
+\begin{scheme}
+  Your scheme graphic would go here: \texttt{.eps} format\\
+  for \LaTeX\, or \texttt{.pdf} (or \texttt{.png}) for pdf\LaTeX\\
+  \textsc{ChemDraw} files are best saved as \texttt{.eps} files:\\
+  these can be scaled without loss of quality, and can be\\
+  converted to \texttt{.pdf} files easily using \texttt{eps2pdf}.\\
+  %\includegraphics{graphic}
+  \caption{An example scheme}
+  \label{sch:example}
+\end{scheme}
+
+\begin{figure}
+  As well as the standard float types \texttt{table}\\
+  and \texttt{figure}, the class also recognises\\
+  \texttt{scheme}, \texttt{chart} and \texttt{graph}.
+  \caption{An example figure}
+  \label{fgr:example}
+\end{figure}
+
+Charts, figures and schemes do not necessarily have to be labelled or
+captioned.  However, tables should always have a title. It is
+possible to include a number and label for a graphic without any
+title, using an empty argument to the \texttt{\textbackslash caption}
+macro.
+
+The use of the different floating environments is not required, but
+it is intended to make document preparation easier for authors. In
+general, you should place your graphics where they make logical
+sense; the production process will move them if needed.
+
+\subsection{Math(s)}
+
+The \textsf{achemso} class does not load any particular additional
+support for mathematics.  If packages such as \textsf{amsmath} are
+required, they should be loaded in the preamble.  However,
+the basic \LaTeX\ math(s) input should work correctly without
+this.  Some inline material \( y = mx + c \) or $ 1 + 1 = 2 $
+followed by some display. \[ A = \pi r^2 \]
+
+It is possible to label equations in the usual way (Eq.~\ref{eqn:example}).
+\begin{equation}
+  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
 \end{equation}
-where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
-and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-In this study, the objective $w_{obj}$ was to maximize the production of circuit output.
+This can also be used to have equations containing graphical
+content. To align the equation number with the middle of the graphic,
+rather than the bottom, a minipage may be used.
+\begin{equation}
+  \begin{minipage}[c]{0.80\linewidth}
+    \centering
+    As illustrated here, the width of \\
+    the minipage needs to allow some  \\
+    space for the number to fit in to.
+    %\includegraphics{graphic}
+  \end{minipage}
+  \label{eqn:graphic}
+\end{equation}
 
-\subsection*{Transcription and translation template reactions.}
+\section{Experimental}
 
-\subsection*{Global sensitivity analysis.}
-We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
-We computed the total sensitivity index of each parameter relative to two performance objectives, the peak thrombin time and the area under the thrombin curve (thrombin exposure).
-We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble.
-We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
-where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 10,000 and $d$ = 22, so the total sensitivity indices were computed from
-460,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
+The usual experimental details should appear here.  This could
+include a table, which can be referenced as Table~\ref{tbl:example}.
+Notice that the caption is positioned at the top of the table.
+\begin{table}
+  \caption{An example table}
+  \label{tbl:example}
+  \begin{tabular}{ll}
+    \hline
+    Header one  & Header two  \\
+    \hline
+    Entry one   & Entry two   \\
+    Entry three & Entry four  \\
+    Entry five  & Entry five  \\
+    Entry seven & Entry eight \\
+    \hline
+  \end{tabular}
+\end{table}
 
-\clearpage
+Adding notes to tables can be complicated.  Perhaps the easiest
+method is to generate these using the basic
+\texttt{\textbackslash textsuperscript} and
+\texttt{\textbackslash emph} macros, as illustrated (Table~\ref{tbl:notes}).
+\begin{table}
+  \caption{A table with notes}
+  \label{tbl:notes}
+  \begin{tabular}{ll}
+    \hline
+    Header one                            & Header two \\
+    \hline
+    Entry one\textsuperscript{\emph{a}}   & Entry two  \\
+    Entry three\textsuperscript{\emph{b}} & Entry four \\
+    \hline
+  \end{tabular}
 
-\section*{Acknowledgements}
-This study was supported by an award from the Army Research Office (ARO \#59155-LS).
-\clearpage
+  \textsuperscript{\emph{a}} Some text;
+  \textsuperscript{\emph{b}} Some more text.
+\end{table}
 
+The example file also loads the optional \textsf{mhchem} package, so
+that formulas are easy to input: \texttt{\textbackslash ce\{H2SO4\}}
+gives \ce{H2SO4}.  See the use in the bibliography file (when using
+titles in the references section).
+
+The use of new commands should be limited to simple things which will
+not interfere with the production process.  For example,
+\texttt{\textbackslash mycommand} has been defined in this example,
+to give italic, mono-spaced text: \mycommand{some text}.
+
+\section{Extra information when writing JACS Communications}
+
+When producing communications for \emph{J.~Am.\ Chem.\ Soc.}, the
+class will automatically lay the text out in the style of the
+journal. This gives a guide to the length of text that can be
+accommodated in such a publication. There are some points to bear in
+mind when preparing a JACS Communication in this way.  The layout
+produced here is a \emph{model} for the published result, and the
+outcome should be taken as a \emph{guide} to the final length. The
+spacing and sizing of graphical content is an area where there is
+some flexibility in the process.  You should not worry about the
+space before and after graphics, which is set to give a guide to the
+published size. This is very dependant on the final published layout.
+
+You should be able to use the same source to produce a JACS
+Communication and a normal article.  For example, this demonstration
+file will work with both \texttt{type=article} and
+\texttt{type=communication}. Sections and any abstract are
+automatically ignored, although you will get warnings to this effect.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The "Acknowledgement" section can be given in all manuscript
+%% classes.  This should be given within the "acknowledgement"
+%% environment, which will make the correct section or running title.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{acknowledgement}
+
+Please use ``The authors thank \ldots'' rather than ``The
+authors would like to thank \ldots''.
+
+The author thanks Mats Dahlgren for version one of \textsf{achemso},
+and Donald Arseneau for the code taken from \textsf{cite} to move
+citations after punctuation. Many users have provided feedback on the
+class, which is reflected in all of the different demonstrations
+shown in this document.
+
+\end{acknowledgement}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The same is true for Supporting Information, which should use the
+%% suppinfo environment.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{suppinfo}
+
+A listing of the contents of each file supplied as Supporting Information
+should be included. For instructions on what should be included in the
+Supporting Information as well as how to prepare this material for
+publications, refer to the journal's Instructions for Authors.
+
+The following files are available free of charge.
+\begin{itemize}
+  \item Filename: brief description
+  \item Filename: brief description
+\end{itemize}
+
+\end{suppinfo}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% The appropriate \bibliography command should be placed here.
+%% Notice that the class file automatically sets \bibliographystyle
+%% and also names the section correctly.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \bibliography{References_v1}
 
-\clearpage
-
-% Supplemental figures -
-% Set the S-
-\renewcommand\thefigure{S\arabic{figure}}
-\renewcommand\thetable{T\arabic{table}}
-\renewcommand\thepage{S-\arabic{page}}
-\renewcommand\theequation{S\arabic{equation}}
-
-% Reset the counters -
-\setcounter{equation}{0}
-\setcounter{table}{0}
-\setcounter{figure}{0}
-\setcounter{page}{1}
-
 \end{document}

### 2016-12-15 Paper_v1.tex
draft_v1

@@
 \usepackage{graphicx}
+\usepackage{textcomp}
+\usepackage{multirow}
+\usepackage{booktabs}
 \usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{tocentry}
-
-Some journals require a graphical entry for the Table of Contents.
-This should be laid out ``print ready'' so that the sizing of the
-text is correct.
-
-Inside the \texttt{tocentry} environment, the font used is Helvetica
-8\,pt, as required by \emph{Journal of the American Chemical
-Society}.
-
-The surrounding frame is 9\,cm by 3.5\,cm, which is the maximum
-permitted for  \emph{Journal of the American Chemical Society}
-graphical table of content entries. The box will not resize if the
-content is too big: instead it will overflow the edge of the box.
-
-This box and the associated title will always be printed on a
-separate page at the end of the document.
-
-\end{tocentry}
 
@@
 \begin{abstract}
-In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} TX-TL system.
-A core \emph{E.~coli} metabolic model, consisting of XX metabolites and YY reactions, was developed from literature [REF].
-This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
-sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
-Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples
-the transcription and translation of circuit components with the availability of metabolic resources.
-Model parameters were largely taken from literature; our approach had very few adjustable parameters thereby allowing the a first principles prediction of circuit performance.
-We tested this approach by first simulating $\sigma_{70}$-induced deGFP expression and then expanded these studies to more complex multicomponent circuits.
+In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis system.
+A core \emph{E.~coli} metabolic model, consisting of glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
+Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
+Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
+We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins
 First principles predictions of circuit performance were consistent with measurements for a variety of cases.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance.
-Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity and yield.
+A sufficient energy supply with oxidative phosphorylation is instrumental for high carbon yields, in addition the translation rate is the next step to optimize for higher productivity and yield. 
+Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
 \end{abstract}
@@
 
-In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} TX-TL system.
-A core \emph{E.~coli} cell free metabolic model, consisting of XX metabolites and YY reactions, was developed from literature [REF].
+In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
+A core \emph{E.~coli} cell free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
 This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
 sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
-Thus, sequence specific constraints based modeling explicitly couples
-the transcription and translation of circuit components with the availability of metabolic resources.
-Model parameters were largely taken from literature; our approach had very few adjustable parameters thereby allowing the a first principles prediction of circuit performance.
-We tested this approach by first simulating $\sigma_{70}$-induced deGFP expression and then expanded these studies to more complex multicomponent circuits.
-First principles predictions of circuit performance were consistent with measurements for a variety of cases.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance.
+Thus, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
+We tested this approach by first simulating $\sigma_{70}$-induced deGFP expression at different plasmid concentrations and then expanded this for the production of CAT and deGFP under two different promoters.
+First principles predictions of circuit performance were consistent with measurements.
+We then investigated the productivity and carbon yield for a range of 8 different proteins for a variety of cases. 
+From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins. 
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate for productivity.
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
@@
 
-\subsection{Outline}
-
-The document layout should follow the style of the journal concerned.
-Where appropriate, sections and subsections should be added in the
-normal way. If the class options are set correctly, warnings will be
-given if these should not be present.
-
-\subsection{References}
-
-The class makes various changes to the way that references are
-handled.  The class loads \textsf{natbib}, and also the
-appropriate bibliography style.  References can be made using
-the normal method; the citation should be placed before any
-punctuation, as the class will move it if using a superscript
-citation style
-The use of \textsf{natbib} allows the use of the various citation
-commands of that package: \citeauthor{Abernethy2003} have shown
-something, in \citeyear{Cotton1999}, or as given by
-Ref.~\citenum{Mena2000}.  Long lists of authors will be
-automatically truncated in most article formats, but not in
-supplementary information or reviews. If you
-encounter problems with the citation macros, please check that
-your copy of \textsf{natbib} is up to date. The demonstration
-database file \texttt{achemso-demo.bib} shows how to complete
-entries correctly. Notice that ``\latin{et al.}'' is auto-formatted
-using the \texttt{\textbackslash latin} command.
-
-Multiple citations to be combined into a list can be given as
-a single citation.  This uses the \textsf{mcitelus} package
-\cite{Johnson1972,*Arduengo1992,*Eisenstein2005,*Arduengo1994}.
-Citations other than the first of the list should be indicated
-with a star. If the \textsf{mcitelus} package is not installed,
-the standard bibliography tools will still work but starred
-references will be ignored. Individual references can be referred
-to using \texttt{\textbackslash mciteSubRef}:
-``ref.~\mciteSubRef{Eisenstein2005}''.
-
-The class also handles notes to be added to the bibliography.  These
-should be given in place in the document \bibnote{This is a note.
-The text will be moved the the references section.  The title of the
-section will change to ``Notes and References''.}.  As with
-citations, the text should be placed before punctuation.  A note is
-also generated if a citation has an optional note.  This assumes that
-the whole work has already been cited: odd numbering will result if
-this is not the case .
-
-\subsection{Floats}
-
-New float types are automatically set up by the class file.  The
-means graphics are included as follows (Scheme~\ref{sch:example}).  As
-illustrated, the float is ``here'' if possible.
-\begin{scheme}
-  Your scheme graphic would go here: \texttt{.eps} format\\
-  for \LaTeX\, or \texttt{.pdf} (or \texttt{.png}) for pdf\LaTeX\\
-  \textsc{ChemDraw} files are best saved as \texttt{.eps} files:\\
-  these can be scaled without loss of quality, and can be\\
-  converted to \texttt{.pdf} files easily using \texttt{eps2pdf}.\\
-  %\includegraphics{graphic}
-  \caption{An example scheme}
-  \label{sch:example}
-\end{scheme}
-
-\begin{figure}
-  As well as the standard float types \texttt{table}\\
-  and \texttt{figure}, the class also recognises\\
-  \texttt{scheme}, \texttt{chart} and \texttt{graph}.
-  \caption{An example figure}
-  \label{fgr:example}
+\section*{Results}
+\subsection{Model Derivation}
+The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
+One mathematical framework that has found wide use in modeling metabolism are constraint based models such as flux balance analysis (FBA).
+FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
+Traditionally, FBA is used to model \textit{in vivo} processes, however cell-free systems do not have growth associated reactions or transport through the cell membrane.
+Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
+The network consisted of 281 reactions and 132 species. 
+We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS. 
+We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
+The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
+The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E.
+coli} extract for a duration of 8 hours under maltose consumption(Fig \ref{fig:CAT_GFP_prod}B).
+The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
+Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
+Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
+\caption{Sequence specific flux balance analysis of protein production vesus time.
+A.
+CAT production under a T7 promoter in CFPS \textit{E.
+coli} extract for 1 hr under glucose consumption.
+B.
+deGFP production under a P70 promoter in TXTL 2.0 \textit{E.
+coli} extract for 8 hr under glucose consumption.
+95\% CI (blue region) over the ensemble of 100 sets.}
+\label{fig:CAT_GFP_prod}
 \end{figure}
 
-Charts, figures and schemes do not necessarily have to be labelled or
-captioned.  However, tables should always have a title. It is
-possible to include a number and label for a graphic without any
-title, using an empty argument to the \texttt{\textbackslash caption}
-macro.
+Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:GFP_plasmid}).
+Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{2016_Garamella_etal_TXTL}.
+The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.  
+However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
+\begin{figure}[t!]
+\centering
+\includegraphics[width=0.6\textwidth]{./Figures/GFP_plasmid.pdf}
+\caption{Predicted deGFP concentration (black line) at different plasmid concentrations versus measurements of deGFP (dots) synthesized in TXTL 2.0.
+95\% CI (blue region) over the ensemble of 100 sets.}
+\label{fig:GFP_plasmid}
+\end{figure}
 
-The use of the different floating environments is not required, but
-it is intended to make document preparation easier for authors. In
-general, you should place your graphics where they make logical
-sense; the production process will move them if needed.
+These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters. 
+It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems. 
+Since the model accurately predicted the behavior of protein production, we wanted to use our mathematical framework to help us understand the performance limits of CFPS and how these shortcomings could be addressed.
 
-\subsection{Math(s)}
+\subsection{Examining CFPS Performance: Productivity}
+Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases.
+Each of the proteins were produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
+In all our cases, CFPS is supplied with glucose.
+The first case, CFPS is supplied with amino acids and the system can synthesis amino acids (control).
+In the second case, CFPS is supplied with amino acids, however the system cannot synthesis amino acids (AA uptake w/o synthesis).
+We turned off these synthesis reactions because during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present.
+In the third case, CFPS is not supplied with amino acids, but the system can synthesis them (AA synthesis w/o uptakte).
+We used ssFBA to estimate the productivity of eight proteins for each case (Fig \ref{fig:Prod_POI}A).
+The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
+This shows the system has sufficient substrates to power the system and synthesis each protein. 
+The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
+The qualitative trend of productivity for the three different cases was the same, however some proteins had higher productivity than others. 
+For instance, in the second case FGF21 had a productivity of 17 ($\mu$M/h) whereas FII had a productivity of 3 ($\mu$M/h).
+To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
+The proteins with the highest productivity had the lowest carbon number whereas proteins with low productivity had higher carbon numbers.     
+This inverse qualitative trend was due to the fact that larger proteins require more amino acids and substrates to assemble them resulting in lower productivity. 
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
+\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey).
+A.
+Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble).
+B.
+Mean productivity versus the carbon number for each corresponding protein.}
+\label{fig:Prod_POI}
+\end{figure}
 
-The \textsf{achemso} class does not load any particular additional
-support for mathematics.  If packages such as \textsf{amsmath} are
-required, they should be loaded in the preamble.  However,
-the basic \LaTeX\ math(s) input should work correctly without
-this.  Some inline material \( y = mx + c \) or $ 1 + 1 = 2 $
-followed by some display. \[ A = \pi r^2 \]
+\subsection{Examining CFPS Performance: Carbon Yield}
+Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A). 
+The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance. 
+The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is now utilized to synthesize the necessary amino acids for each protein as well as power the system.
+We also investigated the carbon contribution of glucose and each amino acid toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
+For the first case, the system is relying on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
+Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield is increased to 22.2\%.
+Only the necessary amount of amino acids are used for the production of CAT, thus it may be hypothesized that all the glucose is used to power CFPS and does not contribute to the carbon yield. 
+If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
+Finally for the third case (without amino acid supplemented), the carbon yield reduces to 14.8\% and the system uses about double the amount of glucose as the first two cases. 
+For this case glucose is being used to synthesize amino acids and power the system and this trend was seen across all proteins except for CAT.
 
-It is possible to label equations in the usual way (Eq.~\ref{eqn:example}).
-\begin{equation}
-  \frac{\mathrm{d}}{\mathrm{d}x} \, r^2 = 2r \label{eqn:example}
-\end{equation}
-This can also be used to have equations containing graphical
-content. To align the equation number with the middle of the graphic,
-rather than the bottom, a minipage may be used.
-\begin{equation}
-  \begin{minipage}[c]{0.80\linewidth}
+\begin{table}[t!]
 \centering
-    As illustrated here, the width of \\
-    the minipage needs to allow some  \\
-    space for the number to fit in to.
-    %\includegraphics{graphic}
-  \end{minipage}
-  \label{eqn:graphic}
-\end{equation}
-
-\section{Experimental}
-
-The usual experimental details should appear here.  This could
-include a table, which can be referenced as Table~\ref{tbl:example}.
-Notice that the caption is positioned at the top of the table.
-\begin{table}
-  \caption{An example table}
-  \label{tbl:example}
-  \begin{tabular}{ll}
-    \hline
-    Header one  & Header two  \\
-    \hline
-    Entry one   & Entry two   \\
-    Entry three & Entry four  \\
-    Entry five  & Entry five  \\
-    Entry seven & Entry eight \\
-    \hline
+    \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
+    \renewcommand{\arraystretch}{0.95}
+    \begin{tabular}{lccc} \toprule
+        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Produced (mM)} & Control & \parbox{2.5 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
+        deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
+        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Consumed (mM)} & \\ \midrule
+        GLC & 37.3 & 33.7 & 66.9 \\ 
+        ALA & 0.0 & 0.2 & -  \\ 
+        ARG & 0.3 & 0.3 & - \\ 
+        ASN & 0.5 & 0.4 & -  \\ 
+        ASP & 0.4 & 0.6 & -  \\ 
+        CYS & 0.2 & 0.1 & -  \\ 
+        GLU & 2.1 & 0.6 & -  \\ 
+        GLN & 0.4 & 0.3 & - \\ 
+        GLY & 0.3 & 0.3 & -  \\
+        HIS & 0.5 & 0.5 & - \\
+        ILE & 0.0 & 0.6 & -  \\ 
+        LEU & 1.0 & 1.0 & - \\ 
+        LYS & 1.0 & 0.9 & - \\ 
+		MET & 0.0 & 0.2 & -  \\ 
+        PHE & 1.0 & 0.9 & - \\ 
+        PRO & 0.4 & 0.4 & -  \\ 
+        SER & 0.0 & 0.2 & - \\ 
+        THR & 1.0 & 0.5 & -  \\
+        TRP & 0.1 & 0.1 & -  \\ 
+        TYR & 0.8 & 0.8 & - \\ 
+        VAL & 0.6 & 0.7 & - \\ \hline
+        Sum & 47.9 & 43.3 & 66.9  \\ \toprule
+        Yield & 20.5\% & 22.2\% & 14.8\% \\ 
+	Yield w/o GLC & 92.5\% & 100\% & - \\ \bottomrule
     \end{tabular}
+\label{tbl:yield_breakdown}
 \end{table}
 
-Adding notes to tables can be complicated.  Perhaps the easiest
-method is to generate these using the basic
-\texttt{\textbackslash textsuperscript} and
-\texttt{\textbackslash emph} macros, as illustrated (Table~\ref{tbl:notes}).
-\begin{table}
-  \caption{A table with notes}
-  \label{tbl:notes}
-  \begin{tabular}{ll}
-    \hline
-    Header one                            & Header two \\
-    \hline
-    Entry one\textsuperscript{\emph{a}}   & Entry two  \\
-    Entry three\textsuperscript{\emph{b}} & Entry four \\
-    \hline
-  \end{tabular}
+Interestingly, CAT carbon yield showed similar performance for all three cases. 
+Each protein has different energy requirements for transcription and translation depending on its sequence.
+Thus, for the case of CAT, energy requirements for transcription or translation are low enough that it's carbon yield is not hindered in the third case.
+We next investigated the effect of the carbon number of each protein to the carbon yield (Fig \ref{fig:Yield_POI}B). 
+The same inverse qualitative trend is observed, however it is less significant. 
+Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
+In addition, ssFBA is solved by setting the production of the protein as the objective function.
+Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction. 
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/Yield_POI.pdf}
+\caption{CFPS carbon yield performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey).
+A.
+Carbon yield (Error bars represent the 95\% CI of the ensemble).
+B.
+Mean carbon yield versus the carbon number for each corresponding protein.}
+\label{fig:Yield_POI}
+\end{figure}
 
-  \textsuperscript{\emph{a}} Some text;
-  \textsuperscript{\emph{b}} Some more text.
-\end{table}
+\subsection{Sensitivity analysis on CFPS Performance}
+To better understand the effect of substrate utilization and transcription/translation parameters on CFPS performance we performed global sensitivity analysis on productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
+In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters were fairly constant across all three cases, with the elongation rate by ribosomes being the most sensitive.
+This shows the rate of translation is instrumental in achieving high productivity, and should be the first step to investigate in order to optimize productivity, prior to examining transcription rates, this is consistent with literature findings.
+Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
+In addition, Li et al.
+have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids was removed from CFPS.
+This makes sense, as amino acid synthesis can only come from glucose and becomes the only way to power protein synthesis.
+On the other hand, when amino acid synthesis is removed, amino acid uptake becomes more important, as it becomes the only source of amino acids.
+\begin{figure}[t!]
+\includegraphics[width=1.00\textwidth]{./Figures/SI_GFP.pdf}
+\caption{Total order sensitivity of deGFP productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).
+(Error bars represent 95\% CI of the ensemble).}
+\label{fig:SI_GFP}
+\end{figure}
 
-The example file also loads the optional \textsf{mhchem} package, so
-that formulas are easy to input: \texttt{\textbackslash ce\{H2SO4\}}
-gives \ce{H2SO4}.  See the use in the bibliography file (when using
-titles in the references section).
+When considering carbon yield performance (Fig \ref{fig:SI_GFP}B), the substrate and oxygen uptake fluxes become much more important while the sensitivity to the transcription/translation parameters decreases slightly.
+The transcription/translation parameters had the same trend as in the productivity trend, where the ribosome elongation rate was the most sensitive compared to the transcription/translation parameters and showed significance across all parameters.
+Thus, in investigating CFPS performance in terms of productivity and yield, the ribosome elongation rate seems to be an important parameter to optimize, as been already shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
+Glucose and oxygen uptake shown significant importance for carbon yield since it determines the mechanism of energy generation, which is critical to efficient protein production.
+Meanwhile, productivity is determined primarily by the rate of the most downstream steps, transcription and translation.
+Glucose is required for both amino acid synthesis and efficient energy generation, both of which are important for a high yield.
+There is a tradeoff between energy generation to power transcription/translation as well as synthesis of amino acids which are required to assemble the protein of interest. 
+Across all three cases, substrate utilization (amino acid uptake and glucose) showed to be the next important as these substrates are contributing to the carbon yield of deGFP.
+Interestingly, in the first two cases, oxygen uptake has a significant effect on carbon yield.
+This is most likely due to the importance of oxidative phosphorylation for efficient energy generation.
+However, in the third case (with no amino acid uptake), oxygen uptake becomes insignificant compared to the other parameters, while glucose uptake was the most siginificant.
+Jewett and coworkers have reported that oxidative phosphorylation still operates in cell free and also saw a decrease in yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions\cite{Jewett:2008aa}.
+However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, thus to investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig \ref{fig:oxphos_yield}).
+Interestingly, oxidative phosphorylation has a somewhat linear correlation with carbon yield for the first two cases (R\textsuperscript{2}=0.77 for the control, R\textsuperscript{2}=0.79 for the second case), whereas for the third case, this same effect was not as prevalent (R\textsuperscript{2}=0.46).
+\begin{figure}[t!]
+\includegraphics[width=0.7\textwidth]{./Figures/OxPhos_yield.pdf}
+\caption{An ensemble of 1000 ssFBA solutions for deGFP carbon yield versus oxidative phosphorylation flux for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).
+}
+\label{fig:oxphos}
+\end{figure}
 
-The use of new commands should be limited to simple things which will
-not interfere with the production process.  For example,
-\texttt{\textbackslash mycommand} has been defined in this example,
-to give italic, mono-spaced text: \mycommand{some text}.
+When the system can utilize glucose for energy generation, the increase in oxidative phosphorylation has a positive effect on carbon yield. 
+However, when glucose is used to synthesize amino acids and power CFPS, there is a limit of how much oxidative phosphorylation can effect carbon yield.
+Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and where the remaining energy is coming from. 
+To investigate the differences in productivity and carbon yields, we compared the flux distributions for deGFP predicted by ssFBA simulations for the three different cases (Fig.\ref{fig:flux}).
+All cases used the first step in the pentose phosphate pathway to generate NADPH;
+the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
+The majority of the flux was distributed between lactate and acetate, representing an inefficiency in energy generation as well as wasteful byproducts.
+For the first two cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
+The second case with no amino acid synthesis showed to be the most efficient process (highest carbon yield).
+All the NADPH produced is interconverted to NADH to power oxidative phosphorylation.
+Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow 
+However, diverting the flux away from acetate accumulation may allow more resources towards the protein of interest.
+Interestingly, for the first case and more for the third case, there is a mixture of aerobic and anearobic energy generation. 
+Fumerate dehydrogenase typically operates in anaerobic conditions and is less efficient than oxidative phosphorylation. 
+Cell-free systems do not have a sense of regulation as \textit{in vivo} systems thus any enzyme present in the mixture will carry out its corresponding reaction.
+In the third case (third row), there is a low flux through oxidative phosphorylation and high flux through fumerate dehydrogenase (fumerate to succinate). 
+One of the reasons for this inefficiency is some of the NADH is interconverted to NADPH which is required for amino acid synthesis, thus the lack of NADH cannot power oxidative phosphorylation as effectively as in the first two cases.
+It is currently unknown how much oxidative phosphorylation is present in cell-free systems, however our model and the sensitivity analysis suggests that efficient energy generation is essential to high carbon yields.
+Thus, removing anaerobic enzymes during the cell-free extract preparation is a potential area for improvement for CFPS performance and protein yield. 
+\begin{figure}[h!]
+\includegraphics[width=0.8\textwidth]{./Figures/flux.pdf}
+\caption{deGFP production flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
+Flux (mean $\pm$ standard deviation) across ensemble normalized to glucose uptake flux.
+Flux distribution for three different cases: control (top row), amino acid uptake without synthesis (second row), and amino acid synthesis without uptake (third row).
+Bold metabolites are allowed to accumulate; grey arrows lead to amino acid synthesis.}
+\label{fig:flux}
+\end{figure}
 
-\section{Extra information when writing JACS Communications}
+All proteins followed a similar trend where for the first two cases (when amino acids were supplied) there was a relatively high yield than compared to the third case, with the exception of CAT. 
+Thus, we also performed a sensitivity analysis on CAT (Fig \ref{fig:SI_CAT}), which did not show the same trend in yield across the three cases as the other proteins.
+For CAT, glucose uptake was as important to productivity as the kinetic parameters.
+The oxygen uptake was the most important factor in yield and maintained this across all three cases.
+The energy requirements for transcription of CAT is lower than the other proteins, thus it can accomodate its resources towards efficient energy generation.
+Thus, the carbon yield is not hindered in the third case, since the system maintains the necessary amino acid synthesis rates and the required energy species.
+Apart from these differences, there was a general trend of transcription and translation holding more significance for productivity and substrate uptake holding more significance for yield.
 
-When producing communications for \emph{J.~Am.\ Chem.\ Soc.}, the
-class will automatically lay the text out in the style of the
-journal. This gives a guide to the length of text that can be
-accommodated in such a publication. There are some points to bear in
-mind when preparing a JACS Communication in this way.  The layout
-produced here is a \emph{model} for the published result, and the
-outcome should be taken as a \emph{guide} to the final length. The
-spacing and sizing of graphical content is an area where there is
-some flexibility in the process.  You should not worry about the
-space before and after graphics, which is set to give a guide to the
-published size. This is very dependant on the final published layout.
+Taken together, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
+We have shown sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate of protein production for deGFP and CAT under two different proteins.
+This modeling approach validated general trends seen in literature for productivity and yield as a function of carbon number. 
+Further, global sensitivity analysis identified oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate is important for productivity.
+The model also identified that cell-free systems can operate with aerobic and anaerobic enzymes which leads to inefficient CFPS systems and should be addressed to optimize carbon yield as seen in the globol sensitivity analysis.
+Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
 
-You should be able to use the same source to produce a JACS
-Communication and a normal article.  For example, this demonstration
-file will work with both \texttt{type=article} and
-\texttt{type=communication}. Sections and any abstract are
-automatically ignored, although you will get warnings to this effect.
+\begin{figure}[h!]
+\includegraphics[width=1.00\textwidth]{./Figures/SI_CAT.pdf}
+\caption{Total order sensitivity of CAT productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).
+(Error bars represent 95\% CI of the ensemble).}
+\label{fig:SI_CAT}
+\end{figure}
+
+\section*{Materials and Methods}
+\subsection*{Formulation and solution of the model equations.}
+The flux balance analysis problem was formulated as:
+\begin{equation}\nonumber
+ \begin{multlined}
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+	\mathrm{Subject \; to:}
+	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
+\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+ \end{multlined}
+\end{equation}
+where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
+
+\subsection*{Sequence-specific flux balance analysis}
+We estimated the theoretical maximum carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+The sequence-specific flux balance analysis problem was formulated as a linear program:
+\begin{equation}
+ \begin{multlined}
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+	\mathrm{Subject \; to:}
+	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
+\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+ \end{multlined}
+\end{equation}
+where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
+The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
+\begin{eqnarray*}
+\mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
+\mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
+\mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
+\mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
+\mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
+&& \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
+\mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
+&& \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
+\end{eqnarray*}
+where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
+$\rm R_{1}$ denotes the concentration of RNA polymerase,
+$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
+$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
+$\rm Pi$ denotes inorganic phosphate,
+$\rm R_{2}$ denotes the ribosome concentration,
+$\rm R_{2}^{*}$ denotes bounded ribosome,
+and $AA_{j}$ denotes $j^{th}$ amino acid.
+
+The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
+\begin{equation}
+	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
+\end{equation}
+where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
+The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
+\begin{equation}
+	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)P\right]
+\end{equation}
+The term $R_{1}$ denotes the RNA polymerase abundance,
+$v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
+$l_{G}$ denotes the gene length in nucleotides, and $P$ describes the promoter activity.
+ The T7 binding promoter was used for CAT production was formulated as:
+\begin{equation}
+	P_{T7} = (\frac{K_{T7}}{1 + K_{T7}}
+\end{equation}
+where $K_{T7}$ denotes a T7 RNA polymerase binding constant \cite{Moon:2012aa}.
+The $\sigma_{70}$ binding promoter was used for all other proteins was formulated as:
+\begin{equation}
+	P_{\sigma_{70}} = \frac{K_{1}+K_{2}f_{p70}}{1 + K_{1}+K_{2}f_{p70}}
+\end{equation}
+where $K_{1}$ denotes the state of RNA polymerase binding, $K_{2}$ is the state of sigma-70 binding along with RNA polymerase, and $f_{p70}$ is the fraction of the transcription factor, sigma-70, bound to the promoter following Hills kinetics.
+
+On the other hand, the translation rate ($w_{TL}$) was bounded by:
+ \begin{equation}
+	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{\rm mRNA_{SS}}{K_{TL}+\rm mRNA_{SS}}\right)
+\end{equation}
+where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance, and $K_{TL}$ denotes the translation saturation constant.
+The maximum translation rate $V_{TL}^{max}$ was formulated as:
+\begin{equation}
+	V_{TL}^{max} \equiv \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]
+\end{equation}
+The term $K_{P}$ denotes the polysome amplification constant,
+$v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
+$l_{P}$ denotes the number of amino acids in the protein of interest,
+and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
+\begin{equation}
+	\rm mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+\end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+
+The objective of the sequence-specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
+The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
+while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
+The protein gene and protein sequences were taken from literature and are available in the Supporting Information.
+The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
+
+\subsubsection*{Quantification of uncertainty.}
+An ensemble of 100 sets of flux distributions was calculated for three different cases: control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
+An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data and randomly sampling oxygen uptake, RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
+
+\subsection*{Calculation of the carbon yield.}
+The carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as the protein of interest divided by the carbon consumed as reactants (glucose and amino acids):
+\begin{equation}\label{eqn:yield-definition}
+	Y_{C}^{POI}=\frac{\Delta\texttt{POI}\cdot C_{POI}}{\displaystyle\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\cdot C_{m_i}}
+\end{equation}
+where $\Delta\texttt{POI}$ denotes the abundance of the protein of interest produced, $C_{POI}$ denotes carbon number of the protein of interest, $\mathcal{R}$ denotes the number of reactants,
+$\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
+
+\subsection*{Global sensitivity analysis.}
+We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
+We computed the total sensitivity index of each parameter relative to two performance objectives, productivity of the protein of interest and the carbon yield.
+We established the sampling bounds for each parameter from literature.
+We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
+where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 1,000 and $d$ = 7, so the total sensitivity indices were computed from
+16,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
 
@@
 
-A listing of the contents of each file supplied as Supporting Information
-should be included. For instructions on what should be included in the
-Supporting Information as well as how to prepare this material for
-publications, refer to the journal's Instructions for Authors.
-
 The following files are available free of charge.

### 2016-12-19 Paper_v1.tex
Inital changes

@@
 \email{jdv27@cornell.edu}
-\phone{+1 607 2554258}
-\fax{+1 607 2559166}
+\phone{+1~(607)~255-4258}
+\fax{+1~(607)~255-9166}
 \affiliation[Cornell University]
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Analysis of Cell-Free Synthetic Circuits using Sequence Specific Constraints Based Modeling}
+\title{Sequence Specific Modeling of Cell-Free Protein Synthesis}
 %\abbreviations{MV,NH,JDV}
@@
 Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
-However, CFPS is not new; CFPS in crude \textit{E. coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
+However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
 Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
@@
 
-\section*{Results}
 \subsection{Model Derivation}
@@
 
-\subsection{Examining CFPS Performance: Productivity}
+\subsection{CFPS Productivity}
 Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases.
@@
 
-\subsection{Examining CFPS Performance: Carbon Yield}
+\subsection{CFPS Carbon Yield}
 Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
@@
 
-\subsection{Sensitivity analysis on CFPS Performance}
+\subsection{Sensitivity analysis on CFPS performance}
 To better understand the effect of substrate utilization and transcription/translation parameters on CFPS performance we performed global sensitivity analysis on productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
@@
 \section*{Materials and Methods}
+
 \subsection*{Formulation and solution of the model equations.}
-The flux balance analysis problem was formulated as:
-\begin{equation}\nonumber
- \begin{multlined}
-	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
-	\mathrm{Subject \; to:}
-	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
- \end{multlined}
-\end{equation}
-where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
-and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-
-\subsection*{Sequence-specific flux balance analysis}
-We estimated the theoretical maximum carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+We estimated the theoretical maximum performance of the cell free protein synthesis system using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
 The sequence-specific flux balance analysis problem was formulated as a linear program:
@@
 \begin{equation}
-	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)P\right]
+	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)\mathcal{P}\right]
 \end{equation}
@@
 $v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
-$l_{G}$ denotes the gene length in nucleotides, and $P$ describes the promoter activity.
- The T7 binding promoter was used for CAT production was formulated as:
+$l_{G}$ denotes the gene length in nucleotides, and the last term $\mathcal{P}$ describes a model of promoter activity.
+In this study, we considered two promoters; the T7 and $\sigma_{70}$ promoters.
+The promoter function for the T7 promoter, $\mathcal{P}_{T7}$ was given by:
 \begin{equation}
@@
 \end{equation}
-where $K_{1}$ denotes the state of RNA polymerase binding, $K_{2}$ is the state of sigma-70 binding along with RNA polymerase, and $f_{p70}$ is the fraction of the transcription factor, sigma-70, bound to the promoter following Hills kinetics.
+where $K_{1}$ denotes the state of RNA polymerase binding,
+$K_{2}$ is the state of $\sigma_{70} $binding along with RNA polymerase,
+and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to the promoter (modeled as a Hill function).
 
@@
 
-The objective of the sequence-specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
+The objective of the sequence-specific flux balance calculation was to maximize the rate of protein translation, $w_{TL}$.
 The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
@@
 
-\subsubsection*{Quantification of uncertainty.}
-An ensemble of 100 sets of flux distributions was calculated for three different cases: control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
-An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data and randomly sampling oxygen uptake, RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
+% The flux balance analysis problem was formulated as:
+% \begin{equation}\nonumber
+%  \begin{multlined}
+% 	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+% 	\mathrm{Subject \; to:}
+% 	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
+% \alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+%  \end{multlined}
+% \end{equation}
+% where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+% and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+% The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
+%
+% \subsection*{Sequence-specific flux balance analysis}
+% We estimated the theoretical maximum carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+% The sequence-specific flux balance analysis problem was formulated as a linear program:
+% \begin{equation}
+%  \begin{multlined}
+% 	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+% 	\mathrm{Subject \; to:}
+% 	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
+% \alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+%  \end{multlined}
+% \end{equation}
+% where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+% and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+% The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
+% The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
+% \begin{eqnarray*}
+% \mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
+% \mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
+% \mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
+% \mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
+% \mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
+% && \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
+% \mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
+% && \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
+% \end{eqnarray*}
+% where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
+% $\rm R_{1}$ denotes the concentration of RNA polymerase,
+% $G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
+% $\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
+% $\rm Pi$ denotes inorganic phosphate,
+% $\rm R_{2}$ denotes the ribosome concentration,
+% $\rm R_{2}^{*}$ denotes bounded ribosome,
+% and $AA_{j}$ denotes $j^{th}$ amino acid.
+%
+% The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
+% \begin{equation}
+% 	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
+% \end{equation}
+% where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
+% The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
+% \begin{equation}
+% 	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)P\right]
+% \end{equation}
+% The term $R_{1}$ denotes the RNA polymerase abundance,
+% $v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
+% $l_{G}$ denotes the gene length in nucleotides, and $P$ describes the promoter activity.
+%  The T7 binding promoter was used for CAT production was formulated as:
+%
+%
+% On the other hand, the translation rate ($w_{TL}$) was bounded by:
+%  \begin{equation}
+% 	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{\rm mRNA_{SS}}{K_{TL}+\rm mRNA_{SS}}\right)
+% \end{equation}
+% where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance, and $K_{TL}$ denotes the translation saturation constant.
+% The maximum translation rate $V_{TL}^{max}$ was formulated as:
+% \begin{equation}
+% 	V_{TL}^{max} \equiv \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]
+% \end{equation}
+% The term $K_{P}$ denotes the polysome amplification constant,
+% $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
+% $l_{P}$ denotes the number of amino acids in the protein of interest,
+% and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
+% \begin{equation}
+% 	\rm mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+% \end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+%
+% The objective of the sequence-specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
+% The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
+% while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
+% The protein gene and protein sequences were taken from literature and are available in the Supporting Information.
+% The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
 
@@
 $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
+
+\subsubsection*{Quantification of uncertainty.}
+An ensemble of 100 sets of flux distributions was calculated for three different cases: control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
+An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data and randomly sampling oxygen uptake, RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
 

### 2016-12-22 Paper_v1.tex
edits and supplement

@@
 \usepackage{xkeyval}
+\usepackage{array}
+\usepackage{listings}
+\usepackage{lmodern}
+\usepackage{mathpazo}
+\usepackage{geometry}
+\usepackage{microtype}
 \usepackage{url}  % Formatting web addresses
@@
 \usepackage{booktabs}
+\usepackage{natmove}
+\usepackage{float}
 \usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
@@
 
-\bibliographystyle{achemso}
+%\bibliographystyle{achemso}
 \newcommand*\mycommand[1]{\texttt{\emph{#1}}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% Meta-data block
-%% ---------------
-%% Each author should be given as a separate \author command.
-%%
-%% Corresponding authors should have an e-mail given after the author
-%% name as an \email command. Phone and fax numbers can be given
-%% using \phone and \fax, respectively; this information is optional.
-%%
-%% The affiliation of authors is given after the authors; each
-%% \affiliation command applies to all preceding authors not already
-%% assigned an affiliation.
-%%
-%% The affiliation takes an option argument for the short name.  This
-%% will typically be something like "University of Somewhere".
-%%
-%% The \altaffiliation macro should be used for new address, etc.
-%% On the other hand, \alsoaffiliation is used on a per author basis
-%% when authors are associated with multiple institutions.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The manuscript does not need to include \maketitle, which is
-%% executed automatically.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%% The "tocentry" environment can be used to create an entry for the
-%% graphical table of contents. It is given here as some journals
-%% require that it is printed as part of the abstract page. It will
-%% be automatically moved as appropriate.
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \section{Introduction}
@@
 sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
-Thus, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
-We tested this approach by first simulating $\sigma_{70}$-induced deGFP expression at different plasmid concentrations and then expanded this for the production of CAT and deGFP under two different promoters.
+Thus, sequence specific constraints based modeling explicitly coupled the transcription and translation of circuit components with the availability of metabolic resources.
+We tested this approach by first simulating the production of CAT and deGFP under two different promoters.
+and then expanded this for $\sigma_{70}$-induced deGFP expression at different plasmid concentrations.
 First principles predictions of circuit performance were consistent with measurements.
@@
 The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E.
-coli} extract for a duration of 8 hours under maltose consumption(Fig \ref{fig:CAT_GFP_prod}B).
+coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:CAT_GFP_prod}B).
 The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
@@
 Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:GFP_plasmid}).
-Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{2016_Garamella_etal_TXTL}.
+Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
 The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
@@
 These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters.
-It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
+It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under two different promoters and cell-free systems.
 Since the model accurately predicted the behavior of protein production, we wanted to use our mathematical framework to help us understand the performance limits of CFPS and how these shortcomings could be addressed.
@@
 In all our cases, CFPS is supplied with glucose.
-The first case, CFPS is supplied with amino acids and the system can synthesis amino acids (control).
-In the second case, CFPS is supplied with amino acids, however the system cannot synthesis amino acids (AA uptake w/o synthesis).
-We turned off these synthesis reactions because during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present.
-In the third case, CFPS is not supplied with amino acids, but the system can synthesis them (AA synthesis w/o uptakte).
-We used ssFBA to estimate the productivity of eight proteins for each case (Fig \ref{fig:Prod_POI}A).
-The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
-This shows the system has sufficient substrates to power the system and synthesis each protein.
+The first case, CFPS is supplied with amino acids and the system can synthesize amino acids (control).
+In the second case, CFPS is supplied with amino acids, however the system cannot synthesize amino acids (AA uptake w/o synthesis).
+We turned off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present.
+In the third case, CFPS is not supplied with amino acids, but the system can synthesize them (AA synthesis w/o uptakte).
+Eight different proteins,ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
+We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod_POI}A).
+The first two cases had very similar performance, with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
+%The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
+This shows the system had sufficient substrates to power the system and synthesize each protein.
 The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
-The qualitative trend of productivity for the three different cases was the same, however some proteins had higher productivity than others.
-For instance, in the second case FGF21 had a productivity of 17 ($\mu$M/h) whereas FII had a productivity of 3 ($\mu$M/h).
+The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.
+The qualitative trend of productivity for the three different cases was the same within each case, however some proteins had higher productivity than others.
+For instance, in the second case FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
 To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
@@
 The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
-The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is now utilized to synthesize the necessary amino acids for each protein as well as power the system.
-We also investigated the carbon contribution of glucose and each amino acid toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
-For the first case, the system is relying on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
-Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield is increased to 22.2\%.
-Only the necessary amount of amino acids are used for the production of CAT, thus it may be hypothesized that all the glucose is used to power CFPS and does not contribute to the carbon yield.
+The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system. 
+To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
+For the first case, the system is relied on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
+Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield increased to 22.2\%.
+Only the necessary amount of amino acids were used for the production of deGFP, thus it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
 If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
-Finally for the third case (without amino acid supplemented), the carbon yield reduces to 14.8\% and the system uses about double the amount of glucose as the first two cases.
-For this case glucose is being used to synthesize amino acids and power the system and this trend was seen across all proteins except for CAT.
+Finally for the third case (without amino acid supplemented), the carbon yield was reduced to 14.8\% and the system used about double the amount of glucose as the first two cases.
+In this case glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
+Interestingly, CAT carbon yield showed similar performance for all three cases, however it still had a lower yield than the first two cases.
+Each protein has different energy requirements for transcription and translation depending on its sequence.
+Thus, for the case of CAT, energy requirements for transcription or translation were low enough that it's carbon yield was not hindered in the third case as significantly as for the other proteins.
 
@@
     \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
-    \renewcommand{\arraystretch}{0.95}
+    \renewcommand{\arraystretch}{0.9}
     \begin{tabular}{lccc} \toprule
@@
 
-Interestingly, CAT carbon yield showed similar performance for all three cases.
-Each protein has different energy requirements for transcription and translation depending on its sequence.
-Thus, for the case of CAT, energy requirements for transcription or translation are low enough that it's carbon yield is not hindered in the third case.
 We next investigated the effect of the carbon number of each protein to the carbon yield (Fig \ref{fig:Yield_POI}B).
-The same inverse qualitative trend is observed, however it is less significant.
-Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
-In addition, ssFBA is solved by setting the production of the protein as the objective function.
-Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
+The same inverse qualitative trend is observed as for productivity, however it was less significant.
+The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
+As the size of the protein to be expressed increases, the productivity and carbon yield decreases, thus proteins that are too large may not be feasible for cell-free production.
+Thus, we wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS systems most effectively.
+%Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
+%In addition, ssFBA is solved by setting the production of the protein as the objective function.
+%Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
 \begin{figure}[t!]
@@
 
-\subsection{Sensitivity analysis on CFPS performance}
-To better understand the effect of substrate utilization and transcription/translation parameters on CFPS performance we performed global sensitivity analysis on productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
+\subsection{CFPS performance sensitivity analysis}
+To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
 In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters were fairly constant across all three cases, with the elongation rate by ribosomes being the most sensitive.
-This shows the rate of translation is instrumental in achieving high productivity, and should be the first step to investigate in order to optimize productivity, prior to examining transcription rates, this is consistent with literature findings.
+This showed the rate of translation is instrumental in achieving high productivity, and should be the first step to investigate in order to optimize productivity, prior to examining transcription rates, this is consistent with literature findings.
 Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
@@
 have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
-In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids was removed from CFPS.
-This makes sense, as amino acid synthesis can only come from glucose and becomes the only way to power protein synthesis.
-On the other hand, when amino acid synthesis is removed, amino acid uptake becomes more important, as it becomes the only source of amino acids.
+In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
+This makes sense, as amino acid synthesis can only come from glucose and became the only way to power protein synthesis.
+On the other hand, when amino acid synthesis is removed, amino acid uptake became more important, as it becomes the only source of amino acids.
 \begin{figure}[t!]
@@
 
-When considering carbon yield performance (Fig \ref{fig:SI_GFP}B), the substrate and oxygen uptake fluxes become much more important while the sensitivity to the transcription/translation parameters decreases slightly.
-The transcription/translation parameters had the same trend as in the productivity trend, where the ribosome elongation rate was the most sensitive compared to the transcription/translation parameters and showed significance across all parameters.
-Thus, in investigating CFPS performance in terms of productivity and yield, the ribosome elongation rate seems to be an important parameter to optimize, as been already shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
-Glucose and oxygen uptake shown significant importance for carbon yield since it determines the mechanism of energy generation, which is critical to efficient protein production.
+When considering carbon yield performance (Fig \ref{fig:SI_GFP}B), the substrate and oxygen uptake fluxes became much more important while the sensitivity to the transcription/translation parameters decreased slightly.
+The transcription/translation parameters had the same trend as for productivity, where the ribosome elongation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all parameters.
+Thus, in investigating CFPS performance in terms of productivity and yield, the ribosome elongation rate is an important parameter to optimize, as been already shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
+Glucose and oxygen uptake shown significant importance for carbon yield since it determined the mechanism of energy generation, which is critical for efficient protein production.
 Meanwhile, productivity is determined primarily by the rate of the most downstream steps, transcription and translation.
-Glucose is required for both amino acid synthesis and efficient energy generation, both of which are important for a high yield.
-There is a tradeoff between energy generation to power transcription/translation as well as synthesis of amino acids which are required to assemble the protein of interest.
-Across all three cases, substrate utilization (amino acid uptake and glucose) showed to be the next important as these substrates are contributing to the carbon yield of deGFP.
-Interestingly, in the first two cases, oxygen uptake has a significant effect on carbon yield.
+Glucose is required for both amino acid synthesis and efficient energy generation, both of which were important for a high yield.
+There is a tradeoff between energy generation to power transcription/translation as well as the synthesis of amino acids which were required to assemble the protein of interest.
+Across all three cases, substrate utilization (amino acid uptake and glucose) showed to be the next important as these substrates contributed to the carbon yield of deGFP.
+Interestingly, in the first two cases, oxygen uptake had also a significant effect on carbon yield.
 This is most likely due to the importance of oxidative phosphorylation for efficient energy generation.
-However, in the third case (with no amino acid uptake), oxygen uptake becomes insignificant compared to the other parameters, while glucose uptake was the most siginificant.
-Jewett and coworkers have reported that oxidative phosphorylation still operates in cell free and also saw a decrease in yield, ranging from 1.5-fold to 4-fold, when knocking out oxidative phosphorylation reactions\cite{Jewett:2008aa}.
+With high oxidative phosphorylation, carbon substrates were not wasted to meet the necessary energy requirements of CFPS, instead the carbon was directed toward the production of deGFP.
+However, in the third case (with no amino acid uptake), oxygen uptake significance decreased, while glucose uptake became the most siginificant.
+When amino acids were not supplied in CFPS, glucose is the only carbon available for amino acid synthesis and energy generation. 
+The same tradeoff was observed between utilizing glucose for amino acid synthesis and efficient energy generation; the significance of oxygen became irrelevant compared to the glucose uptake rate. 
+In addition, NADH was required to power oxidative phosphorylation, however there was a higher demand for NADPH for amino acid syntheis since amino acids were required for the production of the protein of interest.
+Therefore, NADH was interconverted to NADPH, which decreased the significance of the oxygen uptake rate since there was not enough NADH to power oxidative phosphorylation.
+Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free and also reported a decrease in yield, ranging from 1.5-fold to 4-fold, when oxidative phosphorylation reactions were knocked out\cite{Jewett:2008aa}.
 However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, thus to investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig \ref{fig:oxphos_yield}).
@@
 When the system can utilize glucose for energy generation, the increase in oxidative phosphorylation has a positive effect on carbon yield.
-However, when glucose is used to synthesize amino acids and power CFPS, there is a limit of how much oxidative phosphorylation can effect carbon yield.
+However, when glucose is used to synthesize amino acids and power CFPS, there is a limit of how much oxidative phosphorylation can effect carbon yield due to the limitation of NADH to power oxidative phosphorylation.
 Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and where the remaining energy is coming from.
 To investigate the differences in productivity and carbon yields, we compared the flux distributions for deGFP predicted by ssFBA simulations for the three different cases (Fig.\ref{fig:flux}).
-All cases used the first step in the pentose phosphate pathway to generate NADPH;
-the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
-The majority of the flux was distributed between lactate and acetate, representing an inefficiency in energy generation as well as wasteful byproducts.
-For the first two cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
-The second case with no amino acid synthesis showed to be the most efficient process (highest carbon yield).
-All the NADPH produced is interconverted to NADH to power oxidative phosphorylation.
-Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow
-However, diverting the flux away from acetate accumulation may allow more resources towards the protein of interest.
-Interestingly, for the first case and more for the third case, there is a mixture of aerobic and anearobic energy generation.
-Fumerate dehydrogenase typically operates in anaerobic conditions and is less efficient than oxidative phosphorylation.
-Cell-free systems do not have a sense of regulation as \textit{in vivo} systems thus any enzyme present in the mixture will carry out its corresponding reaction.
-In the third case (third row), there is a low flux through oxidative phosphorylation and high flux through fumerate dehydrogenase (fumerate to succinate).
-One of the reasons for this inefficiency is some of the NADH is interconverted to NADPH which is required for amino acid synthesis, thus the lack of NADH cannot power oxidative phosphorylation as effectively as in the first two cases.
-It is currently unknown how much oxidative phosphorylation is present in cell-free systems, however our model and the sensitivity analysis suggests that efficient energy generation is essential to high carbon yields.
-Thus, removing anaerobic enzymes during the cell-free extract preparation is a potential area for improvement for CFPS performance and protein yield.
-\begin{figure}[h!]
-\includegraphics[width=0.8\textwidth]{./Figures/flux.pdf}
+\begin{figure}[t!]
+\includegraphics[width=0.78\textwidth]{./Figures/flux.pdf}
 \caption{deGFP production flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
@@
 
+All cases used the first step in the pentose phosphate pathway to generate NADPH;
+the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
+The majority of the flux was distributed between lactate and acetate, representing an inefficiency in energy generation as well as wasteful byproducts.
+For the first two cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
+The second case with no amino acid synthesis showed to be the most efficient process (highest carbon yield).
+All the NADPH produced was interconverted to NADH to power oxidative phosphorylation.
+Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow
+However, diverting the flux away from acetate accumulation may allow more resources towards the protein of interest.
+Interestingly, for the first case and more for the third case, there was a mixture of aerobic and anearobic energy generation.
+Fumerate dehydrogenase typically operates in anaerobic conditions and was less efficient than oxidative phosphorylation.
+Cell-free systems do not have a sense of regulation as \textit{in vivo} systems, thus any enzyme present in the mixture will carry out its corresponding reaction.
+In the third case (third row), there was a low flux through oxidative phosphorylation and a high flux through fumerate dehydrogenase (fumerate to succinate).
+One of the reasons for this inefficiency was some of the NADH is interconverted to NADPH which is required for amino acid synthesis, thus the lack of NADH cannot power oxidative phosphorylation as effectively as in the first two cases.
+It is currently unknown how much oxidative phosphorylation is present in cell-free systems, however our model and the sensitivity analysis suggests that efficient energy generation is essential to higher carbon yields.
+Thus, removing anaerobic enzymes during the cell-free extract preparation is a potential area for improvement for CFPS performance and protein yield.
+
 All proteins followed a similar trend where for the first two cases (when amino acids were supplied) there was a relatively high yield than compared to the third case, with the exception of CAT.
-Thus, we also performed a sensitivity analysis on CAT (Fig \ref{fig:SI_CAT}), which did not show the same trend in yield across the three cases as the other proteins.
+Thus, we also performed a sensitivity analysis on CAT (Fig S1, available in the Supporting Information), which did not show the same trend in yield across the three cases as the other proteins.
 For CAT, glucose uptake was as important to productivity as the kinetic parameters.
@@
 The energy requirements for transcription of CAT is lower than the other proteins, thus it can accomodate its resources towards efficient energy generation.
-Thus, the carbon yield is not hindered in the third case, since the system maintains the necessary amino acid synthesis rates and the required energy species.
+Thus, the carbon yield was not hindered in the third case, since the system maintained the necessary amino acid synthesis rates and the required energy species.
 Apart from these differences, there was a general trend of transcription and translation holding more significance for productivity and substrate uptake holding more significance for yield.
@@
 Taken together, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
-We have shown sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate of protein production for deGFP and CAT under two different proteins.
+We have shown first principle predictions for protein production of deGFP and CAT under two different promoters and two different cell-free extract systems with no adjustable parameters were in agreement with experimental measurements.
 This modeling approach validated general trends seen in literature for productivity and yield as a function of carbon number.
-Further, global sensitivity analysis identified oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate is important for productivity.
-The model also identified that cell-free systems can operate with aerobic and anaerobic enzymes which leads to inefficient CFPS systems and should be addressed to optimize carbon yield as seen in the globol sensitivity analysis.
-Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
-
-\begin{figure}[h!]
-\includegraphics[width=1.00\textwidth]{./Figures/SI_CAT.pdf}
-\caption{Total order sensitivity of CAT productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).
-(Error bars represent 95\% CI of the ensemble).}
-\label{fig:SI_CAT}
-\end{figure}
+Further, global sensitivity analysis identified oxidative phosphorylation was instrumental for maintaining a high carbon yield and the translation rate was important for productivity.
+The model also identified that cell-free systems can operate with aerobic and anaerobic enzymes which lead to inefficient CFPS systems and should be addressed to optimize carbon yield as seen in the globol sensitivity analysis.
+In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
 
@@
 \subsection*{Calculation of the carbon yield.}
-The carbon yield ($Y_{C}^{CAT}$) was calculated as the ratio of carbon produced as the protein of interest divided by the carbon consumed as reactants (glucose and amino acids):
+The carbon yield ($Y_{C}^{POI}$) was calculated as the ratio of carbon produced as the protein of interest divided by the carbon consumed as reactants (glucose and amino acids):
 \begin{equation}\label{eqn:yield-definition}
@@
 \begin{itemize}
-  \item Filename: brief description
-  \item Filename: brief description
+  \item Protein Sequences: DNA and protein sequences of each protein of interest.
+  \item CAT Sensitivity: Global sensitivity analysis of CAT productivity and carbon yield.
 \end{itemize}

### 2017-01-05 Paper_v1.tex
Nicks Edits

@@
 Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
-We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins
+We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins.
 First principles predictions of circuit performance were consistent with measurements for a variety of cases.
 Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity and yield.
-A sufficient energy supply with oxidative phosphorylation is instrumental for high carbon yields, in addition the translation rate is the next step to optimize for higher productivity and yield.
+A sufficient energy supply with oxidative phosphorylation is instrumental for high carbon yields; in addition, the translation rate is the next step to optimize for higher productivity and yield.
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
@@
 Thus, sequence specific constraints based modeling explicitly coupled the transcription and translation of circuit components with the availability of metabolic resources.
-We tested this approach by first simulating the production of CAT and deGFP under two different promoters.
-and then expanded this for $\sigma_{70}$-induced deGFP expression at different plasmid concentrations.
+We tested this approach by first simulating the production of CAT and deGFP under two different promoters, and then expanded this for $\sigma_{70}$-induced deGFP expression at different plasmid concentrations.
 First principles predictions of circuit performance were consistent with measurements.
@@
 The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
-One mathematical framework that has found wide use in modeling metabolism are constraint based models such as flux balance analysis (FBA).
+One mathematical framework that has found wide use in modeling metabolism is constraint based models such as flux balance analysis (FBA).
 FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
-Traditionally, FBA is used to model \textit{in vivo} processes, however cell-free systems do not have growth associated reactions or transport through the cell membrane.
+Traditionally, FBA is used to model \textit{in vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
 Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
@@
 The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
-Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates or the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
+Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
 Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
@@
 Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases.
-Each of the proteins were produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
-In all our cases, CFPS is supplied with glucose.
-The first case, CFPS is supplied with amino acids and the system can synthesize amino acids (control).
-In the second case, CFPS is supplied with amino acids, however the system cannot synthesize amino acids (AA uptake w/o synthesis).
-We turned off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids, thus the enzymes responsible for amino acid synthesis would not be present.
-In the third case, CFPS is not supplied with amino acids, but the system can synthesize them (AA synthesis w/o uptakte).
-Eight different proteins,ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
+Each of the proteins was produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
+In all cases, CFPS was supplied with glucose.
+In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (control).
+In the second case, CFPS was supplied with amino acids, but the amino acid synthesis was turned off (AA uptake w/o synthesis).
+We shut off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
+In the third case, CFPS was not supplied with amino acids, but the system could synthesize them (AA synthesis w/o uptake).
+Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
 We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod_POI}A).
@@
 This shows the system had sufficient substrates to power the system and synthesize each protein.
-The third case had the lowest productivity for each protein, thus the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
+The third case had the lowest productivity for each protein; thus, the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
 The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.
-The qualitative trend of productivity for the three different cases was the same within each case, however some proteins had higher productivity than others.
-For instance, in the second case FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
+The qualitative trend of productivity for the three different cases was the same within each case; however, some proteins had higher productivity than others.
+For instance, in the second case, FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
 To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
-The proteins with the highest productivity had the lowest carbon number whereas proteins with low productivity had higher carbon numbers.
-This inverse qualitative trend was due to the fact that larger proteins require more amino acids and substrates to assemble them resulting in lower productivity.
+The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
+This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity.
 \begin{figure}[t!]
@@
 To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
-For the first case, the system is relied on a mixture of glucose and some amino acids for the productin of deGFP with a carbon yield of 20.5\%.
-Once amino acid synthesis is removed from the network (2nd case), each amino acid is utilized and the carbon yield increased to 22.2\%.
-Only the necessary amount of amino acids were used for the production of deGFP, thus it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
-If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case its 92.5\%.
-Finally for the third case (without amino acid supplemented), the carbon yield was reduced to 14.8\% and the system used about double the amount of glucose as the first two cases.
-In this case glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
-Interestingly, CAT carbon yield showed similar performance for all three cases, however it still had a lower yield than the first two cases.
-Each protein has different energy requirements for transcription and translation depending on its sequence.
-Thus, for the case of CAT, energy requirements for transcription or translation were low enough that it's carbon yield was not hindered in the third case as significantly as for the other proteins.
+For the first case, the system relied on a mixture of glucose and some amino acids for the production of deGFP with a carbon yield of 20.5\%.
+Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 22.2\%.
+Only the necessary amount of amino acids were used for the production of deGFP; thus, it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
+If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case it is 92.5\%.
+Finally for the third case (without amino acids supplemented), the carbon yield was reduced to 14.8\% and the system used about twice the amount of glucose as in the first two cases.
+In this case, glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
+Interestingly, CAT carbon yield showed similar performance for all three cases; however, it still had a lower yield than the first two cases.
+Each protein had different energy requirements for transcription and translation depending on its sequence.
+In the case of CAT, energy requirements for transcription or translation were low enough that its carbon yield was not hindered in the third case as significantly as for the other proteins.
 
@@
     \begin{tabular}{lccc} \toprule
-        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Produced (mM)} & Control & \parbox{2.5 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
+        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Produced (mM/hr)} & Control & \parbox{2.6 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
         deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
-        \parbox{3.2 cm}{Carbon \phantom{abcdefghi} Consumed (mM)} & \\ \midrule
+        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Consumed (mM/hr)} & \\ \midrule
         GLC & 37.3 & 33.7 & 66.9 \\
@@
 
-We next investigated the effect of the carbon number of each protein to the carbon yield (Fig \ref{fig:Yield_POI}B).
-The same inverse qualitative trend is observed as for productivity, however it was less significant.
+We next investigated the effect of the carbon number of each protein on the carbon yield (Fig \ref{fig:Yield_POI}B).
+The same inverse qualitative trend is observed as for productivity, but it was less significant.
 The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
-As the size of the protein to be expressed increases, the productivity and carbon yield decreases, thus proteins that are too large may not be feasible for cell-free production.
-Thus, we wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS systems most effectively.
+As the protein length increased, the productivity and carbon yield decreased, suggesting that large proteins may not be feasible for cell-free production.
+We thus wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS performance.
 %Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
@@
 To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
-In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters were fairly constant across all three cases, with the elongation rate by ribosomes being the most sensitive.
-This showed the rate of translation is instrumental in achieving high productivity, and should be the first step to investigate in order to optimize productivity, prior to examining transcription rates, this is consistent with literature findings.
+In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the elongation rate by ribosomes being the most significant.
+This showed that translation rate is instrumental for productivity, and should be the first step investigated in its optimization, prior to examining transcription rates.
+This is consistent with literature findings.
 Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
 In addition, Li et al.
-have increased productivity of firefly lucifease by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+have increased productivity of firefly luciferase by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
 In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
-This makes sense, as amino acid synthesis can only come from glucose and became the only way to power protein synthesis.
-On the other hand, when amino acid synthesis is removed, amino acid uptake became more important, as it becomes the only source of amino acids.
+This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
+On the other hand, when amino acid synthesis was removed, amino acid uptake became more important, as it was the only source of amino acids.
 \begin{figure}[t!]
@@
 
-When considering carbon yield performance (Fig \ref{fig:SI_GFP}B), the substrate and oxygen uptake fluxes became much more important while the sensitivity to the transcription/translation parameters decreased slightly.
-The transcription/translation parameters had the same trend as for productivity, where the ribosome elongation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all parameters.
-Thus, in investigating CFPS performance in terms of productivity and yield, the ribosome elongation rate is an important parameter to optimize, as been already shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
-Glucose and oxygen uptake shown significant importance for carbon yield since it determined the mechanism of energy generation, which is critical for efficient protein production.
-Meanwhile, productivity is determined primarily by the rate of the most downstream steps, transcription and translation.
+When considering carbon yield performance (Fig \ref{fig:SI_GFP}B), the substrate and oxygen uptake fluxes became much more important while the sensitivity to transcription/translation parameters decreased slightly.
+The transcription/translation parameters had the same trend as for productivity, where the ribosome elongation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
+Thus, in investigating CFPS performance in terms of productivity and yield, the ribosome elongation rate is an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
+Glucose and oxygen uptake showed significant importance for carbon yield since they determined the mechanism of energy generation, which is critical for efficient protein production.
+Meanwhile, productivity was determined primarily by the rate of the most downstream steps, transcription and translation.
 Glucose is required for both amino acid synthesis and efficient energy generation, both of which were important for a high yield.
-There is a tradeoff between energy generation to power transcription/translation as well as the synthesis of amino acids which were required to assemble the protein of interest.
+This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
 Across all three cases, substrate utilization (amino acid uptake and glucose) showed to be the next important as these substrates contributed to the carbon yield of deGFP.
-Interestingly, in the first two cases, oxygen uptake had also a significant effect on carbon yield.
+Interestingly, in the first two cases, oxygen uptake also had a significant effect on carbon yield.
 This is most likely due to the importance of oxidative phosphorylation for efficient energy generation.
-With high oxidative phosphorylation, carbon substrates were not wasted to meet the necessary energy requirements of CFPS, instead the carbon was directed toward the production of deGFP.
+With high oxidative phosphorylation, carbon substrates were not wasted to meet the necessary energy requirements of CFPS; instead, carbon was directed toward the production of deGFP.
 However, in the third case (with no amino acid uptake), oxygen uptake significance decreased, while glucose uptake became the most siginificant.
-When amino acids were not supplied in CFPS, glucose is the only carbon available for amino acid synthesis and energy generation. 
-The same tradeoff was observed between utilizing glucose for amino acid synthesis and efficient energy generation; the significance of oxygen became irrelevant compared to the glucose uptake rate. 
-In addition, NADH was required to power oxidative phosphorylation, however there was a higher demand for NADPH for amino acid syntheis since amino acids were required for the production of the protein of interest.
-Therefore, NADH was interconverted to NADPH, which decreased the significance of the oxygen uptake rate since there was not enough NADH to power oxidative phosphorylation.
-Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free and also reported a decrease in yield, ranging from 1.5-fold to 4-fold, when oxidative phosphorylation reactions were knocked out\cite{Jewett:2008aa}.
-However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems, thus to investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig \ref{fig:oxphos_yield}).
-Interestingly, oxidative phosphorylation has a somewhat linear correlation with carbon yield for the first two cases (R\textsuperscript{2}=0.77 for the control, R\textsuperscript{2}=0.79 for the second case), whereas for the third case, this same effect was not as prevalent (R\textsuperscript{2}=0.46).
+When amino acids were not supplied in CFPS, glucose was the only carbon source for amino acid synthesis and energy generation. 
+The same tradeoff was observed between utilizing glucose for amino acid synthesis and efficient energy generation; the significance of oxygen was irrelevant compared to that of glucose uptake. 
+In addition, NADH was required to power oxidative phosphorylation; however, the demand for NADPH for amino acid syntheis was higher, since amino acids were required for the production of the protein of interest.
+NADH was thus drawn away from oxidative phosphorylation to be interconverted to NADPH.
+This decreased the significance of oxygen uptake, since there was not enough NADH to power oxidative phosphorylation.
+Jewett and coworkers have reported oxidative phosphorylation still operating in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out\cite{Jewett:2008aa}.
+However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems.
+Thus, to investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig \ref{fig:oxphos_yield}).
+Interestingly, oxidative phosphorylation has a somewhat linear correlation with carbon yield for the first two cases (R\textsuperscript{2}=0.77 for the control, R\textsuperscript{2}=0.79 for the second case), whereas for the third case, this effect was not as prevalent (R\textsuperscript{2}=0.46).
 \begin{figure}[t!]
 \includegraphics[width=0.7\textwidth]{./Figures/OxPhos_yield.pdf}
-\caption{An ensemble of 1000 ssFBA solutions for deGFP carbon yield versus oxidative phosphorylation flux for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).
-}
-\label{fig:oxphos}
+\caption{An ensemble of 1000 ssFBA solutions for deGFP carbon yield versus oxidative phosphorylation flux for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).}
+\label{fig:oxphos_yield}
 \end{figure}
@@
 When the system can utilize glucose for energy generation, the increase in oxidative phosphorylation has a positive effect on carbon yield.
-However, when glucose is used to synthesize amino acids and power CFPS, there is a limit of how much oxidative phosphorylation can effect carbon yield due to the limitation of NADH to power oxidative phosphorylation.
+However, when glucose is used to synthesize amino acids and power CFPS, there is a limit on how much oxidative phosphorylation can effect carbon yield due to the limitation of NADH to power oxidative phosphorylation.
 Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and where the remaining energy is coming from.
@@
 \caption{deGFP production flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
-Flux (mean $\pm$ standard deviation) across ensemble normalized to glucose uptake flux.
+Flux across ensemble (mean $\pm$ standard deviation) normalized to glucose uptake flux.
 Flux distribution for three different cases: control (top row), amino acid uptake without synthesis (second row), and amino acid synthesis without uptake (third row).
@@
 the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
-The majority of the flux was distributed between lactate and acetate, representing an inefficiency in energy generation as well as wasteful byproducts.
+The majority of the flux was distributed between lactate and acetate, representing an inefficiency in energy generation as well as the accumulation of wasteful byproducts.
 For the first two cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
@@
 All the NADPH produced was interconverted to NADH to power oxidative phosphorylation.
-Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate overflow
-However, diverting the flux away from acetate accumulation may allow more resources towards the protein of interest.
-Interestingly, for the first case and more for the third case, there was a mixture of aerobic and anearobic energy generation.
+Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate.
+Diverting flux away from acetate accumulation may allow more resources towards the protein of interest.
+Interestingly, for the first case and especially for the third case, there was simultaneous aerobic and anaerobic energy generation, which is imposible for \textit{in vivo} systems.
+Cell-free systems do not have the same sense of regulation as \textit{in vivo} systems, so any enzyme present in the extract will catalyze its corresponding reaction.
+In the third case (third row), there was a low flux through oxidative phosphorylation and a high flux through fumerate dehydrogenase (fumerate to succinate).
 Fumerate dehydrogenase typically operates in anaerobic conditions and was less efficient than oxidative phosphorylation.
-Cell-free systems do not have a sense of regulation as \textit{in vivo} systems, thus any enzyme present in the mixture will carry out its corresponding reaction.
-In the third case (third row), there was a low flux through oxidative phosphorylation and a high flux through fumerate dehydrogenase (fumerate to succinate).
-One of the reasons for this inefficiency was some of the NADH is interconverted to NADPH which is required for amino acid synthesis, thus the lack of NADH cannot power oxidative phosphorylation as effectively as in the first two cases.
-It is currently unknown how much oxidative phosphorylation is present in cell-free systems, however our model and the sensitivity analysis suggests that efficient energy generation is essential to higher carbon yields.
-Thus, removing anaerobic enzymes during the cell-free extract preparation is a potential area for improvement for CFPS performance and protein yield.
+One of the reasons for this inefficiency was that some of the NADH was interconverted to NADPH, which is required for amino acid synthesis; thus, NADH was depleted and could not power oxidative phosphorylation as effectively as in the first two cases.
+It is currently unknown how much oxidative phosphorylation is present in cell-free systems, but our model and the sensitivity analysis suggest that efficient energy generation is essential to higher carbon yields.
+Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
 
-All proteins followed a similar trend where for the first two cases (when amino acids were supplied) there was a relatively high yield than compared to the third case, with the exception of CAT.
-Thus, we also performed a sensitivity analysis on CAT (Fig S1, available in the Supporting Information), which did not show the same trend in yield across the three cases as the other proteins.
+All proteins with the exception of CAT followed a similar trend, where for the first two cases (when amino acids were supplied) there was a relatively high yield compared to the third case.
+We also performed a sensitivity analysis on CAT (Fig S1, available in the Supporting Information), which did not show the same trend in yield across the three cases as the other proteins.
 For CAT, glucose uptake was as important to productivity as the kinetic parameters.
 The oxygen uptake was the most important factor in yield and maintained this across all three cases.
-The energy requirements for transcription of CAT is lower than the other proteins, thus it can accomodate its resources towards efficient energy generation.
+The energy requirements for transcription of CAT were lower than for the other proteins, so it could divert more resources towards efficient energy generation.
 Thus, the carbon yield was not hindered in the third case, since the system maintained the necessary amino acid synthesis rates and the required energy species.
-Apart from these differences, there was a general trend of transcription and translation holding more significance for productivity and substrate uptake holding more significance for yield.
+Apart from these differences, there was a general trend of transcription and translation being more significant for productivity and substrate uptake being more significant for yield.
 
-Taken together, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
-We have shown first principle predictions for protein production of deGFP and CAT under two different promoters and two different cell-free extract systems with no adjustable parameters were in agreement with experimental measurements.
-This modeling approach validated general trends seen in literature for productivity and yield as a function of carbon number.
-Further, global sensitivity analysis identified oxidative phosphorylation was instrumental for maintaining a high carbon yield and the translation rate was important for productivity.
-The model also identified that cell-free systems can operate with aerobic and anaerobic enzymes which lead to inefficient CFPS systems and should be addressed to optimize carbon yield as seen in the globol sensitivity analysis.
+Taken together, this represents sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} CFPS system.
+We have shown first principle predictions for protein production of deGFP and CAT, under two different promoters and two different cell-free extract systems, with no adjustable parameters and in agreement with experimental measurements.
+This modeling approach validated trends seen in literature for productivity and yield as a function of carbon number.
+Furthermore, global sensitivity analysis identified oxidative phosphorylation as being instrumental for maintaining a high carbon yield and translation rate as being important for productivity.
+The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize carbon yield.
 In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
@@
 $\rm R_{2}^{*}$ denotes bounded ribosome,
-and $AA_{j}$ denotes $j^{th}$ amino acid.
+and $AA_{j}$ denotes the $j^{th}$ amino acid.
 
@@
 \end{equation}
-where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
+where $G$ denotes the gene concentration and $K_{TX}$ denotes a transcription saturation coefficient.
 The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
@@
 $l_{G}$ denotes the gene length in nucleotides, and the last term $\mathcal{P}$ describes a model of promoter activity.
-In this study, we considered two promoters; the T7 and $\sigma_{70}$ promoters.
-The promoter function for the T7 promoter, $\mathcal{P}_{T7}$ was given by:
+In this study, we considered two promoters: T7 and $\sigma_{70}$.
+The promoter function for the T7 promoter, $\mathcal{P}_{T7}$, was given by:
 \begin{equation}
-	P_{T7} = (\frac{K_{T7}}{1 + K_{T7}}
+	P_{T7} = \frac{K_{T7}}{1 + K_{T7}}
 \end{equation}
 where $K_{T7}$ denotes a T7 RNA polymerase binding constant \cite{Moon:2012aa}.
-The $\sigma_{70}$ binding promoter was used for all other proteins was formulated as:
+The $\sigma_{70}$ binding promoter, used for all other proteins, was formulated as:
 \begin{equation}
@@
 where $K_{1}$ denotes the state of RNA polymerase binding,
-$K_{2}$ is the state of $\sigma_{70} $binding along with RNA polymerase,
-and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to the promoter (modeled as a Hill function).
+$K_{2}$ is the state of $\sigma_{70}$ binding along with RNA polymerase, and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to the promoter (modeled as a Hill function).
 
-On the other hand, the translation rate ($w_{TL}$) was bounded by:
+The translation rate ($w_{TL}$) was bounded by:
  \begin{equation}
@@
 \end{equation}
-where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance, and $K_{TL}$ denotes the translation saturation constant.
+where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance and $K_{TL}$ denotes the translation saturation constant.
 The maximum translation rate $V_{TL}^{max}$ was formulated as:
@@
 \begin{equation}
-	\rm mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
-\end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+	 mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+\end{equation}
+where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
 
 The objective of the sequence-specific flux balance calculation was to maximize the rate of protein translation, $w_{TL}$.
-The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
-while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
-The protein gene and protein sequences were taken from literature and are available in the Supporting Information.
+The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data, while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
+Gene and protein sequences were taken from literature and are available in the Supporting Information.
 The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
@@
 We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
-We computed the total sensitivity index of each parameter relative to two performance objectives, productivity of the protein of interest and the carbon yield.
+We computed the total sensitivity index of each parameter relative to two performance objectives: productivity of the protein of interest and carbon yield.
 We established the sampling bounds for each parameter from literature.
 We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
-where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 1,000 and $d$ = 7, so the total sensitivity indices were computed from
-16,000 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
+where $N$ was a parameter proportional to the desired number of model evaluations and $d$ was the number of parameters in the model.
+In our case, $N$ = 1000 and $d$ = 7, so the total sensitivity indices were computed from 16,000 model evaluations.
+The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.
 
@@
 \begin{suppinfo}
-
 The following files are available free of charge.
@@
 \end{itemize}
-
 \end{suppinfo}

### 2017-02-28 Paper_v1.tex
supportinfo

@@
 \author{Michael Vilkhovoy}
+\affiliation[Cornell University]
+{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
 \author{Nicholas Horvath}
+\affiliation[Cornell University]
+{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
+\author{Joseph Wayman}
+\affiliation[Cornell University]
+{Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
+\author{Kara Calhoun}
+\affiliation[Stanford University]
+{School of Chemical Engineering, Stanford University, Stanford, CA 94305}
+\author{James Swartz}
+\affiliation[Stanford University]
+{School of Chemical Engineering, Stanford University, Stanford, CA 94305}
 \author{Jeffrey D. Varner}
@@
 %\abbreviations{MV,NH,JDV}
-\keywords{Synthetic biology, Constraints based modeling, Biochemical modeling}
+\keywords{Synthetic biology, Constraints based modeling, cell-free protein synthesis}
 
@@
 First principles predictions of circuit performance were consistent with measurements for a variety of cases.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity and yield.
-A sufficient energy supply with oxidative phosphorylation is instrumental for high carbon yields; in addition, the translation rate is the next step to optimize for higher productivity and yield.
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity, energy efficiency, and carbon yield.
+A sufficient energy supply with oxidative phosphorylation is instrumental for high energy efficiency and carbon yields; in addition, the translation rate is the next step to optimize for higher productivity.
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
@@
 subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
-Stoichiometric models rely on a pseudo-steady-state assumption to reduce unidentifiable genome-scale kinetic models to an underdetermined linear algebraic system, which can be solved efficiently even for large systems using linear programming.
 Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
@@
 \subsection{Model Derivation}
-The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity and yield for a protein of interest.
+The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity, energy efficiency and carbon yield for a protein of interest.
 One mathematical framework that has found wide use in modeling metabolism is constraint based models such as flux balance analysis (FBA).
@@
 Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
-The network consisted of 281 reactions and 132 species.
+The network consisted of 272 reactions and 145 species (Fig \ref{fig:network}A).
 We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
 We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
-The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:CAT_GFP_prod}A).
+The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:network}B).
 The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E.
-coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:CAT_GFP_prod}B).
+coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:network}C).
 The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
@@
 Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
-\begin{figure}[t!]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/CAT_GFP_prod.pdf}
-\caption{Sequence specific flux balance analysis of protein production vesus time.
-A.
-CAT production under a T7 promoter in CFPS \textit{E.
-coli} extract for 1 hr under glucose consumption.
-B.
-deGFP production under a P70 promoter in TXTL 2.0 \textit{E.
-coli} extract for 8 hr under glucose consumption.
-95\% CI (blue region) over the ensemble of 100 sets.}
-\label{fig:CAT_GFP_prod}
-\end{figure}
-
-Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:GFP_plasmid}).
+Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:network}D).
 Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
@@
 \centering
-\includegraphics[width=0.6\textwidth]{./Figures/GFP_plasmid.pdf}
-\caption{Predicted deGFP concentration (black line) at different plasmid concentrations versus measurements of deGFP (dots) synthesized in TXTL 2.0.
-95\% CI (blue region) over the ensemble of 100 sets.}
-\label{fig:GFP_plasmid}
+\includegraphics[width=1.00\textwidth]{./Figures/ssFBA_network.pdf}
+\caption{Sequence specific flux balance analysis.
+A.
+Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway.
+Thick gray arrows indicate withdrawal of precursors for amino acid synthesis.
+B.
+CAT production under a T7 promoter in CFPS \textit{E.
+coli} extract for 1 hr under glucose consumption.
+C.
+deGFP production under a P70 promoter in TXTL 2.0 \textit{E.
+coli} extract for 8 hr under glucose consumption.
+D.
+Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0.
+95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
+\label{fig:network}
 \end{figure}
 
-These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters.
+These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with few adjustable parameters in the promoter model taken from literature \cite{Moon:2012aa}.
 It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under two different promoters and cell-free systems.
@@
 
-\subsection{CFPS Productivity}
-Our next goal was to examine the performance of CFPS productivity for eight different proteins under three different cases.
+Our next goal was to examine the performance of CFPS for eight different proteins under three different cases. 
 Each of the proteins was produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
 In all cases, CFPS was supplied with glucose.
-In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (control).
-In the second case, CFPS was supplied with amino acids, but the amino acid synthesis was turned off (AA uptake w/o synthesis).
-We shut off these synthesis reactions since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
+In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (AA uptake and synthesis).
+In the second case, CFPS was supplied with amino acids, but the amino acid synthesis were turned off (AA uptake w/o synthesis).
+These amino acid synthesis reactions were blocked since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
 In the third case, CFPS was not supplied with amino acids, but the system could synthesize them (AA synthesis w/o uptake).  
 Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
-We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod_POI}A).
-The first two cases had very similar performance, with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
-%The second case (without amino acid synthesis) showed the highest productivity for each of the proteins, however the control case had very similar performance.
-This shows the system had sufficient substrates to power the system and synthesize each protein.
-The third case had the lowest productivity for each protein; thus, the addition of amino acids to CFPS extract is important for maintaining a relatively high productivity.
-The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.
-The qualitative trend of productivity for the three different cases was the same within each case; however, some proteins had higher productivity than others.
-For instance, in the second case, FGF21 (fibroblast growth factor 21) had a productivity of 17 ($\mu$M/h) whereas FII (prothrombin) had a productivity of 3 ($\mu$M/h).
-To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod_POI}B).
+We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod}).
+An additional case was considered for CAT, since a comprehensive dataset is available \cite{2005_calhoun_BiotechnologyProgress}; thus fluxes were constrained to experimental measurements where avaliable, except for CAT production which was determined by the transcription/translation parameters.
+
+\subsection{CFPS Productivity}
+We evaluated CFPS productivity for all eight proteins and for each case. 
+All cases had very similar performance for each protein and were in the standard deviation of each other (Fig \ref{fig:Prod}A). 
+%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins. 
+The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.  
+This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
+However, each protein had a different level of productivity.
+For instance, BMP10 had a productivity of about 2.5 ($\mu$M/h) whereas CAT had a productivity of about 12 ($\mu$M/h).
+To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod}B).
 The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
 This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity.
+A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information). 
+CAT was an outlier for the trendline, even though it was in the same order of magnitude of the trendline.
+The difference in the T7 and  $\sigma_{70}$ promoters did not alter the qualitative productivity performance of CAT. 
+The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation, however this should be investigated further.
+% a further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated. 
 \begin{figure}[t!]
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/Prod_POI.pdf}
-\caption{CFPS productivity performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey).
+\includegraphics[width=1.00\textwidth]{./Figures/Productivity.pdf}
+\caption{CFPS productivity of eight proteins for four cases.
+Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white).
 A.
-Productivity normalized to the specific glucose uptake rate (Error bars represent the 95\% CI of the ensemble).
+Productivity (Error bars represent the 95\% CI of the ensemble).
 B.
-Mean productivity versus the carbon number for each corresponding protein.}
-\label{fig:Prod_POI}
+Mean productivity versus the carbon number for each corresponding protein.
+Single trendline of productivity versus carbon number (dotted line) for all cases (R\textsuperscript{2} = 0.99).
+Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
+\label{fig:Prod}
+\end{figure}
+
+The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.
+
+\subsection{CFPS Energy Efficiency}
+Following the same outline as in examining the productivity, we calculated the energy efficiency for each protein (Fig \ref{fig:Energy}A).
+The first two cases had comparable performance with having the highest energy efficiency, where amino acids were supplied in the media.
+The third case (with no amino acid uptake) had the lowest energy efficiency; this is because glucose is utilized to synthesize the necessary amino acids for each protein as well as energize the system.
+We next investigated the effect of the carbon number of each protein on the energy efficiency (Fig \ref{fig:Energy}B).
+The same inverse qualitative trend is observed as for productivity, but followed a linear trendline.
+The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for the first two cases.
+Proteins with a high carbon number have a higher transcription and translation cost then smaller proteins, thus it makes sense for larger proteins to have a lower energy efficiency.  
+The first two cases had the same trendline whereas for the third case there was a significant drop in energy efficiency which required its own trendline (available in the Supporting Information). 
+Interestingly, in the third case each protein had a similar energy efficiency of about 25\% regardless of the carbon number of the protien. 
+In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein of interest kept the energy efficiency saturated at a relatively low level.
+However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.9 $\pm$ 1.5\% compared to the theoretical maximum of 72 $\pm$ 7.1\%. 
+This shows CFPS systems have a lot of room for improvement: first, the experimental setup was still producing certain amino acids which could be turned off. 
+Second, the system has a high accumulation of metabolic byproducts of organic acids which is a result of inefficient energy utilization.
+
+%This may be a potential area for improvement in CFPS to optimize for energy utilization. 
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
+\caption{CFPS energy efficiency of eight proteins for four cases.
+Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white).
+A.
+Energy efficiency (Error bars represent the 95\% CI of the ensemble).
+B.
+Mean energy efficiency versus the carbon number for each corresponding protein.
+Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.97) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.79).
+Dagger: constrained by experimental measurements and excluded from trendline.}
+\label{fig:Energy}
 \end{figure}
@@
 \subsection{CFPS Carbon Yield}
-Following the same outline as in examining the productivity, we calculated the carbon yield for each protein (Fig \ref{fig:Yield_POI}A).
-The same trends followed, where the case without amino acid synthesis showed the highest carbon yield and the control case had comparable performance.
+We also calculated the carbon yield for each protein (Fig \ref{fig:Yield}A).
+The same trends followed, where the cases with amino acids supplied in the media showed the highest carbon yield.
 The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system. 
-To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
-For the first case, the system relied on a mixture of glucose and some amino acids for the production of deGFP with a carbon yield of 20.5\%.
-Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 22.2\%.
-Only the necessary amount of amino acids were used for the production of deGFP; thus, it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
-If that is the case, the carbon yield without glucose contribution is 100\%, whereas for the first case it is 92.5\%.
-Finally for the third case (without amino acids supplemented), the carbon yield was reduced to 14.8\% and the system used about twice the amount of glucose as in the first two cases.
-In this case, glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins except for CAT.
-Interestingly, CAT carbon yield showed similar performance for all three cases; however, it still had a lower yield than the first two cases.
-Each protein had different energy requirements for transcription and translation depending on its sequence.
-In the case of CAT, energy requirements for transcription or translation were low enough that its carbon yield was not hindered in the third case as significantly as for the other proteins.
-
-\begin{table}[t!]
-\centering
-    \caption{Carbon contribution from glucose and each amino acid toward deGFP production for three cases: control, amino acid uptake without synthesis, and amino acid synthesis without uptake.}
-    \renewcommand{\arraystretch}{0.9}
-    \begin{tabular}{lccc} \toprule
-        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Produced (mM/hr)} & Control & \parbox{2.6 cm}{AA uptake \phantom{ai} w/o synthesis} & \parbox{2.5 cm}{AA synthesis w/o uptake} \\ \midrule
-        deGFP & 9.8 & 9.6 & 9.9  \\ \toprule
-        \parbox{3.3 cm}{Carbon \phantom{abcdefghi} Consumed (mM/hr)} & \\ \midrule
-        GLC & 37.3 & 33.7 & 66.9 \\
-        ALA & 0.0 & 0.2 & - \\
-        ARG & 0.3 & 0.3 & - \\
-        ASN & 0.5 & 0.4 & - \\
-        ASP & 0.4 & 0.6 & - \\
-        CYS & 0.2 & 0.1 & - \\
-        GLU & 2.1 & 0.6 & - \\
-        GLN & 0.4 & 0.3 & - \\
-        GLY & 0.3 & 0.3 & - \\
-        HIS & 0.5 & 0.5 & - \\
-        ILE & 0.0 & 0.6 & - \\
-        LEU & 1.0 & 1.0 & - \\
-        LYS & 1.0 & 0.9 & - \\
-		MET & 0.0 & 0.2 & - \\
-        PHE & 1.0 & 0.9 & - \\
-        PRO & 0.4 & 0.4 & - \\
-        SER & 0.0 & 0.2 & - \\
-        THR & 1.0 & 0.5 & - \\
-        TRP & 0.1 & 0.1 & - \\
-        TYR & 0.8 & 0.8 & - \\
-        VAL & 0.6 & 0.7 & - \\ \hline
-        Sum & 47.9 & 43.3 & 66.9 \\ \toprule
-        Yield & 20.5\% & 22.2\% & 14.8\% \\
-	Yield w/o GLC & 92.5\% & 100\% & - \\ \bottomrule
-    \end{tabular}
-\label{tbl:yield_breakdown}
-\end{table}
-
-We next investigated the effect of the carbon number of each protein on the carbon yield (Fig \ref{fig:Yield_POI}B).
-The same inverse qualitative trend is observed as for productivity, but it was less significant.
-The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
-As the protein length increased, the productivity and carbon yield decreased, suggesting that large proteins may not be feasible for cell-free production.
-We thus wanted to examine the most sensitive parameters for cell-free productivity and carbon yield in order to optimize CFPS performance.
-%Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
-%In addition, ssFBA is solved by setting the production of the protein as the objective function.
-%Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
+%To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
+For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 56.9 $\pm$ 2.4\% for CAT.
+Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 57.0 $\pm$ 2.2\% for CAT.
+Only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
+If that is the case, the carbon yield without glucose contribution increases to 100\%.
+Finally for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\%  for CAT and the system used about twice the amount of glucose as in the first two cases.
+In this case, glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins.
+The experimentally constrained case for CAT had carbon yield of about 6\% compared to the theoretical maximum of 57\%.
+This loss in carbon yield shows inefficiencies in CFPS that can be potentially improved.
+Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
+In addition, ssFBA is solved by setting the production of the protein as the objective function.
+Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
+In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate.
+The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.    
 \begin{figure}[t!]
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/Yield_POI.pdf}
-\caption{CFPS carbon yield performace for eight proteins for the control (black), AA uptake w/o synthesis (off white), and AA synthesis w/o uptake (grey).
+\includegraphics[width=1.00\textwidth]{./Figures/CarbonYield.pdf}
+\caption{CFPS carbon yield of eight proteins for four cases.
+Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white).
 A.
-Carbon yield (Error bars represent the 95\% CI of the ensemble).
+Energy efficiency (Error bars represent the 95\% CI of the ensemble).
 B.
-Mean carbon yield versus the carbon number for each corresponding protein.}
-\label{fig:Yield_POI}
+Mean energy efficiency versus the carbon number for each corresponding protein.
+Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.92) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.83).
+Dagger: constrained by experimental measurements and excluded from trendline.}
+\label{fig:Yield}
 \end{figure}
+We next investigated the effect of the carbon number of each protein on the carbon yield (Fig \ref{fig:Yield}B).
+The same inverse qualitative trend is observed as for productivity.
+The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case. 
+A single trendline was formulated for the first two cases since they had similar performance and another trendline was formulated for the third cases (available in Supporting Information)
+As the protein size increased, the carbon yield decreased, suggesting that large proteins may not be feasible for cell-free production.
+Thus, we wanted to examine the most parameters that have a significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.
 
 \subsection{CFPS performance sensitivity analysis}
-To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and carbon yield for deGFP, a representative protein (Fig \ref{fig:SI_GFP}).
-In examining productivity performance (Fig \ref{fig:SI_GFP}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the elongation rate by ribosomes being the most significant.
-This showed that translation rate is instrumental for productivity, and should be the first step investigated in its optimization, prior to examining transcription rates.
+To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for deGFP, a representative protein (Fig \ref{fig:SI}), as well as on the carbon yield (available in Supporting Information).
+In examining productivity performance (Fig \ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the translation rate by ribosomes being the most significant.
+As expected, this showed that the translation rate is instrumental for productivity, and should be the first step investigated in its optimization, prior to examining transcription parameters.
 This is consistent with literature findings.
-Underwood and coworkers have also shown that an increase in ribosome levels does not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
+Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
 In addition, Li et al.
@@
 This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
-On the other hand, when amino acid synthesis was removed, amino acid uptake became more important, as it was the only source of amino acids.
+In addition, oxygen uptake had more significance in the third case than the first two cases since this case has to synthesize amino acids and power the system. 
+Also, oxygen determines how effectively glucose is utilized, therefore if the glucose uptake rate effects productivity then oxygen should have a similar effect.
+On the other hand, for the first two cases, amino acid uptake showed significance and was even higher for the second case (without AA synthesis) as it was the only source of amino acids.
 \begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./Figures/SI_GFP.pdf}
-\caption{Total order sensitivity of deGFP productivity (A) and carbon yield (B) to specific uptake rates and transcription/translation parameters for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).
-(Error bars represent 95\% CI of the ensemble).}
-\label{fig:SI_GFP}
+\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
+\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).
+Error bars represent 95\% CI of the sensitivity index.}
+\label{fig:SI}
 \end{figure}
 
-When considering carbon yield performance (Fig \ref{fig:SI_GFP}B), the substrate and oxygen uptake fluxes became much more important while the sensitivity to transcription/translation parameters decreased slightly.
-The transcription/translation parameters had the same trend as for productivity, where the ribosome elongation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
-Thus, in investigating CFPS performance in terms of productivity and yield, the ribosome elongation rate is an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
-Glucose and oxygen uptake showed significant importance for carbon yield since they determined the mechanism of energy generation, which is critical for efficient protein production.
+When considering energy efficiency performance (Fig \ref{fig:SI}B), the oxygen uptake rates were the most important for all three cases while the sensitivity to transcription/translation parameters decreased slightly.
+The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
+Thus, in investigating CFPS performance in terms of productivity and energy efficiency, the translation rate is an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
+Oxygen uptake showed significant importance for energy efficiency since it is responsible for oxidative phosphorylation which is the most efficieny pathway for energy generation.
 Meanwhile, productivity was determined primarily by the rate of the most downstream steps, transcription and translation.
-Glucose is required for both amino acid synthesis and efficient energy generation, both of which were important for a high yield.
-This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
 Across all three cases, substrate utilization (amino acid uptake and glucose) showed to be the next important as these substrates contributed to the carbon yield of deGFP.
-Interestingly, in the first two cases, oxygen uptake also had a significant effect on carbon yield.
-This is most likely due to the importance of oxidative phosphorylation for efficient energy generation.
-With high oxidative phosphorylation, carbon substrates were not wasted to meet the necessary energy requirements of CFPS; instead, carbon was directed toward the production of deGFP.
-However, in the third case (with no amino acid uptake), oxygen uptake significance decreased, while glucose uptake became the most siginificant.
-When amino acids were not supplied in CFPS, glucose was the only carbon source for amino acid synthesis and energy generation. 
-The same tradeoff was observed between utilizing glucose for amino acid synthesis and efficient energy generation; the significance of oxygen was irrelevant compared to that of glucose uptake. 
-In addition, NADH was required to power oxidative phosphorylation; however, the demand for NADPH for amino acid syntheis was higher, since amino acids were required for the production of the protein of interest.
-NADH was thus drawn away from oxidative phosphorylation to be interconverted to NADPH.
-This decreased the significance of oxygen uptake, since there was not enough NADH to power oxidative phosphorylation.
-Jewett and coworkers have reported oxidative phosphorylation still operating in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out\cite{Jewett:2008aa}.
-However, it is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems.
+In the first two cases where amino acids are supplied, there was a significance on amino acid uptake since without it energy would be wasted to synthesize amino acids and for the second case amino acid would not be possible.
+Glucose uptake only showed significance in the third case since this is the only source of carbon for protein synthesis and energy generation.
+%This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
+Jewett and coworkers have reported oxidative phosphorylation was still operating in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in a pyruvate powered CFPS \cite{Jewett:2008aa}.
+The carbon yield sensitivity analysis followed the same trends and is very similar to the energy efficiency sensitivity analysis. 
+It is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems.
 Thus, to investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig \ref{fig:oxphos_yield}).
-Interestingly, oxidative phosphorylation has a somewhat linear correlation with carbon yield for the first two cases (R\textsuperscript{2}=0.77 for the control, R\textsuperscript{2}=0.79 for the second case), whereas for the third case, this effect was not as prevalent (R\textsuperscript{2}=0.46).
+Interestingly, oxidative phosphorylation has a strong effect on the carbon yield for all cases. 
+The first two cases follow the same trend, ranging from a carbon yield of 20\% to 55\% depending on the oxidative phosphorylation activity.
+The third case, followed the same trend, however it's carbon yield ranged from about 10\% to 55\%.  
+The third case was expected to have a lower carbon yield for the same oxidative phosphorylation flux compared to the first two cases, since carbon must be utilized for energy generation and amino acid synthesis.
+In all cases where the carbon yield is below its theoretical maximum, there is an accumulation of acetate and lactate, resulting in the lower carbon yield. 
+In examining the experimental dataset, there is a mixture of acetate and lactate accumulation during CAT synthesis which shows CFPS is not operating at a full aerobic state.
+It is unclear how active oxidative phosphorylation is in CFPS since the reactions rely on electron transport from membrane vesicles . 
+The addition of phosphate showed an increase in CAT yield, however it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
+Interestingly, the addition of NADH did not increase protein synthesis since the ATP concentrations were most likely saturated. 
+Thus, an alternative strategy may be to inhibit anaerobic processes in cell-free in order to minimize unwanted byproducts such as acetate and lactate.
+%Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and how much from anaerobic processes.
 \begin{figure}[t!]
-\includegraphics[width=0.7\textwidth]{./Figures/OxPhos_yield.pdf}
-\caption{An ensemble of 1000 ssFBA solutions for deGFP carbon yield versus oxidative phosphorylation flux for three cases: control (black), amino acid uptake without synthesis (off white), and amino acid synthesis without uptake (gray).}
+\includegraphics[width=0.7\textwidth]{./Figures/Yield_Ox.pdf}
+\caption{An ensemble of 1000 ssFBA solutions for deGFP carbon yield versus oxidative phosphorylation flux for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
 \label{fig:oxphos_yield}
@@
 
-When the system can utilize glucose for energy generation, the increase in oxidative phosphorylation has a positive effect on carbon yield.
-However, when glucose is used to synthesize amino acids and power CFPS, there is a limit on how much oxidative phosphorylation can effect carbon yield due to the limitation of NADH to power oxidative phosphorylation.
-Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and where the remaining energy is coming from.
-To investigate the differences in productivity and carbon yields, we compared the flux distributions for deGFP predicted by ssFBA simulations for the three different cases (Fig.\ref{fig:flux}).
+To investigate the differences for each case, we compared the flux distributions for a representative protein, deGFP, predicted by ssFBA simulations and for the case constrained by experimental measurements for CAT (Fig.\ref{fig:flux}).
 \begin{figure}[t!]
-\includegraphics[width=0.78\textwidth]{./Figures/flux.pdf}
-\caption{deGFP production flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, NADPH/NADH transfer, and oxidative phosphorylation.
-Flux across ensemble (mean $\pm$ standard deviation) normalized to glucose uptake flux.
-Flux distribution for three different cases: control (top row), amino acid uptake without synthesis (second row), and amino acid synthesis without uptake (third row).
-Bold metabolites are allowed to accumulate; grey arrows lead to amino acid synthesis.}
+\includegraphics[width=1.00\textwidth]{./Figures/Flux_all.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation.
+Flux across ensemble normalized to glucose uptake flux.
+Flux distribution for four different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, (C) amino acid synthesis without uptake, and (D) CAT flux distribution constrained by experimental measurements.
+Thick arrows indicate flux to or from amino acids.}
 \label{fig:flux}
 \end{figure}
+In the first case, which was supplied with amino acids and could synthesize them in the network, relied on a combination of glucose and amino acids to power the system (Fig \ref{fig:flux}A).
+Glucose traveled through glycolysis and the Entner-Doudoroff pathway to generate NADH for oxidative phosphorylation as well as utilize pyruvate for amino acid biosynthesis. 
+Interestingly, amino acids powered the TCA cycle instead of glucose via alpha-ketoglutarate and fumerate and utilizing oxaloacetic acid for additional amino acid biosynthesis.
+Thus, ssFBA found the combinational use of glucose and amino acids to energize CFPS as the optimum case for protein synthesis. 
+In the second case, where amino acid synthesis was removed from the network, glucose was solely utilized to provide the necessary energy requirements via glycolysis which generated enough NADH for oxidative phosphorylation.
+Ubiquinone was generated via \textit{nuo} to power oxidative phosphorylation, instead of relying on the TCA cycle. 
+Once the required energy requirements were met for transcription and translation, the amino acids necessary for the assembly of the protein of interest were utilized from the media.
+The first two cases had similar performance in productivity, energy efficiency and carbon yield for all proteins, this is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
+In the third case, where amino acids must be synthesized, there is drop in performance (productivity, energy efficiency, and carbon yield) compared to the first two cases. 
+Glucose undergoes through all the major pathways: glycolysis, pentose phosphate pathway and the tca cycle. 
+First, glucose must be converted into all the necessary amino acids for the assembly of the protein of interest. 
+Second, it must provide the necessary energy requirements for transcription and translation. 
+There is a relatively high flux between the interconversion of NADH to NADPH.
+NADPH is required for the biosynthesis of amino acids, however NADH is required for oxidative phosphorylation, thus requiring glucose to energize CFPS as well as synthesize amino acids results in a decrease in performance.
+Our model and the sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
+Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
+The fourth case, constrained by experimental measurements, had a fairly similar flux distribution as the third case. 
+Even though the experimental setup was supplied with amino acids, glucose travels through all the major pathways and metabolic precursors are utilized for amino acid biosynthesis. 
+Only certain amino acid synthesis reactions were blocked since during the cell-free extract preparation all amino acids were not supplied.  
+It can be seen, there is an accumulation of pyruvate, lactate, acetate, and other organic acids which is an inefficiency of carbon utilization.
+In addition, there is a high flux via the Entner-Douodoroff Pathway which may be an artifact of the optimum solution of ssFBA.
+In order to address the feasibility of which reactions are occuring in CFPS, adding thermodynamic constraints on reactions may result in a better depiction of the intracellular flux distribution \cite{Henry:2007,Hamilton:2013}.
+In this case, it is unclear which carbon (glucose or amino acids) is used to power CFPS and which carbon is used for protein synthesis and may in fact be a combination of glucose and amino acids.
+Thus, it would be interesting to track the carbon with C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA with C\textsuperscript{13} measurements, which has been shown to have a good representation of the flux distribution for \textit{in vivo} process \cite{Zamboni:2009}.
 
-All cases used the first step in the pentose phosphate pathway to generate NADPH;
-the carbon flux then continued through the Entner–Doudoroff pathway toward pyruvate.
-The majority of the flux was distributed between lactate and acetate, representing an inefficiency in energy generation as well as the accumulation of wasteful byproducts.
-For the first two cases, the energy source was primarily oxidative phosphorylation, and to a lesser extent the TCA cycle.
-The second case with no amino acid synthesis showed to be the most efficient process (highest carbon yield).
-All the NADPH produced was interconverted to NADH to power oxidative phosphorylation.
-Oxidative phosphorylation lead to a high redox ratio contributing to the accumulation of acetate.
-Diverting flux away from acetate accumulation may allow more resources towards the protein of interest.
-Interestingly, for the first case and especially for the third case, there was simultaneous aerobic and anaerobic energy generation, which is imposible for \textit{in vivo} systems.
-Cell-free systems do not have the same sense of regulation as \textit{in vivo} systems, so any enzyme present in the extract will catalyze its corresponding reaction.
-In the third case (third row), there was a low flux through oxidative phosphorylation and a high flux through fumerate dehydrogenase (fumerate to succinate).
-Fumerate dehydrogenase typically operates in anaerobic conditions and was less efficient than oxidative phosphorylation.
-One of the reasons for this inefficiency was that some of the NADH was interconverted to NADPH, which is required for amino acid synthesis; thus, NADH was depleted and could not power oxidative phosphorylation as effectively as in the first two cases.
-It is currently unknown how much oxidative phosphorylation is present in cell-free systems, but our model and the sensitivity analysis suggest that efficient energy generation is essential to higher carbon yields.
-Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
-
-All proteins with the exception of CAT followed a similar trend, where for the first two cases (when amino acids were supplied) there was a relatively high yield compared to the third case.
-We also performed a sensitivity analysis on CAT (Fig S1, available in the Supporting Information), which did not show the same trend in yield across the three cases as the other proteins.
-For CAT, glucose uptake was as important to productivity as the kinetic parameters.
-The oxygen uptake was the most important factor in yield and maintained this across all three cases.
-The energy requirements for transcription of CAT were lower than for the other proteins, so it could divert more resources towards efficient energy generation.
-Thus, the carbon yield was not hindered in the third case, since the system maintained the necessary amino acid synthesis rates and the required energy species.
-Apart from these differences, there was a general trend of transcription and translation being more significant for productivity and substrate uptake being more significant for yield.
-
-Taken together, this represents sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} CFPS system.
-We have shown first principle predictions for protein production of deGFP and CAT, under two different promoters and two different cell-free extract systems, with no adjustable parameters and in agreement with experimental measurements.
-This modeling approach validated trends seen in literature for productivity and yield as a function of carbon number.
-Furthermore, global sensitivity analysis identified oxidative phosphorylation as being instrumental for maintaining a high carbon yield and translation rate as being important for productivity.
-The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize carbon yield.
+Taken together, we developed a sequence specific constraints based modeling approach to evaluate the performance of synthetic circuits in an \emph{E.~coli} CFPS system for a range of different proteins and three different cases.
+We have shown first principle predictions for protein production of deGFP and CAT in agreement with experimental measurements, under two different promoters and two different cell-free extract systems, with few adjustable parameters in the promoter models taken from literature.
+This modeling approach suggested trends for productivity, energy efficiency and carbon yield as a function of carbon number.
+Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
+The translation rate was identified as the rate limiting step for productivity.
+The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficinecy and carbon yield.
 In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
@@
 Gene and protein sequences were taken from literature and are available in the Supporting Information.
-The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
+The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.55 \cite{GLPK}.
 
-% The flux balance analysis problem was formulated as:
-% \begin{equation}\nonumber
-%  \begin{multlined}
-% 	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{obj} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
-% 	\mathrm{Subject \; to:}
-% 	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-% \alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
-%  \end{multlined}
-% \end{equation}
-% where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
-% and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-% The flux balance analysis problem was solved using the GNU Linear Programming Kit (v4.52) \cite{GLPK}.
-%
-% \subsection*{Sequence-specific flux balance analysis}
-% We estimated the theoretical maximum carbon yield using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
-% The sequence-specific flux balance analysis problem was formulated as a linear program:
-% \begin{equation}
-%  \begin{multlined}
-% 	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
-% 	\mathrm{Subject \; to:}
-% 	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-% \alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
-%  \end{multlined}
-% \end{equation}
-% where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
-% and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-% The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
-% The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
-% \begin{eqnarray*}
-% \mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
-% \mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
-% \mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
-% \mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
-% \mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
-% && \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
-% \mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
-% && \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
-% \end{eqnarray*}
-% where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
-% $\rm R_{1}$ denotes the concentration of RNA polymerase,
-% $G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
-% $\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
-% $\rm Pi$ denotes inorganic phosphate,
-% $\rm R_{2}$ denotes the ribosome concentration,
-% $\rm R_{2}^{*}$ denotes bounded ribosome,
-% and $AA_{j}$ denotes $j^{th}$ amino acid.
-%
-% The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
-% \begin{equation}
-% 	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
-% \end{equation}
-% where $G$ denotes the gene concentration, and $K_{TX}$ denotes a transcription saturation coefficient.
-% The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
-% \begin{equation}
-% 	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)P\right]
-% \end{equation}
-% The term $R_{1}$ denotes the RNA polymerase abundance,
-% $v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
-% $l_{G}$ denotes the gene length in nucleotides, and $P$ describes the promoter activity.
-%  The T7 binding promoter was used for CAT production was formulated as:
-%
-%
-% On the other hand, the translation rate ($w_{TL}$) was bounded by:
-%  \begin{equation}
-% 	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{\rm mRNA_{SS}}{K_{TL}+\rm mRNA_{SS}}\right)
-% \end{equation}
-% where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance, and $K_{TL}$ denotes the translation saturation constant.
-% The maximum translation rate $V_{TL}^{max}$ was formulated as:
-% \begin{equation}
-% 	V_{TL}^{max} \equiv \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]
-% \end{equation}
-% The term $K_{P}$ denotes the polysome amplification constant,
-% $v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
-% $l_{P}$ denotes the number of amino acids in the protein of interest,
-% and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
-% \begin{equation}
-% 	\rm mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
-% \end{equation}where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
-%
-% The objective of the sequence-specific flux balance calculation was to maximize the rate of CAT translation, $w_{TL}$.
-% The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data;
-% while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
-% The protein gene and protein sequences were taken from literature and are available in the Supporting Information.
-% The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.52 \cite{GLPK}.
+\subsection*{Calculation of energy efficiency.}
+Energy efficiency was calculated as the ratio of protein production to glucose consumption, both in terms of equivalent ATP molecules:
+\begin{equation}\label{eqn:energy-efficiency-definition}
+	Efficiency=\frac{\Delta\texttt{POI}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot\texttt{ATP}_{GLC}}
+\end{equation}
+where $\Delta\texttt{POI}$ denotes the flux of the protein of interest produced, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\Delta\texttt{GLC}$ denotes the glucose flux, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
+The energy species stoichiometric coefficients are available in the Supporting Information.
 
@@
 \end{equation}
-where $\Delta\texttt{POI}$ denotes the abundance of the protein of interest produced, $C_{POI}$ denotes carbon number of the protein of interest, $\mathcal{R}$ denotes the number of reactants,
+where $\Delta\texttt{POI}$ denotes the flux of the protein of interest produced, $C_{POI}$ denotes carbon number of the protein of interest, $\mathcal{R}$ denotes the number of reactants,
 $\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
 
-\subsubsection*{Quantification of uncertainty.}
+\subsection*{Quantification of uncertainty.}
 An ensemble of 100 sets of flux distributions was calculated for three different cases: control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
@@
 We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
-We computed the total sensitivity index of each parameter relative to two performance objectives: productivity of the protein of interest and carbon yield.
+We computed the total sensitivity index of each parameter relative to three performance objectives: productivity of the protein of interest, energy efficiency and carbon yield.
 We established the sampling bounds for each parameter from literature.
@@
   \item Protein Sequences: DNA and protein sequences of each protein of interest.
-  \item CAT Sensitivity: Global sensitivity analysis of CAT productivity and carbon yield.
+  \item Supporting Information: Performance trendlines as a function of carbon number and transcription/translation stoichiometric coefficients of energy species.
+  \item Carbon Yield Sensitivity Analysis: Global sensitivity analysis on deGFP carbon yield.
 \end{itemize}

### 2017-03-03 Paper_v1.tex
updated figs

@@
 %\abbreviations{MV,NH,JDV}
-\keywords{Synthetic biology, Constraints based modeling, cell-free protein synthesis}
+\keywords{Synthetic biology, constraints based modeling, cell-free protein synthesis}
 
@@
 Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
-We tested this approach by first simulating T7 induced CAT production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins.
+We tested this approach by first simulating T7 induced chloramphenicol acetyltransferase production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins.
 First principles predictions of circuit performance were consistent with measurements for a variety of cases.
 Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity, energy efficiency, and carbon yield.
-A sufficient energy supply with oxidative phosphorylation is instrumental for high energy efficiency and carbon yields; in addition, the translation rate is the next step to optimize for higher productivity.
+A sufficient energy supply with oxidative phosphorylation is instrumental for high energy efficiency and carbon yields; in addition, the translation rate could be optimized for higher productivity.
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
@@
 \section{Introduction}
-Cell free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in~vivo} processes.
 Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
@@
 Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
-Interestingly, many of the challenges confronting in-vivo genome-scale kinetic modeling can potentially be overcome in a cell-free system.
+Interestingly, many of the challenges confronting \textit{in~vivo} genome scale kinetic modeling can potentially be overcome in a cell-free system.
 For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
 Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome-scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.
-coli} or \textit{B.
+Theoretically, genome scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.~coli} or \textit{B.
 subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
 
-Stoichiometric reconstructions of microbial metabolism popularized by constraint based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \cite{2012_lewis_palsson_NatRevMicrobio}.
-Since the first genome-scale stoichiometric model of \textit{E.
-coli}, developed by Edwards and Palsson \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.
-coli} \cite{Feist:2007aa} or \textit{B.
-subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
+Stoichiometric reconstructions of microbial metabolism popularized by constraints based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \cite{2012_lewis_palsson_NatRevMicrobio}.
+Since the first genome scale stoichiometric model of \textit{E.~coli}, developed by Edwards and Palsson \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B.
+subtilis} \cite{Oh:2007aa} are now available \cite{2009_feist_palsson_NatRevMicrobio}.
 Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
 Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \cite{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \cite{2007_schuetz_sauer_MolSysBio}.
-Recent advances in constraint-based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}.
-Thus, modern constraint-based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+Recent advances in constraints based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}.
+Thus, modern constraints based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
 
 In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
-A core \emph{E.~coli} cell free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
+A core \emph{E.~coli} cell-free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
 This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
@@
 From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation is instrumental for maintaining a high carbon yield and the translation rate for productivity.
-Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high carbon yield and the translation rate for productivity.
+Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
 
@@
 The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity, energy efficiency and carbon yield for a protein of interest.
-One mathematical framework that has found wide use in modeling metabolism is constraint based models such as flux balance analysis (FBA).
+One mathematical framework that has found wide use in modeling metabolism is constraints based models such as flux balance analysis (FBA).
 FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
-Traditionally, FBA is used to model \textit{in vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
-Thus, to model cell-free metabolism we constructed a cell-free stoichiometric network by removing growth associated reactions from the MG1655 reconstruction \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
-The network consisted of 272 reactions and 145 species (Fig \ref{fig:network}A).
+Traditionally, FBA is used to model \textit{in~vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
+Thus, to model cell-free metabolism we constructed a stoichiometric network by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
+The network consisted of 272 reactions and 145 species.
+The core metabolism is shown in Fig.~\ref{fig:network}A.
 We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
-We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E. coli} extract systems.
-The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig \ref{fig:network}B).
-The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E.
-coli} extract for a duration of 8 hours under maltose consumption (Fig \ref{fig:network}C).
+We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E.~coli} extract systems.
+The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig.~\ref{fig:network}B).
+The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for a duration of 8 hours under maltose consumption (Fig.~\ref{fig:network}C).
 The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
@@
 Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
-Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig \ref{fig:network}D).
+Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
 Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
@@
 B.
-CAT production under a T7 promoter in CFPS \textit{E.
-coli} extract for 1 hr under glucose consumption.
+CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption.
 C.
-deGFP production under a P70 promoter in TXTL 2.0 \textit{E.
-coli} extract for 8 hr under glucose consumption.
+deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption.
 D.
@@
 It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under two different promoters and cell-free systems.
-Since the model accurately predicted the behavior of protein production, we wanted to use our mathematical framework to help us understand the performance limits of CFPS and how these shortcomings could be addressed.
+Since the model accurately predicted the behavior of protein production, we extended our mathematical framework to understand the performance limits of CFPS and how they could be addressed.
 
@@
 In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (AA uptake and synthesis).
-In the second case, CFPS was supplied with amino acids, but the amino acid synthesis were turned off (AA uptake w/o synthesis).
+In the second case, CFPS was supplied with amino acids, but the amino acid synthesis reactions were turned off (AA uptake w/o synthesis).
 These amino acid synthesis reactions were blocked since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
@@
 Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
-We used ssFBA to estimate the productivity for each of these proteins for each case (Fig \ref{fig:Prod}).
-An additional case was considered for CAT, since a comprehensive dataset is available \cite{2005_calhoun_BiotechnologyProgress}; thus fluxes were constrained to experimental measurements where avaliable, except for CAT production which was determined by the transcription/translation parameters.
+We used ssFBA to estimate the productivity for each of these proteins for each case (Fig.~\ref{fig:Prod}).
+An additional case was considered for CAT, since a comprehensive dataset is available \cite{2005_calhoun_BiotechnologyProgress}; in this case, fluxes were constrained to experimental measurements where avaliable, with the exception of CAT production which was determined by the transcription/translation parameters.
 
@@
 We evaluated CFPS productivity for all eight proteins and for each case.
-All cases had very similar performance for each protein and were in the standard deviation of each other (Fig \ref{fig:Prod}A). 
+All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prod}A).
 %with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
@@
 However, each protein had a different level of productivity.
-For instance, BMP10 had a productivity of about 2.5 ($\mu$M/h) whereas CAT had a productivity of about 12 ($\mu$M/h).
-To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig \ref{fig:Prod}B).
+For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
+To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prod}B).
 The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
-This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity.
+This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
 A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
-CAT was an outlier for the trendline, even though it was in the same order of magnitude of the trendline.
+CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
 The difference in the T7 and  $\sigma_{70}$ promoters did not alter the qualitative productivity performance of CAT.
-The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation, however this should be investigated further.
+The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
 % a further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
@@
 \caption{CFPS productivity of eight proteins for four cases.
-Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white).
+Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
 A.
-Productivity (Error bars represent the 95\% CI of the ensemble).
+Productivity across the ensemble (error bars represent 95\% CI).
 B.
-Mean productivity versus the carbon number for each corresponding protein.
-Single trendline of productivity versus carbon number (dotted line) for all cases (R\textsuperscript{2} = 0.99).
+Mean productivity versus carbon number.
+Single trendline (dotted line) calculated across all cases (R\textsuperscript{2} = 0.99).
 Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
@@
 \end{figure}
-
-The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation as well as synthesize each amino acids required for the production of the protein of interest.
+The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation, as well as synthesize each amino acid required for the production of the protein of interest.
 
 \subsection{CFPS Energy Efficiency}
-Following the same outline as in examining the productivity, we calculated the energy efficiency for each protein (Fig \ref{fig:Energy}A).
-The first two cases had comparable performance with having the highest energy efficiency, where amino acids were supplied in the media.
-The third case (with no amino acid uptake) had the lowest energy efficiency; this is because glucose is utilized to synthesize the necessary amino acids for each protein as well as energize the system.
-We next investigated the effect of the carbon number of each protein on the energy efficiency (Fig \ref{fig:Energy}B).
-The same inverse qualitative trend is observed as for productivity, but followed a linear trendline.
+Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein (Fig.~\ref{fig:Energy}A).
+The first two cases, where amino acids were supplied in the media, had comparable performance including having the highest energy efficiencies.
+The third case (with no amino acid uptake) had the lowest energy efficiency; this was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis, in addition to being available for energy generation.
+We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Energy}B).
+The same inverse trend was observed as for productivity, except that it was linear.
 The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for the first two cases.
-Proteins with a high carbon number have a higher transcription and translation cost then smaller proteins, thus it makes sense for larger proteins to have a lower energy efficiency.  
+Proteins with a high carbon number have a higher transcription and translation cost then smaller proteins, leading to a lower energy efficiency of protein synthesis.  
 The first two cases had the same trendline whereas for the third case there was a significant drop in energy efficiency which required its own trendline (available in the Supporting Information).
-Interestingly, in the third case each protein had a similar energy efficiency of about 25\% regardless of the carbon number of the protien. 
-In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein of interest kept the energy efficiency saturated at a relatively low level.
+Interestingly, in the third case each protein had a similar energy efficiency of about 25\% regardless of carbon number.
+In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
 However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.9 $\pm$ 1.5\% compared to the theoretical maximum of 72 $\pm$ 7.1\%.
-This shows CFPS systems have a lot of room for improvement: first, the experimental setup was still producing certain amino acids which could be turned off. 
-Second, the system has a high accumulation of metabolic byproducts of organic acids which is a result of inefficient energy utilization.
+This shows CFPS systems have a lot of room for improvement: first, the experimental setup still produced certain amino acids; these reactions could be turned off.
+Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.
 
@@
 \caption{CFPS energy efficiency of eight proteins for four cases.
-Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white).
+Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
 A.
@@
 \subsection{CFPS Carbon Yield}
-We also calculated the carbon yield for each protein (Fig \ref{fig:Yield}A).
+We also calculated the carbon yield for each protein (Fig.~\ref{fig:Yield}A).
 The same trends followed, where the cases with amino acids supplied in the media showed the highest carbon yield.
@@
 Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 57.0 $\pm$ 2.2\% for CAT.
-Only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and does not contribute to the carbon yield.
-If that is the case, the carbon yield without glucose contribution increases to 100\%.
-Finally for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\%  for CAT and the system used about twice the amount of glucose as in the first two cases.
-In this case, glucose is being used to synthesize amino acids and provide the necessary energy requirements to power transcription and translation; this trend was seen across all proteins.
-The experimentally constrained case for CAT had carbon yield of about 6\% compared to the theoretical maximum of 57\%.
-This loss in carbon yield shows inefficiencies in CFPS that can be potentially improved.
-Sequence specific flux balance analysis assumes a psuedo steady state, thus intermediate metabolites cannot be accumulated within the cell-free extract.
-In addition, ssFBA is solved by setting the production of the protein as the objective function.
-Therefore, carbon flux will travel throughout the network to optimize the flux through the protein synthesis reaction.
+Only the necessary amount of amino acids was used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
+In that case, the carbon yield without glucose contribution would be 100\%.
+Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
+In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
+In the experimentally constrained case, CAT was produced with a carbon yield of 6\% compared to the theoretical maximum of 57\%.
+This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
+ssFBA assumes a psuedo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
+In addition, ssFBA is solved by setting protein production as the objective function.
+Therefore, carbon flux will travel through the network to optimize the flux through the protein synthesis reaction.
 In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate.
@@
 \caption{CFPS carbon yield of eight proteins for four cases.
-Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements for CAT (white).
+Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
 A.
-Energy efficiency (Error bars represent the 95\% CI of the ensemble).
+Energy efficiency across the ensemble (error bars represent the 95\% CI).
 B.
@@
 \end{figure}
-We next investigated the effect of the carbon number of each protein on the carbon yield (Fig \ref{fig:Yield}B).
-The same inverse qualitative trend is observed as for productivity.
+Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
+The same inverse qualitative trend was observed as for productivity.
 The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
-A single trendline was formulated for the first two cases since they had similar performance and another trendline was formulated for the third cases (available in Supporting Information)
-As the protein size increased, the carbon yield decreased, suggesting that large proteins may not be feasible for cell-free production.
-Thus, we wanted to examine the most parameters that have a significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.
+A single trendline was formulated for the first two cases since they had similar performance, while another trendline was formulated for the third cases (available in Supporting Information).
+As the protein size increased, the carbon yield decreased, suggesting that large proteins may be less feasible for cell-free production.
+Thus, we examined the parameters that had the most significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.
 
 \subsection{CFPS performance sensitivity analysis}
-To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for deGFP, a representative protein (Fig \ref{fig:SI}), as well as on the carbon yield (available in Supporting Information).
-In examining productivity performance (Fig \ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the translation rate by ribosomes being the most significant.
-As expected, this showed that the translation rate is instrumental for productivity, and should be the first step investigated in its optimization, prior to examining transcription parameters.
-This is consistent with literature findings.
+To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for deGFP, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield (available in Supporting Information).
+In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation by ribosomes being the most significant.
+As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
 Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
 In addition, Li et al.
-have increased productivity of firefly luciferase by 5-fold in CFPS systems by adding and adjusting factors that affect transcription and translation such as elongation factors, ribosome recyclcing factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
-In examining the substrate utilization, glucose uptake was not very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
-This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
-In addition, oxygen uptake had more significance in the third case than the first two cases since this case has to synthesize amino acids and power the system. 
-Also, oxygen determines how effectively glucose is utilized, therefore if the glucose uptake rate effects productivity then oxygen should have a similar effect.
-On the other hand, for the first two cases, amino acid uptake showed significance and was even higher for the second case (without AA synthesis) as it was the only source of amino acids.
+have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
+This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis in that case.
+In addition, oxygen uptake had more significance in the third case than in the first two cases, since in that case oxygen was required to synthesize amino acids and power the system.
+Also, oxygen determined how effectively glucose was utilized; therefore, if glucose uptake rate effects productivity, then oxygen could be expected to have a similar effect.
+On the other hand, amino acid uptake showed significance for the first two cases, and was even higher for the second case (without AA synthesis), as it was the only source of amino acids.
 \begin{figure}[t!]
@@
 \caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).
-Error bars represent 95\% CI of the sensitivity index.}
+Error bars represent a 95\% CI on the sensitivity index.}
 \label{fig:SI}
@@
 
-When considering energy efficiency performance (Fig \ref{fig:SI}B), the oxygen uptake rates were the most important for all three cases while the sensitivity to transcription/translation parameters decreased slightly.
+When considering energy efficiency performance (Fig.~\ref{fig:SI}B), the oxygen uptake rates were the most important for all three cases while the sensitivity to transcription/translation parameters decreased slightly.
 The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
-Thus, in investigating CFPS performance in terms of productivity and energy efficiency, the translation rate is an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
-Oxygen uptake showed significant importance for energy efficiency since it is responsible for oxidative phosphorylation which is the most efficieny pathway for energy generation.
+Thus, in investigating CFPS performance in terms of productivity and energy efficiency, the translation rate was an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
+Oxygen uptake showed significant importance for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficieny pathway for energy generation.
 Meanwhile, productivity was determined primarily by the rate of the most downstream steps, transcription and translation.
-Across all three cases, substrate utilization (amino acid uptake and glucose) showed to be the next important as these substrates contributed to the carbon yield of deGFP.
-In the first two cases where amino acids are supplied, there was a significance on amino acid uptake since without it energy would be wasted to synthesize amino acids and for the second case amino acid would not be possible.
-Glucose uptake only showed significance in the third case since this is the only source of carbon for protein synthesis and energy generation.
+Across all three cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the carbon yield of deGFP.
+In the first two cases where amino acids were supplied, amino acid uptake was significant, as without it energy was used to synthesize amino acids.
+Glucose uptake was only seen to be significant in the third case, since it was the only source of carbon for protein synthesis and energy generation.
 %This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
-Jewett and coworkers have reported oxidative phosphorylation was still operating in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in a pyruvate powered CFPS \cite{Jewett:2008aa}.
-The carbon yield sensitivity analysis followed the same trends and is very similar to the energy efficiency sensitivity analysis. 
-It is unknown how active oxidative phosphorylation is compared to that of \textit{in vivo} systems.
-Thus, to investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig \ref{fig:oxphos_yield}).
-Interestingly, oxidative phosphorylation has a strong effect on the carbon yield for all cases. 
-The first two cases follow the same trend, ranging from a carbon yield of 20\% to 55\% depending on the oxidative phosphorylation activity.
-The third case, followed the same trend, however it's carbon yield ranged from about 10\% to 55\%.  
+Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in pyruvate-powered CFPS \cite{Jewett:2008aa}.
+We also investigated the sensitivity of carbon yield to network fluxes; it followed the same trends as the energy efficiency sensitivity analysis.
+It is unknown how active oxidative phosphorylation is compared to in \textit{in~vivo} systems.
+To investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
+Interestingly, oxidative phosphorylation has a strong effect on the carbon yield in all cases.
+The first two cases follow the same trend, ranging from a carbon yield of 20\% to 55\%, depending on the oxidative phosphorylation activity.
+The third case, followed the same trend; however, its carbon yield ranged from about 10\% to 55\%.  
 The third case was expected to have a lower carbon yield for the same oxidative phosphorylation flux compared to the first two cases, since carbon must be utilized for energy generation and amino acid synthesis.
-In all cases where the carbon yield is below its theoretical maximum, there is an accumulation of acetate and lactate, resulting in the lower carbon yield. 
-In examining the experimental dataset, there is a mixture of acetate and lactate accumulation during CAT synthesis which shows CFPS is not operating at a full aerobic state.
-It is unclear how active oxidative phosphorylation is in CFPS since the reactions rely on electron transport from membrane vesicles . 
-The addition of phosphate showed an increase in CAT yield, however it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
-Interestingly, the addition of NADH did not increase protein synthesis since the ATP concentrations were most likely saturated. 
-Thus, an alternative strategy may be to inhibit anaerobic processes in cell-free in order to minimize unwanted byproducts such as acetate and lactate.
+In all three cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of acetate and lactate, resulting in the lower carbon yield.
+The experimental dataset exhibits a mixture of acetate and lactate accumulation during CAT synthesis, which shows that CFPS is not operating in a fully aerobic state.
+It is unclear how active oxidative phosphorylation is in CFPS, since the reactions rely on electron transport from membrane vesicles.
+The addition of phosphate showed an increase in CAT yield; however, it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
+Interestingly, the addition of NADH did not increase the rate of protein synthesis, since the concentration of ATP was most likely saturated.
+An alternative strategy may be the inhibition of anaerobic processes in cell-free, in order to minimize unwanted byproducts such as acetate and lactate.
 %Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and how much from anaerobic processes.
@@
 \includegraphics[width=0.7\textwidth]{./Figures/Yield_Ox.pdf}
-\caption{An ensemble of 1000 ssFBA solutions for deGFP carbon yield versus oxidative phosphorylation flux for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
+\caption{deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 ssFBA solutions, for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
 \label{fig:oxphos_yield}
@@
 
-To investigate the differences for each case, we compared the flux distributions for a representative protein, deGFP, predicted by ssFBA simulations and for the case constrained by experimental measurements for CAT (Fig.\ref{fig:flux}).
+To investigate the differences between the three cases, we compared the flux distributions for a representative protein, deGFP, predicted by ssFBA simulations, as well as the case constrained by experimental measurements for CAT (Fig.~\ref{fig:flux}).
 \begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./Figures/Flux_all.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation.
-Flux across ensemble normalized to glucose uptake flux.
-Flux distribution for four different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, (C) amino acid synthesis without uptake, and (D) CAT flux distribution constrained by experimental measurements.
+\includegraphics[width=1.00\textwidth]{./Figures/Flux.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation, for three different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, and (C) amino acid synthesis without uptake.
+Mean flux across the ensemble, normalized to glucose uptake flux.
 Thick arrows indicate flux to or from amino acids.}
@@
 \end{figure}
-In the first case, which was supplied with amino acids and could synthesize them in the network, relied on a combination of glucose and amino acids to power the system (Fig \ref{fig:flux}A).
+The first case, which was supplied with amino acids and could synthesize them in the network, relied on a combination of glucose and amino acids to power the system (Fig.~\ref{fig:flux}A).
 Glucose traveled through glycolysis and the Entner-Doudoroff pathway to generate NADH for oxidative phosphorylation as well as utilize pyruvate for amino acid biosynthesis.
-Interestingly, amino acids powered the TCA cycle instead of glucose via alpha-ketoglutarate and fumerate and utilizing oxaloacetic acid for additional amino acid biosynthesis.
-Thus, ssFBA found the combinational use of glucose and amino acids to energize CFPS as the optimum case for protein synthesis. 
+Interestingly, amino acids rather than glucose powered the TCA cycle, via alpha-ketoglutarate and fumarate, and by utilizing oxaloacetic acid for additional amino acid biosynthesis.
+Thus, ssFBA found that the combined utilization of glucose and amino acids to energize CFPS was the optimal case for protein synthesis.
 In the second case, where amino acid synthesis was removed from the network, glucose was solely utilized to provide the necessary energy requirements via glycolysis which generated enough NADH for oxidative phosphorylation.
 Ubiquinone was generated via \textit{nuo} to power oxidative phosphorylation, instead of relying on the TCA cycle.
-Once the required energy requirements were met for transcription and translation, the amino acids necessary for the assembly of the protein of interest were utilized from the media.
-The first two cases had similar performance in productivity, energy efficiency and carbon yield for all proteins, this is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
-In the third case, where amino acids must be synthesized, there is drop in performance (productivity, energy efficiency, and carbon yield) compared to the first two cases. 
-Glucose undergoes through all the major pathways: glycolysis, pentose phosphate pathway and the tca cycle. 
-First, glucose must be converted into all the necessary amino acids for the assembly of the protein of interest. 
-Second, it must provide the necessary energy requirements for transcription and translation. 
-There is a relatively high flux between the interconversion of NADH to NADPH.
-NADPH is required for the biosynthesis of amino acids, however NADH is required for oxidative phosphorylation, thus requiring glucose to energize CFPS as well as synthesize amino acids results in a decrease in performance.
-Our model and the sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
+Once the energy requirements for transcription and translation were met, amino acids were taken from the media to assemble the protein of interest.
+The first two cases had similar performance in terms of productivity, energy efficiency and carbon yield for all proteins.
+This is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
+In the third case, where amino acids must be synthesized, there is drop in all performance metrics (productivity, energy efficiency, and carbon yield) compared to the first two cases.
+%Glucose passes through all the major pathways: glycolysis, pentose phosphate pathway and the TCA cycle.
+%First, glucose must be converted into all the necessary amino acids for the assembly of the protein of interest.
+%Second, it must provide the necessary energy requirements for transcription and translation.
+This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
+This leads to the relatively high flux for the conversion of NADH to NADPH.
+Thus, less NADH is available for oxidative phosphorylation.
+%, thus requiring glucose to energize CFPS as well as synthesize amino acids results in a decrease in performance.
+The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
 Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
-The fourth case, constrained by experimental measurements, had a fairly similar flux distribution as the third case. 
-Even though the experimental setup was supplied with amino acids, glucose travels through all the major pathways and metabolic precursors are utilized for amino acid biosynthesis. 
-Only certain amino acid synthesis reactions were blocked since during the cell-free extract preparation all amino acids were not supplied.  
-It can be seen, there is an accumulation of pyruvate, lactate, acetate, and other organic acids which is an inefficiency of carbon utilization.
-In addition, there is a high flux via the Entner-Douodoroff Pathway which may be an artifact of the optimum solution of ssFBA.
-In order to address the feasibility of which reactions are occuring in CFPS, adding thermodynamic constraints on reactions may result in a better depiction of the intracellular flux distribution \cite{Henry:2007,Hamilton:2013}.
-In this case, it is unclear which carbon (glucose or amino acids) is used to power CFPS and which carbon is used for protein synthesis and may in fact be a combination of glucose and amino acids.
-Thus, it would be interesting to track the carbon with C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA with C\textsuperscript{13} measurements, which has been shown to have a good representation of the flux distribution for \textit{in vivo} process \cite{Zamboni:2009}.
+
+The fourth case, constrained by experimental measurements (Fig. \ref{fig:flux_exp}), had a fairly similar flux distribution as the third case.
+The central carbon organic acids show good agreement with the data (Fig. \ref{fig:flux_exp}B).
+Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour which constrained the solution space of ssFBA to have a more realistic depiction of the flux distribution.
+Unlike the second case, only certain amino acid synthesis reactions were blocked since during the growth of \emph{E.~coli} not all amino acids were supplied.
+During the cell-free reaction all amino acids were supplied, however glucose still traveled through all the major pathways, and the same metabolic precursors were still utilized for amino acid biosynthesis.
+Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig. \ref{fig:flux_exp}B), implying an inefficiency of carbon utilization.
+In addition, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
+To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the intracellular flux distribution \cite{Henry:2007,Hamilton:2013}.
+In this case, it is unclear which substrate (glucose or amino acids) is used to power CFPS and may in fact be a combination of both.
+Thus, it would be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
+\begin{figure}[t!]
+\includegraphics[width=1.00\textwidth]{./Figures/ssFBA_exp.pdf}
+\caption{ssFBA simulation of CAT production for an experimentally constrained case.
+(A) Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation.
+Mean flux across the ensemble, normalized to glucose uptake flux.
+Thick arrows indicate flux to or from amino acids.
+(B) Central carbon metabolite measurements versus ssFBA simulations over a one hour time course.}
+\label{fig:flux_exp}
+\end{figure}
 
@@
 The translation rate was identified as the rate limiting step for productivity.
-The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficinecy and carbon yield.
-In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell free synthetic circuits.
+The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficiency and carbon yield.
+In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
 
@@
 \subsection*{Formulation and solution of the model equations.}
-We estimated the theoretical maximum performance of the cell free protein synthesis system using sequence-specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
-The sequence-specific flux balance analysis problem was formulated as a linear program:
+We estimated the theoretical maximum performance of the cell-free protein synthesis system using sequence specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+The sequence specific flux balance analysis problem was formulated as a linear program:
 \begin{equation}
@@
 The term $R_{1}$ denotes the RNA polymerase abundance,
-$v_{TX}$ denotes the RNA polymerase elongation rate (nt/hr),
+$v_{TX}$ denotes the RNA polymerase elongation rate (nt/h),
 $l_{G}$ denotes the gene length in nucleotides, and the last term $\mathcal{P}$ describes a model of promoter activity.
@@
 
-The objective of the sequence-specific flux balance calculation was to maximize the rate of protein translation, $w_{TL}$.
+The objective of the sequence specific flux balance calculation was to maximize the rate of protein translation, $w_{TL}$.
 The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data, while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
 Gene and protein sequences were taken from literature and are available in the Supporting Information.
-The sequence-specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.55 \cite{GLPK}.
+The sequence specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.55 \cite{GLPK}.
+For all cases, amino acid degradation reactions were blocked since these enzymes were knocked out during the cell-free extract preparation \cite{2005_calhoun_BiotechnologyProgress, Garamella:2016aa}.
+In the second case, all amino acid synthesis reactions were set to 0 mM/hr since \textit{E.
+coli} was grown in the presence of amino acids, thus these enzymes would not be present in the cell-free extract media.
+In the third case, amino acid uptake reactions were set to 0 mM/hr.   
+In the experimental constrained case, \textit{E.
+coli} was grown in the presence of 13 amino acids (alanine, arginine, cysteine, serine, aspartate, glutamate, and glutamine were excluded) \cite{Zawada:2003}, thus the synthesis reactions responsible for those 13 amino acid were set to 0 mM/hr.
 
@@
 where $\Delta\texttt{POI}$ denotes the flux of the protein of interest produced, $C_{POI}$ denotes carbon number of the protein of interest, $\mathcal{R}$ denotes the number of reactants,
-$\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (never allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
+$\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (not allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
 
 \subsection*{Quantification of uncertainty.}
-An ensemble of 100 sets of flux distributions was calculated for three different cases: control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
-An ensemble of flux distributions was then calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 30 to 40 mM/h, determined from experimental data and randomly sampling oxygen uptake, RNAP polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+An ensemble of 100 sets of flux distributions was calculated for each of the three different cases: control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
+The ensemble was calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 0 to 30 mM/h, determined from experimental data and randomly sampling RNA polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
 RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
@@
 \subsection*{Global sensitivity analysis.}
-We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
+We conducted a global sensitivity analysis using the variance-based method of Sobol to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
 We computed the total sensitivity index of each parameter relative to three performance objectives: productivity of the protein of interest, energy efficiency and carbon yield.
@@
 \end{document}
+\grid

### 2017-03-06 Paper_v1.tex
MM updates

@@
 CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption.
+Error bars denote the standard deviation of experimental measurements.
 C.
 deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption.
+Error bars denote a 10\% coefficient of variation.
 D.
@@
 
+\subsection*{Cell-free protein synthesis.}
+The glucose/NMP cell-free protein synthesis reactions were performed at a volume of 15 $\mu$L in 1.5-mL eppendorf tubes and incubated in a humidified incubator. 
+Plasmid pK7CAT was used as the DNA template for chloramphenical acetyl transferase (CAT) expression by placing the \emph{cat} gene between the T7 promoter and the T7 terminator \cite{Kigawa1995}.
+The plasmid was isolated and purified using a Plasmid Maxi Kit (Qiagen, Valencia CA). 
+Cell-free protein synthesis of CAT protein was performed at 37 $^{\circ}$C for 3 h.
+
+S30 extract was prepared from E. coli strain KC6 (A19 $\delta$tonA $\delta$tnaA $\delta$speA $\delta$endA $\delta$sdaA $\delta$sdaB $\delta$gshA met+). 
+This K12-derivative has several gene deletions to stabilize amino acid concentrations during the cell-free reaction. 
+The KC6 strain was grown to approximately 3 OD595 in a 10-L fermenter (B.
+Braun, Allentown PA) on defined media with glucose as the carbon source and with the addition of 13 amino acids (alanine, arginine, cysteine, serine, aspartate, glutamate, and glutamine were excluded) \cite{Zawada:2003}.
+Crude S30 extract was made as previously described \cite{Jewett:2002}.
+
+The standard protein synthesis reaction was conducted according to the PANOxSP protocol with slight modifications from that described previously \cite{BIT:BIT20026}. 
+Unless otherwise noted, all reagents were purchased from Sigma (St. Louis, MO). 
+The initial mixture includes 1.2 mM ATP; 0.85 mM each of GTP, UTP, and CTP; 30 mM phosphoenolpyruvate (Roche, Indianapolis IN); 130 mM potassium glutamate; 10 mM ammonium glutamate; 16 mM magnesium glutamate; 50 mM HEPES-KOH buffer (pH 7.5); 1.5 mM spermidine; 1.0 mM putrescine; 34 $\mu$g/mL folinic acid; 170.6 $\mu$g/mL E.
+coli tRNA mixture (Roche, Indianapolis IN); 13.3 $\mu$g/mL pK7CAT plasmid; 100 $\mu$g/mL T7 RNA polymerase; 20 unlabeled amino acids at 2 mM each; 5 $\mu$M l-[U-\textsuperscript{14}C]-leucine (Amersham Pharmacia, Uppsala Sweden); 0.33 mM nicotinamide adenine dinucleotide (NAD); 0.26 mM coenzyme A (CoA); 2.7 mM sodium oxalate; and 0.24 volumes of E.
+coli S30 extract.
+This reaction was modified for the energy source used such that glucose reactions have 30 mM glucose in place of PEP, and glutamate reactions do not have a secondary energy source added since glutamate is already present in high concentrations from the reaction salts.
+Sodium oxalate was not added since it has a detrimental effect on protein synthesis and ATP concentrations when using glucose or an early glycolytic intermediate as the energy source \cite{BIT:BIT1121}.
+The HEPES buffer (pKa $\sim$ 7.5) was replaced with Bis-Tris (pKa $\sim$ 6.5).
+In addition, the magnesium glutamate concentration was reduced to 8 mM for the glucose and glutamate reactions since a lower magnesium optimum is found when using a nonphosphorylated energy source \cite{BIT:BIT20026}.
+Finally, 10 mM phosphate was added to some reactions as described below in the form of potassium phosphate dibasic adjusted to pH 7.2 with acetic acid.
+
+\subsection*{Measurements of protein product and metabolites.}
+Cell-free reaction samples were quenched at specific timepoints with equal volumes of ice-cold 150 mM sulfuric acid to precipitate proteins. 
+Protein synthesis of CAT was determined from the total amount of \textsuperscript{14}C-leucine-labeled product by trichloroacetic acid precipitation followed by scintillation counting as described previously \cite{2005_calhoun_BiotechnologyProgress}.
+Samples were centrifuged for 10 min at 12,000g and 4$^{\circ}$C. 
+The supernatant was collected for high performance liquid chromatography (HPLC) analysis. 
+HPLC analysis (Agilent 1100 HPLC, Palo Alto CA) was used to separate nucleotides and organic acids, including glucose.
+Compounds were identified and quantified by comparison to known standards for retention time and UV absorbance (260 nm for nucleotides and 210 nm for organic acids) as described previously \cite{2005_calhoun_BiotechnologyProgress}.
+Calibration standards were routinely run for improved quantification of compounds. 
+The standard compounds quantified with a refractive index detector included inorganic phosphate, glucose, and acetate. 
+The compounds quantified with the UV detector included pyruvate, malate, succinate, and lactate.
+The stability of the amino acids in the cell extract was determined using a Dionex Amino Acid Analysis (AAA) HPLC System (Sunnyvale, CA) that separates amino acids by gradient anion exchange (AminoPac PA10 column).
+Compounds were identified with pulsed amperometric electrochemical detection and by comparison to known standards.
+
 \subsection*{Formulation and solution of the model equations.}

### 2017-03-07 Paper_v1.tex
Working on results

@@
 \section{Introduction}
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in~vivo} processes.
-Central amongst these advantages is direct access to metabolites and the microbial biosynthetic machinery without the interference of a cell wall.
-This allows us to control as well as interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
-Second, cell-free systems also allow us to study biological processes without the complications associated with cell growth.
+Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized protein production.
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
+This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
 Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
-However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamentally important biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
+However, CFPS is not new; CFPS in crude \textit{E.~coli} extracts has been used since the 1960s to explore fundamental biological mechanisms \cite{MATTHAEI:1961aa,NIRENBERG:1961aa}.
 Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
-Interestingly, many of the challenges confronting \textit{in~vivo} genome scale kinetic modeling can potentially be overcome in a cell-free system.
-For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
-Thus, cell-free operation holds several significant advantages for model development, identification and validation.
-Theoretically, genome scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.~coli} or \textit{B.
+However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing, we must first understand the performance limits of these systems.
+One tool to address this question is mathematical modeling.
+
+% Interestingly, many of the challenges confronting \textit{in~vivo} genome scale kinetic modeling can potentially be overcome in a cell-free system.
+% For example, there is no complex transcriptional regulation to consider, transient metabolic measurements are easier to obtain, and we no longer have to consider cell growth.
+% Thus, cell-free operation holds several significant advantages for model development, identification and validation.
+% Theoretically, genome scale cell-free kinetic models may be possible for industrially important organisms, such as \textit{E.~coli} or \textit{B.
 subtilis}, if a simple, tractable framework for integrating allosteric regulation with enzyme kinetics can be formulated.
-
-Stoichiometric reconstructions of microbial metabolism popularized by constraints based modeling techniques such as flux balance analysis (FBA) have become standard tools to interrogate biological networks \cite{2012_lewis_palsson_NatRevMicrobio}.
-Since the first genome scale stoichiometric model of \textit{E.~coli}, developed by Edwards and Palsson \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B.
-subtilis} \cite{Oh:2007aa} are now available \cite{2009_feist_palsson_NatRevMicrobio}.
-Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
+% Traditionally, stoichiometric models have also neglected explicit descriptions of metabolic regulation and control mechanisms, instead opting to describe the choice of pathways by prescribing an objective function on metabolism.
 Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \cite{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \cite{2007_schuetz_sauer_MolSysBio}.
-Recent advances in constraints based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}.
+% Recent advances in constraints based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}.
 Thus, modern constraints based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
 
-In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis (CFPS) system.
-A core \emph{E.~coli} cell-free metabolic model, consisting of 132 metabolites and 281 reactions, was developed from literature \cite{Feist:2007aa}.
+Stoichiometric reconstructions of microbial metabolism, popularized by flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
+Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B.
+subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
+In this study, we used sequence specific constraints based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
+A core \emph{E.~coli} cell-free metabolic model was developed from literature \cite{Feist:2007aa}.
 This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
-sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
-Thus, sequence specific constraints based modeling explicitly coupled the transcription and translation of circuit components with the availability of metabolic resources.
-We tested this approach by first simulating the production of CAT and deGFP under two different promoters, and then expanded this for $\sigma_{70}$-induced deGFP expression at different plasmid concentrations.
-First principles predictions of circuit performance were consistent with measurements.
-We then investigated the productivity and carbon yield for a range of 8 different proteins for a variety of cases.
+sequence specific descriptions of promoter function, transcription and translation processes.
+Thus, sequence specific constraints based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
+We tested this approach by simulating the production of two model proteins, and then investigated the productivity and carbon yield for eight different proteins.
 From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins.
@@
 
+\clearpage
+
 \section{Results and discussion}
 
-\subsection{Model Derivation}
-The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity, energy efficiency and carbon yield for a protein of interest.
-One mathematical framework that has found wide use in modeling metabolism is constraints based models such as flux balance analysis (FBA).
-FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
-Traditionally, FBA is used to model \textit{in~vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
-Thus, to model cell-free metabolism we constructed a stoichiometric network by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}, and incorporating amino acid synthesis and transcription/translation associated reactions \cite{Allen:2003aa} for a protein of interest to be expressed.
-The network consisted of 272 reactions and 145 species.
-The core metabolism is shown in Fig.~\ref{fig:network}A.
-We developed a cell-free sequence specific flux balance analysis (ssFBA) with a detailed promoter model \cite{Moon:2012aa} to examine the performance of CFPS.
-We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two different cell-free \textit{E.~coli} extract systems.
-The first protein, chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for a duration of 1 hour under glucose consumption (Fig.~\ref{fig:network}B).
-The second protein, dual emission green fluorescent protein (deGFP), was produced under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for a duration of 8 hours under maltose consumption (Fig.~\ref{fig:network}C).
-The ssFBA simulations predicted the production of both proteins for the duration of both CFPS batch reactions.
-Uncertainty in experimental factors such as RNA polymerase, ribosome concentrations, elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
-Thus, the metabolic network and molecular description of trancription and translation were consistent with experimental measurements.
-Next, ssFBA predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
+% The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity, energy efficiency and carbon yield for a protein of interest.
+% One mathematical framework that has found wide use in modeling metabolism is constraints based models such as flux balance analysis (FBA).
+% FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
+% Traditionally, FBA is used to model \textit{in~vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
+% Thus,
+% sequence specific flux balance analysis (ssFBA) with a detailed promoter model  to examine the performance of CFPS.
+% We first validated the ssFBA approach by comparing simulated and measured concentrations of two proteins from two
+
+\subsection{Model derivation and validation}
+The cell free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}.
+We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
+A schematic of the metabolic network, which consisted of 272 reactions and 145 species, is shown in Fig.~\ref{fig:network}A.
+Using this network, in combination with detailed promoter models, and literature values for cell free culture parameters (Table ZZ),
+we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
+different cell-free \textit{E.~coli} extracts. The cell free metabolic network, model parameters, model code,
+and each protein sequence are available in the supplemental materials.
+
+Cell free simulations predicted CAT and deGFP production for the duration of the CFPS batch reactions (Fig. ~\ref{fig:network}B and C).
+Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
+With the exception of the first 10-15 min, the cell free prediction of CAT abundance was consistent with the measured values.
+On the other hand, deGFP was produced under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
+The cell free simulation captured the overall trend of deGFP abundance, but was not able to capture saturation at the end of the CFPS culture.
+Uncertainty in experimental factors such as the concentration of RNA polymerase, and ribosomes,
+transcription and translation elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
+Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.
+
+Next, we predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
 Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
@@
 \end{figure}
+These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters.
+It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
+Since the model accurately predicted protein production, we used our mathematical framework to understand the performance limits of CFPS.
 
-These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with few adjustable parameters in the promoter model taken from literature \cite{Moon:2012aa}.
-It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under two different promoters and cell-free systems.
-Since the model accurately predicted the behavior of protein production, we extended our mathematical framework to understand the performance limits of CFPS and how they could be addressed.
-
+\subsection{Analysis of CFPS performance}
 Our next goal was to examine the performance of CFPS for eight different proteins under three different cases.
@@
 
-\subsection{CFPS Productivity}
+\subsubsection{Productivity}
 We evaluated CFPS productivity for all eight proteins and for each case.
@@
 
-\subsection{CFPS Energy Efficiency}
+\subsubsection{Energy efficiency}
 Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein (Fig.~\ref{fig:Energy}A).
@@
 
-\subsection{CFPS Carbon Yield}
+\subsubsection{Yield}
 We also calculated the carbon yield for each protein (Fig.~\ref{fig:Yield}A).
@@
 
-\subsection{CFPS performance sensitivity analysis}
+\subsection{Sensitivity analysis}
 To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for deGFP, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield (available in Supporting Information).
@@
 
+\clearpage
+
 \section*{Materials and Methods}
 
-\subsection*{Cell-free protein synthesis.}
+\subsection*{Glucose/NMP cell-free protein synthesis.}
 The glucose/NMP cell-free protein synthesis reactions were performed at a volume of 15 $\mu$L in 1.5-mL eppendorf tubes and incubated in a humidified incubator.
@@
 
+\clearpage
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-03-08 Paper_v1.tex
working through the MM

@@
 \subsection*{Formulation and solution of the model equations.}
-We estimated the theoretical maximum performance of the cell-free protein synthesis system using sequence specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
-The sequence specific flux balance analysis problem was formulated as a linear program:
+The sequence-specific flux balance analysis problem was formulated as a linear program:
 \begin{equation}
@@
 	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
-\alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+\mathcal{L}_{i} \leq w_i \leq \mathcal{U}_{i}  \qquad i=1,2,\hdots,\mathcal{R}
  \end{multlined}
@@
 where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
-and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
-The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
+and $\mathcal{L}_{i}$ and $\mathcal{U}_{i}$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions of Allen and Palsson \cite{Allen:2003aa}:
 \begin{eqnarray*}
-\mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
-\mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{1}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
-\mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
-\mathrm{mRNA}+\mathrm{R_{2}} &\longrightarrow& \mathrm{R_{2}^{*}} \\
-\mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
-&& \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
-\mathrm{R_{2}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{2}+mRNA}+\\
-&& \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
+{G}_{\mathcal{P}}+{R}_{T} &\longrightarrow& G_{\mathcal{P}}^{*} \\
+G_{\mathcal{P}}^{*}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left\{k\right\}TP &\overset{\rm TX}{\longrightarrow}& mRNA+G_{\mathcal{P}}+R_{T}+ \sum_{k\in\left\{A,C,G,U\right\}}2 \eta_{k} \cdot Pi\\
+mRNA &\longrightarrow& \sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP \\
+mRNA+R_{X} &\longrightarrow& R_{X}^{*} \\
+\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP &\longrightarrow& \alpha_{j}\cdot AA_{j}-tRNA_{j}+\\
+&& \qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi \qquad{j=1,2,\hdots,20}\\
+R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big) &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+R_{X}+mRNA+\\
+&& \quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)
 \end{eqnarray*}
 where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
-$\rm R_{1}$ denotes the concentration of RNA polymerase,
-$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
+$R_{T}$ denotes the concentration of RNA polymerase,
+$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase (open complex),
 $\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
-$\rm Pi$ denotes inorganic phosphate,
-$\rm R_{2}$ denotes the ribosome concentration,
-$\rm R_{2}^{*}$ denotes bounded ribosome,
-and $AA_{j}$ denotes the $j^{th}$ amino acid.
-
-The transcription rate ($w_{TX}$) was fixed in the ssFBA calculation at:
-\begin{equation}
-	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
-\end{equation}
-where $G$ denotes the gene concentration and $K_{TX}$ denotes a transcription saturation coefficient.
-The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
-\begin{equation}
-	V_{TX}^{max} \equiv \left[R_{1}\left(\frac{v_{TX}}{l_{G}}\right)\mathcal{P}\right]
-\end{equation}
-The term $R_{1}$ denotes the RNA polymerase abundance,
-$v_{TX}$ denotes the RNA polymerase elongation rate (nt/h),
-$l_{G}$ denotes the gene length in nucleotides, and the last term $\mathcal{P}$ describes a model of promoter activity.
-In this study, we considered two promoters: T7 and $\sigma_{70}$.
-The promoter function for the T7 promoter, $\mathcal{P}_{T7}$, was given by:
-\begin{equation}
-	P_{T7} = \frac{K_{T7}}{1 + K_{T7}}
-\end{equation}
-where $K_{T7}$ denotes a T7 RNA polymerase binding constant \cite{Moon:2012aa}.
-The $\sigma_{70}$ binding promoter, used for all other proteins, was formulated as:
-\begin{equation}
-	P_{\sigma_{70}} = \frac{K_{1}+K_{2}f_{p70}}{1 + K_{1}+K_{2}f_{p70}}
-\end{equation}
-where $K_{1}$ denotes the state of RNA polymerase binding,
-$K_{2}$ is the state of $\sigma_{70}$ binding along with RNA polymerase, and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to the promoter (modeled as a Hill function).
-
-The translation rate ($w_{TL}$) was bounded by:
- \begin{equation}
-	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{\rm mRNA_{SS}}{K_{TL}+\rm mRNA_{SS}}\right)
-\end{equation}
-where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance and $K_{TL}$ denotes the translation saturation constant.
-The maximum translation rate $V_{TL}^{max}$ was formulated as:
-\begin{equation}
-	V_{TL}^{max} \equiv \left[K_{P} R_{2}\left(\frac{v_{TL}}{l_{P}}\right)\right]
-\end{equation}
-The term $K_{P}$ denotes the polysome amplification constant,
-$v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
-$l_{P}$ denotes the number of amino acids in the protein of interest,
-and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
-\begin{equation}
-	 mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
-\end{equation}
-where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+$Pi$ denotes inorganic phosphate,
+$R_{X}$ denotes the ribosome concentration,
+$R_{X}^{*}$ denotes bounded ribosome,
+and $AA_{j}$ denotes $j^{th}$ amino acid.
 
@@
 
+% We estimated the theoretical maximum performance of the cell-free protein synthesis system using sequence specific flux balance analysis (ssFBA) \cite{Allen:2003aa}.
+% The sequence specific flux balance analysis problem was formulated as a linear program:
+% \begin{equation}
+%  \begin{multlined}
+% 	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+% 	\mathrm{Subject \; to:}
+% 	 \; \; \mathbf{S}\mathbf{w}=\mathbf{0} \\
+% \alpha_i \leq w_i \leq \beta_i  \qquad i=1,2,\hdots,\mathcal{R}
+%  \end{multlined}
+% \end{equation}
+% where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+% and $\alpha_i$ and $\beta_i$ denote the lower and upper bounds on flux $w_{i}$, respectively.
+% The stoichiometry of the kinetic model was used for the ssFBA calculations, with the execpetion of the transcription and translation rates.
+% The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions taken from Allen and Palsson \cite{Allen:2003aa}:
+% \begin{eqnarray*}
+% \mathrm{{G}_{\mathcal{P}}}+\mathrm{{R}_{1}} &\longrightarrow& \mathrm{G_{\mathcal{P}}^{*}} \\
+% \mathrm{G_{\mathcal{P}}^{*}}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \mathrm{\left\{k\right\}TP} &\overset{\rm TX}{\longrightarrow}& \mathrm{mRNA}+\mathrm{G_{\mathcal{P}}}+\mathrm{R_{T}}+ \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}2\eta_{k}\cdot Pi}\\
+% \mathrm{mRNA} &\longrightarrow& \mathrm{\sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP} \\
+% \mathrm{mRNA}+\mathrm{R_{X}} &\longrightarrow& \mathrm{R_{X}^{*}} \\
+% \mathrm{\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP} &\longrightarrow& \mathrm{\alpha_{j}\cdot AA_{j}-tRNA_{j}+}\\
+% && \mathrm{\qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi} \qquad{j=1,2,\hdots,20}\\
+% \mathrm{R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big)} &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+\mathrm{R_{X}+mRNA}+\\
+% && \mathrm{\quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)}
+% \end{eqnarray*}
+% where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
+% $\rm R_{T}$ denotes the concentration of RNA polymerase,
+% $G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase,
+% $\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
+% $\rm Pi$ denotes inorganic phosphate,
+% $\rm R_{X}$ denotes the ribosome concentration,
+% $\rm R_{X}^{*}$ denotes bounded ribosome,
+% and $AA_{j}$ denotes the $j^{th}$ amino acid.
+
+The bounds on the transcription rate ($\mathcal{L}_{TX}=w_{TX}=\mathcal{U}_{TX}$) was modeled as:
+\begin{equation}
+	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
+\end{equation}
+where $G$ denotes the gene concentration and $K_{TX}$ denotes a transcription saturation coefficient.
+The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
+\begin{equation}
+	V_{TX}^{max} \equiv \left[R_{T}\left(\frac{v_{TX}}{l_{G}}\right)u\left(\kappa\right)\right]
+\end{equation}
+The term $R_{T}$ denotes the RNA polymerase concentration (nM),
+$v_{TX}$ denotes the RNA polymerase elongation rate (nt/h),
+$l_{G}$ denotes the gene length in nucleotides (nt).
+The term $u\left(\kappa\right)$ (dimensionless, $0\leq u\left(\kappa\right)\leq 1$) describes a model of promoter activity, where $\kappa$ denotes promoter specific parameters.
+The forms for the promoter models were taken from Moon $\emph{et~al.}$ \cite{Moon:2012ab}.
+In this study, we considered two promoters: T7 and $\sigma_{70}$.
+The promoter function for the T7 promoter, $u_{T7}$, was given by:
+\begin{equation}
+	u_{T7} = \frac{K_{T7}}{1 + K_{T7}}
+\end{equation}
+where $K_{T7}$ denotes a T7 RNA polymerase binding constant.
+The $\sigma_{70}$ promoter function $u_{\sigma_{70}}$ (which was used for all other proteins) was formulated as:
+\begin{equation}
+	u_{\sigma_{70}} = \frac{K_{1}+K_{2}f_{\sigma_{70}}}{1 + K_{1}+K_{2}f_{\sigma_{70}}}
+\end{equation}
+where $K_{1}$ denotes the weight of RNA polymerase binding alone,
+$K_{2}$ denotes the weight of RNAP-$\sigma_{70}$ bound to the promoter,
+and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to RNAP (modeled as a Hill function).
+The values for $K_{T7}, K_{1}$, $K_{2}$ and $f_{\sigma_{70}}$, along with all other promoter parameters, are given in Table ZZ.
+
+The translation rate ($w_{TL}$) was bounded by:
+ \begin{equation}
+	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{mRNA_{SS}}{K_{TL}+mRNA_{SS}}\right)
+\end{equation}
+where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance and $K_{TL}$ denotes the translation saturation constant.
+The maximum translation rate $V_{TL}^{max}$ was formulated as:
+\begin{equation}
+	V_{TL}^{max} \equiv \left[K_{P} R_{X}\left(\frac{v_{TL}}{l_{P}}\right)\right]
+\end{equation}
+The term $K_{P}$ denotes the polysome amplification constant,
+$v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
+$l_{P}$ denotes the number of amino acids in the protein of interest,
+and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
+\begin{equation}
+	 mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+\end{equation}
+where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+All translation parameters are given in Table ZZ.
+
 \subsection*{Calculation of energy efficiency.}
-Energy efficiency was calculated as the ratio of protein production to glucose consumption, both in terms of equivalent ATP molecules:
+Energy efficiency ($\mathcal{E}$) was calculated as the ratio of protein production to glucose consumption, both in terms of equivalent ATP molecules:
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	Efficiency=\frac{\Delta\texttt{POI}\cdot \left(2\left({ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot\texttt{ATP}_{GLC}}
+	\mathcal{E}=\frac{\Delta\texttt{POI}\cdot \left(2\left(\rm {ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot\texttt{ATP}_{GLC}}
 \end{equation}
-where $\Delta\texttt{POI}$ denotes the flux of the protein of interest produced, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\Delta\texttt{GLC}$ denotes the glucose flux, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
+where $\Delta\texttt{POI}=w_{TX}$ denotes the translation rate for the protein of interest, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\Delta\texttt{GLC}=w_{GLC}$ denotes the glucose flux, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
 The energy species stoichiometric coefficients are available in the Supporting Information.
@@
 \subsection*{Quantification of uncertainty.}
-An ensemble of 100 sets of flux distributions was calculated for each of the three different cases: control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
-The ensemble was calculated by randomly sampling the maximum specific glucose uptake rate from within a range of 0 to 30 mM/h, determined from experimental data and randomly sampling RNA polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/sec \cite{2005_underwood_biotech, Garamella:2016aa}.
+Experimental factors taken from literature, for example macromolecular concentrations or elongation rates, have uncertainty associated with their values.
+To quantify the influence of this uncertainty on model performance, we randomly sampled the expected physiological ranges for these parameters as determined from literature.
+An ensemble of N = 100 flux distributions was calculated for the three different cases we considered:
+control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
+The flux ensemble was calculated by randomly sampling the maximum glucose consumption rate within a range of 0 to 30 mM/h, (determined from experimental data)
+and randomly sampling RNA polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/s \cite{2005_underwood_biotech, Garamella:2016aa}.
 
 \subsection*{Global sensitivity analysis.}
-We conducted a global sensitivity analysis using the variance-based method of Sobol to estimate which parameters controlled the performance of synthetic circuits \citep{SOBOL_METHOD}.
+We conducted a global sensitivity analysis using the variance-based method of Sobol to estimate which parameters controlled the performance of the cell free protein synthesis reaction \citep{SOBOL_METHOD}.
 We computed the total sensitivity index of each parameter relative to three performance objectives: productivity of the protein of interest, energy efficiency and carbon yield.

### 2017-03-08 Paper_v1.tex
Changes to MM

@@
 \subsubsection{Yield}
-We also calculated the carbon yield for each protein (Fig.~\ref{fig:Yield}A).
+The mean carbon yield was inversely proportional to protein size and varied between 40-55\% for the proteins sampled (Fig.~\ref{fig:Yield}).
 The same trends followed, where the cases with amino acids supplied in the media showed the highest carbon yield.
@@
 
-The bounds on the transcription rate ($\mathcal{L}_{TX}=w_{TX}=\mathcal{U}_{TX}$) was modeled as:
+The bounds on the transcription rate ($\mathcal{L}_{TX}=w_{TX}=\mathcal{U}_{TX}$) were modeled as:
 \begin{equation}

### 2017-03-09 Paper_v1.tex
minor updates

@@
 Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
-We tested this approach by first simulating T7 induced chloramphenicol acetyltransferase production and $\sigma_{70}$-induced deGFP expression; we then expanded these studies for a range of different proteins.
+We tested this approach by first simulating T7 induced chloramphenicol acetyltransferase production and P70a-induced deGFP expression; we then expanded these studies for a range of different proteins.
 First principles predictions of circuit performance were consistent with measurements for a variety of cases.
@@
 Thus, sequence specific constraints based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
-We tested this approach by simulating the production of two model proteins, and then investigated the productivity and carbon yield for eight different proteins.
-From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high carbon yield and the translation rate for productivity.
+We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight different proteins.
+From this, higher carbon number proteins typically had lower productivity rates, energy efficiency, and carbon yields than that of the lower carbon number proteins.
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
@@
 \subsection{Model derivation and validation}
-The cell free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}.
+The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}.
 We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
-A schematic of the metabolic network, which consisted of 272 reactions and 145 species, is shown in Fig.~\ref{fig:network}A.
-Using this network, in combination with detailed promoter models, and literature values for cell free culture parameters (Table ZZ),
+A schematic of the metabolic network, which consisted of 270 reactions and 145 species, is shown in Fig.~\ref{fig:network}A.
+Using this network, in combination with detailed promoter models, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
 we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
-different cell-free \textit{E.~coli} extracts. The cell free metabolic network, model parameters, model code,
+different cell-free \textit{E.~coli} extracts. The cell-free metabolic network, model parameters, model code,
 and each protein sequence are available in the supplemental materials.
 
-Cell free simulations predicted CAT and deGFP production for the duration of the CFPS batch reactions (Fig. ~\ref{fig:network}B and C).
+Cell-free simulations predicted CAT and deGFP production for the duration of the CFPS batch reactions (Fig. ~\ref{fig:network}B and C).
 Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
-With the exception of the first 10-15 min, the cell free prediction of CAT abundance was consistent with the measured values.
-On the other hand, deGFP was produced under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
-The cell free simulation captured the overall trend of deGFP abundance, but was not able to capture saturation at the end of the CFPS culture.
-Uncertainty in experimental factors such as the concentration of RNA polymerase, and ribosomes,
+With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values.
+On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
+The cell-free simulation captured the overall trend of deGFP abundance, but was not able to capture saturation at the end of the CFPS culture.
+Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
 transcription and translation elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
@@
 \end{figure}
-These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with no adjustable parameters.
+These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
 It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
@@
 Our next goal was to examine the performance of CFPS for eight different proteins under three different cases.
-Each of the proteins was produced under a P70 promoter, except for CAT which was produced under a T7 promoter.
+Each of the proteins was produced under a P70a promoter, except for CAT which was produced under a T7 promoter.
 In all cases, CFPS was supplied with glucose.
@@
 Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
-We used ssFBA to estimate the productivity for each of these proteins for each case (Fig.~\ref{fig:Prod}).
-An additional case was considered for CAT, since a comprehensive dataset is available \cite{2005_calhoun_BiotechnologyProgress}; in this case, fluxes were constrained to experimental measurements where avaliable, with the exception of CAT production which was determined by the transcription/translation parameters.
+We used ssFBA to estimate the productivity, energy efficiency, and carbon yield for each of these proteins for each case.
+An additional case was considered for CAT, since a comprehensive dataset is available (Supporting Information); in this case, fluxes were constrained to experimental measurements where avaliable, with the exception of CAT production which was determined by the transcription/translation parameters.
 
@@
 CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
-The difference in the T7 and  $\sigma_{70}$ promoters did not alter the qualitative productivity performance of CAT.
+The difference in the T7 and  P70a promoters did not alter the qualitative productivity performance of CAT.
 The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
@@
 The initial mixture includes 1.2 mM ATP; 0.85 mM each of GTP, UTP, and CTP; 30 mM phosphoenolpyruvate (Roche, Indianapolis IN); 130 mM potassium glutamate; 10 mM ammonium glutamate; 16 mM magnesium glutamate; 50 mM HEPES-KOH buffer (pH 7.5); 1.5 mM spermidine; 1.0 mM putrescine; 34 $\mu$g/mL folinic acid; 170.6 $\mu$g/mL E.
-coli tRNA mixture (Roche, Indianapolis IN); 13.3 $\mu$g/mL pK7CAT plasmid; 100 $\mu$g/mL T7 RNA polymerase; 20 unlabeled amino acids at 2 mM each; 5 $\mu$M l-[U-\textsuperscript{14}C]-leucine (Amersham Pharmacia, Uppsala Sweden); 0.33 mM nicotinamide adenine dinucleotide (NAD); 0.26 mM coenzyme A (CoA); 2.7 mM sodium oxalate; and 0.24 volumes of E.
+coli tRNA mixture (Roche, Indianapolis IN); 13.3 $\mu$g/mL pK7CAT plasmid; 100 $\mu$g/mL T7 RNA polymerase; 20 unlabeled amino acids at 2-3 mM each; 5 $\mu$M l-[U-\textsuperscript{14}C]-leucine (Amersham Pharmacia, Uppsala Sweden); 0.33 mM nicotinamide adenine dinucleotide (NAD); 0.26 mM coenzyme A (CoA); 2.7 mM sodium oxalate; and 0.24 volumes of E.
 coli S30 extract.
-This reaction was modified for the energy source used such that glucose reactions have 30 mM glucose in place of PEP, and glutamate reactions do not have a secondary energy source added since glutamate is already present in high concentrations from the reaction salts.
+This reaction was modified for the energy source used such that glucose reactions have 30-40 mM glucose in place of PEP, and glutamate reactions do not have a secondary energy source added since glutamate is already present in high concentrations from the reaction salts.
 Sodium oxalate was not added since it has a detrimental effect on protein synthesis and ATP concentrations when using glucose or an early glycolytic intermediate as the energy source \cite{BIT:BIT1121}.
@@
 The forms for the promoter models were taken from Moon $\emph{et~al.}$ \cite{Moon:2012ab}.
-In this study, we considered two promoters: T7 and $\sigma_{70}$.
+In this study, we considered two promoters: T7 and P70a.
 The promoter function for the T7 promoter, $u_{T7}$, was given by:
@@
 where $K_{T7}$ denotes a T7 RNA polymerase binding constant.
-The $\sigma_{70}$ promoter function $u_{\sigma_{70}}$ (which was used for all other proteins) was formulated as:
+The P70a promoter function $u_P70a$ (which was used for all other proteins) was formulated as:
 \begin{equation}
-	u_{\sigma_{70}} = \frac{K_{1}+K_{2}f_{\sigma_{70}}}{1 + K_{1}+K_{2}f_{\sigma_{70}}}
+	u_{P70a} = \frac{K_{1}+K_{2}f_{\sigma_{70}}}{1 + K_{1}+K_{2}f_{\sigma_{70}}}
 \end{equation}
@@
 $K_{2}$ denotes the weight of RNAP-$\sigma_{70}$ bound to the promoter,
-and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to RNAP (modeled as a Hill function).
-The values for $K_{T7}, K_{1}$, $K_{2}$ and $f_{\sigma_{70}}$, along with all other promoter parameters, are given in Table ZZ.
+and $f_{p70}$ denotes the fraction of the $\sigma_{70}$ transcription factor bound to RNAP, modeled as a Hill function: 
+\begin{equation}
+	f_{\sigma_{70}} = \frac{\sigma_{70}^{n}}{K_{D}^{n} + \sigma_{70}^{n}}
+\end{equation}
+where $\sigma_{70}$ denotes the sigma-factor 70 ($\sigma_{70}$) concentration, $K_{D}$ denotes the dissociation constant, and $n$ denotes the hill coefficient.
+The values for all promoter parameters are given in Table ~\ref{tbl:parameters}.
 
@@
 where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
-All translation parameters are given in Table ZZ.
+All translation parameters are given in Table \ref{tbl:parameters}.
+
+\begin{table}[!]
+\centering
+    \caption{Parameters for sequence specific flux balance analysis}
+    \renewcommand{\arraystretch}{1}
+    \begin{tabular}{lcccc} \toprule
+        \textbf{Description} & \textbf{Parameter} & \textbf{Value} & \textbf{Units} & \textbf{Reference} \\ \toprule
+        RNA polymerase concentration & $R_{T}$ & 75 & nM & \cite{Garamella:2016aa} \\
+        Transcription rate & $v_{TX}$ & 25 & nt/sec & \cite{Garamella:2016aa} \\
+        Ribsome concentration & $R_{X}$ & 1.6 & $\mu$M & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
+        Translation rate & $v_{TL}$ & 2 & AA/sec & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
+        Transcription saturation coefficient & $K_{TX}$ & 3.5 & nM & estimated \\
+        Translation saturation coefficient & $K_{TL}$ & 0.045 & mM & estimated \\
+        Polysome amplification constant & $K_{P}$ & 10 & constant & estimated \\
+        mRNA degradation rate & $\lambda$ & 5.2 & hr\textsuperscript{-1} & \cite{Garamella:2016aa} \\
+        T7 promoter & $K_{T7}$ & 10 & constant & estimated \\
+        P70a promoter & $K_{1}$ & 0.014 & constant & estimated \\
+        P70a promoter & $K_{2}$ & 10 & constant & estimated \\
+        $\sigma_{70}$ concentration & $\sigma_{70}$ & 35 & nM & \cite{Garamella:2016aa} \\
+        $\sigma_{70}$ dissociation constant & $K_{D}$ & 130 & nM & estimated \\
+	$\sigma_{70}$ hill coefficient & n & 1 & constant & estimated \\
+	Gene concentration & $G$ & 5 & nM &  \cite{Garamella:2016aa} \\
+        Gene length of CAT & $l_{G}$ & 683 & nt & \cite{Kigawa1995} \\
+        Gene length of deGFP & $l_{G}$ & 660 & nt & \cite{Garamella:2016aa} \\
+        Protein length of CAT & $l_{P}$ & 229 & AA & \cite{Kigawa1995} \\
+        Protein length of deGFP & $l_{P}$ & 219 & AA & \cite{Garamella:2016aa} \\ \bottomrule
+    \end{tabular}
+\label{tbl:parameters}
+\end{table}
+
+\clearpage
 
@@
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\mathcal{E}=\frac{\Delta\texttt{POI}\cdot \left(2\left(\rm {ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\Delta\texttt{GLC}\cdot\texttt{ATP}_{GLC}}
+	\mathcal{E}=\frac{\texttt{q}_{POI}\cdot \left(2\left(\rm {ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}
 \end{equation}
-where $\Delta\texttt{POI}=w_{TX}$ denotes the translation rate for the protein of interest, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\Delta\texttt{GLC}=w_{GLC}$ denotes the glucose flux, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
+where $\texttt{q}_{POI}=w_{TX}$ denotes the production rate for the protein of interest, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
 The energy species stoichiometric coefficients are available in the Supporting Information.
@@
 \begin{equation}\label{eqn:yield-definition}
-	Y_{C}^{POI}=\frac{\Delta\texttt{POI}\cdot C_{POI}}{\displaystyle\sum_{i=1}^{\mathcal{R}}\max(\Delta m_{i},0)\cdot C_{m_i}}
+	Y_{C}^{POI}=\frac{\texttt{q}_{POI}\cdot C_{POI}}{\displaystyle\sum_{i=1}^{\mathcal{R}} \texttt{q}_{m_{i}}\cdot C_{m_i}}
 \end{equation}
-where $\Delta\texttt{POI}$ denotes the flux of the protein of interest produced, $C_{POI}$ denotes carbon number of the protein of interest, $\mathcal{R}$ denotes the number of reactants,
-$\Delta m_{i}$ denotes the amount of the $i^{th}$ reactant consumed (not allowed to be negative), and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
+where $\texttt{q}_{POI}$ denotes the flux of the protein of interest produced, $C_{POI}$ denotes carbon number of the protein of interest, $\mathcal{R}$ denotes the number of reactants,
+$\texttt{q}_{m_{i}}$ denotes the uptake flux of the $i^{th}$ reactant, and $C_{m_i}$ denotes the carbon number of the $i^{th}$ reactant.
 
@@
 and randomly sampling RNA polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 7 and 16 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/s \cite{2005_underwood_biotech, Garamella:2016aa}.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 12 and 18 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/s \cite{2005_underwood_biotech, Garamella:2016aa}.
 
 \subsection*{Global sensitivity analysis.}
-We conducted a global sensitivity analysis using the variance-based method of Sobol to estimate which parameters controlled the performance of the cell free protein synthesis reaction \citep{SOBOL_METHOD}.
+We conducted a global sensitivity analysis using the variance-based method of Sobol to estimate which parameters controlled the performance of the cell-free protein synthesis reaction \citep{SOBOL_METHOD}.
 We computed the total sensitivity index of each parameter relative to three performance objectives: productivity of the protein of interest, energy efficiency and carbon yield.
@@
   \item Protein Sequences: DNA and protein sequences of each protein of interest.
-  \item Supporting Information: Performance trendlines as a function of carbon number and transcription/translation stoichiometric coefficients of energy species.
+  \item Supporting Information: Performance trendlines as a function of carbon number, transcription/translation stoichiometric coefficients of energy species, and experimental measurements of CAT production.
   \item Carbon Yield Sensitivity Analysis: Global sensitivity analysis on deGFP carbon yield.
+  \item Metabolites and reactions of the cell-free stoichiometric network.
 \end{itemize}

### 2017-03-10 Paper_v1.tex
draft MV1

@@
 \subsubsection{Productivity}
-We evaluated CFPS productivity for all eight proteins and for each case.
+The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prod}).
 All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prod}A).
@@
 This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
-However, each protein had a different level of productivity.
+However, each individual protein had a different level of productivity.
 For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
@@
 The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
-% a further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
+A further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
 \begin{figure}[t!]
@@
 \end{figure}
-The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation, as well as synthesize each amino acid required for the production of the protein of interest.
+%The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation, as well as synthesize each amino acid required for the production of the protein of interest.
 
 \subsubsection{Energy efficiency}
-Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein (Fig.~\ref{fig:Energy}A).
-The first two cases, where amino acids were supplied in the media, had comparable performance including having the highest energy efficiencies.
-The third case (with no amino acid uptake) had the lowest energy efficiency; this was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis, in addition to being available for energy generation.
+The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 30\% when amino acids were not available (Fig.~\ref{fig:Energy}).
+Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
+When amino acids were supplied in the media (first two cases), there was a comparable performance in having the highest energy efficiency, despite the network's ability to synthesis amino acids.
+When amino acids were removed from the media, protein production resulted in a low energy efficiency; below 30\% depending on the protein. 
+This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis, in addition to being available for energy generation.
 We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Energy}B).
 The same inverse trend was observed as for productivity, except that it was linear.
-The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for the first two cases.
+The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
 Proteins with a high carbon number have a higher transcription and translation cost then smaller proteins, leading to a lower energy efficiency of protein synthesis.
 The first two cases had the same trendline whereas for the third case there was a significant drop in energy efficiency which required its own trendline (available in the Supporting Information).
-Interestingly, in the third case each protein had a similar energy efficiency of about 25\% regardless of carbon number.
+Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 25\% regardless of carbon number.
 In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
-However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.9 $\pm$ 1.5\% compared to the theoretical maximum of 72 $\pm$ 7.1\%.
+However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.3 $\pm$ 1.4\% compared to the theoretical maximum of 72 $\pm$ 7.1\%.
 This shows CFPS systems have a lot of room for improvement: first, the experimental setup still produced certain amino acids; these reactions could be turned off.
@@
 \subsubsection{Yield}
-The mean carbon yield was inversely proportional to protein size and varied between 40-55\% for the proteins sampled (Fig.~\ref{fig:Yield}).
+The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Yield}).
 The same trends followed, where the cases with amino acids supplied in the media showed the highest carbon yield.
@@
 ssFBA assumes a psuedo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
-In addition, ssFBA is solved by setting protein production as the objective function.
-Therefore, carbon flux will travel through the network to optimize the flux through the protein synthesis reaction.
-In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate.
+In addition, ssFBA is solved by setting protein production as the objective function to maximize.
+Therefore, carbon flux will travel through the network to optimize the maximum flux through the protein synthesis reaction.
+In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate, pyruvate and lactate.
 The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.
@@
 Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
-The same inverse qualitative trend was observed as for productivity.
+The same inverse qualitative trend was observed as for energy efficiency.
 The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
-A single trendline was formulated for the first two cases since they had similar performance, while another trendline was formulated for the third cases (available in Supporting Information).
+A single trendline was formulated for the cases where amino acids were supplied in the media (first two cases) since they had similar performance, while another trendline was formulated for when amino acids were not available (available in Supporting Information).
 As the protein size increased, the carbon yield decreased, suggesting that large proteins may be less feasible for cell-free production.
@@
 \subsection{Sensitivity analysis}
+The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield (Fig.~\ref{fig:SI}).
 To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for deGFP, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield (available in Supporting Information).
@@
 Also, oxygen determined how effectively glucose was utilized; therefore, if glucose uptake rate effects productivity, then oxygen could be expected to have a similar effect.
-On the other hand, amino acid uptake showed significance for the first two cases, and was even higher for the second case (without AA synthesis), as it was the only source of amino acids.
+On the other hand, amino acid uptake showed significance for the cases where amino acids were available, and was even higher for the second case (without AA synthesis), as it was the only source of amino acids.
 \begin{figure}[t!]
@@
 Thus, in investigating CFPS performance in terms of productivity and energy efficiency, the translation rate was an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
-Oxygen uptake showed significant importance for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficieny pathway for energy generation.
+Oxygen uptake showed significant importance for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficient pathway for energy generation.
 Meanwhile, productivity was determined primarily by the rate of the most downstream steps, transcription and translation.
-Across all three cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the carbon yield of deGFP.
-In the first two cases where amino acids were supplied, amino acid uptake was significant, as without it energy was used to synthesize amino acids.
-Glucose uptake was only seen to be significant in the third case, since it was the only source of carbon for protein synthesis and energy generation.
+Across all three cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the energy efficiency and carbon yield of deGFP.
+In the first two cases where amino acids were supplied, amino acid uptake was significant, as without it, energy was required to synthesize amino acids for the protein of interest.
+Glucose uptake was only seen to be significant when amino acids were not available in the media (third case), since it was the only source of carbon for protein synthesis and energy generation.
 %This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
@@
 Interestingly, oxidative phosphorylation has a strong effect on the carbon yield in all cases.
-The first two cases follow the same trend, ranging from a carbon yield of 20\% to 55\%, depending on the oxidative phosphorylation activity.
-The third case, followed the same trend; however, its carbon yield ranged from about 10\% to 55\%.
-The third case was expected to have a lower carbon yield for the same oxidative phosphorylation flux compared to the first two cases, since carbon must be utilized for energy generation and amino acid synthesis.
-In all three cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of acetate and lactate, resulting in the lower carbon yield.
+The first two cases where amino acids are supplied followed the same trend: ranging from a carbon yield of 20\% to 55\%, depending on the oxidative phosphorylation activity.
+Once amino acids were removed from the media the carbon yield dropped to about 10\%, but still reached the same maximum of 55\% with higher oxidative phosphorylation flux.
+When amino acids are not avialable in the media, a lower carbon yield was expected for the same oxidative phosphorylation flux, since carbon must be utilized for energy generation and amino acid synthesis.
+In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of acetate and lactate, resulting in the lower carbon yield.
 The experimental dataset exhibits a mixture of acetate and lactate accumulation during CAT synthesis, which shows that CFPS is not operating in a fully aerobic state.
@@
 
-To investigate the differences between the three cases, we compared the flux distributions for a representative protein, deGFP, predicted by ssFBA simulations, as well as the case constrained by experimental measurements for CAT (Fig.~\ref{fig:flux}).
+To investigate the differences between the three cases, we compared the flux distributions for a representative protein, deGFP, predicted by ssFBA simulations (Fig.~\ref{fig:flux}).
 \begin{figure}[t!]
@@
 Once the energy requirements for transcription and translation were met, amino acids were taken from the media to assemble the protein of interest.
-The first two cases had similar performance in terms of productivity, energy efficiency and carbon yield for all proteins.
+These first two cases where amino acids were available had similar performance with a correlation of 0.99 between their flux distributions and were comparable in terms of productivity, energy efficiency and carbon yield for all proteins.
 This is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
-In the third case, where amino acids must be synthesized, there is drop in all performance metrics (productivity, energy efficiency, and carbon yield) compared to the first two cases.
+In the third case, where amino acids must be synthesized, there is drop in the performance metrics of energy efficiency and carbon yield compared to the cases where amino acids are avialble.
+The flux distribution had a correlation fo 0.95 when compared to both cases where amino acids were avaialbe. 
 %Glucose passes through all the major pathways: glycolysis, pentose phosphate pathway and the TCA cycle.
@@
 
-The fourth case, constrained by experimental measurements (Fig. \ref{fig:flux_exp}), had a fairly similar flux distribution as the third case.
+The fourth case, constrained by experimental measurements (Fig.
+\ref{fig:flux_exp}), had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.6 with the cases supplied with amino acids.
+This shows the case constrained by experimental measurements has some differences in the flux distribution compared to the optimum solutions which may provide some insight to improve CFPS performance.
+Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour which constrained the solution space of ssFBA to have a more realistic depiction of the flux distribution.
 The central carbon organic acids show good agreement with the data (Fig. \ref{fig:flux_exp}B).
-Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour which constrained the solution space of ssFBA to have a more realistic depiction of the flux distribution.
-Unlike the second case, only certain amino acid synthesis reactions were blocked since during the growth of \emph{E.~coli} not all amino acids were supplied.
+Only certain amino acid synthesis reactions were blocked since during the growth of \emph{E.~coli} not all amino acids were supplied (see Materials and Methods).
 During the cell-free reaction all amino acids were supplied, however glucose still traveled through all the major pathways, and the same metabolic precursors were still utilized for amino acid biosynthesis.
-Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig. \ref{fig:flux_exp}B), implying an inefficiency of carbon utilization.
+The optimum solutions only produced the required amount of amino acids necessary, however in examining the measurements there is an accumulation of alanine and glutamine which may explain some of the differences in the correlations of the flux distributions.
+Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig. \ref{fig:flux_exp}B) and shows an inefficiency of carbon utilization.
 In addition, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.

### 2017-03-21 Paper_v1.tex
new figs and text updates

@@
 We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
-A schematic of the metabolic network, which consisted of 270 reactions and 145 species, is shown in Fig.~\ref{fig:network}A.
+A schematic of the metabolic network, which consisted of 264 reactions and 146 species, is shown in Fig.~\ref{fig:network}A.
 Using this network, in combination with detailed promoter models, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
@@
 Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.
-
-Next, we predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
-Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
-The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
-However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
 \begin{figure}[t!]
@@
 \end{figure}
+
+Next, we predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
+Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
+The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
+However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
 These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
@@
 CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
-The difference in the T7 and  P70a promoters did not alter the qualitative productivity performance of CAT.
-The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
-A further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
+The relative high productivity of CAT was due to its T7 promoter.
+CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h. 
+Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
+%The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
+% further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
 \begin{figure}[t!]
@@
 Mean productivity versus carbon number.
-Single trendline (dotted line) calculated across all cases (R\textsuperscript{2} = 0.99).
+Single trendline (dotted line) calculated across all cases (y = 6015004x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99).
 Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
@@
 \end{figure}
-%The drop in productivity in the third case is due to the fact that glucose is the only substrate supplied and must be used to provide the necessary energy requirements for transcription and translation, as well as synthesize each amino acid required for the production of the protein of interest.
 
 \subsubsection{Energy efficiency}
-The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 30\% when amino acids were not available (Fig.~\ref{fig:Energy}).
+The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 32\% when amino acids were not available (Fig.~\ref{fig:Energy}).
 Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
 When amino acids were supplied in the media (first two cases), there was a comparable performance in having the highest energy efficiency, despite the network's ability to synthesis amino acids.
-When amino acids were removed from the media, protein production resulted in a low energy efficiency; below 30\% depending on the protein. 
-This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis, in addition to being available for energy generation.
+For the case when amino acids were removed from the media, protein production resulted in a low energy efficiency; below 32\% depending on the protein. 
+This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
 We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Energy}B).
 The same inverse trend was observed as for productivity, except that it was linear.
-The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
-Proteins with a high carbon number have a higher transcription and translation cost then smaller proteins, leading to a lower energy efficiency of protein synthesis.
-The first two cases had the same trendline whereas for the third case there was a significant drop in energy efficiency which required its own trendline (available in the Supporting Information).
-Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 25\% regardless of carbon number.
+%The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
+Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation cost then smaller proteins.
+The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
+Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 28\% regardless of carbon number.
 In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
-However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.3 $\pm$ 1.4\% compared to the theoretical maximum of 72 $\pm$ 7.1\%.
-This shows CFPS systems have a lot of room for improvement: first, the experimental setup still produced certain amino acids; these reactions could be turned off.
+However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.6 $\pm$ 1.1\% compared to the theoretical maximum of 77.3 $\pm$ 7.3\%.
+This shows CFPS systems have a lot of room for improvement: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potential could be diverted towards a protein of interest.
 Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.
@@
 Mean energy efficiency versus the carbon number for each corresponding protein.
-Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.97) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.79).
+Trendline of energy efficiency versus carbon number (black dotted line) for first two cases (y = -1.65$\cdot$10\textsuperscript{-2}x+90.95; R\textsuperscript{2} = 0.91) and trendline for AA synthesis without uptake (grey dotted line; y = -3.18$\cdot$10\textsuperscript{-3}x+33.48; R\textsuperscript{2} = 0.78).
 Dagger: constrained by experimental measurements and excluded from trendline.}
@@
 The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Yield}).
-The same trends followed, where the cases with amino acids supplied in the media showed the highest carbon yield.
-The third case (with no amino acid uptake) had the lowest yields; this is most likely because glucose is utilized to synthesize the necessary amino acids for each protein as well as power the system.
+The same inverse qualitative trend was observed as for energy efficiency.
+%The cases supplemented with amino acids in the media showed the highest carbon yield.
+There was a drop in carbon yield by about 7\% once amino acids were not available; this is most likely because glucose was utilized to synthesize the necessary amino acids for each protein as well as power the system.
 %To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
-For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 56.9 $\pm$ 2.4\% for CAT.
-Once amino acid synthesis was removed from the network (second case), each amino acid was utilized and the carbon yield increased to 57.0 $\pm$ 2.2\% for CAT.
+For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 58.2 $\pm$ 2.3\% for CAT.
+Once amino acid synthesis reactions were blocked in the network (second case), the carbon yield dropped to 57.1 $\pm$ 2.2\% for CAT.
 Only the necessary amount of amino acids was used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
 In that case, the carbon yield without glucose contribution would be 100\%.
-Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
-In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
-In the experimentally constrained case, CAT was produced with a carbon yield of 6\% compared to the theoretical maximum of 57\%.
+%Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
+%In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
+In the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58.2\%.
 This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
 ssFBA assumes a psuedo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
-In addition, ssFBA is solved by setting protein production as the objective function to maximize.
+In addition, ssFBA is solved by maximizing the flux through the protein production reaction.
 Therefore, carbon flux will travel through the network to optimize the maximum flux through the protein synthesis reaction.
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/CarbonYield.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Yield.pdf}
 \caption{CFPS carbon yield of eight proteins for four cases.
@@
 Mean energy efficiency versus the carbon number for each corresponding protein.
-Trendline of energy efficiency versus carbon number (black dotted line) for first two cases  (R\textsuperscript{2} = 0.92) and trendline for AA synthesis without uptake (grey dotted line; R\textsuperscript{2} = 0.83).
+Trendline of energy efficiency versus carbon number (black dotted line) for first two cases (y = -8.03$\cdot$10\textsuperscript{-3}x+63.83; R\textsuperscript{2} = 0.95) and trendline for AA synthesis without uptake (grey dotted line; y = -6.98$\cdot$10\textsuperscript{-3}x+58.86; R\textsuperscript{2} = 0.90).
 Dagger: constrained by experimental measurements and excluded from trendline.}
@@
 \end{figure}
-Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
-The same inverse qualitative trend was observed as for energy efficiency.
-The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
-A single trendline was formulated for the cases where amino acids were supplied in the media (first two cases) since they had similar performance, while another trendline was formulated for when amino acids were not available (available in Supporting Information).
-As the protein size increased, the carbon yield decreased, suggesting that large proteins may be less feasible for cell-free production.
+%Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
+%The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
+A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}B).
+The trendlines showed good predictability for estimating carbon yield for a certain range of potein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
+However, as the protein size increased, the carbon yield decreased, suggesting that larger proteins may be less feasible for cell-free production.
 Thus, we examined the parameters that had the most significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.
@@
 The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield (Fig.~\ref{fig:SI}).
-To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for deGFP, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield (available in Supporting Information).
-In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation by ribosomes being the most significant.
+To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for CAT, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield which followed the same trend of significance as energy efficiency (available in Supporting Information).
+In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.
 As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
@@
 have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
-In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the first two cases, but its significance increased when amino acids were removed from CFPS.
-This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis in that case.
-In addition, oxygen uptake had more significance in the third case than in the first two cases, since in that case oxygen was required to synthesize amino acids and power the system.
-Also, oxygen determined how effectively glucose was utilized; therefore, if glucose uptake rate effects productivity, then oxygen could be expected to have a similar effect.
-On the other hand, amino acid uptake showed significance for the cases where amino acids were available, and was even higher for the second case (without AA synthesis), as it was the only source of amino acids.
+In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the cases with amino acid supplementation, but its significance increased when amino acids were not available.
+This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
+On the other hand, amino acid uptake only showed significance for the case where amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
 \begin{figure}[t!]
@@
 
-When considering energy efficiency performance (Fig.~\ref{fig:SI}B), the oxygen uptake rates were the most important for all three cases while the sensitivity to transcription/translation parameters decreased slightly.
+When considering energy efficiency performance (Fig.~\ref{fig:SI}B), the oxygen uptake rates were the most important for all cases while the sensitivity of the transcription/translation parameters decreased.
 The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
-Thus, in investigating CFPS performance in terms of productivity and energy efficiency, the translation rate was an important parameter to optimize, as has already been shown in literature \cite{2005_underwood_biotech, 2014_li_PlosOne}.
-Oxygen uptake showed significant importance for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficient pathway for energy generation.
-Meanwhile, productivity was determined primarily by the rate of the most downstream steps, transcription and translation.
-Across all three cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the energy efficiency and carbon yield of deGFP.
-In the first two cases where amino acids were supplied, amino acid uptake was significant, as without it, energy was required to synthesize amino acids for the protein of interest.
-Glucose uptake was only seen to be significant when amino acids were not available in the media (third case), since it was the only source of carbon for protein synthesis and energy generation.
+Oxygen uptake was the most significant for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficient pathway for energy generation.
+Across all cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the energy efficiency of CAT synthesis.
+In cases where amino acids were supplied, amino acid uptake was significant, as without it, energy was required to synthesize amino acids for the protein of interest.
+Glucose uptake was only important when amino acids were not available, since it was the only source of carbon for protein synthesis and energy generation.
 %This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
 Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in pyruvate-powered CFPS \cite{Jewett:2008aa}.
-We also investigated the sensitivity of carbon yield to network fluxes; it followed the same trends as the energy efficiency sensitivity analysis.
 It is unknown how active oxidative phosphorylation is compared to in \textit{in~vivo} systems.
-To investigate this further we compared deGFP carbon yield to oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
-Interestingly, oxidative phosphorylation has a strong effect on the carbon yield in all cases.
-The first two cases where amino acids are supplied followed the same trend: ranging from a carbon yield of 20\% to 55\%, depending on the oxidative phosphorylation activity.
-Once amino acids were removed from the media the carbon yield dropped to about 10\%, but still reached the same maximum of 55\% with higher oxidative phosphorylation flux.
-When amino acids are not avialable in the media, a lower carbon yield was expected for the same oxidative phosphorylation flux, since carbon must be utilized for energy generation and amino acid synthesis.
+To investigate this further we compared CAT carbon yield to oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
+Interestingly, oxidative phosphorylation had a strong effect on the carbon yield in all cases.
+The cases where amino acids were supplied followed the same trend: ranging from a carbon yield of 20\% to 58\%, depending on the oxidative phosphorylation activity.
+Once amino acids were removed from the media, the carbon yield dropped to about 10\%, and reached a maximum of 52\%.
+When amino acids are not avialable in the media, a lower carbon yield was expected for the same oxidative phosphorylation flux, since carbon must be utilized for energy generation and amino acid biosynthesis.
 In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of acetate and lactate, resulting in the lower carbon yield.
 The experimental dataset exhibits a mixture of acetate and lactate accumulation during CAT synthesis, which shows that CFPS is not operating in a fully aerobic state.
-It is unclear how active oxidative phosphorylation is in CFPS, since the reactions rely on electron transport from membrane vesicles.
+Glucose can not be fully oxidized and therefore fermentation pathways are used. 
+Oxidative phosphorylation relies on electron transport from membrane vesicles, however CFPS has no cell membrane, thus it is expected CFPS has limited oxidative phosphorylation activity.
 The addition of phosphate showed an increase in CAT yield; however, it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
@@
 \begin{figure}[t!]
-\includegraphics[width=0.7\textwidth]{./Figures/Yield_Ox.pdf}
+\includegraphics[width=0.7\textwidth]{./Figures/Ox_yield.pdf}
 \caption{deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 ssFBA solutions, for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
@@
 
-To investigate the differences between the three cases, we compared the flux distributions for a representative protein, deGFP, predicted by ssFBA simulations (Fig.~\ref{fig:flux}).
+\subsection{Flux distribution}
+To investigate the differences between the three cases, we compared the flux distributions for CAT production predicted by ssFBA simulations (Fig.~\ref{fig:flux}).
 \begin{figure}[t!]
@@
 \end{figure}
-The first case, which was supplied with amino acids and could synthesize them in the network, relied on a combination of glucose and amino acids to power the system (Fig.~\ref{fig:flux}A).
-Glucose traveled through glycolysis and the Entner-Doudoroff pathway to generate NADH for oxidative phosphorylation as well as utilize pyruvate for amino acid biosynthesis.
-Interestingly, amino acids rather than glucose powered the TCA cycle, via alpha-ketoglutarate and fumarate, and by utilizing oxaloacetic acid for additional amino acid biosynthesis.
-Thus, ssFBA found that the combined utilization of glucose and amino acids to energize CFPS was the optimal case for protein synthesis.
-In the second case, where amino acid synthesis was removed from the network, glucose was solely utilized to provide the necessary energy requirements via glycolysis which generated enough NADH for oxidative phosphorylation.
-Ubiquinone was generated via \textit{nuo} to power oxidative phosphorylation, instead of relying on the TCA cycle.
+In the cases supplied with amino acids (Fig.~\ref{fig:flux}A-B), glucose traveled to acetyl-coenzyme A that generated NADH for oxidative phosphorylation to energize CFPS.
+Interestingly, in the first case with amino acid synthesis reactions, ssFBA relied on a combination of glucose and amino acids to power the system
+Amino acids rather than glucose powered the TCA cycle via glutamate to alpha-ketoglutarate which traveled to oxaloacetic acid and pyruvate for additional amino acid biosynthesis.
+In the case where amino acid synthesis reactions were blocked (Fig.~\ref{fig:flux}B), ubiquinone was generated via \textit{nuo} to power oxidative phosphorylation, instead of relying on the TCA cycle.
 Once the energy requirements for transcription and translation were met, amino acids were taken from the media to assemble the protein of interest.
 These first two cases where amino acids were available had similar performance with a correlation of 0.99 between their flux distributions and were comparable in terms of productivity, energy efficiency and carbon yield for all proteins.
-This is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
-In the third case, where amino acids must be synthesized, there is drop in the performance metrics of energy efficiency and carbon yield compared to the cases where amino acids are avialble.
-The flux distribution had a correlation fo 0.95 when compared to both cases where amino acids were avaialbe. 
-%Glucose passes through all the major pathways: glycolysis, pentose phosphate pathway and the TCA cycle.
-%First, glucose must be converted into all the necessary amino acids for the assembly of the protein of interest.
-%Second, it must provide the necessary energy requirements for transcription and translation.
+%This is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
+In the case where amino acids must be synthesized, there was a drop in the performance metrics of energy efficiency and carbon yield compared to the cases where amino acids are avialble.
+The flux distribution had a correlation of 0.90 when compared to both cases where amino acids were avaialbe. 
 This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
-This leads to the relatively high flux for the conversion of NADH to NADPH.
-Thus, less NADH is available for oxidative phosphorylation.
-%, thus requiring glucose to energize CFPS as well as synthesize amino acids results in a decrease in performance.
+This leads to the relatively high flux in the first step of the pentose phosphate pathway to generate NADPH, thus, less NADH is available for oxidative phosphorylation.
 The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
-Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
+%Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
 
-The fourth case, constrained by experimental measurements (Fig.
-\ref{fig:flux_exp}), had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.6 with the cases supplied with amino acids.
-This shows the case constrained by experimental measurements has some differences in the flux distribution compared to the optimum solutions which may provide some insight to improve CFPS performance.
+The case constrained by experimental measurements (Fig.
+\ref{fig:flux_exp}), had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.52 with the cases supplied with amino acids.
+Thus there are some differences in the flux distribution compared to the optimum solutions which may provide some insight to improve CFPS performance. 
 Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour which constrained the solution space of ssFBA to have a more realistic depiction of the flux distribution.
-The central carbon organic acids show good agreement with the data (Fig. \ref{fig:flux_exp}B).
+The central carbon organic acids showed good agreement with the data (Fig. \ref{fig:flux_exp}B).
 Only certain amino acid synthesis reactions were blocked since during the growth of \emph{E.~coli} not all amino acids were supplied (see Materials and Methods).
 During the cell-free reaction all amino acids were supplied, however glucose still traveled through all the major pathways, and the same metabolic precursors were still utilized for amino acid biosynthesis.
+In this case, it is unclear which substrate (glucose or amino acids) is used to power CFPS and may in fact be a combination of both.
 The optimum solutions only produced the required amount of amino acids necessary, however in examining the measurements there is an accumulation of alanine and glutamine which may explain some of the differences in the correlations of the flux distributions.
@@
 In addition, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
-To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the intracellular flux distribution \cite{Henry:2007,Hamilton:2013}.
-In this case, it is unclear which substrate (glucose or amino acids) is used to power CFPS and may in fact be a combination of both.
-Thus, it would be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
+Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
+To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
+It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
 \begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./Figures/ssFBA_exp.pdf}
+\includegraphics[width=1.00\textwidth]{./Figures/Exp_flux.pdf}
 \caption{ssFBA simulation of CAT production for an experimentally constrained case.
@@
 
-S30 extract was prepared from E. coli strain KC6 (A19 $\delta$tonA $\delta$tnaA $\delta$speA $\delta$endA $\delta$sdaA $\delta$sdaB $\delta$gshA met+).
+S30 extract was prepared from E. coli strain KC6 (A19 $\Delta$tonA $\Delta$tnaA $\Delta$speA $\Delta$endA $\Delta$sdaA $\Delta$sdaB $\Delta$gshA met+).
 This K12-derivative has several gene deletions to stabilize amino acid concentrations during the cell-free reaction.
@@
 and $\mathcal{L}_{i}$ and $\mathcal{U}_{i}$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The transcription (TX) and translation (TL) stoichiometry was modeled using the template reactions of Allen and Palsson \cite{Allen:2003aa}:
+The transcription (TX) and translation (TL) stoichiometry was modeled based on the template reactions of Allen and Palsson \cite{Allen:2003aa}:
 \begin{eqnarray*}
 {G}_{\mathcal{P}}+{R}_{T} &\longrightarrow& G_{\mathcal{P}}^{*} \\
-G_{\mathcal{P}}^{*}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left\{k\right\}TP &\overset{\rm TX}{\longrightarrow}& mRNA+G_{\mathcal{P}}+R_{T}+ \sum_{k\in\left\{A,C,G,U\right\}}2 \eta_{k} \cdot Pi\\
+G_{\mathcal{P}}^{*}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right) &\overset{\rm TX}{\longrightarrow}& mRNA+G_{\mathcal{P}}+R_{T}+ \sum_{k\in\left\{A,C,G,U\right\}} \eta_{k} \cdot PPi\\
 mRNA &\longrightarrow& \sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP \\
 mRNA+R_{X} &\longrightarrow& R_{X}^{*} \\
-\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP &\longrightarrow& \alpha_{j}\cdot AA_{j}-tRNA_{j}+\\
-&& \qquad \alpha_{j}\cdot AMP+2\alpha_{j}\cdot Pi \qquad{j=1,2,\hdots,20}\\
-R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}-tRNA_{j}+2\cdot GTP\Big) &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+R_{X}+mRNA+\\
-&& \quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2\cdot GDP+2\cdot Pi\Big)
+%\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP +\alpha_{j}\cdot H_{2}O &\longrightarrow& \alpha_{j}\cdot AA_{j}\text{-}tRNA_{j}+\\
+%&& \qquad \alpha_{j}\cdot AMP+\alpha_{j}\cdot PPi \qquad{j=1,2,\hdots,20}\\
+\alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big) &\longrightarrow& \alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)\\
+&& \qquad \qquad \qquad \qquad {j=1,2,\hdots,20}\\
+R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big) &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+R_{X}+mRNA\\
+&& \quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)
 \end{eqnarray*}

### 2017-03-23 Paper_v1.tex
Norm figure

@@
 Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour which constrained the solution space of ssFBA to have a more realistic depiction of the flux distribution.
-The central carbon organic acids showed good agreement with the data (Fig. \ref{fig:flux_exp}B).
+The central carbon organic acids showed good agreement with the data (Fig.~\ref{fig:flux_exp}B).
 Only certain amino acid synthesis reactions were blocked since during the growth of \emph{E.~coli} not all amino acids were supplied (see Materials and Methods).
@@
 The optimum solutions only produced the required amount of amino acids necessary, however in examining the measurements there is an accumulation of alanine and glutamine which may explain some of the differences in the correlations of the flux distributions.
-Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig. \ref{fig:flux_exp}B) and shows an inefficiency of carbon utilization.
-In addition, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
-Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
-To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
-It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
+Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig.~\ref{fig:flux_exp}B) and shows an inefficiency of carbon utilization.
 \begin{figure}[t!]
@@
 \label{fig:flux_exp}
+\end{figure}
+
+Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm}) 
+For example, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
+A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
+In addition, pairwise knockouts of Entner-Doudoroff and most subgroups in the network result in the same optimal solution of CAT productivity.
+However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B), ssFBA will reroute the flux to optimize the objective function.
+Interestingly, a single group knockout of glycolysis/gluconoeogenesis, glutamate/glutamine biosynthesis, alanine/aspartate/asparagine biosynthesis was detrimental to CAT productivity.
+Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
+In our case, ssFBA reached the same optimal solution of CAT productivity for 73\% of the pairwise group knockouts. 
+To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
+It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
+\begin{figure}[t!]
+\includegraphics[width=1.0\textwidth]{./Figures/Norm.pdf}
+\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case.
+(B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
+\label{fig:norm}
 \end{figure}
@@
 
+\subsection*{Pairwise group knockouts.}
+Pairwise and single group knouckouts were simulated in ssFBA by setting the flux bounds for the all the reactions in a group to zero. 
+We grouped reactions in the cell-free network into 19 subgroups (available in Supporting Information).
+We computed the norm of the productivity of CAT for each pairwise knockout compared to the productivity of CAT with no knockouts.
+We also computed the norm of the flux distribution for each pairwise knockout compared to the flux distribution with no knockouts.
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2017-03-30 Paper_v1.tex
Some initial changes

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\title{Sequence Specific Modeling of Cell-Free Protein Synthesis}
+\title{Sequence Specific Modeling of \emph{E.~coli} Cell-Free Protein Synthesis}
 %\abbreviations{MV,NH,JDV}
@@
 \begin{abstract}
-In this study, we used sequence specific constraints based modeling to evaluate the performance of synthetic circuits in an \emph{E.~coli} cell-free protein synthesis system.
-A core \emph{E.~coli} metabolic model, consisting of glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with sequence specific descriptions of genetic circuits which included mechanistic models of promoter function, transcription and translation.
-Thus, unlike other synthetic biology modeling efforts, sequence specific constraints based modeling explicitly couples the transcription and translation of circuit components with the availability of metabolic resources.
-Model parameters for transcription and translation were taken from literature, allowing a first principles prediction of circuit performance.
-We tested this approach by first simulating T7 induced chloramphenicol acetyltransferase production and P70a-induced deGFP expression; we then expanded these studies for a range of different proteins.
-First principles predictions of circuit performance were consistent with measurements for a variety of cases.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance in terms of productivity, energy efficiency, and carbon yield.
-A sufficient energy supply with oxidative phosphorylation is instrumental for high energy efficiency and carbon yields; in addition, the translation rate could be optimized for higher productivity.
-Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
+Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized medicine.
+In this study, we used sequence specific constraint based modeling to evaluate the performance of an \emph{E.~coli} cell-free protein synthesis system.
+A core \emph{E.~coli} metabolic model, describing glycolysis, the pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism,
+was augmented with sequence specific descriptions of transcription and translation processes, and effective models of promoter function.
+Thus, sequence specific constraint based modeling explicitly couples transcription and translation processes, and the regulation of gene expression,
+with the availability of metabolic resources.
+We tested this approach by simulating the expression of two model proteins chloramphenicol acetyltransferase and dual emission green fluorescent protein for which we have training data sets;
+we then expanded the simulations to a range of therapeutically relevant proteins.
+Protein expression simulations were consistent with measurements for a variety of cases.
+Further, global sensitivity analysis identified the key metabolic processes that controlled the productivity, energy efficiency, and carbon yield of the process.
+Taken together, sequence specific constraint based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
 \end{abstract}
+
+% A sufficient energy supply with oxidative phosphorylation is instrumental for high energy efficiency and carbon yields; in addition, the translation rate could be optimized for higher productivity.
 
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./Figures/ssFBA_network.pdf}
+\includegraphics[width=1.00\textwidth]{./figs/Fig-1-Network-Validation-Simulations.pdf}
 \caption{Sequence specific flux balance analysis.
@@
 
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Analysis.pdf}
+\caption{CFPS productivity of eight proteins for four cases.
+Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
+A.
+Productivity across the ensemble (error bars represent 95\% CI).
+B.
+Mean productivity versus carbon number.
+Single trendline (dotted line) calculated across all cases (y = 6015004x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99).
+Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
+\label{fig:Prod}
+\end{figure}
+
 \subsubsection{Productivity}
@@
 % further study on the nucleotide and amino acid requirements of each protein and its effect on CFPS performance should be investigated.
-\begin{figure}[t!]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/Productivity.pdf}
-\caption{CFPS productivity of eight proteins for four cases.
-Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
-A.
-Productivity across the ensemble (error bars represent 95\% CI).
-B.
-Mean productivity versus carbon number.
-Single trendline (dotted line) calculated across all cases (y = 6015004x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99).
-Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
-\label{fig:Prod}
-\end{figure}
 
@@
 %This may be a potential area for improvement in CFPS to optimize for energy utilization.
-\begin{figure}[t!]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
-\caption{CFPS energy efficiency of eight proteins for four cases.
+% \begin{figure}[t!]
+% \centering
+% \includegraphics[width=1.00\textwidth]{./Figures/Energy.pdf}
+% \caption{CFPS energy efficiency of eight proteins for four cases.
 Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
@@
 Dagger: constrained by experimental measurements and excluded from trendline.}
-\label{fig:Energy}
-\end{figure}
+% \label{fig:Energy}
+% \end{figure}
 
@@
 The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.
-\begin{figure}[t!]
-\centering
-\includegraphics[width=1.00\textwidth]{./Figures/Yield.pdf}
-\caption{CFPS carbon yield of eight proteins for four cases.
+% \begin{figure}[t!]
+% \centering
+% \includegraphics[width=1.00\textwidth]{./Figures/Yield.pdf}
+% \caption{CFPS carbon yield of eight proteins for four cases.
 Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
@@
 Dagger: constrained by experimental measurements and excluded from trendline.}
-\label{fig:Yield}
-\end{figure}
+% \label{fig:Yield}
+% \end{figure}
 %Next we investigated the effect of the carbon number of each protein on the carbon yield (Fig.~\ref{fig:Yield}B).
@@
 \begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./Figures/Sensitivity.pdf}
+\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
 \caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).
@@
 \begin{figure}[t!]
-\includegraphics[width=0.7\textwidth]{./Figures/Ox_yield.pdf}
+\includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
 \caption{deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 ssFBA solutions, for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
@@
 \begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./Figures/Flux.pdf}
+\includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
 \caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation, for three different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, and (C) amino acid synthesis without uptake.
@@
 \begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./Figures/Exp_flux.pdf}
+\includegraphics[width=1.00\textwidth]{./figs/Fig-6-FluxDistribition-Experimental.pdf}
 \caption{ssFBA simulation of CAT production for an experimentally constrained case.
@@
 \begin{figure}[t!]
-\includegraphics[width=1.0\textwidth]{./Figures/Norm.pdf}
+\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribition-Analysis.pdf}
 \caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case.
@@
 \subsection*{Glucose/NMP cell-free protein synthesis.}
-The glucose/NMP cell-free protein synthesis reactions were performed at a volume of 15 $\mu$L in 1.5-mL eppendorf tubes and incubated in a humidified incubator.
-Plasmid pK7CAT was used as the DNA template for chloramphenical acetyl transferase (CAT) expression by placing the \emph{cat} gene between the T7 promoter and the T7 terminator \cite{Kigawa1995}.
+The glucose/NMP cell-free protein synthesis reaction was
+performed using the S30 extract in 1.5-mL Eppendorf tubes (working volume of 15 $\mu$L) and incubated in a humidified incubator.
+The S30 extract was prepared from \textit{E.~coli} strain KC6 (A19 $\Delta$tonA $\Delta$tnaA $\Delta$speA $\Delta$endA $\Delta$sdaA $\Delta$sdaB $\Delta$gshA met+).
+This K12-derivative has several gene deletions to stabilize amino acid concentrations during the cell-free reaction.
+The KC6 strain was grown to approximately 3.0 OD595 in a 10-L fermenter (B.
+Braun, Allentown PA) on defined media with glucose as the carbon source and with the addition of 13 amino acids (alanine, arginine, cysteine, serine, aspartate, glutamate, and glutamine were excluded) \cite{Zawada:2003}.
+Crude S30 extract was prepared as described previously \cite{Jewett:2002}.
+Plasmid pK7CAT was used as the DNA template for chloramphenical acetyl transferase (CAT) expression by placing the \emph{cat} gene
+between the T7 promoter and the T7 terminator \cite{Kigawa1995}.
 The plasmid was isolated and purified using a Plasmid Maxi Kit (Qiagen, Valencia CA).
-Cell-free protein synthesis of CAT protein was performed at 37 $^{\circ}$C for 3 h.
+Cell-free CAT synthesis was performed at 37 $^{\circ}$C.
 
-S30 extract was prepared from E. coli strain KC6 (A19 $\Delta$tonA $\Delta$tnaA $\Delta$speA $\Delta$endA $\Delta$sdaA $\Delta$sdaB $\Delta$gshA met+).
-This K12-derivative has several gene deletions to stabilize amino acid concentrations during the cell-free reaction.
-The KC6 strain was grown to approximately 3 OD595 in a 10-L fermenter (B.
-Braun, Allentown PA) on defined media with glucose as the carbon source and with the addition of 13 amino acids (alanine, arginine, cysteine, serine, aspartate, glutamate, and glutamine were excluded) \cite{Zawada:2003}.
-Crude S30 extract was made as previously described \cite{Jewett:2002}.
-
-The standard protein synthesis reaction was conducted according to the PANOxSP protocol with slight modifications from that described previously \cite{BIT:BIT20026}.
+The protein synthesis reaction was conducted according to the PANOxSP protocol with slight modifications from that described previously \cite{BIT:BIT20026}.
 Unless otherwise noted, all reagents were purchased from Sigma (St. Louis, MO).
-The initial mixture includes 1.2 mM ATP; 0.85 mM each of GTP, UTP, and CTP; 30 mM phosphoenolpyruvate (Roche, Indianapolis IN); 130 mM potassium glutamate; 10 mM ammonium glutamate; 16 mM magnesium glutamate; 50 mM HEPES-KOH buffer (pH 7.5); 1.5 mM spermidine; 1.0 mM putrescine; 34 $\mu$g/mL folinic acid; 170.6 $\mu$g/mL E.
-coli tRNA mixture (Roche, Indianapolis IN); 13.3 $\mu$g/mL pK7CAT plasmid; 100 $\mu$g/mL T7 RNA polymerase; 20 unlabeled amino acids at 2-3 mM each; 5 $\mu$M l-[U-\textsuperscript{14}C]-leucine (Amersham Pharmacia, Uppsala Sweden); 0.33 mM nicotinamide adenine dinucleotide (NAD); 0.26 mM coenzyme A (CoA); 2.7 mM sodium oxalate; and 0.24 volumes of E.
-coli S30 extract.
-This reaction was modified for the energy source used such that glucose reactions have 30-40 mM glucose in place of PEP, and glutamate reactions do not have a secondary energy source added since glutamate is already present in high concentrations from the reaction salts.
-Sodium oxalate was not added since it has a detrimental effect on protein synthesis and ATP concentrations when using glucose or an early glycolytic intermediate as the energy source \cite{BIT:BIT1121}.
+The initial mixture included 1.2 mM ATP; 0.85 mM each of GTP, UTP, and CTP; 30 mM phosphoenolpyruvate (Roche, Indianapolis IN); 130 mM potassium glutamate; 10 mM ammonium glutamate; 16 mM magnesium glutamate; 50 mM HEPES-KOH buffer (pH 7.5); 1.5 mM spermidine; 1.0 mM putrescine; 34 $\mu$g/mL folinic acid; 170.6 $\mu$g/mL \textit{E.~coli} tRNA mixture (Roche, Indianapolis IN); 13.3 $\mu$g/mL pK7CAT plasmid; 100 $\mu$g/mL T7 RNA polymerase; 20 unlabeled amino acids at 2-3 mM each; 5 $\mu$M l-[U-\textsuperscript{14}C]-leucine (Amersham Pharmacia, Uppsala Sweden); 0.33 mM nicotinamide adenine dinucleotide (NAD); 0.26 mM coenzyme A (CoA); 2.7 mM sodium oxalate; and 0.24 volumes of \textit{E.~coli} S30 extract.
+This reaction was modified for the energy source used such that glucose reactions have 30-40 mM glucose in place of PEP.
+Sodium oxalate was not added since it has a detrimental effect on protein synthesis and ATP concentrations when using glucose or other early glycolytic intermediate energy sources \cite{BIT:BIT1121}.
 The HEPES buffer (pKa $\sim$ 7.5) was replaced with Bis-Tris (pKa $\sim$ 6.5).
-In addition, the magnesium glutamate concentration was reduced to 8 mM for the glucose and glutamate reactions since a lower magnesium optimum is found when using a nonphosphorylated energy source \cite{BIT:BIT20026}.
-Finally, 10 mM phosphate was added to some reactions as described below in the form of potassium phosphate dibasic adjusted to pH 7.2 with acetic acid.
+In addition, the magnesium glutamate concentration was reduced to 8 mM for the glucose reaction since a lower magnesium optimum was found when using a nonphosphorylated energy source \cite{BIT:BIT20026}.
+Finally, 10 mM phosphate was added in the form of potassium phosphate dibasic adjusted to pH 7.2 with acetic acid.
 
@@
 Compounds were identified and quantified by comparison to known standards for retention time and UV absorbance (260 nm for nucleotides and 210 nm for organic acids) as described previously \cite{2005_calhoun_BiotechnologyProgress}.
-Calibration standards were routinely run for improved quantification of compounds.
 The standard compounds quantified with a refractive index detector included inorganic phosphate, glucose, and acetate.
-The compounds quantified with the UV detector included pyruvate, malate, succinate, and lactate.
+Pyruvate, malate, succinate, and lactate were quantified with the UV detector.
 The stability of the amino acids in the cell extract was determined using a Dionex Amino Acid Analysis (AAA) HPLC System (Sunnyvale, CA) that separates amino acids by gradient anion exchange (AminoPac PA10 column).
@@
  \begin{multlined}
-	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{TL} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
+	\qquad \qquad \qquad \max_{\boldsymbol{w}}{} \! \left( w_{X} = \mathbf{\boldsymbol{\theta}}^T \boldsymbol{w} \right) \\
 	\mathrm{Subject \; to:}
@@
 and $\mathcal{L}_{i}$ and $\mathcal{U}_{i}$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The transcription (TX) and translation (TL) stoichiometry was modeled based on the template reactions of Allen and Palsson \cite{Allen:2003aa}:
-\begin{eqnarray*}
-{G}_{\mathcal{P}}+{R}_{T} &\longrightarrow& G_{\mathcal{P}}^{*} \\
-G_{\mathcal{P}}^{*}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right) &\overset{\rm TX}{\longrightarrow}& mRNA+G_{\mathcal{P}}+R_{T}+ \sum_{k\in\left\{A,C,G,U\right\}} \eta_{k} \cdot PPi\\
-mRNA &\longrightarrow& \sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP \\
-mRNA+R_{X} &\longrightarrow& R_{X}^{*} \\
-%\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP +\alpha_{j}\cdot H_{2}O &\longrightarrow& \alpha_{j}\cdot AA_{j}\text{-}tRNA_{j}+\\
-%&& \qquad \alpha_{j}\cdot AMP+\alpha_{j}\cdot PPi \qquad{j=1,2,\hdots,20}\\
-\alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big) &\longrightarrow& \alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)\\
-&& \qquad \qquad \qquad \qquad {j=1,2,\hdots,20}\\
-R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big) &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+R_{X}+mRNA\\
-&& \quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)
-\end{eqnarray*}
+The transcription (T) and translation (X) stoichiometry was modeled based on the template reactions of Allen and Palsson \cite{Allen:2003aa}:
+
+\begin{table}[!]
+\centering
+    \caption{Transcription and translation template reactions for protein production.}
+    \renewcommand{\arraystretch}{1.2}
+    \resizebox{\textwidth}{!}{\begin{tabular}{llcl} \toprule
+        \textbf{Description} & \multicolumn{3}{l}{\textbf{Template reaction}} \\ \toprule
+        Transcription initiation & $G_\mathcal{P}+R_{T}$ & $\longrightarrow$& $G_{\mathcal{P}}^{*}$ \\
+        Transcription ($w_{T}$) & $G_{\mathcal{P}}^{*}+\quad\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right)$ & $\longrightarrow$& $mRNA+G_{\mathcal{P}}+R_{T}+\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}} \eta_{k} \cdot PPi$ \\
+        mRNA degradation &  $mRNA$ &$\longrightarrow$& \quad $\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left\{k\right\}MP$ \\
+	Translation initiation & $mRNA+R_{X}$ &$\longrightarrow$& $R_{X}^{*}$ \\
+        tRNA charging & $\alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big)$ &$\longrightarrow$& $\alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)$ \\
+ 	&&& $\qquad \qquad \qquad \qquad \quad  {j=1,2,\hdots,20}$\\
+        Translation ($w_{X}$) &  $R_{X}^{*}+\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big)$ &$\longrightarrow$& $\mathcal{P}+R_{X}+mRNA$\\
+&&& $\quad +\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)$ \\ \bottomrule
+    \end{tabular}}
+\label{tbl:parameters}
+\end{table}
 where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
@@
 $R_{X}$ denotes the ribosome concentration,
-$R_{X}^{*}$ denotes bounded ribosome,
+$R_{X}^{*}$ denotes bound ribosome,
 and $AA_{j}$ denotes $j^{th}$ amino acid.
 
-The objective of the sequence specific flux balance calculation was to maximize the rate of protein translation, $w_{TL}$.
+% \begin{eqnarray*}
+% {G}_{\mathcal{P}}+{R}_{T} &\longrightarrow& G_{\mathcal{P}}^{*} \\
+% G_{\mathcal{P}}^{*}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right) &\overset{\rm TX}{\longrightarrow}& mRNA+G_{\mathcal{P}}+R_{T}+ \sum_{k\in\left\{A,C,G,U\right\}} \eta_{k} \cdot PPi\\
+% mRNA &\longrightarrow& \sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP \\
+% mRNA+R_{X} &\longrightarrow& R_{X}^{*} \\
+% %\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP +\alpha_{j}\cdot H_{2}O &\longrightarrow& \alpha_{j}\cdot AA_{j}\text{-}tRNA_{j}+\\
+% %&& \qquad \alpha_{j}\cdot AMP+\alpha_{j}\cdot PPi \qquad{j=1,2,\hdots,20}\\
+% \alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big) &\longrightarrow& \alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)\\
+% && \qquad \qquad \qquad \qquad {j=1,2,\hdots,20}\\
+% R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big) &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+R_{X}+mRNA\\
+% && \quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)
+% \end{eqnarray*}
+
+The objective of the sequence specific flux balance calculation was to maximize the rate of protein translation, $w_{X}$.
 The total glucose uptake rate was bounded by [0,40 mM/h] according to experimental data, while the amino acid uptake rates were bounded by [0,30 mM/h], but did not reach the maximum flux.
@@
 The sequence specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.55 \cite{GLPK}.
-For all cases, amino acid degradation reactions were blocked since these enzymes were knocked out during the cell-free extract preparation \cite{2005_calhoun_BiotechnologyProgress, Garamella:2016aa}.
+For all cases, amino acid degradation reactions were blocked as these enzymes were knocked out during the cell-free extract preparation \cite{2005_calhoun_BiotechnologyProgress, Garamella:2016aa}.
 In the second case, all amino acid synthesis reactions were set to 0 mM/hr since \textit{E.
@@
 
-The bounds on the transcription rate ($\mathcal{L}_{TX}=w_{TX}=\mathcal{U}_{TX}$) were modeled as:
+The bounds on the transcription rate ($\mathcal{L}_{T}=w_{T}=\mathcal{U}_{T}$) were modeled as:
 \begin{equation}
-	w_{TX} = V_{TX}^{max}\left(\frac{G}{K_{TX}+G}\right)
+	w_{T} = V_{T}^{max}\left(\frac{G}{K_{T}+G}\right)
 \end{equation}
-where $G$ denotes the gene concentration and $K_{TX}$ denotes a transcription saturation coefficient.
-The maximum rate of transcription $V_{TX}^{max}$ was formulated as:
+where $G$ denotes the gene concentration and $K_{T}$ denotes a transcription saturation coefficient.
+The maximum transcription rate $V_{T}^{max}$ was formulated as:
 \begin{equation}
-	V_{TX}^{max} \equiv \left[R_{T}\left(\frac{v_{TX}}{l_{G}}\right)u\left(\kappa\right)\right]
+	V_{T}^{max} \equiv \left[R_{T}\left(\frac{\dot{v}_{T}}{l_{G}}\right)u\left(\kappa\right)\right]
 \end{equation}
 The term $R_{T}$ denotes the RNA polymerase concentration (nM),
-$v_{TX}$ denotes the RNA polymerase elongation rate (nt/h),
+$\dot{v}_{T}$ denotes the RNA polymerase elongation rate (nt/h),
 $l_{G}$ denotes the gene length in nucleotides (nt).
-The term $u\left(\kappa\right)$ (dimensionless, $0\leq u\left(\kappa\right)\leq 1$) describes a model of promoter activity, where $\kappa$ denotes promoter specific parameters.
-The forms for the promoter models were taken from Moon $\emph{et~al.}$ \cite{Moon:2012ab}.
+The term $u\left(\kappa\right)$ (dimensionless, $0\leq u\left(\kappa\right)\leq 1$)
+is an effective model of promoter activity, where $\kappa$ denotes promoter specific parameters.
+The general form for the promoter models was taken from Moon $\emph{et~al.}$ \cite{Moon:2012ab}.
 In this study, we considered two promoters: T7 and P70a.
@@
 where $K_{T7}$ denotes a T7 RNA polymerase binding constant.
-The P70a promoter function $u_P70a$ (which was used for all other proteins) was formulated as:
+The P70a promoter function $u_{P70a}$ (which was used for all other proteins) was formulated as:
 \begin{equation}
@@
 \end{equation}
-where $\sigma_{70}$ denotes the sigma-factor 70 ($\sigma_{70}$) concentration, $K_{D}$ denotes the dissociation constant, and $n$ denotes the hill coefficient.
+where $\sigma_{70}$ denotes the sigma-factor 70 concentration, $K_{D}$ denotes the dissociation constant, and $n$ denotes a cooperativity coefficient.
 The values for all promoter parameters are given in Table ~\ref{tbl:parameters}.
 
-The translation rate ($w_{TL}$) was bounded by:
+The translation rate ($w_{X}$) was bounded by:
  \begin{equation}
-	0\leq w_{TL} \leq V_{TL}^{max}\left(\frac{mRNA_{SS}}{K_{TL}+mRNA_{SS}}\right)
+	0\leq w_{X} \leq V_{X}^{max}\left(\frac{mRNA^{*}}{K_{X}+mRNA^{*}}\right)
 \end{equation}
-where $\rm mRNA_{SS}$ denotes the steady state mRNA abundance and $K_{TL}$ denotes the translation saturation constant.
-The maximum translation rate $V_{TL}^{max}$ was formulated as:
+where $\rm mRNA^{*}$ denotes the steady state mRNA abundance and $K_{X}$ denotes a translation saturation constant.
+The maximum translation rate $V_{X}^{max}$ was formulated as:
 \begin{equation}
-	V_{TL}^{max} \equiv \left[K_{P} R_{X}\left(\frac{v_{TL}}{l_{P}}\right)\right]
+	V_{X}^{max} \equiv \left[K_{P} R_{X}\left(\frac{\dot{v}_{X}}{l_{P}}\right)\right]
 \end{equation}
 The term $K_{P}$ denotes the polysome amplification constant,
-$v_{TL}$ denotes the ribosome elongation rate (amino acids per hour),
-$l_{P}$ denotes the number of amino acids in the protein of interest,
-and $\rm mRNA_{SS}$ denotes the steady-state mRNA concentration:
+$\dot{v}_{X}$ denotes the ribosome elongation rate (amino acids per hour),
+$l_{P}$ denotes the number of amino acids in the protein of interest.
+The steady-state mRNA abundance $\rm mRNA^{*}$ was estimated as:
 \begin{equation}
-	 mRNA_{SS}\simeq\frac{w_{TX}}{\lambda}
+	 mRNA^{*}\simeq\frac{w_{T}}{\lambda}
 \end{equation}
@@
         RNA polymerase concentration & $R_{T}$ & 75 & nM & \cite{Garamella:2016aa} \\
-        Transcription rate & $v_{TX}$ & 25 & nt/sec & \cite{Garamella:2016aa} \\
-        Ribsome concentration & $R_{X}$ & 1.6 & $\mu$M & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
-        Translation rate & $v_{TL}$ & 2 & AA/sec & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
+        Ribosome concentration & $R_{X}$ & 1.6 & $\mu$M & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
+
+        & & & & \\
+        Transcription elongation rate & $\dot{v}_{TX}$ & 25 & nt/sec & \cite{Garamella:2016aa} \\
+        Translation elongation rate & $\dot{v}_{TL}$ & 2 & aa/sec & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
         Transcription saturation coefficient & $K_{TX}$ & 3.5 & nM & estimated \\
@@
         mRNA degradation rate & $\lambda$ & 5.2 & hr\textsuperscript{-1} & \cite{Garamella:2016aa} \\
+
+        & & & & \\
         T7 promoter & $K_{T7}$ & 10 & constant & estimated \\
@@
 	      $\sigma_{70}$ hill coefficient & n & 1 & constant & estimated \\
+
+        & & & & \\
 	      Gene concentration & $G$ & 5 & nM &  \cite{Garamella:2016aa} \\
@@
         Gene length of deGFP & $l_{G}$ & 660 & nt & \cite{Garamella:2016aa} \\
-        Protein length of CAT & $l_{P}$ & 229 & AA & \cite{Kigawa1995} \\
-        Protein length of deGFP & $l_{P}$ & 219 & AA & \cite{Garamella:2016aa} \\ \bottomrule
+        Protein length of CAT & $l_{P}$ & 229 & aa & \cite{Kigawa1995} \\
+        Protein length of deGFP & $l_{P}$ & 219 & aa & \cite{Garamella:2016aa} \\ \bottomrule
     \end{tabular}

### 2017-03-31 Paper_v1.tex
Changes in the materials and methods

@@
 
-The protein synthesis reaction was conducted according to the PANOxSP protocol with slight modifications from that described previously \cite{BIT:BIT20026}.
+The protein synthesis reaction was conducted using the PANOxSP protocol with slight modifications from that described previously \cite{BIT:BIT20026}.
 Unless otherwise noted, all reagents were purchased from Sigma (St. Louis, MO).
@@
 
-\subsection*{Measurements of protein product and metabolites.}
+\subsection*{Protein product and metabolite measurements.}
 Cell-free reaction samples were quenched at specific timepoints with equal volumes of ice-cold 150 mM sulfuric acid to precipitate proteins.
@@
 \end{equation}
-where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective selection vector
+where $\mathbf{S}$ denotes the stoichiometric matrix, $\mathbf{w}$ denotes the unknown flux vector, $\boldsymbol{\theta}$ denotes the objective cost vector
 and $\mathcal{L}_{i}$ and $\mathcal{U}_{i}$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The transcription (T) and translation (X) stoichiometry was modeled based on the template reactions of Allen and Palsson \cite{Allen:2003aa}:
+The transcription (T) and translation (X) stoichiometry was modeled based upon the template reactions of Allen and Palsson \cite{Allen:2003aa}:
 
@@
 \end{equation}
-where $\lambda$ denotes the rate constant controlling the mRNA degradation rate.
+where $\lambda$ denotes the rate constant controlling the mRNA degradation rate (hr$^{-1}$).
 All translation parameters are given in Table \ref{tbl:parameters}.
@@
         Transcription saturation coefficient & $K_{TX}$ & 3.5 & nM & estimated \\
-        Translation saturation coefficient & $K_{TL}$ & 0.045 & mM & estimated \\
+        Translation saturation coefficient & $K_{TL}$ & 45.0 & $\mu$M & estimated \\
         Polysome amplification constant & $K_{P}$ & 10 & constant & estimated \\
@@
         T7 promoter & $K_{T7}$ & 10 & constant & estimated \\
-        P70a promoter & $K_{1}$ & 0.014 & constant & estimated \\
-        P70a promoter & $K_{2}$ & 10 & constant & estimated \\
+        Weight RNA polymerase binding alone P70a & $K_{1}$ & 0.014 & constant & estimated \\
+        Weight bound RNAP-$\sigma_{70}$ P70a & $K_{2}$ & 10 & constant & estimated \\
         $\sigma_{70}$ concentration & $\sigma_{70}$ & 35 & nM & \cite{Garamella:2016aa} \\

### 2017-03-31 Paper_v1.tex
updated intro

@@
 Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
-However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing, we must first understand the performance limits of these systems.
+However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
 One tool to address this question is mathematical modeling.
@@
 Stoichiometric reconstructions of microbial metabolism, popularized by flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
+Constraints based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
+These constraints based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} of arbitrary complexity, including genome scale networks, using linear programming\cite{Covert:2004aa}.
 Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B.
 subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
+Stoichiometric reconstructions have been expanded to include metabolism integrated with detailed gene expression (ME-Model) \cite{Allen:2003aa, Brien693}, and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220} which has increased the scope of simulating genetic and environmental perturbations.
+%However, FBA has been shown to have alternative solutions that lead to the same desired optimal phenotype using flux variability analysis \cite{Mahadevan2003264} or mixed-integer linear programming based algorithms \cite{LEE2000711}.
+Constraints based methods are great tools to estimate the performance of metabolic networks with none to very few parameters. 
+However, constraints based methods are typically used to model \textit{in~vivo} processes and have not been applied to cell-free metabolism.
+
 In this study, we used sequence specific constraints based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
@@
 \includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Analysis.pdf}
-\caption{CFPS productivity of eight proteins for four cases.
+\caption{CFPS performance of eight proteins for four cases.
 Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
 A.
-Productivity across the ensemble (error bars represent 95\% CI).
+Productivity.
 B.
 Mean productivity versus carbon number.
-Single trendline (dotted line) calculated across all cases (y = 6015004x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99).
-Asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
-\label{fig:Prod}
+Single trendline (dotted line) calculated across all cases (y = 6.02$\cdot$10\textsuperscript{6}x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99).
+C.
+Energy efficiency.
+D.
+Mean energy efficiency versus carbon number.
+Trendline for cases with amino acids (black dotted line; y = -1.65$\cdot$10\textsuperscript{-2}x+90.95; R\textsuperscript{2} = 0.91) and trendline for without amino acids (grey dotted line; y = -3.18$\cdot$10\textsuperscript{-3}x+33.48; R\textsuperscript{2} = 0.78).
+E.
+Carbon yield.
+F.
+Mean carbon yield versus carbon number.
+Trendline for cases with amino acids (black dotted line; y = -8.03$\cdot$10\textsuperscript{-3}x+63.83; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; y = -6.98$\cdot$10\textsuperscript{-3}x+58.86; R\textsuperscript{2} = 0.90).
+Error bars: 95\% CI; asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
+\label{fig:Prof}
 \end{figure}
@@
 \subsubsection{Productivity}
-The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prod}).
-All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prod}A).
+The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
+All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prof}A).
 %with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
@@
 For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
-To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prod}B).
+To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
 The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
@@
 \subsubsection{Energy efficiency}
-The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 32\% when amino acids were not available (Fig.~\ref{fig:Energy}).
+The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 32\% when amino acids were not available (Fig.~\ref{fig:Prof}C-D).
 Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
@@
 This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
-We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Energy}B).
+We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Prof}D).
 The same inverse trend was observed as for productivity, except that it was linear.
@@
 \subsubsection{Yield}
-The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Yield}).
+The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
 The same inverse qualitative trend was observed as for energy efficiency.
@@
 %The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
-A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}B).
+A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}F).
 The trendlines showed good predictability for estimating carbon yield for a certain range of potein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
@@
 A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
-In addition, pairwise knockouts of Entner-Doudoroff and most subgroups in the network result in the same optimal solution of CAT productivity.
-However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B), ssFBA will reroute the flux to optimize the objective function.
-Interestingly, a single group knockout of glycolysis/gluconoeogenesis, glutamate/glutamine biosynthesis, alanine/aspartate/asparagine biosynthesis was detrimental to CAT productivity.
+In addition, pairwise knockouts of Entner-Doudoroff and other subgroups (i.e.
+pentose phosphate pathway, cofactors, folate metabolism, etc.) in the network result in the same optimal solution of CAT productivity.
+However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); ssFBA will reroute the flux to optimize the objective function.
+Interestingly, there are crucial subgroups that are detrimental to CAT productivity. 
+For example, a single group knockout of glycolysis/gluconoeogenesis or oxidative phosphorylation leads to no CAT production. 
+There were also cases of group knockouts that had no effect on productivity or its flux distribution such as isoleucine, leucine, and valine biosynthesis knocked out with histidine biosynthesis. 
 Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
-In our case, ssFBA reached the same optimal solution of CAT productivity for 73\% of the pairwise group knockouts.
+In our case, ssFBA reached the same optimal solution of CAT productivity for 39.9\% of the pairwise group knockouts, but had a different flux distribution than compared to no knockouts.
 To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.

### 2017-04-01 Paper_v1.tex
Working my way through the results

@@
 However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
-One tool to address this question is mathematical modeling.
+One tool to address these questions is mathematical modeling.
 
@@
 
-Stoichiometric reconstructions of microbial metabolism, popularized by flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
-Constraints based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
-These constraints based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} of arbitrary complexity, including genome scale networks, using linear programming\cite{Covert:2004aa}.
+Stoichiometric reconstructions of microbial metabolism, popularized by constraint based approaches such as flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
+Constraint based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
+Constraint based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} for biochemical networks of arbitrary complexity, including genome scale networks, using linear programming \cite{Covert:2004aa}.
 Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B.
 subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
-Stoichiometric reconstructions have been expanded to include metabolism integrated with detailed gene expression (ME-Model) \cite{Allen:2003aa, Brien693}, and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220} which has increased the scope of simulating genetic and environmental perturbations.
+Stoichiometric reconstructions have been expanded to include the integration of with detailed descriptions of gene expression (ME-Model) \cite{Allen:2003aa, Brien693},
+and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220}.
+These expansion have greatly expanded the scope of questions these models can explore.
 %However, FBA has been shown to have alternative solutions that lead to the same desired optimal phenotype using flux variability analysis \cite{Mahadevan2003264} or mixed-integer linear programming based algorithms \cite{LEE2000711}.
-Constraints based methods are great tools to estimate the performance of metabolic networks with none to very few parameters. 
-However, constraints based methods are typically used to model \textit{in~vivo} processes and have not been applied to cell-free metabolism.
+Constraint based methods are powerful tools to estimate the performance of metabolic networks with very few adjustable parameters.
+However, constraint based methods are typically used to model \textit{in~vivo} processes, and have not yet been applied to cell-free metabolism.
 
-In this study, we used sequence specific constraints based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
+In this study, we used sequence specific constraint based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
 A core \emph{E.~coli} cell-free metabolic model was developed from literature \cite{Feist:2007aa}.
@@
 \subsection{Model derivation and validation}
-The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa}.
+The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa},
+and adding deletions associated with the specific cell free system (see materials and methods).
 We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
 A schematic of the metabolic network, which consisted of 264 reactions and 146 species, is shown in Fig.~\ref{fig:network}A.
-Using this network, in combination with detailed promoter models, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
+The network described the major carbon and energy pathways, as well as amino acid biosynthesis and degradation pathways.
+Using this network, in combination with effective promoter models taken from Moon et al. [REF], and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
 we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
-different cell-free \textit{E.~coli} extracts. The cell-free metabolic network, model parameters, model code,
-and each protein sequence are available in the supplemental materials.
+different \textit{E.~coli} cell-free extracts.
+The cell free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab GitHub site [REF].
 
-Cell-free simulations predicted CAT and deGFP production for the duration of the CFPS batch reactions (Fig. ~\ref{fig:network}B and C).
-Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
-With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values.
+Cell free simulations of the time evolution of CAT and deGFP production were consistent with experimental measurements (Fig. ~\ref{fig:network}B and C).
+Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress}
+for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
+With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values (Correlation?).
 On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
-The cell-free simulation captured the overall trend of deGFP abundance, but was not able to capture saturation at the end of the CFPS culture.
+The cell-free simulation captured the overall trend of deGFP abundance, but failed to capture saturation at the end of the CFPS culture (Correlation?).
 Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
-transcription and translation elongation rates and the upper bounds for oxygen and glucose consumption rates did not alter the qualitative performance of the model.
-Thus, the metabolic network and molecular description of transcription and translation were consistent with experimental measurements.
+transcription and translation elongation rates or the upper bounds for oxygen and glucose consumption rates (modeled as being normally distributed around the parameter values shown in Table ~\ref{tbl:parameters})
+did not qualitatively alter the performance of the model (blue region, 95\% confidence estimate).
+Together, these simulations suggested the description of transcription and translation, and its integration with metabolism encoded in the cell free metabolic model, were consistent with experimental measurements.
+However, these simulations were conducted at a single plasmid concentration.
+Thus, it was unclear if the model could capture cell free protein synthesis for a range of plasmid concentrations.
+
 \begin{figure}[t!]
@@
 
-Next, we predicted deGFP production as a function of plasmid concentration (Fig.~\ref{fig:network}D).
-Concentration of deGFP at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
-The mean of the ensemble shows a good prediction against the measured deGFP levels, even though it under predicted deGFP concentration at the saturating point of 5 nM of plasmid concentration.
-However, the ensemble and the mean of the ensemble captured the overall saturating dynamics of deGFP production as a function of plasmid concentration.
+Cell free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
+The cell free deGFP titer at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
+The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration.
+However, while the mean and 95\% confidence estimate of the ensemble was consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating point of 5 nM of plasmid concentration.
 These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
@@
 
+%These amino acid synthesis reactions were blocked since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
+% We used ssFBA to estimate the productivity, energy efficiency, and carbon yield for each of these proteins for each case.
+
 \subsection{Analysis of CFPS performance}
-Our next goal was to examine the performance of CFPS for eight different proteins under three different cases.
-Each of the proteins was produced under a P70a promoter, except for CAT which was produced under a T7 promoter.
-In all cases, CFPS was supplied with glucose.
-In the first case, CFPS was supplied with amino acids, and the system was allowed to synthesize amino acids (AA uptake and synthesis).
-In the second case, CFPS was supplied with amino acids, but the amino acid synthesis reactions were turned off (AA uptake w/o synthesis).
-These amino acid synthesis reactions were blocked since during the cell-free extract preparation the cells are often supplied with amino acids; thus, the enzymes responsible for amino acid synthesis would not be present.
-In the third case, CFPS was not supplied with amino acids, but the system could synthesize them (AA synthesis w/o uptake).
-Eight different proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
-We used ssFBA to estimate the productivity, energy efficiency, and carbon yield for each of these proteins for each case.
-An additional case was considered for CAT, since a comprehensive dataset is available (Supporting Information); in this case, fluxes were constrained to experimental measurements where avaliable, with the exception of CAT production which was determined by the transcription/translation parameters.
+To better understand the performance of CFPS reactions, we predicted the productivity, energy efficiency and carbon yield for the cell free production of eight model proteins with and without amino acid supplementation (Fig.
+\ref{fig:Prof}).
+The expression of each of these proteins was under a P70a promoter, with the exception of CAT which we assumed was expressed using a T7 promoter.
+In all cases, CFPS was supplied with glucose, however, we also considered different scenarios for amino acid supplementation.
+First, the CFPS reaction was supplied with amino acids, and the system was also allowed to synthesize amino acids from glucose (AA uptake and synthesis).
+Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AA uptake w/o synthesis).
+This scenario was consistent with common cell free extract preparation protocols which often involve amino acid supplementation; thus, we would expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction.
+Lastly, the system was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
+Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
+chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP),
+prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
+An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements (see supplemental materials).
 
@@
 \subsubsection{Productivity}
-The mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
+The theoretical maximum mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
 All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prof}A).
@@
 
+%To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
+%The cases supplemented with amino acids in the media showed the highest carbon yield.
+%Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
+%In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
+
 \subsubsection{Yield}
-The mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
+The theoretical maximum mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
 The same inverse qualitative trend was observed as for energy efficiency.
-%The cases supplemented with amino acids in the media showed the highest carbon yield.
 There was a drop in carbon yield by about 7\% once amino acids were not available; this is most likely because glucose was utilized to synthesize the necessary amino acids for each protein as well as power the system.
-%To determine the carbon contribution from each substrate (glucose and amino acids) we examined the carbon flux going toward the production of deGFP for all three cases (Table \ref{tbl:yield_breakdown}).
 For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 58.2 $\pm$ 2.3\% for CAT.
@@
 In that case, the carbon yield without glucose contribution would be 100\%.
-%Finally, for the third case (without amino acids supplemented), the carbon yield was reduced to 51.1 $\pm$ 1.8\% for CAT, and the system used about twice the amount of glucose as in the first two cases.
-%In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
 In the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58.2\%.
@@
 %The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
-A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Yield}F).
-The trendlines showed good predictability for estimating carbon yield for a certain range of potein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
+A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Prof}F).
+The trendlines showed good predictability for estimating carbon yield for a certain range of protein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
 However, as the protein size increased, the carbon yield decreased, suggesting that larger proteins may be less feasible for cell-free production.
-Thus, we examined the parameters that had the most significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.
+% Thus, we examined the parameters that had the most significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.
 
 \subsection{Sensitivity analysis}
-The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield (Fig.~\ref{fig:SI}).
-To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance we performed global sensitivity analysis on the productivity and energy efficiency for CAT, a representative protein (Fig.~\ref{fig:SI}), as well as on the carbon yield which followed the same trend of significance as energy efficiency (available in Supporting Information).
+To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for CAT, a representative protein (Fig.~\ref{fig:SI}).
+The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield.
 In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.

### 2017-04-06 Paper_v1.tex
Need a break ...

@@
 The expression of each of these proteins was under a P70a promoter, with the exception of CAT which we assumed was expressed using a T7 promoter.
-In all cases, CFPS was supplied with glucose, however, we also considered different scenarios for amino acid supplementation.
-First, the CFPS reaction was supplied with amino acids, and the system was also allowed to synthesize amino acids from glucose (AA uptake and synthesis).
+In all cases, CFPS was supplied with glucose; however, we considered different scenarios for amino acid supplementation.
+First, the CFPS reaction was supplied with amino acids, and simultaneously the system was also allowed to synthesize amino acids from glucose (AA uptake and synthesis).
 Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AA uptake w/o synthesis).
-This scenario was consistent with common cell free extract preparation protocols which often involve amino acid supplementation; thus, we would expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction.
-Lastly, the system was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
+This scenario was consistent with common cell free extract preparation protocols which often involve amino acid supplementation; thus, we expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction in these cases.
+Lastly, the cell free reaction was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
 Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
@@
 The theoretical maximum mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
-All cases had very similar performance for each protein and were within a standard deviation of each other (Fig.~\ref{fig:Prof}A).
-%with the second case (without amino acid synthesis) having a slightly higher productivity for most proteins.
-The model framework is setup to optimize for the production of each protein and is constrained by the translation rate.
-This shows the system had sufficient substrates and metabolic precursors to power CFPS and synthesize each protein of interest with the same productivity, regardless of the case.
-However, each individual protein had a different level of productivity.
-For instance, BMP10 had a productivity of about 2.5 $\mu$M/h whereas CAT had a productivity of about 12 $\mu$M/h.
+The theoretical maximum productivity was within a standard deviation with and without
+amino acid supplementation for each protein, but varied significantly between proteins.
+This suggested the cell free reaction had sufficient metabolic precursors to power CFPS and synthesize each protein of interest with similar productivities.
+However, the theoretical productivity was inversely proportional to protein size;
+for instance, BMP10 (XX aa) had a productivity of about 2.5 $\mu$M/h whereas CAT (ZZ aa) had a productivity of about 12 $\mu$M/h.
 To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
-The proteins with the highest productivity had the lowest carbon number, whereas proteins with low productivity had higher carbon numbers.
-This inverse trend was due to the fact that larger proteins require more amino acids and substrates to assemble them, resulting in lower productivity given the same resources.
-A single trendline for all cases shows the expected productivity in CFPS depending on the carbon number of the protein of interest (available in the Supporting Information).
-CAT was an outlier for the trendline, even though it was in the same order of magnitude as the trendline.
-The relative high productivity of CAT was due to its T7 promoter.
-CAT on a P70a promoter frollowed the same trendline as the other proteins and had a productivity of about 8.8 $\mu$M/h.
-Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a P70a promoter in CFPS, given the size of the protein.
+The productivity and protein size were related by the power-law relationship $\alpha\times(L)^{\beta}$, where $\alpha = 6.0\times 10^{6}$ and $\beta=-1.9$.
+Interestingly, CAT did not obey the trendline; the relatively high productivity of CAT was due to its T7 promoter.
+CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.8 $\mu$M/h.
+Taken together, the maximum productivity of a cell free reaction was found to be inversely proportional to protein size,
+following a power-law relationship for proteins expressed under a P70a promoter.
+
+% Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a  in CFPS, given the size of the protein.
 %The higher productivity of CAT compared to all other proteins was most likely due to the lower transcription requirement of cytidine triphosphate which allowed a higher flux for translation.
@@
 \subsection{Sensitivity analysis}
-To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for CAT, a representative protein (Fig.~\ref{fig:SI}).
+To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI}).
 The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield.
@@
 
-\subsection{Flux distribution}
-To investigate the differences between the three cases, we compared the flux distributions for CAT production predicted by ssFBA simulations (Fig.~\ref{fig:flux}).
+\subsection{Optimal Metabolic Flux Distribution}
+Amino acid supplementation altered the optimal metabolic flux distribution predicted for CAT production (Fig.~\ref{fig:flux}).
+To investigate the influence of amino acid supplementation, we compared the simulated metabolic flux distributions for CAT production with and without external amino acids.
 \begin{figure}[t!]
@@
 \end{figure}
-In the cases supplied with amino acids (Fig.~\ref{fig:flux}A-B), glucose traveled to acetyl-coenzyme A that generated NADH for oxidative phosphorylation to energize CFPS.
-Interestingly, in the first case with amino acid synthesis reactions, ssFBA relied on a combination of glucose and amino acids to power the system
-Amino acids rather than glucose powered the TCA cycle via glutamate to alpha-ketoglutarate which traveled to oxaloacetic acid and pyruvate for additional amino acid biosynthesis.
-In the case where amino acid synthesis reactions were blocked (Fig.~\ref{fig:flux}B), ubiquinone was generated via \textit{nuo} to power oxidative phosphorylation, instead of relying on the TCA cycle.
-Once the energy requirements for transcription and translation were met, amino acids were taken from the media to assemble the protein of interest.
-These first two cases where amino acids were available had similar performance with a correlation of 0.99 between their flux distributions and were comparable in terms of productivity, energy efficiency and carbon yield for all proteins.
-%This is most likely due to the efficient utilization of carbon to energize CFPS without the burden of amino acid biosynthesis.
-In the case where amino acids must be synthesized, there was a drop in the performance metrics of energy efficiency and carbon yield compared to the cases where amino acids are avialble.
-The flux distribution had a correlation of 0.90 when compared to both cases where amino acids were avaialbe.
-This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
-This leads to the relatively high flux in the first step of the pentose phosphate pathway to generate NADPH, thus, less NADH is available for oxidative phosphorylation.
-The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
-%Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
+In the presence of amino acid supplementation, and \textit{de~novo} amino acid synthesis, there was an incomplete TCA cycle, where a combination
+of glucose and amino acids powered protein expression (Fig.~\ref{fig:flux}A).
+Glucose was consumed to produce acetyl-coenzyme A, and associated by-products,
+while glutamate was converted to alpha-ketoglutarate which traveled to oxaloacetic acid and pyruvate for additional amino acid biosynthesis.
+On the other hand, in the presence of amino acid supplementation, but without \textit{de~novo} amino acid biosynthesis, there was no TCA cycle flux.
+In this case, ATP was produced by a combination of substrate level and oxidative phosphorylation, where
+ubiquinone was regenerated via \textit{nuo} activity, without relying on succinate dehydrogenase in the TCA cycle (Fig.~\ref{fig:flux}B).
+These first two cases where amino acids were available had similar performance, and their respective metabolic flux distributions had a 99\% correlation for all proteins.
+However, in the absence of amino acid supplementation (where all amino acids were synthesized \textit{de~novo} from glucose),
+the energy efficiency and carbon yield decreased; in this case the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C).
+[Mike: we need a link sentence to the experimentally constrained case.]
+
+% to both cases where amino acids were avaialbe.
+% This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
+% This leads to the relatively high flux in the first step of the pentose phosphate pathway to generate NADPH, thus, less NADH is available for oxidative phosphorylation.
+% The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
+% %Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
 
@@
 
+\subsubsection{Potential Alternative Metabolic Optima}
 Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
@@
 
-Taken together, we developed a sequence specific constraints based modeling approach to evaluate the performance of synthetic circuits in an \emph{E.~coli} CFPS system for a range of different proteins and three different cases.
-We have shown first principle predictions for protein production of deGFP and CAT in agreement with experimental measurements, under two different promoters and two different cell-free extract systems, with few adjustable parameters in the promoter models taken from literature.
-This modeling approach suggested trends for productivity, energy efficiency and carbon yield as a function of carbon number.
+\subsection{Summary and conclusions}
+In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell free protein synthesis reactions for a range of proteins.
+We showed first principle predictions for protein production of deGFP and CAT were in agreement with experimental measurements, under two different promoters and two different cell-free extract systems.
+This modeling approach suggested correlations for productivity, energy efficiency and carbon yield as a function of the size of the protein.
 Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
@@
 The transcription (T) and translation (X) stoichiometry was modeled based upon the template reactions of Allen and Palsson \cite{Allen:2003aa}:
-
 \begin{table}[!]
@@
 and $AA_{j}$ denotes $j^{th}$ amino acid.
-
-% \begin{eqnarray*}
-% {G}_{\mathcal{P}}+{R}_{T} &\longrightarrow& G_{\mathcal{P}}^{*} \\
-% G_{\mathcal{P}}^{*}+\sum_{\mathrm{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right) &\overset{\rm TX}{\longrightarrow}& mRNA+G_{\mathcal{P}}+R_{T}+ \sum_{k\in\left\{A,C,G,U\right\}} \eta_{k} \cdot PPi\\
-% mRNA &\longrightarrow& \sum_{k\in\left\{A,C,G,U\right\}}\eta_{k}\cdot \left\{k\right\}MP \\
-% mRNA+R_{X} &\longrightarrow& R_{X}^{*} \\
-% %\alpha_{j}\cdot AA_{j}+\alpha_{j}\cdot tRNA+\alpha_{j}\cdot ATP +\alpha_{j}\cdot H_{2}O &\longrightarrow& \alpha_{j}\cdot AA_{j}\text{-}tRNA_{j}+\\
-% %&& \qquad \alpha_{j}\cdot AMP+\alpha_{j}\cdot PPi \qquad{j=1,2,\hdots,20}\\
-% \alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big) &\longrightarrow& \alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)\\
-% && \qquad \qquad \qquad \qquad {j=1,2,\hdots,20}\\
-% R_{X}^{*}+\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big) &\stackrel{\rm TL}{\longrightarrow}& \mathcal{P}+R_{X}+mRNA\\
-% && \quad +\sum_{j\in\left\{AA\right\}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)
-% \end{eqnarray*}
 

### 2017-04-06 Paper_v1.tex
Working on sensitivity results

@@
 \subsubsection{Productivity}
-The theoretical maximum mean productivity was inversely proportional to the protein size and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
+The theoretical maximum mean productivity was inversely proportional to the protein length ($l_{P}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
 The theoretical maximum productivity was within a standard deviation with and without
@@
 To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
-The productivity and protein size were related by the power-law relationship $\alpha\times(L)^{\beta}$, where $\alpha = 6.0\times 10^{6}$ and $\beta=-1.9$.
-Interestingly, CAT did not obey the trendline; the relatively high productivity of CAT was due to its T7 promoter.
+The productivity and protein size were related by the power-law relationship $\alpha\times(l_{P})^{\beta}$, where $\alpha = 6.0\times 10^{6}$ and $\beta=-1.9$.
+Interestingly, CAT did not obey the power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
 CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.8 $\mu$M/h.
@@
 To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI}).
-The translation rate had the highest effect on protein productivity, whereas oxygen followed by substrate uptake had the highest effect on energy efficiency and carbon yield.
-In examining productivity performance (Fig.~\ref{fig:SI}A), the significance of transcription/translation parameters was fairly constant across all three cases, with the rate of translation being the most significant.
-As expected, this showed that the translation rate was instrumental for productivity, and should be the first step investigated during optimization, prior to examining transcription parameters.
-Underwood and coworkers have also shown that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
+Surprisingly, RNAP and ribosome abundance had only a modest effect;
+the translation elongation rate had the largest effect on protein productivity.
+On the other hand, oxygen followed by glucose consumption had the largest influence on the energy efficiency and carbon yield, respectively.
+The significance of transcription/translation parameters was robust to amino acid supplementation, with the translation rate being the most sensitive across all cases (Fig.~\ref{fig:SI}A).
+This suggested that the translation elongation rate, and not transcriptional parameters, controlled productivity.
+Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
 In addition, Li et al.
-have increased productivity of firefly luciferase by 5-fold in CFPS systems by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
-In examining substrate utilization, glucose uptake was not seen to be very important for productivity in the cases with amino acid supplementation, but its significance increased when amino acids were not available.
-This makes sense, as amino acid synthesis from glucose became the only way to power protein synthesis.
-On the other hand, amino acid uptake only showed significance for the case where amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
+increased the productivity of firefly luciferase production by 5-fold by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+In examining substrate utilization, glucose consumption was not important for productivity in the presence of amino acid supplementation.
+However, its importance increased significantly when amino acids were not available.
+On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
 \begin{figure}[t!]
@@
 
-When considering energy efficiency performance (Fig.~\ref{fig:SI}B), the oxygen uptake rates were the most important for all cases while the sensitivity of the transcription/translation parameters decreased.
+The oxygen consumption rate was the most important factor controlling the energy efficiency of cell free protein synthesis (Fig.~\ref{fig:SI}B).
 The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
@@
 Glucose uptake was only important when amino acids were not available, since it was the only source of carbon for protein synthesis and energy generation.
-%This resulted in a tradeoff between energy generation to power transcription/translation and amino acid synthesis required to assemble the protein of interest.
 Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in pyruvate-powered CFPS \cite{Jewett:2008aa}.

### 2017-04-07 Paper_v1.tex
Working my way thru the results .... still ...

@@
 \subsubsection{Productivity}
-The theoretical maximum mean productivity was inversely proportional to the protein length ($l_{P}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
+The theoretical maximum mean productivity ($\mu$M/h) was inversely proportional to the protein length ($l_{P}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
 The theoretical maximum productivity was within a standard deviation with and without
 amino acid supplementation for each protein, but varied significantly between proteins.
-This suggested the cell free reaction had sufficient metabolic precursors to power CFPS and synthesize each protein of interest with similar productivities.
-However, the theoretical productivity was inversely proportional to protein size;
-for instance, BMP10 (XX aa) had a productivity of about 2.5 $\mu$M/h whereas CAT (ZZ aa) had a productivity of about 12 $\mu$M/h.
-To examine this further, the mean productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
-The productivity and protein size were related by the power-law relationship $\alpha\times(l_{P})^{\beta}$, where $\alpha = 6.0\times 10^{6}$ and $\beta=-1.9$.
+For instance, BMP10 (XX aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas CAT (ZZ aa) had an optimal productivity of approximately 12 $\mu$M/h.
+To examine the influence of protein length further, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
+The optimal productivity and protein length were related by the power-law relationship $\alpha\times(l_{P})^{\beta}$, where $\alpha = 6.0\times 10^{6}$ [UNITS?] and $\beta=-1.9$.
 Interestingly, CAT did not obey the power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
-CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.8 $\mu$M/h.
-Taken together, the maximum productivity of a cell free reaction was found to be inversely proportional to protein size,
+However, CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.8 $\mu$M/h (predicted to be XX by the optimal productivity correlation).
+Taken together, the maximum optimal productivity of a cell free reaction was found to be inversely proportional to protein size,
 following a power-law relationship for proteins expressed under a P70a promoter.
@@
 %In this case, glucose was used to synthesize amino acids and provide the energy necessary to power transcription and translation; this trend was seen across all proteins.
+% W, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 58.2 $\pm$ 2.3\% for CAT.
+% Once amino acid synthesis reactions were blocked in the network (second case), the carbon yield dropped to 57.1 $\pm$ 2.2\% for CAT.
 
-\subsubsection{Yield}
-The theoretical maximum mean carbon yield was inversely proportional to protein size and varied between 40-57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
-The same inverse qualitative trend was observed as for energy efficiency.
-There was a drop in carbon yield by about 7\% once amino acids were not available; this is most likely because glucose was utilized to synthesize the necessary amino acids for each protein as well as power the system.
-For the first case, the system relied on a mixture of glucose and some amino acids for each protein with a carbon yield of 58.2 $\pm$ 2.3\% for CAT.
-Once amino acid synthesis reactions were blocked in the network (second case), the carbon yield dropped to 57.1 $\pm$ 2.2\% for CAT.
-Only the necessary amount of amino acids was used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
-In that case, the carbon yield without glucose contribution would be 100\%.
-In the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58.2\%.
+\subsubsection{Carbon Yield}
+The theoretical maximum mean carbon yield was inversely proportional to protein length and varied between 40\% and 57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
+The relationship between the optimal carbon yield and protein length was linear; $m_{Y}\times(l_{P})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ [UNITS] with supplementation, and $m_{Y} = -6.98\times10^{-3}$ [UNITS] without.
+The y-intercept $b_{Y} = 63.83$ [UNITS] with supplementation, and $b_{Y} = 58.86$ [UNITS] without.
+The linear carbon yield models showed good predictability for estimating the optimal carbon yield for a range of protein lengths ($R^{2} = 0.95$ with amino acids
+and $R^{2} = 0.90$ without supplementation), irrespective of the promoter used to control protein expression.
+The effective yield models were approximately parallel, with a drop in carbon yield of approximately 7\% without supplementation.
+Only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
+In that case, the carbon yield without the glucose contribution would be 100\%.
+However, in the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58\%.
 This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
@@
 The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.
+
 % \begin{figure}[t!]
@@
 %The proteins with the lowest carbon number had the highest yield and the higher carbon number proteins had a lower carbon yield within each case.
-A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Prof}F).
-The trendlines showed good predictability for estimating carbon yield for a certain range of protein size (R\textsuperscript{2} = 0.95 with amino acids and R\textsuperscript{2} = 0.90 without amino acids), regardless of the type of promoters used.
-However, as the protein size increased, the carbon yield decreased, suggesting that larger proteins may be less feasible for cell-free production.
+% A single trendline was formulated for the cases where amino acids were supplied in the media since they had similar performance, while another trendline was formulated for when amino acids were not available (Fig.~\ref{fig:Prof}F).
+
 % Thus, we examined the parameters that had the most significant effect on cell-free productivity, energy efficiency, and carbon yield in order to optimize CFPS performance.
 
-\subsection{Sensitivity analysis}
+\subsection{Global sensitivity analysis}
 To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI}).
@@
 
+% The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
+
 The oxygen consumption rate was the most important factor controlling the energy efficiency of cell free protein synthesis (Fig.~\ref{fig:SI}B).
-The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
-Oxygen uptake was the most significant for energy efficiency, since it was responsible for oxidative phosphorylation, the most efficient pathway for energy generation.
-Across all cases, substrate utilization (amino acid uptake and glucose) was shown to be the next most important as these substrates contributed to the energy efficiency of CAT synthesis.
-In cases where amino acids were supplied, amino acid uptake was significant, as without it, energy was required to synthesize amino acids for the protein of interest.
-Glucose uptake was only important when amino acids were not available, since it was the only source of carbon for protein synthesis and energy generation.
-Jewett and coworkers have reported that oxidative phosphorylation still operated in cell-free systems, and that yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were knocked out in pyruvate-powered CFPS \cite{Jewett:2008aa}.
-It is unknown how active oxidative phosphorylation is compared to in \textit{in~vivo} systems.
-To investigate this further we compared CAT carbon yield to oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
-Interestingly, oxidative phosphorylation had a strong effect on the carbon yield in all cases.
-The cases where amino acids were supplied followed the same trend: ranging from a carbon yield of 20\% to 58\%, depending on the oxidative phosphorylation activity.
-Once amino acids were removed from the media, the carbon yield dropped to about 10\%, and reached a maximum of 52\%.
-When amino acids are not avialable in the media, a lower carbon yield was expected for the same oxidative phosphorylation flux, since carbon must be utilized for energy generation and amino acid biosynthesis.
-In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of acetate and lactate, resulting in the lower carbon yield.
-The experimental dataset exhibits a mixture of acetate and lactate accumulation during CAT synthesis, which shows that CFPS is not operating in a fully aerobic state.
-Glucose can not be fully oxidized and therefore fermentation pathways are used.
-Oxidative phosphorylation relies on electron transport from membrane vesicles, however CFPS has no cell membrane, thus it is expected CFPS has limited oxidative phosphorylation activity.
-The addition of phosphate showed an increase in CAT yield; however, it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
-Interestingly, the addition of NADH did not increase the rate of protein synthesis, since the concentration of ATP was most likely saturated.
-An alternative strategy may be the inhibition of anaerobic processes in cell-free, in order to minimize unwanted byproducts such as acetate and lactate.
+In addition to the oxygen consumption, amino acid utilization was also sensitive.
+On the other hand, glucose consumption was only sensitive in the absence of amino acid supplementation,
+since in the absence of amino acids, glucose acted as the sole source of carbon and energy for protein synthesis.
+In the model, we assumed that ATP could be produced by both substrate level and oxidative phosphorylation.
+Jewett and coworkers reported that oxidative phosphorylation still operated in cell free systems, and that the protein
+yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
+However, it is unknown how active oxidative phosphorylation is for a glucose powered cell free system.
+Moreover, the connection between oxidative phosphorylation activity and other performance metrics, such as the carbon yield, is also unclear.
+
+To investigate the connection between carbon yield and oxidative phosphorylation further,
+we calculated the CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
+Oxidative phosphorylation had a strong effect on the carbon yield, both with and without amino acid supplementation.
+In the presence of amino acid supplementation, the carbon yield ranged from 20\% to approximately 60\%,
+depending upon the oxidative phosphorylation flux.
+However, without amino acid supplementation, the carbon yield dropped to approximately 10\%, and reached a maximum of 50\%.
+In the absence of supplementation, a lower carbon yield was expected for the same oxidative phosphorylation flux, as glucose was utilized for both energy generation and amino acid biosynthesis.
+In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of both acetate and lactate.
+The experimental dataset also exhibited a mixture of acetate and lactate accumulation during CAT synthesis, which suggested the CFPS reaction was not
+operating with optimal oxidative phosphorylation activity.
+Oxidative phosphorylation is a membrane associated process, however CFPS has no cell membrane.
+Jewett et al hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation [REF];
+however, the number, size, protein loading and lifetime of these vesicles remains an open area of study.
+Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity.
+
+% The addition of phosphate showed an increase in CAT yield; however, it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
+% Interestingly, the addition of NADH did not increase the rate of protein synthesis, since the concentration of ATP was most likely saturated.
+% An alternative strategy may be the inhibition of anaerobic processes in cell-free, in order to minimize unwanted byproducts such as acetate and lactate.
 %Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and how much from anaerobic processes.
@@
 \subsubsection{Potential Alternative Metabolic Optima}
+One of the interesting features of constraint based approaches is the ability to compute potentially alternative optimal solutions.
+These solutions have the same objective value, e.g., the productivity but different underlying metabolic flux distributions.
 Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})

### 2017-04-07 Paper_v1.tex
Working my way thru the results .... still ...

@@
 The network described the major carbon and energy pathways, as well as amino acid biosynthesis and degradation pathways.
-Using this network, in combination with effective promoter models taken from Moon et al. [REF], and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
+Using this network, in combination with effective promoter models taken from Moon et al. [TODO:REF], and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
 we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
 different \textit{E.~coli} cell-free extracts.
-The cell free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab GitHub site [REF].
+The cell free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab GitHub site [TODO:REF].
 
@@
 To examine the influence of protein length further, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
-The optimal productivity and protein length were related by the power-law relationship $\alpha\times(l_{P})^{\beta}$, where $\alpha = 6.0\times 10^{6}$ [UNITS?] and $\beta=-1.9$.
+The optimal productivity and protein length were related by the power-law relationship $\alpha\times(l_{P})^{\beta}$, where $\alpha = 6.0\times 10^{6}$ [TODO: UNITS?] and $\beta=-1.9$.
 Interestingly, CAT did not obey the power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
@@
 The theoretical maximum mean carbon yield was inversely proportional to protein length and varied between 40\% and 57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
-The relationship between the optimal carbon yield and protein length was linear; $m_{Y}\times(l_{P})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ [UNITS] with supplementation, and $m_{Y} = -6.98\times10^{-3}$ [UNITS] without.
-The y-intercept $b_{Y} = 63.83$ [UNITS] with supplementation, and $b_{Y} = 58.86$ [UNITS] without.
+The relationship between the optimal carbon yield and protein length was linear; $m_{Y}\times(l_{P})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ [TODO: UNITS] with supplementation, and $m_{Y} = -6.98\times10^{-3}$ [UNITS] without.
+The y-intercept $b_{Y} = 63.83$ [UNITS] with supplementation, and $b_{Y} = 58.86$ [TODO:UNITS] without.
 The linear carbon yield models showed good predictability for estimating the optimal carbon yield for a range of protein lengths ($R^{2} = 0.95$ with amino acids
@@
 This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
-ssFBA assumes a psuedo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
-In addition, ssFBA is solved by maximizing the flux through the protein production reaction.
+Constraint based calculations assume a metabolic pseudo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
+In addition, the flux balance analysis calculation is solved by maximizing the flux through the protein production reaction.
 Therefore, carbon flux will travel through the network to optimize the maximum flux through the protein synthesis reaction.
@@
 Oxidative phosphorylation is a membrane associated process, however CFPS has no cell membrane.
-Jewett et al hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation [REF];
+Jewett et al hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation [TODO: REF];
 however, the number, size, protein loading and lifetime of these vesicles remains an open area of study.
@@
 \includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
-\caption{deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 ssFBA solutions, for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
+\caption{The deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 flux balance solutions, for amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
 \label{fig:oxphos_yield}
@@
 the energy efficiency and carbon yield decreased; in this case the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C).
-[Mike: we need a link sentence to the experimentally constrained case.]
+[TODO: We need a link sentence to the experimentally constrained case.]
 
@@
 Thus there are some differences in the flux distribution compared to the optimum solutions which may provide some insight to improve CFPS performance.
-Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour which constrained the solution space of ssFBA to have a more realistic depiction of the flux distribution.
+Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour
+which constrained the feasible solution space.
 The central carbon organic acids showed good agreement with the data (Fig.~\ref{fig:flux_exp}B).
@@
 \includegraphics[width=1.00\textwidth]{./figs/Fig-6-FluxDistribition-Experimental.pdf}
-\caption{ssFBA simulation of CAT production for an experimentally constrained case.
+\caption{Constraint based simulation of CAT production for an experimentally constrained case.
 (A) Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation.
@@
 Thick arrows indicate flux to or from amino acids.
-(B) Central carbon metabolite measurements versus ssFBA simulations over a one hour time course.}
+(B) Central carbon metabolite measurements versus simulations over a one hour time course.}
 \label{fig:flux_exp}
@@
 \subsubsection{Potential Alternative Metabolic Optima}
-One of the interesting features of constraint based approaches is the ability to compute potentially alternative optimal solutions.
+One of the interesting features of constraint based approaches is the ability to compute alternative optimal solutions.
 These solutions have the same objective value, e.g., the productivity but different underlying metabolic flux distributions.
 Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
-For example, there is a high flux through the Entner-Douodoroff pathway, but this is likely non-physiological, and simply an artifact of the optimal solution of ssFBA.
+For example, there is a high flux through the Entner-Douodoroff pathway, but this may be an artifact of the optimal solution.
 A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
@@
 pentose phosphate pathway, cofactors, folate metabolism, etc.) in the network result in the same optimal solution of CAT productivity.
-However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); ssFBA will reroute the flux to optimize the objective function.
+However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B);
+the flux balance approach will reroute the flux to optimize the objective function.
 Interestingly, there are crucial subgroups that are detrimental to CAT productivity.
@@
 Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
-In our case, ssFBA reached the same optimal solution of CAT productivity for 39.9\% of the pairwise group knockouts, but had a different flux distribution than compared to no knockouts.
+In our case, the constraint based simulation reached the same optimal solution of CAT productivity for 40\% of the pairwise group knockouts, but had a different flux distribution than compared to no knockouts.
 To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
-It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS and constrain branch reactions in ssFBA to the resulting measurements, a method that has been shown to represent the flux distribution for \emph{in~vivo} processes well \cite{Zamboni:2009}.
+It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS to better constrain the flux distribution.
+However, while C\textsuperscript{13} labeling techniques are well established for \emph{in~vivo} processes \cite{Zamboni:2009},
+they are an active area of research for cell free systems.
 \begin{figure}[t!]
@@
 \subsection*{Pairwise group knockouts.}
-Pairwise and single group knouckouts were simulated in ssFBA by setting the flux bounds for the all the reactions in a group to zero.
+Pairwise and single group knockouts were simulated by setting the flux bounds for the all the reactions in a group to zero.
 We grouped reactions in the cell-free network into 19 subgroups (available in Supporting Information).

### 2017-04-13 Paper_v1.tex
text updates MV

@@
 subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
-Stoichiometric reconstructions have been expanded to include the integration of with detailed descriptions of gene expression (ME-Model) \cite{Allen:2003aa, Brien693},
+Stoichiometric reconstructions have been expanded to include the integration of metabolism with detailed descriptions of gene expression (ME-Model) \cite{Allen:2003aa, Brien693},
 and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220}.
-These expansion have greatly expanded the scope of questions these models can explore.
+These expansions have greatly increased the scope of questions these models can explore.
 %However, FBA has been shown to have alternative solutions that lead to the same desired optimal phenotype using flux variability analysis \cite{Mahadevan2003264} or mixed-integer linear programming based algorithms \cite{LEE2000711}.
@@
 The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa},
-and adding deletions associated with the specific cell free system (see materials and methods).
+and adding deletions associated with the specific cell-free system (see materials and methods).
 We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
@@
 The network described the major carbon and energy pathways, as well as amino acid biosynthesis and degradation pathways.
-Using this network, in combination with effective promoter models taken from Moon et al. [TODO:REF], and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
+Using this network, in combination with effective promoter models taken from Moon et al. \cite{Moon:2012ab}, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
 we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
 different \textit{E.~coli} cell-free extracts.
-The cell free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab GitHub site [TODO:REF].
+The cell-free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab GitHub site [TODO:REF].
 
-Cell free simulations of the time evolution of CAT and deGFP production were consistent with experimental measurements (Fig. ~\ref{fig:network}B and C).
+Cell-free simulations of the time evolution of CAT and deGFP production were consistent with experimental measurements (Fig. ~\ref{fig:network}B and C).
 Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress}
 for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
-With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values (Correlation?).
+With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values (coefficient of determination, R\textsuperscript{2} = 0.86).
 On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
-The cell-free simulation captured the overall trend of deGFP abundance, but failed to capture saturation at the end of the CFPS culture (Correlation?).
+The cell-free simulation captured the overall trend of deGFP abundance, but failed to capture saturation at the end of the CFPS culture (R\textsuperscript{2} = 0.84).
 Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
@@
 did not qualitatively alter the performance of the model (blue region, 95\% confidence estimate).
-Together, these simulations suggested the description of transcription and translation, and its integration with metabolism encoded in the cell free metabolic model, were consistent with experimental measurements.
-However, these simulations were conducted at a single plasmid concentration.
-Thus, it was unclear if the model could capture cell free protein synthesis for a range of plasmid concentrations.
+Together, these simulations suggested the description of transcription and translation, and its integration with metabolism encoded in the cell-free metabolic model, were consistent with experimental measurements.
+However, these simulations were conducted at a single plasmid concentration of 5 nM.
+Thus, it was unclear if the model could capture cell-free protein synthesis for a range of plasmid concentrations.
 
@@
 
-Cell free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
-The cell free deGFP titer at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
-The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration.
+Cell-free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
+The cell-free deGFP titer at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
+The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration  (R\textsuperscript{2} = 0.97).
 However, while the mean and 95\% confidence estimate of the ensemble was consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating point of 5 nM of plasmid concentration.
@@
 \subsection{Analysis of CFPS performance}
-To better understand the performance of CFPS reactions, we predicted the productivity, energy efficiency and carbon yield for the cell free production of eight model proteins with and without amino acid supplementation (Fig.
+To better understand the performance of CFPS reactions, we predicted the productivity, energy efficiency and carbon yield for the cell-free production of eight model proteins with and without amino acid supplementation (Fig.
 \ref{fig:Prof}).
@@
 Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AA uptake w/o synthesis).
-This scenario was consistent with common cell free extract preparation protocols which often involve amino acid supplementation; thus, we expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction in these cases.
-Lastly, the cell free reaction was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
+This scenario was consistent with common cell-free extract preparation protocols which often involve amino acid supplementation; thus, we expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction in these cases.
+Lastly, the cell-free reaction was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
 Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
@@
 An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements (see supplemental materials).
+We also considered maltose binding protein (MBP) as a first principle prediction against the trends formulated by the other proteins.
 
@@
 \centering
-\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Analysis.pdf}
+\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance.pdf}
+%\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Predicted.pdf}
 \caption{CFPS performance of eight proteins for four cases.
@@
 Mean productivity versus carbon number.
-Single trendline (dotted line) calculated across all cases (y = 6.02$\cdot$10\textsuperscript{6}x\textsuperscript{-1.9}; R\textsuperscript{2} = 0.99).
+Single trendline (black dotted line) calculated across all cases (R\textsuperscript{2} = 0.99) and maximum productivity trendline (grey dotted line; R\textsuperscript{2} = 0.99) .
 C.
@@
 Mean energy efficiency versus carbon number.
-Trendline for cases with amino acids (black dotted line; y = -1.65$\cdot$10\textsuperscript{-2}x+90.95; R\textsuperscript{2} = 0.91) and trendline for without amino acids (grey dotted line; y = -3.18$\cdot$10\textsuperscript{-3}x+33.48; R\textsuperscript{2} = 0.78).
+Trendline for cases with amino acids (black dotted line; R\textsuperscript{2} = 0.91) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.78).
 E.
@@
 Mean carbon yield versus carbon number.
-Trendline for cases with amino acids (black dotted line; y = -8.03$\cdot$10\textsuperscript{-3}x+63.83; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; y = -6.98$\cdot$10\textsuperscript{-3}x+58.86; R\textsuperscript{2} = 0.90).
-Error bars: 95\% CI; asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline.}
+Trendline for cases with amino acids (black dotted line; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.90).
+Error bars: 95\% CI; asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline; traingles: first principle prediction and excluded from formulating trendline.}
 \label{fig:Prof}
@@
 \subsubsection{Productivity}
-The theoretical maximum mean productivity ($\mu$M/h) was inversely proportional to the protein length ($l_{P}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
+The theoretical maximum mean productivity ($\mu$M/h) was inversely proportional to the carbon number ($C_{POI}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
 The theoretical maximum productivity was within a standard deviation with and without
 amino acid supplementation for each protein, but varied significantly between proteins.
-For instance, BMP10 (XX aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas CAT (ZZ aa) had an optimal productivity of approximately 12 $\mu$M/h.
+For instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas CAT (219 aa) had an optimal productivity of approximately 12 $\mu$M/h.
 To examine the influence of protein length further, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
-The optimal productivity and protein length were related by the power-law relationship $\alpha\times(l_{P})^{\beta}$, where $\alpha = 6.0\times 10^{6}$ [TODO: UNITS?] and $\beta=-1.9$.
+The optimal productivity and protein length were related by the power-law relationship $\alpha\times(C_{POI})^{\beta}$, where $\alpha = 6.02\times 10^{6}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.93$.
 Interestingly, CAT did not obey the power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
-However, CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.8 $\mu$M/h (predicted to be XX by the optimal productivity correlation).
-Taken together, the maximum optimal productivity of a cell free reaction was found to be inversely proportional to protein size,
+The higher demand on transcription by the T7 promoter led to a higher mRNA steady state level resulting in a higher productivity.
+Thus, we also calculated the theoretical productivity correlation given maximum promoter activity using all eight proteins which resulted with $\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$.
+However, CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.2 $\pm$ 2.2 $\mu$M/h (predicted to be 7.2 $\mu$M/h by the optimal productivity correlation).
+In addition, we predicted the productivity of MBP to be 4.6 $\mu$M/h whereas the power-law relationship predicted a productivity of 5.0 $\mu$M/h.
+This difference in prediction resulted in less than 8\% error in estimating the productivity of an \emph{a~priori} protein in CFPS.  
+Taken together, the maximum optimal productivity of a cell-free reaction was found to be inversely proportional to protein size,
 following a power-law relationship for proteins expressed under a P70a promoter.
@@
 This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
-We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Prof}D).
-The same inverse trend was observed as for productivity, except that it was linear.
+%We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Prof}D).
+The same inverse trend was observed as for productivity for the effect of protein carbon number on energy efficiency, except that it was linear (Fig.~\ref{fig:Prof}D).
+The relationship between the optimal energy efficiency and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -1.65\times10^{-2}$ energy efficiency (\%)/carbon number with supplementation, and $m_{Y} = -3.18\times10^{-3}$ energy efficiency (\%)/carbon number without.
+The y-intercept $b_{Y} = 90.96$ energy efficiency (\%) with supplementation, and $b_{Y} = 33.48$ energy efficiency (\%) without.
 %The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
 Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation cost then smaller proteins.
+MBP showed good agreement with the formulated relationship between energy efficiency and carbon number for each case. 
+The estimated MBP energy efficiency had a maximum error of 7.1\% for the case where amino acid biosynthesis reactions were blocked and an error of 0.2\% for both remaining cases.  
 The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
-Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency of about 28\% regardless of carbon number.
+Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency ranging between 24-32\% regardless of carbon number.
 In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
 However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.6 $\pm$ 1.1\% compared to the theoretical maximum of 77.3 $\pm$ 7.3\%. 
-This shows CFPS systems have a lot of room for improvement: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potential could be diverted towards a protein of interest.
+This shows CFPS systems are not fully optimized and could be improved: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potentially be diverted towards a protein of interest.
 Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.
@@
 The theoretical maximum mean carbon yield was inversely proportional to protein length and varied between 40\% and 57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
-The relationship between the optimal carbon yield and protein length was linear; $m_{Y}\times(l_{P})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ [TODO: UNITS] with supplementation, and $m_{Y} = -6.98\times10^{-3}$ [UNITS] without.
-The y-intercept $b_{Y} = 63.83$ [UNITS] with supplementation, and $b_{Y} = 58.86$ [TODO:UNITS] without.
-The linear carbon yield models showed good predictability for estimating the optimal carbon yield for a range of protein lengths ($R^{2} = 0.95$ with amino acids
+The relationship between the optimal carbon yield and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ carbon yield (\%)/carbon number with supplementation, and $m_{Y} = -6.98\times10^{-3}$ carbon yield (\%)/carbon number  without.
+The y-intercept $b_{Y} = 63.83$ carbon yield (\%) with supplementation, and $b_{Y} = 58.86$ carbon yield (\%) without.
+The linear carbon yield models showed good predictability for estimating the optimal carbon yield for a range of carbon numbers ($R^{2} = 0.95$ with amino acids
 and $R^{2} = 0.90$ without supplementation), irrespective of the promoter used to control protein expression.
-The effective yield models were approximately parallel, with a drop in carbon yield of approximately 7\% without supplementation.
+MBP also showed good agreement for its predicted carbon yield with the formulated relationship with an error of 3.2-4.2\% with amino acid supplementation and 4.4\% error without supplementation.
+The effective yield models were approximately parallel, with a drop in carbon yield of approximately 7\% without amino acids.
+%In this case, there is a higher uptake of glucose and an increase in CO\textsubscript{2} production by 37-40\% and acetate production by 4-14\%. 
+In this case, there was a higher uptake of glucose and an increase of CO\textsubscript{2} production in the carbon yield by 7\%. 
 Only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
-In that case, the carbon yield without the glucose contribution would be 100\%.
+The carbon yield without the glucose contribution would be 100\%.
 However, in the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58\%.
@@
 the translation elongation rate had the largest effect on protein productivity.
-On the other hand, oxygen followed by glucose consumption had the largest influence on the energy efficiency and carbon yield, respectively.
+On the other hand, oxygen followed by substrate consumption had the largest influence on the energy efficiency and carbon yield, respectively.
 The significance of transcription/translation parameters was robust to amino acid supplementation, with the translation rate being the most sensitive across all cases (Fig.~\ref{fig:SI}A).
 This suggested that the translation elongation rate, and not transcriptional parameters, controlled productivity.
-Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
+Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased protein synthesis rates by 33\% and yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
 In addition, Li et al.
-increased the productivity of firefly luciferase production by 5-fold by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+increased the productivity of firefly luciferase production in PURE CFPS by 5-fold by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
 In examining substrate utilization, glucose consumption was not important for productivity in the presence of amino acid supplementation.
@@
 
-The oxygen consumption rate was the most important factor controlling the energy efficiency of cell free protein synthesis (Fig.~\ref{fig:SI}B).
+The oxygen consumption rate was the most important factor controlling the energy efficiency of cell-free protein synthesis (Fig.~\ref{fig:SI}B).
 In addition to the oxygen consumption, amino acid utilization was also sensitive.
 On the other hand, glucose consumption was only sensitive in the absence of amino acid supplementation,
-since in the absence of amino acids, glucose acted as the sole source of carbon and energy for protein synthesis.
+since glucose acted as the sole source of carbon and energy for protein synthesis.
 In the model, we assumed that ATP could be produced by both substrate level and oxidative phosphorylation.
-Jewett and coworkers reported that oxidative phosphorylation still operated in cell free systems, and that the protein
+Jewett and coworkers reported that oxidative phosphorylation still operated in cell-free systems, and that the protein
 yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
-However, it is unknown how active oxidative phosphorylation is for a glucose powered cell free system.
+However, it is unknown how active oxidative phosphorylation is for a glucose powered cell-free system.
 Moreover, the connection between oxidative phosphorylation activity and other performance metrics, such as the carbon yield, is also unclear.
@@
 Oxidative phosphorylation is a membrane associated process, however CFPS has no cell membrane.
-Jewett et al hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation [TODO: REF];
-however, the number, size, protein loading and lifetime of these vesicles remains an open area of study.
-Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity.
+Jewett et al. hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation \cite{Jewett:2008aa}.
+In addition, they were able to enhance the production yields of CAT by 33\% when the reaction was augmented with 10mM phosphate. 
+However, it is unclear why the additional phosphate increased the CAT yield, but Jewett et al. believe it either enhances oxidative phosphorylation activity or inhibits phosphatase reactions.
+Taken together, the number, size, protein loading and lifetime of these vesicles remains an open area of study.
+Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity and remains an important area to optimize for improved CFPS performance.
 
-% The addition of phosphate showed an increase in CAT yield; however, it is unclear whether the addition of phosphate enhances oxidative phosphorylation, inhibits phosphatase reactions, or both \cite{Jewett:2008aa}.
-% Interestingly, the addition of NADH did not increase the rate of protein synthesis, since the concentration of ATP was most likely saturated.
-% An alternative strategy may be the inhibition of anaerobic processes in cell-free, in order to minimize unwanted byproducts such as acetate and lactate.
-%Thus, it is interesting to see how much of the energetic needs of the system are met by oxidative phosphorylation and how much from anaerobic processes.
 \begin{figure}[t!]
@@
 These first two cases where amino acids were available had similar performance, and their respective metabolic flux distributions had a 99\% correlation for all proteins. 
-However, in the absence of amino acid supplementation (where all amino acids were synthesized \textit{de~novo} from glucose),
+In the absence of amino acid supplementation (where all amino acids were synthesized \textit{de~novo} from glucose),
 the energy efficiency and carbon yield decreased; in this case the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C).
-[TODO: We need a link sentence to the experimentally constrained case.]
+However, each of these three cases represent the theoretical optimum under different conditions; in order to more accurately describe system performance, we constrained the feasible solution space with experimental measurements.
 
@@
 The case constrained by experimental measurements (Fig.
-\ref{fig:flux_exp}), had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.52 with the cases supplied with amino acids.
+\ref{fig:flux_exp}) had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.52 with the cases supplied with amino acids.
 Thus there are some differences in the flux distribution compared to the optimum solutions which may provide some insight to improve CFPS performance.
@@
 which constrained the feasible solution space.
-The central carbon organic acids showed good agreement with the data (Fig.~\ref{fig:flux_exp}B).
-Only certain amino acid synthesis reactions were blocked since during the growth of \emph{E.~coli} not all amino acids were supplied (see Materials and Methods).
+The central carbon organic acids showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}=0.92 (Fig.~\ref{fig:flux_exp}B).
+Only certain amino acid synthesis reactions were blocked, since during the growth of \emph{E.~coli} not all amino acids were supplied (see Materials and Methods).
 During the cell-free reaction all amino acids were supplied, however glucose still traveled through all the major pathways, and the same metabolic precursors were still utilized for amino acid biosynthesis.
@@
 The optimum solutions only produced the required amount of amino acids necessary, however in examining the measurements there is an accumulation of alanine and glutamine which may explain some of the differences in the correlations of the flux distributions.
-Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig.~\ref{fig:flux_exp}B) and shows an inefficiency of carbon utilization.
+Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig.~\ref{fig:flux_exp}B) and is an inefficiency of carbon utilization.
 \begin{figure}[t!]
@@
 However, while C\textsuperscript{13} labeling techniques are well established for \emph{in~vivo} processes \cite{Zamboni:2009},
-they are an active area of research for cell free systems.
+they are an active area of research for cell-free systems.
 \begin{figure}[t!]
-\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribition-Analysis.pdf}
+\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribution-Analysis.pdf}
 \caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case.
@@
 
-\subsection{Summary and conclusions}
-In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell free protein synthesis reactions for a range of proteins.
-We showed first principle predictions for protein production of deGFP and CAT were in agreement with experimental measurements, under two different promoters and two different cell-free extract systems.
+\subsubsection{Summary and conclusions}
+In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell-free protein synthesis reactions for a range of proteins.
+We showed first principle predictions for protein production of deGFP and CAT which were in agreement with experimental measurements, under two different promoters and two different cell-free extract systems.
+We are currently working on expanding our library of promoters that can provide good estimates of performance for different proteins.
 This modeling approach suggested correlations for productivity, energy efficiency and carbon yield as a function of the size of the protein.
+Maltose binding protein showed good agreement with these performance metrics with the established model correlations.
 Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
@@
 The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficiency and carbon yield.
-In conclusion, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
+In addition, our modeling approach could be expanded with a more detailed description of gene expression that has been utilized in a \emph{E. coli} genome scale model \cite{Brien693}.
+We would also like to expand our model to include post translation modifications that proteins would require for proper folding and functionality. 
+In conclusion, sequence specific constraint based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
 
@@
         $\sigma_{70}$ concentration & $\sigma_{70}$ & 35 & nM & \cite{Garamella:2016aa} \\
-        $\sigma_{70}$ dissociation constant & $K_{D}$ & 130 & nM & estimated \\
-	      $\sigma_{70}$ hill coefficient & n & 1 & constant & estimated \\
+        $\sigma_{70}$ dissociation constant & $K_{D}$ & 130 & nM & \cite{Mauri2014} \\
+	$\sigma_{70}$ hill coefficient & n & 1 & constant & \cite{Mauri2014} \\
 

### 2017-04-19 Paper_v1.tex
Changes to alternative optima

@@
 These expansions have greatly increased the scope of questions these models can explore.
-%However, FBA has been shown to have alternative solutions that lead to the same desired optimal phenotype using flux variability analysis \cite{Mahadevan2003264} or mixed-integer linear programming based algorithms \cite{LEE2000711}.
 Constraint based methods are powerful tools to estimate the performance of metabolic networks with very few adjustable parameters.
@@
 different \textit{E.~coli} cell-free extracts.
-The cell-free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab GitHub site [TODO:REF].
+The cell-free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab website \cite{varnerlab}.
 
@@
 \subsubsection{Potential Alternative Metabolic Optima}
-One of the interesting features of constraint based approaches is the ability to compute alternative optimal solutions.
-These solutions have the same objective value, e.g., the productivity but different underlying metabolic flux distributions.
-Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
-For example, there is a high flux through the Entner-Douodoroff pathway, but this may be an artifact of the optimal solution.
-A knockout of the Entner-Doudoroff pathway has no effect on the norm productivity of CAT compared to no knockouts (Fig.~\ref{fig:norm}A).
-In addition, pairwise knockouts of Entner-Doudoroff and other subgroups (i.e.
-pentose phosphate pathway, cofactors, folate metabolism, etc.) in the network result in the same optimal solution of CAT productivity.
-However, there is a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B);
-the flux balance approach will reroute the flux to optimize the objective function.
-Interestingly, there are crucial subgroups that are detrimental to CAT productivity.
-For example, a single group knockout of glycolysis/gluconoeogenesis or oxidative phosphorylation leads to no CAT production.
-There were also cases of group knockouts that had no effect on productivity or its flux distribution such as isoleucine, leucine, and valine biosynthesis knocked out with histidine biosynthesis.
-Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
-In our case, the constraint based simulation reached the same optimal solution of CAT productivity for 40\% of the pairwise group knockouts, but had a different flux distribution than compared to no knockouts.
-To determine which reactions occur in CFPS, adding thermodynamic feasibility constraints to reactions may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}.
-It would also be interesting to track the carbon flux using C\textsuperscript{13} labeling in CFPS to better constrain the flux distribution.
-However, while C\textsuperscript{13} labeling techniques are well established for \emph{in~vivo} processes \cite{Zamboni:2009},
-they are an active area of research for cell-free systems.
+Constraint based approaches can be used to compute alternative optimal solutions.
+These solutions have the same objective value, e.g., productivity, but different underlying metabolic flux distributions.
+Several techniques have been developed to estimate alternative optima, e.g., flux variability analysis \cite{Mahadevan2003264,Schuetz119} or mixed-integer approaches \cite{LEE2000711}.
+In this study, we used reaction group flux variability analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).
+There were crucial pathways required for CAT production; for example, removal of the glycolysis/gluconoeogenesis or oxidative phosphorylation pathways resulted in no CAT production.
+Likewise, there were also group knockouts that had no effect on productivity or the metabolic flux distribution, such as removal of isoleucine, leucine, and valine biosynthesis combined with histidine biosynthesis.
+Globally, the constraint based simulation reached the same optimal CAT productivity for 40\% of the pairwise group knockouts, but had a different flux distribution when compared to the absence of knockouts.
+For example, one of the features of the predicted optimal metabolic flux distribution was a high flux through the Entner-Douodoroff (ED) pathway.
+Removal of the ED pathway had no effect on the CAT productivity compared to the absence of knockouts (Fig.~\ref{fig:norm}A).
+Pairwise knockouts of the ED pathway and other subgroups (i.e. pentose phosphate pathway, cofactors, folate metabolism, etc.) also resulted in the same optimal CAT productivity.
+However, there was a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); thus, alternative optimal metabolic flux distributions exist for CAT production, despite experimental constraints.
+Ultimately, to determine the metabolic flux distribution occurring in CFPS, we need to add additional constraints to the flux estimation calculation.
+For example, thermodynamic feasibility constraints may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013},
+and $^{13}$C labeling in CFPS could provide significant insight.
+However, while $^{13}$C labeling techniques are well established for \emph{in~vivo} processes \cite{Zamboni:2009},
+application of these techniques to CFPS remains an active area of research.
 \begin{figure}[t!]
@@
 \end{figure}
+
+% Despite the constraints by experimental measurements, it is difficult to calculate the physiological flux distribution of metabolism (Fig.~\ref{fig:norm})
+% For example, there is a high flux through the Entner-Douodoroff pathway, but this may be an artifact of the optimal solution.
+% Flux balance analysis has been shown to have multiple alternative optimal solutions with flux variablitiy analysis and mixed-integer linear programming resulting in a poor depiction of the physiological flux distribution \cite{LEE2000711, Mahadevan2003264, Schuetz119}.
 
@@
 \begin{equation}
-	w_{T} = V_{T}^{max}\left(\frac{G}{K_{T}+G}\right)
+	w_{T} = V_{T}^{max}\left(\frac{G_{\mathcal{P}}}{K_{T}+G_{\mathcal{P}}}\right)
 \end{equation}
-where $G$ denotes the gene concentration and $K_{T}$ denotes a transcription saturation coefficient.
+where $G_{\mathcal{P}}$ denotes the gene concentration of the protein of interest, and $K_{T}$ denotes a transcription saturation coefficient.
 The maximum transcription rate $V_{T}^{max}$ was formulated as:
@@
         & & & & \\
-        Transcription elongation rate & $\dot{v}_{TX}$ & 25 & nt/sec & \cite{Garamella:2016aa} \\
-        Translation elongation rate & $\dot{v}_{TL}$ & 2 & aa/sec & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
-        Transcription saturation coefficient & $K_{TX}$ & 3.5 & nM & estimated \\
-        Translation saturation coefficient & $K_{TL}$ & 45.0 & $\mu$M & estimated \\
+        Transcription elongation rate & $\dot{v}_{T}$ & 25 & nt/sec & \cite{Garamella:2016aa} \\
+        Translation elongation rate & $\dot{v}_{X}$ & 2 & aa/sec & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
+        Transcription saturation coefficient & $K_{T}$ & 3.5 & nM & estimated \\
+        Translation saturation coefficient & $K_{X}$ & 45.0 & $\mu$M & estimated \\
         Polysome amplification constant & $K_{P}$ & 10 & constant & estimated \\
@@
 	      $\sigma_{70}$ hill coefficient & n & 1 & constant & \cite{Mauri2014} \\
+        Gene concentration & $G_{\mathcal{P}}$ & 5 & nM &  \cite{Garamella:2016aa} \\ \bottomrule
 
-        & & & & \\
-	      Gene concentration & $G$ & 5 & nM &  \cite{Garamella:2016aa} \\
-        Gene length of CAT & $l_{G}$ & 683 & nt & \cite{Kigawa1995} \\
-        Gene length of deGFP & $l_{G}$ & 660 & nt & \cite{Garamella:2016aa} \\
-        Protein length of CAT & $l_{P}$ & 229 & aa & \cite{Kigawa1995} \\
-        Protein length of deGFP & $l_{P}$ & 219 & aa & \cite{Garamella:2016aa} \\ \bottomrule
+        % & & & & \\
+	      %
+        % Gene length of CAT & $l_{G}$ & 683 & nt & \cite{Kigawa1995} \\
+        % Gene length of deGFP & $l_{G}$ & 660 & nt & \cite{Garamella:2016aa} \\
+        % Protein length of CAT & $l_{P}$ & 229 & aa & \cite{Kigawa1995} \\
+        % Protein length of deGFP & $l_{P}$ & 219 & aa & \cite{Garamella:2016aa} \\
     \end{tabular}
@@
 and randomly sampling RNA polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 12 and 18 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/s \cite{2005_underwood_biotech, Garamella:2016aa}.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 12 and 18 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/s
+\cite{2005_underwood_biotech, Garamella:2016aa}.
 
@@
 
-\subsection*{Pairwise group knockouts.}
-Pairwise and single group knockouts were simulated by setting the flux bounds for the all the reactions in a group to zero.
+\subsection*{Identification of potential alternative optimal metabolic flux solutions.}
+We identified potential alternative optimal metabolic flux distributions by performing single and pairwise reaction group knockout simulations [TODO:REF].
+Reaction group knockouts were simulated by setting the flux bounds for all the reactions involved in a group to zero.
 We grouped reactions in the cell-free network into 19 subgroups (available in Supporting Information).
-We computed the norm of the productivity of CAT for each pairwise knockout compared to the productivity of CAT with no knockouts.
-We also computed the norm of the flux distribution for each pairwise knockout compared to the flux distribution with no knockouts.
+We computed the difference ($l^{2}$-norm) for CAT productivity in the presence and absence of pairwise reaction knockouts.
+Simultaneously, we computed the difference in the flux distribution ($l^{2}$-norm) for each pairwise reaction knockout compared to the flux distribution with no knockouts.
+Those solutions with the same or similar productivity but large changes in the metabolic flux distribution represent alternative optimal soluations.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-Please use ``The authors thank \ldots'' rather than ``The
-authors would like to thank \ldots''.
-
-The author thanks Mats Dahlgren for version one of \textsf{achemso},
-and Donald Arseneau for the code taken from \textsf{cite} to move
-citations after punctuation. Many users have provided feedback on the
-class, which is reflected in all of the different demonstrations
-shown in this document.
+This study was supported by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V. for the support of M.V.
 

### 2017-04-19 Paper_v1.tex
Working through the performance section

@@
 Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress}
-for 1 hour using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
+using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
 With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values (coefficient of determination, R\textsuperscript{2} = 0.86).
-On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for 8 hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
+On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for eight hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
 The cell-free simulation captured the overall trend of deGFP abundance, but failed to capture saturation at the end of the CFPS culture (R\textsuperscript{2} = 0.84).
@@
 The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration  (R\textsuperscript{2} = 0.97).
-However, while the mean and 95\% confidence estimate of the ensemble was consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating point of 5 nM of plasmid concentration.
-These results validated our mathematical framework to model CFPS systems and predict the production of two proteins with very few adjustable parameters.
-It also showed that the sequence specific reactions were sufficient to predict the production of two different proteins under different promoters and cell-free systems.
-Since the model accurately predicted protein production, we used our mathematical framework to understand the performance limits of CFPS.
+However, while the mean and 95\% confidence estimate of the ensemble were consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating point of 5 nM of plasmid concentration.
+These results, in combination with the time-dependent CAT and deGFP simulations, validated our CFPS modeling approach which required very few adjustable parameters.
+It also showed that the sequence specific template reactions and metabolic network were sufficient to predict the production of different proteins under different promoters, and cell-free systems.
+Next, we used our modeling approach to better understand the theoretical performance limits of CFPS.
 
@@
 \ref{fig:Prof}).
-The expression of each of these proteins was under a P70a promoter, with the exception of CAT which we assumed was expressed using a T7 promoter.
-In all cases, CFPS was supplied with glucose; however, we considered different scenarios for amino acid supplementation.
-First, the CFPS reaction was supplied with amino acids, and simultaneously the system was also allowed to synthesize amino acids from glucose (AA uptake and synthesis).
-Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AA uptake w/o synthesis).
-This scenario was consistent with common cell-free extract preparation protocols which often involve amino acid supplementation; thus, we expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction in these cases.
-Lastly, the cell-free reaction was not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (AA synthesis w/o uptake).
+The expression of each of these proteins was under a P70a promoter, with the exception of CAT which was expressed using a T7 promoter.
+In all cases, the CFPS reaction was supplied with glucose; however, we considered different scenarios for amino acid supplementation.
+First, the CFPS reaction was supplied with amino acids, and simultaneously was also allowed to synthesize amino acids from glucose (AAs supplied and \textit{de~novo} synthesis).
+Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AAs supplied w/o \textit{de~novo} synthesis).
+This scenario was consistent with common cell-free extract preparation protocols which often involve amino acid supplementation;
+thus, we expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction in these cases.
+Lastly, the cell-free reaction was given glucose but not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (\textit{de~novo} synthesis only).
 Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
@@
 prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
-An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements (see supplemental materials).
-We also considered maltose binding protein (MBP) as a first principle prediction against the trends formulated by the other proteins.
+An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements of glucose, organic and amino acid consumption/production rates (see supplemental materials).
+Finally, we considered maltose binding protein (MBP) as a first principle prediction against the trends formulated by the other proteins.
 
@@
 Trendline for cases with amino acids (black dotted line; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.90).
-Error bars: 95\% CI; asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline; traingles: first principle prediction and excluded from formulating trendline.}
+Error bars: 95\% CI; asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline; triangles: first principle prediction and excluded from formulating trendline.}
 \label{fig:Prof}
@@
 \subsubsection{Productivity}
-The theoretical maximum mean productivity ($\mu$M/h) was inversely proportional to the carbon number ($C_{POI}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
-The theoretical maximum productivity was within a standard deviation with and without
-amino acid supplementation for each protein, but varied significantly between proteins.
-For instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas CAT (219 aa) had an optimal productivity of approximately 12 $\mu$M/h.
+The theoretical maximum productivity for proteins expressed using a P70a promoter ($\mu$M/h) was inversely proportional to the carbon number ($C_{POI}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
+The theoretical maximum productivity was within a standard deviation with and without amino acid supplementation for each protein, but varied significantly between proteins.
+Moreover, productivity was a non-linear function of protein length; for instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas CAT (219 aa) had an optimal productivity of approximately 12 $\mu$M/h.
 To examine the influence of protein length further, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
@@
 Interestingly, CAT did not obey the power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
-The higher demand on transcription by the T7 promoter led to a higher mRNA steady state level resulting in a higher productivity.
+The higher transcription rate of the T7 promoter led to a [XX\%\ higher mRNA steady state level, resulting in a higher productivity.
 Thus, we also calculated the theoretical productivity correlation given maximum promoter activity using all eight proteins which resulted with $\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$.
@@
 \subsubsection{Energy efficiency}
-The energy efficiency of protein production remained relatively high for when amino acids were supplied in the media, however energy efficiency dropped below 32\% when amino acids were not available (Fig.~\ref{fig:Prof}C-D).
-Following the same outline as in examining the productivity, we calculated the energy efficiency of production for each protein.
-When amino acids were supplied in the media (first two cases), there was a comparable performance in having the highest energy efficiency, despite the network's ability to synthesis amino acids.
-For the case when amino acids were removed from the media, protein production resulted in a low energy efficiency; below 32\% depending on the protein.
-This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
-%We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Prof}D).
-The same inverse trend was observed as for productivity for the effect of protein carbon number on energy efficiency, except that it was linear (Fig.~\ref{fig:Prof}D).
-The relationship between the optimal energy efficiency and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -1.65\times10^{-2}$ energy efficiency (\%)/carbon number with supplementation, and $m_{Y} = -3.18\times10^{-3}$ energy efficiency (\%)/carbon number without.
+The energy efficiency of protein synthesis was linearly related to protein carbon number, both with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
+However, the dependence of energy efficiency on carbon number was less pronounced in the absence of amino acid supplementation.
+The relationship between the optimal energy efficiency and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -1.65\times10^{-2}$ energy efficiency (\%)/carbon number with supplementation,
+and $m_{Y} = -3.18\times10^{-3}$ energy efficiency (\%)/carbon number without supplementation.
 The y-intercept $b_{Y} = 90.96$ energy efficiency (\%) with supplementation, and $b_{Y} = 33.48$ energy efficiency (\%) without.
-%The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
-Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation cost then smaller proteins.
+Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation costs compared with smaller proteins.
+However, this effect was less pronounced in the absence of supplementation, where the optimal energy efficiency was a weaker function of the protein carbon number.
 MBP showed good agreement with the formulated relationship between energy efficiency and carbon number for each case.
@@
 
+% With amino acid supplementation, there was a comparable energy efficiency, despite the network's ability to synthesis amino acids.
+% For the case when amino acids were removed from the media, protein production resulted in a low energy efficiency; below 32\% depending on the protein.
+% This was because glucose had to be utilized to synthesize the amino acids necessary for protein synthesis and meet the necessary energy demands of CFPS.
+% %We next investigated the effect of protein carbon number on energy efficiency (Fig.~\ref{fig:Prof}D).
+% The same inverse trend was observed as for productivity for the effect of protein carbon number on energy efficiency, except that it was linear
+%
+%
+% %The proteins with the lowest carbon number had the highest energy efficiency and the higher carbon number proteins had a lower energy efficiency for when amino acids were available in the media (first two cases).
 %This may be a potential area for improvement in CFPS to optimize for energy utilization.
@@
 \subsubsection{Carbon Yield}
-The theoretical maximum mean carbon yield was inversely proportional to protein length and varied between 40\% and 57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
+The theoretical maximum carbon yield was inversely proportional to protein length and varied between 40\% and 57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
 The relationship between the optimal carbon yield and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ carbon yield (\%)/carbon number with supplementation, and $m_{Y} = -6.98\times10^{-3}$ carbon yield (\%)/carbon number  without.

### 2017-04-19 Paper_v1.tex
text and fig updates MV

@@
 Interestingly, CAT did not obey the power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
-The higher transcription rate of the T7 promoter led to a [XX\%\ higher mRNA steady state level, resulting in a higher productivity.
+The higher transcription rate of the T7 promoter led to a 34\% higher mRNA steady state level, resulting in a higher productivity.
 Thus, we also calculated the theoretical productivity correlation given maximum promoter activity using all eight proteins which resulted with $\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$.
 However, CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.2 $\pm$ 2.2 $\mu$M/h (predicted to be 7.2 $\mu$M/h by the optimal productivity correlation).
-In addition, we predicted the productivity of MBP to be 4.6 $\mu$M/h whereas the power-law relationship predicted a productivity of 5.0 $\mu$M/h.
-This difference in prediction resulted in less than 8\% error in estimating the productivity of an \emph{a~priori} protein in CFPS.
+Different promoters have been shown to have different strengths which have effected the productivity of proteins in CFPS \cite{Moon:2012ab,Garamella:2016aa}; this is consistent with our results for CAT productivity on a T7 and P70a promoter.
+In addition, to evaluate the accuracy  of our productivity P70a model, we predicted the productivity of MBP to be 4.6 $\mu$M/h whereas the power-law relationship predicted a productivity of 5.0 $\mu$M/h.
+This difference in prediction was less than 8\% error in estimating the productivity of an \emph{a~priori} protein in CFPS.
 Taken together, the maximum optimal productivity of a cell-free reaction was found to be inversely proportional to protein size,
@@
 The y-intercept $b_{Y} = 90.96$ energy efficiency (\%) with supplementation, and $b_{Y} = 33.48$ energy efficiency (\%) without.
-Proteins with a high carbon number had a lower energy efficiency since they have a higher transcription and translation costs compared with smaller proteins.
+Proteins with a high carbon number had a lower energy efficiency since they have higher transcription and translation costs compared with smaller proteins.
 However, this effect was less pronounced in the absence of supplementation, where the optimal energy efficiency was a weaker function of the protein carbon number.
@@
 The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
-Interestingly, without supplemented amino acids (third case) each protein had a similar energy efficiency ranging between 24-32\% regardless of carbon number.
-In this case, the energy burden of synthesizing each amino acid required for the assembly of the protein kept the energy efficiency saturated at a relatively low level.
+Interestingly, without supplemented amino acids each protein had a similar energy efficiency ranging between 24-32\% regardless of carbon number.
+In this case, the glucose uptake relatively doubles (64\% increase for CAT) compared to cases supplemented with amino acids, meanwhile the productivity stays relatively the same for each protein (Fig.~\ref{fig:Prof}D).
+Therefore the energy burden (i.e.
+higher glucose uptake) for synthesizing each amino acid and powering CFPS has kept the energy efficiency saturated at a relatively low level, regardless of the protein.
 However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.6 $\pm$ 1.1\% compared to the theoretical maximum of 77.3 $\pm$ 7.3\%.
 This shows CFPS systems are not fully optimized and could be improved: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potentially be diverted towards a protein of interest.
-Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which is a result of inefficient energy utilization.
+Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which was a result of inefficient energy utilization.
 
@@
 However, there was a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); thus, alternative optimal metabolic flux distributions exist for CAT production, despite experimental constraints.
+In addition, knockouts of amino acid biosynthesis reactions had no effect on the productivity with the exception of ALA, ASP, ASN and GLU, GLN biosynthesis reactions, since amino acids were supplemented in the media.
+ssFBA rerouted the metabolic flux to have the same productivity but an alterantive flux distribution (alternative optima), as can be seen for a pairwise knockout of GLY, SER biosynthesis with HIS biosynthesis (Fig.~\ref{fig:norm}B).
 Ultimately, to determine the metabolic flux distribution occurring in CFPS, we need to add additional constraints to the flux estimation calculation.
@@
 
-\subsubsection{Summary and conclusions}
+\subsection{Summary and conclusions}
 In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell-free protein synthesis reactions for a range of proteins.
@@
 We are currently working on expanding our library of promoters that can provide good estimates of performance for different proteins.
+Different promoter will help design \emph{de~novo} circuits for optimal functionality and performance. 
 This modeling approach suggested correlations for productivity, energy efficiency and carbon yield as a function of the size of the protein.

### 2017-04-20 Paper_v1.tex
A few changes ... more to come" "

@@
 \subsection*{Calculation of energy efficiency.}
-Energy efficiency ($\mathcal{E}$) was calculated as the ratio of protein production to glucose consumption, both in terms of equivalent ATP molecules:
+Energy efficiency ($\mathcal{E}$) was calculated as the ratio of protein production to glucose consumption, written in terms of equivalent ATP molecules:
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\mathcal{E}=\frac{\texttt{q}_{POI}\cdot \left(2\left(\rm {ATP}_{TX}+\rm {CTP}_{TX}+\rm {GTP}_{TX}+\rm {UTP}_{TX}\right)+\rm2\cdot\rm {ATP}_{TL}+\rm {GTP}_{TL}\right)}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}
+	\mathcal{E}\equiv\Bigl(\frac{\texttt{q}_{POI}}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}\Bigr)\cdot \Bigl(2\times\left(\rm {ATP}_{T}+\rm {CTP}_{T}+\rm {GTP}_{T}+\rm {UTP}_{T}\right)+\rm2\cdot\rm {ATP}_{X}+\rm {GTP}_{X}\Bigl)
 \end{equation}
-where $\texttt{q}_{POI}=w_{TX}$ denotes the production rate for the protein of interest, $\rm {ATP}_{TX}$, $\rm {CTP}_{TX}$, $\rm {GTP}_{TX}$, $\rm {UTP}_{TX}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{TL}$, $\rm {GTP}_{TL}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate, and $\rm {ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
+where $\texttt{q}_{POI}=w_{X}$ denotes the production rate for the protein of interest, $\rm {ATP}_{X}$, $\rm {CTP}_{X}$, $\rm {GTP}_{X}$, $\rm {UTP}_{X}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{X}$, $\rm {GTP}_{X}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate,
+and $\texttt{ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
 The energy species stoichiometric coefficients are available in the Supporting Information.
@@
 \begin{equation}\label{eqn:yield-definition}
-	Y_{C}^{POI}=\frac{\texttt{q}_{POI}\cdot C_{POI}}{\displaystyle\sum_{i=1}^{\mathcal{R}} \texttt{q}_{m_{i}}\cdot C_{m_i}}
+	Y_{C}^{POI}\equiv\frac{\texttt{q}_{POI}\cdot C_{POI}}{\displaystyle\sum_{i=1}^{\mathcal{R}} \texttt{q}_{m_{i}}\cdot C_{m_i}}
 \end{equation}

### 2017-04-20 Paper_v1.tex
Updated the KO figure

@@
 
-\subsection*{Identification of potential alternative optimal metabolic flux solutions.}
+\subsection*{Potential alternative optimal metabolic flux solutions.}
 We identified potential alternative optimal metabolic flux distributions by performing single and pairwise reaction group knockout simulations [TODO:REF].

### 2017-04-20 Paper_v1.tex
Some more small changes

@@
 \subsubsection{Potential Alternative Metabolic Optima}
-Constraint based approaches can be used to compute alternative optimal solutions.
-These solutions have the same objective value, e.g., productivity, but different underlying metabolic flux distributions.
-Several techniques have been developed to estimate alternative optima, e.g., flux variability analysis \cite{Mahadevan2003264,Schuetz119} or mixed-integer approaches \cite{LEE2000711}.
-In this study, we used reaction group flux variability analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).
-There were crucial pathways required for CAT production; for example, removal of the glycolysis/gluconoeogenesis or oxidative phosphorylation pathways resulted in no CAT production.
-Likewise, there were also group knockouts that had no effect on productivity or the metabolic flux distribution, such as removal of isoleucine, leucine, and valine biosynthesis combined with histidine biosynthesis.
-Globally, the constraint based simulation reached the same optimal CAT productivity for 40\% of the pairwise group knockouts, but had a different flux distribution when compared to the absence of knockouts.
+Alternative optimal solutions have the same objective value, e.g., productivity, but different metabolic flux distributions.
+Several techniques, e.g., flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz119} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
+In this study, we used flux variability analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).
+FVA identified pathways required for CAT production; for example, deletion of the glycolysis/gluconoeogenesis or oxidative phosphorylation pathways resulted in no CAT production.
+Likewise, there were pathway knockouts that had no effect on productivity or the metabolic flux distribution, such as removal of isoleucine, leucine, histidine and valine biosynthesis.
+Globally, the constraint based simulation reached the same optimal CAT productivity for 40\% of the pairwise knockouts, but had different flux distributions compared with the absence of knockouts.
 For example, one of the features of the predicted optimal metabolic flux distribution was a high flux through the Entner-Douodoroff (ED) pathway.

### 2017-04-21 Paper_v1.tex
new energy efficiency text and updated figs

@@
 Mean energy efficiency versus carbon number.
-Trendline for cases with amino acids (black dotted line; R\textsuperscript{2} = 0.91) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.78).
+Trendline for cases with amino acids (black dotted line) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.65).
 E.
@@
 The theoretical maximum productivity was within a standard deviation with and without amino acid supplementation for each protein, but varied significantly between proteins.
-Moreover, productivity was a non-linear function of protein length; for instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas CAT (219 aa) had an optimal productivity of approximately 12 $\mu$M/h.
+Moreover, productivity was a non-linear function of protein length; for instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas deGFP (229 aa) had an optimal productivity of approximately 8.4 $\mu$M/h.
 To examine the influence of protein length further, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
@@
 \subsubsection{Energy efficiency}
-The energy efficiency of protein synthesis was linearly related to protein carbon number, both with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
-However, the dependence of energy efficiency on carbon number was less pronounced in the absence of amino acid supplementation.
-The relationship between the optimal energy efficiency and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -1.65\times10^{-2}$ energy efficiency (\%)/carbon number with supplementation,
-and $m_{Y} = -3.18\times10^{-3}$ energy efficiency (\%)/carbon number without supplementation.
-The y-intercept $b_{Y} = 90.96$ energy efficiency (\%) with supplementation, and $b_{Y} = 33.48$ energy efficiency (\%) without.
-Proteins with a high carbon number had a lower energy efficiency since they have higher transcription and translation costs compared with smaller proteins.
-However, this effect was less pronounced in the absence of supplementation, where the optimal energy efficiency was a weaker function of the protein carbon number.
+%The energy efficiency of protein synthesis was linearly related to protein carbon number, both with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
+Energy efficiency of protein synthesis with amino acid supplentation was largely independently related to protein carbon number, both with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
+However, the dependence of energy efficiency on carbon number was more pronounced in the absence of amino acid supplementation.
+The relationship between the optimal energy efficiency and carbon number had little dependence; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -4.01\times10^{-4}$ energy efficiency (\%)/carbon number with supplementation,
+and $m_{Y} = 3.03\times10^{-3}$ energy efficiency (\%)/carbon number without supplementation.
+The y-intercept $b_{Y} = 86.20$ energy efficiency (\%) with supplementation, and $b_{Y} = 67.40$ energy efficiency (\%) without.
+In the presence of amino acid supplementation, energy was utilized to power CFPS instead of synthesizing amino acids, thus a constant level of energy efficiency was observed regardless of the protein size.
+In the absence of supplementation there was a drop in energy efficiency, ranging from 68-75.8\%.
+In this case, the glucose uptake relatively doubles (64\% increase for CAT) compared to cases supplemented with amino acids, meanwhile the productivity stayed relatively the same for each protein (Fig.~\ref{fig:Prof}D).
+Therefore, the energy burden (i.e. higher glucose uptake) for synthesizing each amino acid and powering CFPS had lowered the energy efficiency.
+Interestingly, without amino acid supplementation, proteins with a high carbon number had a slightly higher energy efficiency, however, this linear trend was mostly independent of protein size with R\textsuperscript{2} = 0.65.
+%However, this effect was less pronounced in the absence of supplementation, where the optimal energy efficiency was a weaker function of the protein carbon number.
 MBP showed good agreement with the formulated relationship between energy efficiency and carbon number for each case.
-The estimated MBP energy efficiency had a maximum error of 7.1\% for the case where amino acid biosynthesis reactions were blocked and an error of 0.2\% for both remaining cases.
+The estimated MBP energy efficiency had a maximum error of 5\% for without amino acid supplementation and an error of 1-2.7\% for both remaining cases.
 The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
-Interestingly, without supplemented amino acids each protein had a similar energy efficiency ranging between 24-32\% regardless of carbon number.
-In this case, the glucose uptake relatively doubles (64\% increase for CAT) compared to cases supplemented with amino acids, meanwhile the productivity stays relatively the same for each protein (Fig.~\ref{fig:Prof}D).
-Therefore the energy burden (i.e.
-higher glucose uptake) for synthesizing each amino acid and powering CFPS has kept the energy efficiency saturated at a relatively low level, regardless of the protein.
-However, the experimentally constrained case of CAT production showed even a lower energy efficiency of 3.6 $\pm$ 1.1\% compared to the theoretical maximum of 77.3 $\pm$ 7.3\%.
-This shows CFPS systems are not fully optimized and could be improved: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potentially be diverted towards a protein of interest.
+The experimentally constrained case of CAT production showed even a lower energy efficiency of 13.3 $\pm$ 4.7\% compared to the theoretical maximum of 84.2 $\pm$ 0.1\%.
+This showed CFPS systems are not fully optimized and could be improved: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potentially be diverted towards a protein of interest.
 Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which was a result of inefficient energy utilization.
@@
 Energy efficiency ($\mathcal{E}$) was calculated as the ratio of protein production to glucose consumption, written in terms of equivalent ATP molecules:
+%\begin{equation}\label{eqn:energy-efficiency-definition}
+%	\mathcal{E}\equiv\Bigl(\frac{\texttt{q}_{POI}}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}\Bigr)\cdot \Bigl(2\times\left(\rm {ATP}_{T}+\rm {CTP}_{T}+\rm {GTP}_{T}+\rm {UTP}_{T}\right)+\rm2\cdot\rm {ATP}_{X}+\rm {GTP}_{X}\Bigl)
+%\end{equation}
+
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\mathcal{E}\equiv\Bigl(\frac{\texttt{q}_{POI}}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}\Bigr)\cdot \Bigl(2\times\left(\rm {ATP}_{T}+\rm {CTP}_{T}+\rm {GTP}_{T}+\rm {UTP}_{T}\right)+\rm2\cdot\rm {ATP}_{X}+\rm {GTP}_{X}\Bigl)
+	\mathcal{E}\equiv\frac{w_{X}\left(2\cdot\rm {ATP}_{X}+\rm {GTP}_{X}\right)+w_{T}\left(2\times\left(\rm {ATP}_{T}+\rm {CTP}_{T}+\rm {GTP}_{T}+\rm {UTP}_{T}\right) \right)}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}
 \end{equation}
-where $\texttt{q}_{POI}=w_{X}$ denotes the production rate for the protein of interest, $\rm {ATP}_{X}$, $\rm {CTP}_{X}$, $\rm {GTP}_{X}$, $\rm {UTP}_{X}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{X}$, $\rm {GTP}_{X}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate,
-and $\texttt{ATP}_{GLC}$ denotes the equivalent ATP number for glucose.
+
+where $\rm {ATP}_{T}$, $\rm {CTP}_{T}$, $\rm {GTP}_{T}$, $\rm {UTP}_{T}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{X}$, $\rm {GTP}_{X}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate,
+and $\texttt{ATP}_{GLC}$ denotes the equivalent ATP number for glucose produced.
 The energy species stoichiometric coefficients are available in the Supporting Information.

### 2017-04-21 Paper_v1.tex
Some changes from JV in results

@@
 Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized protein production.
-Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+From the perspective of research, cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
 Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
@@
 Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
-However, if CFPS is to become a mainstream technology for applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
+However, if CFPS is to become a mainstream technology for advanced applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
 One tool to address these questions is mathematical modeling.
@@
 Stoichiometric reconstructions of microbial metabolism, popularized by constraint based approaches such as flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
-Constraint based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
+Constraint based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and convex network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
 Constraint based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} for biochemical networks of arbitrary complexity, including genome scale networks, using linear programming \cite{Covert:2004aa}.
@@
 From this, higher carbon number proteins typically had lower productivity rates, energy efficiency, and carbon yields than that of the lower carbon number proteins.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency
+and carbon yield and the translation rate for productivity.
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
@@
 \subsection{Analysis of CFPS performance}
-To better understand the performance of CFPS reactions, we predicted the productivity, energy efficiency and carbon yield for the cell-free production of eight model proteins with and without amino acid supplementation (Fig.
+To better understand the performance of CFPS reactions, we analyzed the productivity, energy efficiency and carbon yield for the cell-free production of eight model proteins with and without amino acid supplementation (Fig.
 \ref{fig:Prof}).
@@
 An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements of glucose, organic and amino acid consumption/production rates (see supplemental materials).
-Finally, we considered maltose binding protein (MBP) as a first principle prediction against the trends formulated by the other proteins.
+Using these model proteins, we developed effective correlation models that predicted the productivity, energy efficiency and carbon yield given the carbon number of the model protein.
+Finally, we independently validated the effective correlations with a protein not in our original data set, maltose binding protein (MBP).
 
@@
 %\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Predicted.pdf}
-\caption{CFPS performance of eight proteins for four cases.
-Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
-A.
-Productivity.
-B.
-Mean productivity versus carbon number.
-Single trendline (black dotted line) calculated across all cases (R\textsuperscript{2} = 0.99) and maximum productivity trendline (grey dotted line; R\textsuperscript{2} = 0.99) .
-C.
-Energy efficiency.
-D.
-Mean energy efficiency versus carbon number.
+\caption{The CFPS performance for eight model proteins with and without amino acid supplementation.
+A. Mean CFPS productivity for a panel of model proteins with and without amino acid supplementation.
+B. Mean CFPS productivity versus carbon number for a panel of model proteins with and without amino acid supplementation.
+Trendline (black dotted line) was calculated across all cases for a P70a promoter (R\textsuperscript{2} = 0.99) and maximum productivity trendline assumed $u\left(\kappa\right)$ = 1 (grey dotted line; R\textsuperscript{2} = 0.99).
+C. Mean CFPS energy efficiency for a panel of model proteins with and without amino acid supplementation.
+D. Mean CFPS energy efficiency versus carbon number for a panel of model proteins with and without amino acid supplementation.
 Trendline for cases with amino acids (black dotted line) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.65).
-E.
-Carbon yield.
-F.
-Mean carbon yield versus carbon number.
+E. Mean CFPS carbon yield for a panel of model proteins with and without amino acid supplementation.
+F. Mean CFPS carbon yield versus carbon number.
 Trendline for cases with amino acids (black dotted line; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.90).
-Error bars: 95\% CI; asterisk: protein excluded from trendline; dagger: constrained by experimental measurements and excluded from trendline; triangles: first principle prediction and excluded from formulating trendline.}
+Error bars: 95\% CI calculated by sampling;
+asterisk: protein excluded from trendline;
+dagger: constrained by experimental measurements and excluded from trendline;
+triangles: first principle prediction and excluded from formulating trendline.}
 \label{fig:Prof}
 \end{figure}
+
+%Amino acid uptake and synthesis (black), AA uptake without synthesis (dark grey), AA synthesis without uptake (light grey), and constrained by experimental measurements, for CAT only (white).
 
@@
 The theoretical maximum productivity was within a standard deviation with and without amino acid supplementation for each protein, but varied significantly between proteins.
-Moreover, productivity was a non-linear function of protein length; for instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas deGFP (229 aa) had an optimal productivity of approximately 8.4 $\mu$M/h.
-To examine the influence of protein length further, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
-The optimal productivity and protein length were related by the power-law relationship $\alpha\times(C_{POI})^{\beta}$, where $\alpha = 6.02\times 10^{6}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.93$.
-Interestingly, CAT did not obey the power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
+Productivity was a non-linear function of protein length; for instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas the optimal productivity of deGFP (229 aa) was approximately 8.4 $\mu$M/h.
+To examine the influence of protein length, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
+The optimal productivity and protein length were related by the power-law relationship $\alpha\times(C_{POI})^{\beta}$, where $\alpha = 6.02\times 10^{6}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.93$ for a P70a promoter.
+Interestingly, CAT did not obey the P70a power-law relationship; the relatively high productivity of CAT was due to its T7 promoter.
 The higher transcription rate of the T7 promoter led to a 34\% higher mRNA steady state level, resulting in a higher productivity.
-Thus, we also calculated the theoretical productivity correlation given maximum promoter activity using all eight proteins which resulted with $\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$.
-However, CAT expressed under a P70a promoter followed the power-law correlation with a productivity of approximately 8.2 $\pm$ 2.2 $\mu$M/h (predicted to be 7.2 $\mu$M/h by the optimal productivity correlation).
-Different promoters have been shown to have different strengths which have effected the productivity of proteins in CFPS \cite{Moon:2012ab,Garamella:2016aa}; this is consistent with our results for CAT productivity on a T7 and P70a promoter.
-In addition, to evaluate the accuracy  of our productivity P70a model, we predicted the productivity of MBP to be 4.6 $\mu$M/h whereas the power-law relationship predicted a productivity of 5.0 $\mu$M/h.
-This difference in prediction was less than 8\% error in estimating the productivity of an \emph{a~priori} protein in CFPS.
+However, CAT expressed under a P70a promoter followed the P70a power-law correlation with a productivity of approximately 8.2 $\pm$ 2.2 $\mu$M/h (predicted to be 7.2 $\mu$M/h by the optimal productivity correlation).
+
+We independently validated the productivity correlation by calculating the optimal productivity of MBP using the full model and the effective correlation model (Fig.~\ref{fig:Prof}B).
+The prediction error was less than 8\% for an \emph{a~priori} prediction of CFPS productivity.
+Thus, the effective productivity correlation could be used as a parameter free method to estimate optimal productivity for cell free protein production using a P70a promoter.
+For CFPS using other promoters, a similar correlation model could be developed.
+For example, maximal transcription occurs when the promoter model coefficient $u\left(\kappa\right)$ = 1;
+the theoretical maximum productivity correlation for maximum promoter activity also followed a power-law distribution ($\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$) (Fig.~\ref{fig:Prof}B, gray).
+The CAT value under a T7 promoter was similar to the maximal possible productivity as $u_{T7}\left(\kappa\right)\simeq~$0.95 given the T7 promoter model parameters used in this study (Table \ref{tbl:parameters}).
 Taken together, the maximum optimal productivity of a cell-free reaction was found to be inversely proportional to protein size,
@@
 
+%We also calculated the theoretical productivity correlation given maximum promoter activity using all eight proteins which resulted with $\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$.
+% In addition, to evaluate the accuracy  of our productivity P70a model,
+%The choice of promoter has been shown to effect CFPS productivity \cite{Moon:2012ab,Garamella:2016aa}; this is consistent with our results for CAT productivity on a T7 and P70a promoter.
 % Taken together, a single trendline showed good agreement (R\textsuperscript{2} = 0.99) with estimating the expected productivity of a protein on a  in CFPS, given the size of the protein.
@@
 %The energy efficiency of protein synthesis was linearly related to protein carbon number, both with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
-Energy efficiency of protein synthesis with amino acid supplentation was largely independently related to protein carbon number, both with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
-However, the dependence of energy efficiency on carbon number was more pronounced in the absence of amino acid supplementation.
-The relationship between the optimal energy efficiency and carbon number had little dependence; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -4.01\times10^{-4}$ energy efficiency (\%)/carbon number with supplementation,
+%However, the dependence of energy efficiency on carbon number was more pronounced in the absence of amino acid supplementation.
+The optimal energy efficiency of protein synthesis was independent of protein length, with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
+The optimal energy efficiency was approximately 86\% for the model proteins sampled.
+The optimal energy efficiency was independent of the protein length; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -4.01\times10^{-4}$ energy efficiency (\%)/carbon number with supplementation,
 and $m_{Y} = 3.03\times10^{-3}$ energy efficiency (\%)/carbon number without supplementation.
-The y-intercept $b_{Y} = 86.20$ energy efficiency (\%) with supplementation, and $b_{Y} = 67.40$ energy efficiency (\%) without.
-In the presence of amino acid supplementation, energy was utilized to power CFPS instead of synthesizing amino acids, thus a constant level of energy efficiency was observed regardless of the protein size.
-In the absence of supplementation there was a drop in energy efficiency, ranging from 68-75.8\%.
-In this case, the glucose uptake relatively doubles (64\% increase for CAT) compared to cases supplemented with amino acids, meanwhile the productivity stayed relatively the same for each protein (Fig.~\ref{fig:Prof}D).
-Therefore, the energy burden (i.e. higher glucose uptake) for synthesizing each amino acid and powering CFPS had lowered the energy efficiency.
-Interestingly, without amino acid supplementation, proteins with a high carbon number had a slightly higher energy efficiency, however, this linear trend was mostly independent of protein size with R\textsuperscript{2} = 0.65.
-%However, this effect was less pronounced in the absence of supplementation, where the optimal energy efficiency was a weaker function of the protein carbon number.
-MBP showed good agreement with the formulated relationship between energy efficiency and carbon number for each case.
-The estimated MBP energy efficiency had a maximum error of 5\% for without amino acid supplementation and an error of 1-2.7\% for both remaining cases.
-The cases supplemented with amino acids had the same trendline whereas when amino acids were not available there was a significant drop in energy efficiency.
+The y-intercept $b_{Y} = 86.20$ energy efficiency (\%) with supplementation, and $b_{Y} = 67.40$ energy efficiency (\%) without supplementation.
+In the presence of amino acids, energy was utilized to power CFPS instead of synthesizing amino acids, thus a constant level of energy efficiency was observed regardless of the protein size.
+In the absence of supplementation, there was a drop in energy efficiency, ranging from 68\% to 76\%.
+In this case, glucose consumption more than doubled (64\% increase for CAT) compared to cases supplemented with amino acids, meanwhile the productivity was similar for each protein (Fig.~\ref{fig:Prof}D).
+Therefore, the energy burden required for synthesizing each amino acid, and powering CFPS, lowered the energy efficiency.
+Surprisingly, without amino acid supplementation, proteins with a higher carbon number had a marginally higher energy efficiency;
+however, this linear trend was mostly independent of protein size (R\textsuperscript{2} = 0.65).
+MBP showed good agreement with the linear efficiency model with and without amino acid supplementation.
+The estimated MBP energy efficiency had a maximum error of 5\% without supplementation, and an error of 1\% - 3\% in the presence of amino acid.
 The experimentally constrained case of CAT production showed even a lower energy efficiency of 13.3 $\pm$ 4.7\% compared to the theoretical maximum of 84.2 $\pm$ 0.1\%.
-This showed CFPS systems are not fully optimized and could be improved: first, the experimental measurements showed accumulation of certain amino acids; this carbon could potentially be diverted towards a protein of interest.
-Second, the system had a high accumulation of metabolic byproducts, specifically organic acids, which was a result of inefficient energy utilization.
+This suggested CFPS was not energy optimized and could be improved: first, the experimental measurements showed accumulation of certain amino acids; second,
+there was significant accumulation of organic acids, suggesting inefficient energy utilization.
 
@@
 \subsubsection{Carbon Yield}
-The theoretical maximum carbon yield was inversely proportional to protein length and varied between 40\% and 57\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
-The relationship between the optimal carbon yield and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ carbon yield (\%)/carbon number with supplementation, and $m_{Y} = -6.98\times10^{-3}$ carbon yield (\%)/carbon number  without.
-The y-intercept $b_{Y} = 63.83$ carbon yield (\%) with supplementation, and $b_{Y} = 58.86$ carbon yield (\%) without.
-The linear carbon yield models showed good predictability for estimating the optimal carbon yield for a range of carbon numbers ($R^{2} = 0.95$ with amino acids
+The theoretical maximum carbon yield was inversely proportional to protein length and varied between 40\% to 64\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
+The relationship between the optimal carbon yield and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -8.03\times10^{-3}$ carbon yield (\%)/carbon number with supplementation, and $m_{Y} = -6.98\times10^{-3}$ carbon yield (\%)/carbon number  without supplementation.
+The y-intercept $b_{Y} = 63.83$ carbon yield (\%) with supplementation, and $b_{Y} = 58.86$ carbon yield (\%) without supplementation.
+The linear yield models showed good predictability for estimating the optimal carbon yield for a range of carbon numbers ($R^{2} = 0.95$ with amino acids
 and $R^{2} = 0.90$ without supplementation), irrespective of the promoter used to control protein expression.
-MBP also showed good agreement for its predicted carbon yield with the formulated relationship with an error of 3.2-4.2\% with amino acid supplementation and 4.4\% error without supplementation.
+The MBP prediction also showed good agreement with the constraint based model, with an error between 3\% to 4\% with and without amino acid supplementation.
 The effective yield models were approximately parallel, with a drop in carbon yield of approximately 7\% without amino acids.
-%In this case, there is a higher uptake of glucose and an increase in CO\textsubscript{2} production by 37-40\% and acetate production by 4-14\%.
-In this case, there was a higher uptake of glucose and an increase of CO\textsubscript{2} production in the carbon yield by 7\%.
-Only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
+In the absence of amino acids, glucose consumption increased, with the change in the carbon yield being largely accounted for by increased CO\textsubscript{2} production.
+If only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
 The carbon yield without the glucose contribution would be 100\%.

### 2017-04-22 Paper_v1.tex
Working on changes in the results

@@
 
-\begin{figure}[t!]
-\centering
-\includegraphics[width=1.00\textwidth]{./figs/Fig-1-Network-Validation-Simulations.pdf}
-\caption{Sequence specific flux balance analysis.
-A.
-Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway.
-Thick gray arrows indicate withdrawal of precursors for amino acid synthesis.
-B.
-CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption.
-Error bars denote the standard deviation of experimental measurements.
-C.
-deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption.
-Error bars denote a 10\% coefficient of variation.
-D.
-Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0.
-95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
-\label{fig:network}
-\end{figure}
-
 Cell-free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
@@
 Finally, we independently validated the effective correlations with a protein not in our original data set, maltose binding protein (MBP).
-
-\begin{figure}[t!]
-\centering
-\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance.pdf}
-%\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Predicted.pdf}
-\caption{The CFPS performance for eight model proteins with and without amino acid supplementation.
-A. Mean CFPS productivity for a panel of model proteins with and without amino acid supplementation.
-B. Mean CFPS productivity versus carbon number for a panel of model proteins with and without amino acid supplementation.
-Trendline (black dotted line) was calculated across all cases for a P70a promoter (R\textsuperscript{2} = 0.99) and maximum productivity trendline assumed $u\left(\kappa\right)$ = 1 (grey dotted line; R\textsuperscript{2} = 0.99).
-C. Mean CFPS energy efficiency for a panel of model proteins with and without amino acid supplementation.
-D. Mean CFPS energy efficiency versus carbon number for a panel of model proteins with and without amino acid supplementation.
-Trendline for cases with amino acids (black dotted line) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.65).
-E. Mean CFPS carbon yield for a panel of model proteins with and without amino acid supplementation.
-F. Mean CFPS carbon yield versus carbon number.
-Trendline for cases with amino acids (black dotted line; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.90).
-Error bars: 95\% CI calculated by sampling;
-asterisk: protein excluded from trendline;
-dagger: constrained by experimental measurements and excluded from trendline;
-triangles: first principle prediction and excluded from formulating trendline.}
-\label{fig:Prof}
-\end{figure}
 
@@
 The estimated MBP energy efficiency had a maximum error of 5\% without supplementation, and an error of 1\% - 3\% in the presence of amino acid.
-The experimentally constrained case of CAT production showed even a lower energy efficiency of 13.3 $\pm$ 4.7\% compared to the theoretical maximum of 84.2 $\pm$ 0.1\%.
-This suggested CFPS was not energy optimized and could be improved: first, the experimental measurements showed accumulation of certain amino acids; second,
-there was significant accumulation of organic acids, suggesting inefficient energy utilization.
+
+Experimentally constrained CAT simulations showed suboptimal energy efficiency (Fig.~\ref{fig:Prof}D, dagger).
+CAT production was simulated using the constraint based model in combination with experimental measurements of glucose consumption, organic and amino acid consumption and production rates (Fig.
+~\ref{fig:network}B).
+The experimentally constrained energy efficiency was 13.3 $\pm$ 5.0\% compared to the theoretical maximum of approximately 84 $\pm$ 0.1\%, however, the
+CAT productivity was similar between the simulated and measured systems.
+This suggested CFPS was not energy optimized.
+There were two factors responsible for the difference between the optimal and experimental system,
+the glucose consumption rate and the ATP yield per glucose.
+The glucose consumption rate was approximately 30 mM/h in the experimental system (even in the presence of amino acids).
+However, the constraint based simulation suggested the optimal glucose consumption rate was significantly less than the observed rate, approximately 2 - 4 mM/h.
+In the constraint based simulation, the CFPS reaction produced only acetate as a by-product, but in the experimental system acetate, lactate, pyruvate, succinate and malate accumulated during the first hour of production.
+Thus, the constraint based simulation was more carbon efficient.
+Next, the energy produced per unit glucose was different between the optimal and experimental constrained system.
+In the optimal simulation, 12 ATPs were produced per glucose (the theoretical maximum for this network was 21),
+while the experimentally constrained simulation produced only 4 ATPs per glucose.
+[TODO JV: Think more about this].
 
@@
 On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
-\begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
-\caption{Total order sensitivity of deGFP productivity (A) and energy efficiency (B) to specific uptake rates and transcription/translation parameters for three cases: amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).
-Error bars represent a 95\% CI on the sensitivity index.}
-\label{fig:SI}
-\end{figure}
 
@@
 
-\begin{figure}[t!]
-\includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
-\caption{The deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 flux balance solutions, for amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
-\label{fig:oxphos_yield}
-\end{figure}
-
 \subsection{Optimal Metabolic Flux Distribution}
@@
 To investigate the influence of amino acid supplementation, we compared the simulated metabolic flux distributions for CAT production with and without external amino acids.
-\begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation, for three different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, and (C) amino acid synthesis without uptake.
-Mean flux across the ensemble, normalized to glucose uptake flux.
-Thick arrows indicate flux to or from amino acids.}
-\label{fig:flux}
-\end{figure}
 In the presence of amino acid supplementation, and \textit{de~novo} amino acid synthesis, there was an incomplete TCA cycle, where a combination
@@
 Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig.~\ref{fig:flux_exp}B) and is an inefficiency of carbon utilization.
-\begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./figs/Fig-6-FluxDistribition-Experimental.pdf}
-\caption{Constraint based simulation of CAT production for an experimentally constrained case.
-(A) Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation.
-Mean flux across the ensemble, normalized to glucose uptake flux.
-Thick arrows indicate flux to or from amino acids.
-(B) Central carbon metabolite measurements versus simulations over a one hour time course.}
-\label{fig:flux_exp}
-\end{figure}
 
@@
 application of these techniques to CFPS remains an active area of research.
-\begin{figure}[t!]
-\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribution-Analysis.pdf}
-\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network (A) CAT productivity norm compared to no knockouts in the experimentally constrained case.
-(B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
-\label{fig:norm}
-\end{figure}
 
@@
 and $\mathcal{L}_{i}$ and $\mathcal{U}_{i}$ denote the lower and upper bounds on flux $w_{i}$, respectively.
-The transcription (T) and translation (X) stoichiometry was modeled based upon the template reactions of Allen and Palsson \cite{Allen:2003aa}:
-\begin{table}[!]
-\centering
-    \caption{Transcription and translation template reactions for protein production.}
-    \renewcommand{\arraystretch}{1.2}
-    \resizebox{\textwidth}{!}{\begin{tabular}{llcl} \toprule
-        \textbf{Description} & \multicolumn{3}{l}{\textbf{Template reaction}} \\ \toprule
-        Transcription initiation & $G_\mathcal{P}+R_{T}$ & $\longrightarrow$& $G_{\mathcal{P}}^{*}$ \\
-        Transcription ($w_{T}$) & $G_{\mathcal{P}}^{*}+\quad\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right)$ & $\longrightarrow$& $mRNA+G_{\mathcal{P}}+R_{T}+\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}} \eta_{k} \cdot PPi$ \\
-        mRNA degradation &  $mRNA$ &$\longrightarrow$& \quad $\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left\{k\right\}MP$ \\
-	Translation initiation & $mRNA+R_{X}$ &$\longrightarrow$& $R_{X}^{*}$ \\
-        tRNA charging & $\alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big)$ &$\longrightarrow$& $\alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)$ \\
- 	&&& $\qquad \qquad \qquad \qquad \quad  {j=1,2,\hdots,20}$\\
-        Translation ($w_{X}$) &  $R_{X}^{*}+\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big)$ &$\longrightarrow$& $\mathcal{P}+R_{X}+mRNA$\\
-&&& $\quad +\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)$ \\ \bottomrule
-    \end{tabular}}
-\label{tbl:parameters}
-\end{table}
-where $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
-$R_{T}$ denotes the concentration of RNA polymerase,
-$G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase (open complex),
-$\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
-$Pi$ denotes inorganic phosphate,
-$R_{X}$ denotes the ribosome concentration,
-$R_{X}^{*}$ denotes bound ribosome,
-and $AA_{j}$ denotes $j^{th}$ amino acid.
-
+The transcription (T) and translation (X) stoichiometry was modeled based upon the template reactions of Allen and Palsson \cite{Allen:2003aa} (Table \ref{tbl:template-reactions}).
 The objective of the sequence specific flux balance calculation was to maximize the rate of protein translation, $w_{X}$.
@@
 
-\begin{table}[!]
-\centering
-    \caption{Parameters for sequence specific flux balance analysis}
-    \renewcommand{\arraystretch}{1}
-    \begin{tabular}{lcccc} \toprule
-        \textbf{Description} & \textbf{Parameter} & \textbf{Value} & \textbf{Units} & \textbf{Reference} \\ \toprule
-        RNA polymerase concentration & $R_{T}$ & 75 & nM & \cite{Garamella:2016aa} \\
-        Ribosome concentration & $R_{X}$ & 1.6 & $\mu$M & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
-
-        & & & & \\
-        Transcription elongation rate & $\dot{v}_{T}$ & 25 & nt/sec & \cite{Garamella:2016aa} \\
-        Translation elongation rate & $\dot{v}_{X}$ & 2 & aa/sec & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
-        Transcription saturation coefficient & $K_{T}$ & 3.5 & nM & estimated \\
-        Translation saturation coefficient & $K_{X}$ & 45.0 & $\mu$M & estimated \\
-        Polysome amplification constant & $K_{P}$ & 10 & constant & estimated \\
-        mRNA degradation rate & $\lambda$ & 5.2 & hr\textsuperscript{-1} & \cite{Garamella:2016aa} \\
-
-        & & & & \\
-        T7 promoter & $K_{T7}$ & 10 & constant & estimated \\
-        Weight RNA polymerase binding alone P70a & $K_{1}$ & 0.014 & constant & estimated \\
-        Weight bound RNAP-$\sigma_{70}$ P70a & $K_{2}$ & 10 & constant & estimated \\
-        $\sigma_{70}$ concentration & $\sigma_{70}$ & 35 & nM & \cite{Garamella:2016aa} \\
-        $\sigma_{70}$ dissociation constant & $K_{D}$ & 130 & nM & \cite{Mauri2014} \\
-	      $\sigma_{70}$ hill coefficient & n & 1 & constant & \cite{Mauri2014} \\
-        Gene concentration & $G_{\mathcal{P}}$ & 5 & nM &  \cite{Garamella:2016aa} \\ \bottomrule
-
-        % & & & & \\
-	      %
-        % Gene length of CAT & $l_{G}$ & 683 & nt & \cite{Kigawa1995} \\
-        % Gene length of deGFP & $l_{G}$ & 660 & nt & \cite{Garamella:2016aa} \\
-        % Protein length of CAT & $l_{P}$ & 229 & aa & \cite{Kigawa1995} \\
-        % Protein length of deGFP & $l_{P}$ & 219 & aa & \cite{Garamella:2016aa} \\
-    \end{tabular}
-\label{tbl:parameters}
-\end{table}
-
-\clearpage
-
 \subsection*{Calculation of energy efficiency.}
 Energy efficiency ($\mathcal{E}$) was calculated as the ratio of protein production to glucose consumption, written in terms of equivalent ATP molecules:
-%\begin{equation}\label{eqn:energy-efficiency-definition}
-%	\mathcal{E}\equiv\Bigl(\frac{\texttt{q}_{POI}}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}\Bigr)\cdot \Bigl(2\times\left(\rm {ATP}_{T}+\rm {CTP}_{T}+\rm {GTP}_{T}+\rm {UTP}_{T}\right)+\rm2\cdot\rm {ATP}_{X}+\rm {GTP}_{X}\Bigl)
-%\end{equation}
-
 \begin{equation}\label{eqn:energy-efficiency-definition}
@@
 \end{equation}
-
 where $\rm {ATP}_{T}$, $\rm {CTP}_{T}$, $\rm {GTP}_{T}$, $\rm {UTP}_{T}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{X}$, $\rm {GTP}_{X}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate,
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\centering
+    \caption{Transcription and translation template reactions for protein production. The symbol $G_{\mathcal{P}}$ denotes the gene encoding protein product $\mathcal{P}$,
+    $R_{T}$ denotes the concentration of RNA polymerase,
+    $G_{\mathcal{P}}^{*}$ denotes the gene bounded by the RNA polymerase (open complex),
+    $\eta_{i}$ and $ \alpha_{j}$ denote the stoichiometric coefficients for nucleotide and amino acid, respectively,
+    $Pi$ denotes inorganic phosphate,
+    $R_{X}$ denotes the ribosome concentration,
+    $R_{X}^{*}$ denotes bound ribosome,
+    and $AA_{j}$ denotes $j^{th}$ amino acid.}
+    \renewcommand{\arraystretch}{1.2}
+    \resizebox{\textwidth}{!}{\begin{tabular}{llcl} \toprule
+        \textbf{Description} & \multicolumn{3}{l}{\textbf{Template reaction}} \\ \toprule
+        Transcription initiation & $G_\mathcal{P}+R_{T}$ & $\longrightarrow$& $G_{\mathcal{P}}^{*}$ \\
+        Transcription ($w_{T}$) & $G_{\mathcal{P}}^{*}+\quad\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left(\left\{k\right\}TP + H_{2}O\right)$ & $\longrightarrow$& $mRNA+G_{\mathcal{P}}+R_{T}+\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}} \eta_{k} \cdot PPi$ \\
+        mRNA degradation &  $mRNA$ &$\longrightarrow$& \quad $\displaystyle\sum_{\mathclap{k\in\left\{A,C,G,U\right\}}}\eta_{k}\cdot \left\{k\right\}MP$ \\
+	Translation initiation & $mRNA+R_{X}$ &$\longrightarrow$& $R_{X}^{*}$ \\
+        tRNA charging & $\alpha_{j}\cdot \Big(AA_{j}+tRNA+ATP+H_{2}O\Big)$ &$\longrightarrow$& $\alpha_{j}\cdot \Big(AA_{j}\text{-}tRNA_{j} +AMP+PPi\Big)$ \\
+ 	&&& $\qquad \qquad \qquad \qquad \quad  {j=1,2,\hdots,20}$\\
+        Translation ($w_{X}$) &  $R_{X}^{*}+\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot \Big(AA_{j}\text{-} tRNA_{j}+2GTP+2H_{2}O\Big)$ &$\longrightarrow$& $\mathcal{P}+R_{X}+mRNA$\\
+&&& $\quad +\displaystyle\sum_{\mathclap{j\in\left\{AA\right\}}}\alpha_{j}\cdot\Big(tRNA+2GDP+2Pi\Big)$ \\ \bottomrule
+    \end{tabular}}
+\label{tbl:template-reactions}
+\end{table}
+
+\clearpage
+
+\begin{table}
+\centering
+    \caption{Parameters for sequence specific flux balance analysis}
+    \renewcommand{\arraystretch}{1}
+    \begin{tabular}{lcccc} \toprule
+        \textbf{Description} & \textbf{Parameter} & \textbf{Value} & \textbf{Units} & \textbf{Reference} \\ \toprule
+        RNA polymerase concentration & $R_{T}$ & 75 & nM & \cite{Garamella:2016aa} \\
+        Ribosome concentration & $R_{X}$ & 1.6 & $\mu$M & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
+
+        & & & & \\
+        Transcription elongation rate & $\dot{v}_{T}$ & 25 & nt/sec & \cite{Garamella:2016aa} \\
+        Translation elongation rate & $\dot{v}_{X}$ & 2 & aa/sec & \cite{Garamella:2016aa, 2005_underwood_biotech} \\
+        Transcription saturation coefficient & $K_{T}$ & 3.5 & nM & estimated \\
+        Translation saturation coefficient & $K_{X}$ & 45.0 & $\mu$M & estimated \\
+        Polysome amplification constant & $K_{P}$ & 10 & constant & estimated \\
+        mRNA degradation rate & $\lambda$ & 5.2 & hr\textsuperscript{-1} & \cite{Garamella:2016aa} \\
+
+        & & & & \\
+        T7 promoter & $K_{T7}$ & 10 & constant & estimated \\
+        Weight RNA polymerase binding alone P70a & $K_{1}$ & 0.014 & constant & estimated \\
+        Weight bound RNAP-$\sigma_{70}$ P70a & $K_{2}$ & 10 & constant & estimated \\
+        $\sigma_{70}$ concentration & $\sigma_{70}$ & 35 & nM & \cite{Garamella:2016aa} \\
+        $\sigma_{70}$ dissociation constant & $K_{D}$ & 130 & nM & \cite{Mauri:2014aa} \\
+	      $\sigma_{70}$ hill coefficient & n & 1 & constant & \cite{Mauri:2014aa} \\
+        Gene concentration & $G_{\mathcal{P}}$ & 5 & nM &  \cite{Garamella:2016aa} \\ \bottomrule
+
+        % & & & & \\
+	      %
+        % Gene length of CAT & $l_{G}$ & 683 & nt & \cite{Kigawa1995} \\
+        % Gene length of deGFP & $l_{G}$ & 660 & nt & \cite{Garamella:2016aa} \\
+        % Protein length of CAT & $l_{P}$ & 229 & aa & \cite{Kigawa1995} \\
+        % Protein length of deGFP & $l_{P}$ & 219 & aa & \cite{Garamella:2016aa} \\
+    \end{tabular}
+\label{tbl:parameters}
+\end{table}
+
+\clearpage
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.00\textwidth]{./figs/Fig-1-Network-Validation-Simulations.pdf}
+\caption{Sequence specific flux balance analysis.
+A.
+Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway.
+Thick gray arrows indicate withdrawal of precursors for amino acid synthesis.
+B.
+CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption.
+Error bars denote the standard deviation of experimental measurements.
+C.
+deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption.
+Error bars denote a 10\% coefficient of variation.
+D.
+Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0.
+95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
+\label{fig:network}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[t!]
+\centering
+\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance.pdf}
+%\includegraphics[width=1.00\textwidth]{./figs/Fig-2-Performance-Predicted.pdf}
+\caption{The CFPS performance for eight model proteins with and without amino acid supplementation.
+A. Mean CFPS productivity for a panel of model proteins with and without amino acid supplementation.
+B. Mean CFPS productivity versus carbon number for a panel of model proteins with and without amino acid supplementation.
+Trendline (black dotted line) was calculated across all cases for a P70a promoter (R\textsuperscript{2} = 0.99) and maximum productivity trendline assumed $u\left(\kappa\right)$ = 1 (grey dotted line; R\textsuperscript{2} = 0.99).
+C. Mean CFPS energy efficiency for a panel of model proteins with and without amino acid supplementation.
+D. Mean CFPS energy efficiency versus carbon number for a panel of model proteins with and without amino acid supplementation.
+Trendline for cases with amino acids (black dotted line) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.65).
+E. Mean CFPS carbon yield for a panel of model proteins with and without amino acid supplementation.
+F. Mean CFPS carbon yield versus carbon number.
+Trendline for cases with amino acids (black dotted line; R\textsuperscript{2} = 0.95) and trendline for without amino acids (grey dotted line; R\textsuperscript{2} = 0.90).
+Error bars: 95\% CI calculated by sampling;
+asterisk: protein excluded from trendline;
+dagger: constrained by experimental measurements and excluded from trendline;
+triangles: first principle prediction and excluded from formulating trendline.}
+\label{fig:Prof}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[t!]
+\includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation, for three different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, and (C) amino acid synthesis without uptake.
+Mean flux across the ensemble, normalized to glucose uptake flux.
+Thick arrows indicate flux to or from amino acids.}
+\label{fig:flux}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[t!]
+\includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
+\caption{The deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 flux balance solutions, for amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
+\label{fig:oxphos_yield}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[t!]
+\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
+\caption{Total order sensitivity analysis of CFPS.
+A. Total order sensitivity of the optimal CAT productivity with respect to metabolic and transcription/translation parameters.
+B. Total order sensitivity of the optimal CAT energy efficiency.
+Metabolic and transcription/translation parameters were varied for amino acid supplementation and synthesis (black), amino acid supplementation without synthesis (dark grey) and amino acid synthesis without supplementation (light gray).
+Error bars represent a 95\% CI on the total-oder sensitivity index.}
+\label{fig:SI}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[t!]
+\includegraphics[width=1.00\textwidth]{./figs/Fig-6-FluxDistribition-Experimental.pdf}
+\caption{Constraint based simulation of CAT production for an experimentally constrained case.
+(A) Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation.
+Mean flux across the ensemble, normalized to glucose uptake flux. Thick arrows indicate flux to or from amino acids.
+(B) Central carbon metabolite measurements versus simulations over a one hour time course.}
+\label{fig:flux_exp}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[t!]
+\includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribution-Analysis.pdf}
+\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network.
+(A) CAT productivity norm compared to no knockouts in the experimentally constrained case.
+(B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
+\label{fig:norm}
+\end{figure}
+
+\clearpage
+
 \bibliography{References_v1}

### 2017-04-27 Paper_v1.tex
Some changed to ee paragraph

@@
 We independently validated the productivity correlation by calculating the optimal productivity of MBP using the full model and the effective correlation model (Fig.~\ref{fig:Prof}B).
-The prediction error was less than 8\% for an \emph{a~priori} prediction of CFPS productivity.
+The prediction error was less than 8\% for an \emph{a~priori} prediction of CFPS productivity using the effective correlation.
 Thus, the effective productivity correlation could be used as a parameter free method to estimate optimal productivity for cell free protein production using a P70a promoter.
@@
 ~\ref{fig:network}B).
-The experimentally constrained energy efficiency was 13.3 $\pm$ 5.0\% compared to the theoretical maximum of approximately 84 $\pm$ 0.1\%, however, the
-CAT productivity was similar between the simulated and measured systems.
-This suggested CFPS was not energy optimized.
-There were two factors responsible for the difference between the optimal and experimental system,
-the glucose consumption rate and the ATP yield per glucose.
-The glucose consumption rate was approximately 30 mM/h in the experimental system (even in the presence of amino acids).
-However, the constraint based simulation suggested the optimal glucose consumption rate was significantly less than the observed rate, approximately 2 - 4 mM/h.
+The experimentally constrained energy efficiency was 13.3 $\pm$ 5.0\% compared to the theoretical maximum of approximately 84 $\pm$ 0.1\%.
+Given that the CAT productivity was similar between the simulated and measured systems, differences in
+the glucose consumption rate and the ATP yield per glucose were likely responsible for the difference between the optimal and experimental systems.
+The glucose consumption rate was approximately 30 - 40 mM/h in the experimental system (even in the presence of amino acids).
+On the other hand, the constraint based simulation suggested the optimal glucose consumption rate was significantly less than the observed rate, approximately 1 - 7 mM/h (depending upon amino acid supplementation).
 In the constraint based simulation, the CFPS reaction produced only acetate as a by-product, but in the experimental system acetate, lactate, pyruvate, succinate and malate accumulated during the first hour of production.
@@
 while the experimentally constrained simulation produced only 4 ATPs per glucose.
-[TODO JV: Think more about this].
+Thus, approximately 120 - 160 mM ATP/hr was produced in the experimental system, in contrast to 12 - 84 mM ATP/hr for the optimal simulations.
+We know from measurements that ATP did not accumulate in the experimental system, rather it was consumed by a variety of pathways that were not active
+in the optimal simulation. This suggested that CFPS retained a growth like operational memory that led to over consumption of glucose, and the over production of ATP.
 

### 2017-04-27 Paper_v1.tex
More changes to results

@@
 
+% If only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
+% The carbon yield without the glucose contribution would be 100\%.
+
 \subsubsection{Carbon Yield}
@@
 and $R^{2} = 0.90$ without supplementation), irrespective of the promoter used to control protein expression.
-The MBP prediction also showed good agreement with the constraint based model, with an error between 3\% to 4\% with and without amino acid supplementation.
+The MBP prediction also showed good agreement with the constraint based model, with a prediction error between 3\% to 4\% with and without amino acid supplementation.
 The effective yield models were approximately parallel, with a drop in carbon yield of approximately 7\% without amino acids.
-In the absence of amino acids, glucose consumption increased, with the change in the carbon yield being largely accounted for by increased CO\textsubscript{2} production.
-If only the necessary amount of amino acids were used for the production of the protein of interest; thus, it may be hypothesized that all the glucose was used to power CFPS and did not contribute to the carbon yield.
-The carbon yield without the glucose contribution would be 100\%.
-However, in the experimentally constrained case, CAT was produced with a carbon yield of 6.2\% compared to the theoretical maximum of 58\%.
+The difference between the cases was associated with glucose consumption.
+In the absence of amino acids, glucose consumption increased, with the change in the carbon yield being
+largely accounted for by increased CO\textsubscript{2} production with supplementation.
+
+The carbon yield for experimentally constrained CAT production was 6.2\% compared to the theoretical maximum of 58\% (Fig.~\ref{fig:Prof}F, dagger).
 This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
@@
 \subsection{Global sensitivity analysis}
-To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI}).
+To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI} and Fig.
+XX?).
 Surprisingly, RNAP and ribosome abundance had only a modest effect;
@@
 However, its importance increased significantly when amino acids were not available.
-On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
-
-% The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
-
-The oxygen consumption rate was the most important factor controlling the energy efficiency of cell-free protein synthesis (Fig.~\ref{fig:SI}B).
-In addition to the oxygen consumption, amino acid utilization was also sensitive.
-On the other hand, glucose consumption was only sensitive in the absence of amino acid supplementation,
-since glucose acted as the sole source of carbon and energy for protein synthesis.
+On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.The oxygen consumption rate was the most important factor controlling the energy efficiency of cell-free protein synthesis (Fig.~\ref{fig:SI}B).
 In the model, we assumed that ATP could be produced by both substrate level and oxidative phosphorylation.
@@
 
+% The transcription/translation parameters had the same trend as for productivity, where the translation rate was the most sensitive compared to the other transcription/translation parameters and showed significance across all cases.
+% In addition to the oxygen consumption, amino acid utilization was also sensitive.
+% On the other hand, glucose consumption was only sensitive in the absence of amino acid supplementation,
+% since glucose acted as the sole source of carbon and energy for protein synthesis.
+
 To investigate the connection between carbon yield and oxidative phosphorylation further,
-we calculated the CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
+we calculated the optimal CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
 Oxidative phosphorylation had a strong effect on the carbon yield, both with and without amino acid supplementation.
@@
 In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of both acetate and lactate.
-The experimental dataset also exhibited a mixture of acetate and lactate accumulation during CAT synthesis, which suggested the CFPS reaction was not
+The experimental dataset exhibited a mixture of acetate and lactate accumulation during CAT synthesis, which suggested the CFPS reaction was not
 operating with optimal oxidative phosphorylation activity.
@@
 In addition, they were able to enhance the production yields of CAT by 33\% when the reaction was augmented with 10mM phosphate.
-However, it is unclear why the additional phosphate increased the CAT yield, but Jewett et al. believe it either enhances oxidative phosphorylation activity or inhibits phosphatase reactions.
-Taken together, the number, size, protein loading and lifetime of these vesicles remains an open area of study.
+However, it is unclear why the additional phosphate increased the CAT yield.
+Jewett et al. believe it either enhanced oxidative phosphorylation activity or inhibited phosphatase reactions.
+The number, size, protein loading and lifetime of these vesicles remains an open area of study.
 Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity and remains an important area to optimize for improved CFPS performance.
+
+% to both cases where amino acids were avaialbe.
+% This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
+% This leads to the relatively high flux in the first step of the pentose phosphate pathway to generate NADPH, thus, less NADH is available for oxidative phosphorylation.
+% The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
+% %Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
 
@@
 the energy efficiency and carbon yield decreased; in this case the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C).
-However, each of these three cases represent the theoretical optimum under different conditions; in order to more accurately describe system performance, we constrained the feasible solution space with experimental measurements.
+However, these simulations represent the theoretical optimum under different conditions.
+To more accurately describe system performance, we constrained the feasible solution space with experimental measurements
+and estimated the optimal CAT flux distribution.
 
-% to both cases where amino acids were avaialbe.
-% This drop in performance is due to the burden of synthesizing amino acids, which require NADPH.
-% This leads to the relatively high flux in the first step of the pentose phosphate pathway to generate NADPH, thus, less NADH is available for oxidative phosphorylation.
-% The performance metrics and sensitivity analysis suggest that efficient energy generation via oxygen uptake is essential to higher energy efficiency and carbon yields.
-% %Thus, removing anaerobic enzymes during the cell-free extract preparation could potentially improve CFPS performance and protein yield.
-
-The case constrained by experimental measurements (Fig.
-\ref{fig:flux_exp}) had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.52 with the cases supplied with amino acids.
-Thus there are some differences in the flux distribution compared to the optimum solutions which may provide some insight to improve CFPS performance.
-Metabolic fluxes were constrained by experimental measurements (available in Supporting Information) where available for the first hour
-which constrained the feasible solution space.
+The experimentally constrained optimal flux distribution had a 52\% similarity with the theoretically optimal flux distribution (Fig. \ref{fig:flux_exp}).
+Metabolic fluxes were constrained by experimental measurements of glucose, amino and organic acid consumption and production rates for the first hour of the reaction  (supporting information).
+The low similarity suggested several differences between the solutions which could be exploited to improve CFPS performance.
 The central carbon organic acids showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}=0.92 (Fig.~\ref{fig:flux_exp}B).
@@
 
+% in the flux distribution compared to the optimum solutions
+% The case constrained by experimental measurements  had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.52 with the cases supplied with amino acids.
+
 \subsubsection{Potential Alternative Metabolic Optima}
+Optimal flux distributions predicted using constraint based approaches may not always be unique.
 Alternative optimal solutions have the same objective value, e.g., productivity, but different metabolic flux distributions.
@@
 However, there was a difference in the flux distribution with these knockouts (Fig.~\ref{fig:norm}B); thus, alternative optimal metabolic flux distributions exist for CAT production, despite experimental constraints.
-In addition, knockouts of amino acid biosynthesis reactions had no effect on the productivity with the exception of ALA, ASP, ASN and GLU, GLN biosynthesis reactions, since amino acids were supplemented in the media.
-ssFBA rerouted the metabolic flux to have the same productivity but an alterantive flux distribution (alternative optima), as can be seen for a pairwise knockout of GLY, SER biosynthesis with HIS biosynthesis (Fig.~\ref{fig:norm}B).
+In addition, knockouts of amino acid biosynthesis reactions had no effect on the productivity with the exception of alanine, aspartate, asparagine, glutamate and glutamine biosynthesis reactions, since amino acids were available in the media.
 Ultimately, to determine the metabolic flux distribution occurring in CFPS, we need to add additional constraints to the flux estimation calculation.
@@
 \begin{figure}[t!]
+\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
+\caption{Total order sensitivity analysis of CFPS.
+A. Total order sensitivity of the optimal CAT productivity with respect to metabolic and transcription/translation parameters.
+B. Total order sensitivity of the optimal CAT energy efficiency.
+Metabolic and transcription/translation parameters were varied for amino acid supplementation and synthesis (black), amino acid supplementation without synthesis (dark grey) and amino acid synthesis without supplementation (light gray).
+Error bars represent a 95\% CI on the total-oder sensitivity index.}
+\label{fig:SI}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[t!]
 \includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
@@
 \label{fig:oxphos_yield}
-\end{figure}
-
-\clearpage
-
-\begin{figure}[t!]
-\includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
-\caption{Total order sensitivity analysis of CFPS.
-A. Total order sensitivity of the optimal CAT productivity with respect to metabolic and transcription/translation parameters.
-B. Total order sensitivity of the optimal CAT energy efficiency.
-Metabolic and transcription/translation parameters were varied for amino acid supplementation and synthesis (black), amino acid supplementation without synthesis (dark grey) and amino acid synthesis without supplementation (light gray).
-Error bars represent a 95\% CI on the total-oder sensitivity index.}
-\label{fig:SI}
 \end{figure}

### 2017-04-27 Paper_v1.tex
final changes... hopefully

@@
 The carbon yield for experimentally constrained CAT production was 6.2\% compared to the theoretical maximum of 58\% (Fig.~\ref{fig:Prof}F, dagger).
-This decrease in carbon yield suggests inefficiencies in CFPS that can potentially be improved.
-Constraint based calculations assume a metabolic pseudo steady state; thus, intermediate metabolites cannot accumulate within the cell-free extract.
-In addition, the flux balance analysis calculation is solved by maximizing the flux through the protein production reaction.
-Therefore, carbon flux will travel through the network to optimize the maximum flux through the protein synthesis reaction.
-In examining the experimental dataset, there is a high accumulation of organic acids, especially acetate, pyruvate and lactate.
-The experimental performance could be improved by diverting this carbon toward the protein of interest by knockouts during the cell-free extract preparation.
+Given that the CAT productivity was similar between the simulated and measured systems, differences in
+the glucose consumption rate and byproduct accumulation were likely responsible for the difference between the optimal and experimental systems.
+The translation rate has been identified as the rate-limiting step for cell-free protein synthesis \cite{2005_underwood_biotech,2014_li_PlosOne}, thus increased glucose consumption only hindered the carbon yield of CAT.
+The experimental system was not as energy efficient as the optimal system which resulted in the increased glucose consumption. 
+Since CAT productivity could not be improved, the increased carbon consumption accumulated in byproducts such as pyruvate, acetate, lactate, succinate and malate. 
+The accumulation of acetate and lactate also indicated that oxidative phosphorylation activity was limited.
+Thus to improve carbon yield for CFPS, energy efficiency should be improved to lower the glucose consumption rate, or the translation rate should be improved for increased protein productivity.  
+%%%%%%%%%%%%%%%%%%%%
+However, it may also be hypothesized that only the amino acids supplied contribute to the carbon yield of the protein of interest, while glucose is used to energize CFPS. 
+In this case, the carbon yield for the optimal and experimental systems was approximately 100\%.
+Therefore, the energy efficiency of CFPS is left to be improved. 
+%%%%%%%%%%%%%%%%%%%%%
 
@@
 To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI} and Fig.
-XX?).
+1 in supporting information).
 Surprisingly, RNAP and ribosome abundance had only a modest effect;
@@
 This suggested that the translation elongation rate, and not transcriptional parameters, controlled productivity.
-Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased protein synthesis rates by 33\% and yields by 23\% at 30 minutes \cite{2005_underwood_biotech}.
+Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased protein synthesis rates by 27\% \cite{2005_underwood_biotech}.
 In addition, Li et al.
-increased the productivity of firefly luciferase production in PURE CFPS by 5-fold by adjusting factors that affect transcription and translation such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+increased the productivity by 5-fold of firefly luciferase in PURE CFPS by first improving the rate-limiting step, translation, followed by transcription by adjusting elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
+%In addition, Li et al.
+increased the productivity of firefly luciferase production in PURE CFPS by 5-fold by adjusting factors that first affected translation and then transcription such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
 In examining substrate utilization, glucose consumption was not important for productivity in the presence of amino acid supplementation.
 However, its importance increased significantly when amino acids were not available.
-On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.The oxygen consumption rate was the most important factor controlling the energy efficiency of cell-free protein synthesis (Fig.~\ref{fig:SI}B).
+On the other hand, amino acid consumption was only sensitive when amino acids synthesis reactions were blocked, as it was the only source of amino acids for CAT synthesis.
+The oxygen consumption rate was the most important factor controlling the energy efficiency of cell-free protein synthesis (Fig.~\ref{fig:SI}B).
 In the model, we assumed that ATP could be produced by both substrate level and oxidative phosphorylation.
 Jewett and coworkers reported that oxidative phosphorylation still operated in cell-free systems, and that the protein
-yield decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
+titer decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
 However, it is unknown how active oxidative phosphorylation is for a glucose powered cell-free system.
@@
 The low similarity suggested several differences between the solutions which could be exploited to improve CFPS performance.
+The largest discrepency was in oxidative phosphorylation reactions, where the experimental system relied heavily on \textit{cyd} rather than \textit{cyo} which was less efficient for energy generation.
+In addition, the \textit{pnt}1 reaction, which interconverts NADPH to NADH, was active in the experimental system and not in the optimal systems; this suggests CFPS may be limited by NADH.
+The next largest discrepency was in folate, purine and pyrimidine metabolism which were inactive in the optimum systems followed by amino acid biosynthesis. 
+The experimentally constrained system had a high flux for alanine and glutamine biosynthesis which accumulated in the media, whereas in the optimum cases there was no accumulation of amino acids. 
 The central carbon organic acids showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}=0.92 (Fig.~\ref{fig:flux_exp}B).
@@
 In this case, it is unclear which substrate (glucose or amino acids) is used to power CFPS and may in fact be a combination of both.
-The optimum solutions only produced the required amount of amino acids necessary, however in examining the measurements there is an accumulation of alanine and glutamine which may explain some of the differences in the correlations of the flux distributions.
-Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig.~\ref{fig:flux_exp}B) and is an inefficiency of carbon utilization.
+The optimum solutions only produced the required amount of amino acids necessary; however, in examining the measurements there was an accumulation of alanine and glutamine which may explain some of the differences in the correlations of the flux distributions.
+Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig.~\ref{fig:flux_exp}B) and may contribute to the inefficiency of carbon utilization.
 
@@
 where $\rm {ATP}_{T}$, $\rm {CTP}_{T}$, $\rm {GTP}_{T}$, $\rm {UTP}_{T}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{X}$, $\rm {GTP}_{X}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate,
-and $\texttt{ATP}_{GLC}$ denotes the equivalent ATP number for glucose produced.
+and $\texttt{ATP}_{GLC}$ denotes the equivalent ATP number produced per glucose.
 The energy species stoichiometric coefficients are available in the Supporting Information.

### 2017-04-28 Paper_v1.tex
Small changes

@@
 the energy efficiency and carbon yield decreased; in this case the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C).
-However, these simulations represent the theoretical optimum under different conditions.
-To more accurately describe system performance, we constrained the feasible solution space with experimental measurements
+However, these simulations represent the theoretical optimum.
+To more accurately describe actual metabolic flux, we constrained the feasible solution space with experimental measurements
 and estimated the optimal CAT flux distribution.
@@
 The low similarity suggested several differences between the solutions which could be exploited to improve CFPS performance.
-The largest discrepency was in oxidative phosphorylation reactions, where the experimental system relied heavily on \textit{cyd} rather than \textit{cyo} which was less efficient for energy generation.
-In addition, the \textit{pnt}1 reaction, which interconverts NADPH to NADH, was active in the experimental system and not in the optimal systems; this suggests CFPS may be limited by NADH.
-The next largest discrepency was in folate, purine and pyrimidine metabolism which were inactive in the optimum systems followed by amino acid biosynthesis. 
-The experimentally constrained system had a high flux for alanine and glutamine biosynthesis which accumulated in the media, whereas in the optimum cases there was no accumulation of amino acids. 
-The central carbon organic acids showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}=0.92 (Fig.~\ref{fig:flux_exp}B).
+The largest discrepancy was in oxidative phosphorylation,
+where the experimental system relied heavily on \textit{cyd} rather than \textit{cyo} which was less efficient for energy generation.
+In addition, the \textit{pnt}1 reaction, which interconverts NADPH to NADH, was active in the experimental system and not in the optimal systems;
+this suggested CFPS may be limited by NADH.
+Next, folate, purine and pyrimidine metabolism and amino acid biosynthesis were inactive in the optimum systems, but active in the experimental system.
+In particular, the experimentally system had high alanine and glutamine biosynthetic flux (both accumulated in the media),
+while there was no accumulation of amino acids in the optimal simulations.
+Estimated organic acid flux showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}~=~0.92 (Fig.~\ref{fig:flux_exp}B).
 Only certain amino acid synthesis reactions were blocked, since during the growth of \emph{E.~coli} not all amino acids were supplied (see Materials and Methods).

### 2017-04-28 Paper_v1.tex
final updates... hopefully

@@
 Protein expression simulations were consistent with measurements for a variety of cases.
+We compared the theoretical optimum solutions with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to over consumption of glucose, and low energy efficiency due to suboptimal oxidative phosphorylation activity.
 Further, global sensitivity analysis identified the key metabolic processes that controlled the productivity, energy efficiency, and carbon yield of the process.
@@
 We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight different proteins.
-From this, higher carbon number proteins typically had lower productivity rates, energy efficiency, and carbon yields than that of the lower carbon number proteins.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency
-and carbon yield and the translation rate for productivity.
+From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins; whereas energy efficiency was independent of protein size.
+Our model correlations were independently validated with a protein not in our original data set.
+We compared the theoretical optimum solutions of ssFBA with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to over consumption of glucose, and over production of ATP.
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
 Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
@@
 We know from measurements that ATP did not accumulate in the experimental system, rather it was consumed by a variety of pathways that were not active
-in the optimal simulation. This suggested that CFPS retained a growth like operational memory that led to over consumption of glucose, and the over production of ATP.
+in the optimal simulation. This suggested that CFPS retained a growth-like operational memory that led to over consumption of glucose, and the over production of ATP.
 
@@
 the glucose consumption rate and byproduct accumulation were likely responsible for the difference between the optimal and experimental systems.
-The translation rate has been identified as the rate-limiting step for cell-free protein synthesis \cite{2005_underwood_biotech,2014_li_PlosOne}, thus increased glucose consumption only hindered the carbon yield of CAT.
-The experimental system was not as energy efficient as the optimal system which resulted in the increased glucose consumption.
+The translation rate has been identified as the rate-limiting step for cell-free protein synthesis \cite{2005_underwood_biotech,2014_li_PlosOne}, which was confirmed by the translation rate flux always hitting the upper bound in the ssFBA simulation, regardless of protein.
 Since CAT productivity could not be improved, the increased carbon consumption accumulated in byproducts such as pyruvate, acetate, lactate, succinate and malate.
-The accumulation of acetate and lactate also indicated that oxidative phosphorylation activity was limited.
-Thus to improve carbon yield for CFPS, energy efficiency should be improved to lower the glucose consumption rate, or the translation rate should be improved for increased protein productivity.
+In addition, the accumulation of acetate and lactate suggested that oxidative phosphorylation activity was limited.
+Thus to improve carbon yield for CFPS, the translation rate should be improved for increased protein productivity, or energy efficiency should be improved to lower the glucose consumption rate.
 %%%%%%%%%%%%%%%%%%%%
-However, it may also be hypothesized that only the amino acids supplied contribute to the carbon yield of the protein of interest, while glucose is used to energize CFPS.
-In this case, the carbon yield for the optimal and experimental systems was approximately 100\%.
-Therefore, the energy efficiency of CFPS is left to be improved.
+%However, it may also be hypothesized that only the amino acids supplied contribute to the carbon yield of the protein of interest, while glucose is used to energize CFPS.
+%If this is the case, the carbon yield for the optimal and experimental systems resulted in approximately 100\%.
 %%%%%%%%%%%%%%%%%%%%%
@@
 increased the productivity by 5-fold of firefly luciferase in PURE CFPS by first improving the rate-limiting step, translation, followed by transcription by adjusting elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
-%In addition, Li et al.
-increased the productivity of firefly luciferase production in PURE CFPS by 5-fold by adjusting factors that first affected translation and then transcription such as elongation factors, ribosome recycling factor, release factors, chaperones, BSA, and tRNAs \cite{2014_li_PlosOne}.
 In examining substrate utilization, glucose consumption was not important for productivity in the presence of amino acid supplementation.
@@
 Jewett et al. hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation \cite{Jewett:2008aa}.
-In addition, they were able to enhance the production yields of CAT by 33\% when the reaction was augmented with 10mM phosphate.
+In addition, they were able to enhance the production titer of CAT by 33\% when the reaction was augmented with 10mM phosphate.
 However, it is unclear why the additional phosphate increased the CAT yield.
@@
 The experimentally constrained optimal flux distribution had a 52\% similarity with the theoretically optimal flux distribution (Fig. \ref{fig:flux_exp}).
-Metabolic fluxes were constrained by experimental measurements of glucose, amino and organic acid consumption and production rates for the first hour of the reaction (supporting information).
+Metabolic fluxes were constrained by experimental measurements of glucose, amino and organic acid consumption and production rates for the first hour of the reaction (supporting information) and showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}~=~0.92 (Fig.~\ref{fig:flux_exp}B).
 The low similarity suggested several differences between the solutions which could be exploited to improve CFPS performance.
@@
 where the experimental system relied heavily on \textit{cyd} rather than \textit{cyo} which was less efficient for energy generation.
-In addition, the \textit{pnt}1 reaction, which interconverts NADPH to NADH, was active in the experimental system and not in the optimal systems;
-this suggested CFPS may be limited by NADH.
-Next, folate, purine and pyrimidine metabolism and amino acid biosynthesis were inactive in the optimum systems, but active in the experimental system.
-In particular, the experimentally system had high alanine and glutamine biosynthetic flux (both accumulated in the media),
+In addition, the experimental system had a high flux through the \textit{zwf} yielding an abundance of NADPH used for amino acid biosynthesis.
+The remaining NADPH was interconverted to NADH via the \textit{pnt}1 reaction which was consumed to convert pyruvate to lactate.
+In contrast, the optimal solutions with amino acid supplementation had low activity in \textit{zwf} and \textit{pnt}1 reactions. 
+Next, folate, purine, and pyrimidine metabolism along with amino acid biosynthesis were inactive in the optimal system, but active in the experimental system.
+In particular, the experimental system had high alanine and glutamine biosynthetic flux (both accumulated in the media),
 while there was no accumulation of amino acids in the optimal simulations.
-Estimated organic acid flux showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}~=~0.92 (Fig.~\ref{fig:flux_exp}B).
-Only certain amino acid synthesis reactions were blocked, since during the growth of \emph{E.~coli} not all amino acids were supplied (see Materials and Methods).
-During the cell-free reaction all amino acids were supplied, however glucose still traveled through all the major pathways, and the same metabolic precursors were still utilized for amino acid biosynthesis.
-In this case, it is unclear which substrate (glucose or amino acids) is used to power CFPS and may in fact be a combination of both.
-The optimum solutions only produced the required amount of amino acids necessary; however, in examining the measurements there was an accumulation of alanine and glutamine which may explain some of the differences in the correlations of the flux distributions.
-Accumulation of pyruvate, lactate, acetate, and other organic acids can be seen (Fig.~\ref{fig:flux_exp}B) and may contribute to the inefficiency of carbon utilization.
-
-% in the flux distribution compared to the optimum solutions
-% The case constrained by experimental measurements  had the highest correlation of 0.66 with the flux distribution from the case supplied with no amino acids and a correlation of 0.52 with the cases supplied with amino acids.
+Even though all amino acids were supplied in the reaction, glucose traveled through the major pathways and was utilized for amino acid biosynthesis.
+Thus, it is unclear which substrate (glucose or amino acids) was used to power CFPS and may in fact be a combination of both.
+The optimum solutions only produced the required amount of amino acids necessary; meanwhile, in the experimental system there was an accumulation of alanine, glutamine, pyruvate, lactate, acetate, malate, and succinate which contributed to some of the differences in the flux distributions.
+% and may have contributed to the inefficiency of carbon utilization.
 
@@
 Likewise, there were pathway knockouts that had no effect on productivity or the metabolic flux distribution, such as removal of isoleucine, leucine, histidine and valine biosynthesis.
-Globally, the constraint based simulation reached the same optimal CAT productivity for 40\% of the pairwise knockouts, but had different flux distributions compared with the absence of knockouts.
+Globally, the constraint based simulation reached the same optimal CAT productivity for 40\% of the pairwise knockouts, while 92\% of these solutions had different flux distributions compared with the wild-type.
 For example, one of the features of the predicted optimal metabolic flux distribution was a high flux through the Entner-Douodoroff (ED) pathway.
@@
 \subsection*{Potential alternative optimal metabolic flux solutions.}
-We identified potential alternative optimal metabolic flux distributions by performing single and pairwise reaction group knockout simulations [TODO:REF].
+We identified potential alternative optimal metabolic flux distributions by performing single and pairwise reaction group knockout simulations \cite{Mahadevan2003264}.
 Reaction group knockouts were simulated by setting the flux bounds for all the reactions involved in a group to zero.
@@
 
+We thank Che-Hsiao Shih (Purdue University) for identification of DNA and protein sequences. 
 This study was supported by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V. for the support of M.V.

### 2017-05-05 Paper_v1.tex
Small changes

@@
 However, if CFPS is to become a mainstream technology for advanced applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
-One tool to address these questions is mathematical modeling.
+One tool to address these questions is constraint based modeling.
 
@@
 different \textit{E.~coli} cell-free extracts.
+In particular, we calculated the transcription rate using effective promoter models in combination with a maximum transcription rate estimated from literature.
+We then maximized the rate of translation, which was bounded from above by an estimate of the maximum translation rate from literature.
+The parameters in the transcription and translation models were largely derived from literature (Table ~\ref{tbl:parameters}).
+Both the transcription and translation rates were subject to resource constraints encoded by the metabolic network.
 The cell-free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab website \cite{varnerlab}.

### 2017-05-05 Paper_v1.tex
Small changes

@@
 Our model correlations were independently validated with a protein not in our original data set.
-We compared the theoretical optimum solutions of ssFBA with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to over consumption of glucose, and over production of ATP.
+We compared the theoretical optimal solutions with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to over consumption of glucose, and over production of ATP.
 Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
@@
 the glucose consumption rate and byproduct accumulation were likely responsible for the difference between the optimal and experimental systems.
-The translation rate has been identified as the rate-limiting step for cell-free protein synthesis \cite{2005_underwood_biotech,2014_li_PlosOne}, which was confirmed by the translation rate flux always hitting the upper bound in the ssFBA simulation, regardless of protein.
+The translation rate has been identified as the rate-limiting step for cell-free protein synthesis \cite{2005_underwood_biotech,2014_li_PlosOne}, which was confirmed by the translation rate flux always hitting the upper bound in the simulation, regardless of protein.
 Since CAT productivity could not be improved, the increased carbon consumption accumulated in byproducts such as pyruvate, acetate, lactate, succinate and malate.

### 2017-05-05 Paper_v1.tex
final edits

@@
 {Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY 14853}
+\author{Che-Hsiao Shih}
+\affiliation[Purdue University]
+{Davidson School of Chemical Engineering, Purdue University, West Lafayette, IN 47907}
 \author{Joseph Wayman}
@@
 %\abbreviations{MV,NH,JDV}
-\keywords{Synthetic biology, constraints based modeling, cell-free protein synthesis}
+\keywords{Synthetic biology, constraint based modeling, cell-free protein synthesis}
 
@@
 \begin{abstract}
-Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized medicine.
+Cell-free protein synthesis (CFPS) has become a widely used research tool in systems and synthetic biology.
 In this study, we used sequence specific constraint based modeling to evaluate the performance of an \emph{E.~coli} cell-free protein synthesis system.
-A core \emph{E.~coli} metabolic model, describing glycolysis, the pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism,
-was augmented with sequence specific descriptions of transcription and translation processes, and effective models of promoter function.
-Thus, sequence specific constraint based modeling explicitly couples transcription and translation processes, and the regulation of gene expression,
-with the availability of metabolic resources.
-We tested this approach by simulating the expression of two model proteins chloramphenicol acetyltransferase and dual emission green fluorescent protein for which we have training data sets;
-we then expanded the simulations to a range of therapeutically relevant proteins.
+A core \emph{E.~coli} metabolic model, describing glycolysis, the pentose phosphate pathway, energy metabolism, amino acid biosynthesis and degradation was augmented with sequence specific descriptions of transcription and translation with effective models of promoter function.
+Thus, sequence specific constraint based modeling explicitly couples transcription and translation processes and the regulation of gene expression with the availability of metabolic resources.
+We tested this approach by simulating the expression of two model proteins: chloramphenicol acetyltransferase and dual emission green fluorescent protein, for which we have training data sets; we then expanded the simulations to a range of therapeutically relevant proteins.
 Protein expression simulations were consistent with measurements for a variety of cases.
-We compared the theoretical optimum solutions with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to over consumption of glucose, and low energy efficiency due to suboptimal oxidative phosphorylation activity.
+We compared the theoretical optimal solutions with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to overconsumption of glucose, and low energy efficiency due to suboptimal oxidative phosphorylation activity.
 Further, global sensitivity analysis identified the key metabolic processes that controlled the productivity, energy efficiency, and carbon yield of the process.
@@
 Cell-free protein expression has become a widely used research tool in systems and synthetic biology, and a promising technology for personalized protein production.
-From the perspective of research, cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
-Central amongst these, is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall, or complications associated with cell growth.
-This allows us to interrogate the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
+Cell-free systems offer many advantages for the study, manipulation and modeling of metabolism compared to \textit{in vivo} processes.
+Central amongst these is direct access to metabolites and the biosynthetic machinery without the interference of a cell wall or the complications associated with cell growth.
+This allows interrogation of the chemical environment while the biosynthetic machinery is operating, potentially at a fine time resolution.
 Cell-free protein synthesis (CFPS) systems are arguably the most prominent examples of cell-free systems used today \cite{Jewett:2008aa}.
@@
 Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
-However, if CFPS is to become a mainstream technology for advanced applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and cost of these systems \cite{Jewett:2008aa}.
+However, if CFPS is to become a mainstream technology for advanced applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and costs of these systems \cite{Jewett:2008aa}.
 One tool to address these questions is constraint based modeling.
@@
 Interestingly, similar to early cybernetic models, the most common metabolic objective function has been the optimization of biomass formation \cite{2002_ibarra_edwards_palsson_Nat}, although other metabolic objectives have also been estimated \cite{2007_schuetz_sauer_MolSysBio}.
-% Recent advances in constraints based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}.
-Thus, modern constraints based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
+% Recent advances in constraint based modeling have overcome the early shortcomings of the platform, including capturing metabolic regulation and control \cite{2013_hyduke_lewis_palsson_MolBioSys}.
+Thus, modern constraint based approaches are extremely useful for the discovery of metabolic engineering strategies and represent the state of the art in metabolic modeling \cite{2013_mccloskey_palsson_feist_MolSysBio, 2012_zomorrodi_maranas_MetaEng}.
 
 Stoichiometric reconstructions of microbial metabolism, popularized by constraint based approaches such as flux balance analysis (FBA), have become standard tools to interrogate metabolism \cite{2012_lewis_palsson_NatRevMicrobio}.
-Constraint based methods such as FBA, metabolic flux analysis (MFA) \cite{WIECHERT2001195}, and convex network decomposition approaches such as elementary modes \cite{Schuster:2000aa} or extreme pathways \cite{Schilling:2000aa} model intracellular metabolism using biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
-Constraint based approaches have predicted productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} for biochemical networks of arbitrary complexity, including genome scale networks, using linear programming \cite{Covert:2004aa}.
-Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} or \textit{B.
+FBA and metabolic flux analysis (MFA) \cite{WIECHERT2001195}, as well as convex network decomposition approaches such as elementary modes \cite{Schuster:2000aa} and extreme pathways \cite{Schilling:2000aa}, model intracellular metabolism using the biochemical stoichiometry and other constraints such as thermodynamical feasibility \cite{Henry:2007,Hamilton:2013} under pseudo steady state conditions.
+Constraint based approaches have used linear programming \cite{Covert:2004aa} to predict productivity \cite{Varma01101994, BTPR:BTPR2132}, yield \cite{Varma01101994}, mutant behavior \cite{Edwards2000}, and growth phenotypes \cite{Brien693} for biochemical networks of varying complexity, including genome scale networks.
+Since the first genome scale stoichiometric model of \textit{E.~coli} \cite{2000_edwards_palsson_PNAS}, stoichiometric reconstructions of hundreds of organisms, including industrially important prokaryotes such as \textit{E.~coli} \cite{Feist:2007aa} and \textit{B.
 subtilis} \cite{Oh:2007aa}, are now available \cite{2009_feist_palsson_NatRevMicrobio}.
-Stoichiometric reconstructions have been expanded to include the integration of metabolism with detailed descriptions of gene expression (ME-Model) \cite{Allen:2003aa, Brien693},
-and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220}.
+Stoichiometric reconstructions have been expanded to include the integration of metabolism with detailed descriptions of gene expression (ME-Model) \cite{Allen:2003aa, Brien693} and protein structures (GEM-PRO) \cite{Zhang1544,Chang1220}.
 These expansions have greatly increased the scope of questions these models can explore.
@@
 
-In this study, we used sequence specific constraint based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis (CFPS).
-A core \emph{E.~coli} cell-free metabolic model was developed from literature \cite{Feist:2007aa}.
-This model, which described glycolysis, pentose phosphate pathway, amino acid biosynthesis and degradation and energy metabolism, was then augmented with
-sequence specific descriptions of promoter function, transcription and translation processes.
-Thus, sequence specific constraints based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
-We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight different proteins.
-From this, higher carbon number proteins typically had lower productivity rates and carbon yields than that of the lower carbon number proteins; whereas energy efficiency was independent of protein size.
-Our model correlations were independently validated with a protein not in our original data set.
-We compared the theoretical optimal solutions with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to over consumption of glucose, and over production of ATP.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation as instrumental for maintaining a high energy efficiency and carbon yield and the translation rate for productivity.
-Taken together, sequence specific constraints based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
+In this study, we used sequence specific constraint based modeling to evaluate the performance of \emph{E.~coli} cell-free protein synthesis.
+A core \emph{E.~coli} cell-free metabolic model describing glycolysis, pentose phosphate pathway, energy metabolism, amino acid biosynthesis and degradation was developed from literature \cite{Feist:2007aa}; it was then augmented with sequence specific descriptions of promoter function, transcription and translation processes.
+Thus, sequence specific constraint based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
+We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield of production for eight different proteins.
+It was observed that higher carbon number proteins typically had lower productivity rates and carbon yields than those of the lower carbon number proteins, whereas energy efficiency was independent of protein size.
+Model correlations were then independently validated with a protein not in the original data set.
+We compared the theoretical optimal solutions with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to overconsumption of glucose.
+Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation vital to energy efficiency and carbon yield, and translation rate most important for productivity.
+Taken together, sequence specific constraint based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
 
@@
 % The goal of this work was first to construct a modeling framework to describe cell-free protein synthesis systems and to examine its performance in productivity, energy efficiency and carbon yield for a protein of interest.
-% One mathematical framework that has found wide use in modeling metabolism is constraints based models such as flux balance analysis (FBA).
-% FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady-state conditions.
+% One mathematical framework that has found wide use in modeling metabolism is constraint based models such as flux balance analysis (FBA).
+% FBA can predict how cells utilize nutrients to produce products by using the biochemical stoichiometry and thermodynamical feasibility under pseudo steady state conditions.
 % Traditionally, FBA is used to model \textit{in~vivo} processes; however, cell-free systems do not have growth associated reactions or transport through the cell membrane.
@@
 The cell-free stoichiometric network was constructed by removing growth associated reactions from the \textit{i}AF1260 reconstruction of K-12 MG1655 \textit{E.~coli} \cite{Feist:2007aa},
-and adding deletions associated with the specific cell-free system (see materials and methods).
+and adding deletions associated with the specific cell-free system (see Materials and Methods).
 We then added the transcription and translation template reactions of Allen and Palsson for the specific proteins of interest \cite{Allen:2003aa}.
-A schematic of the metabolic network, which consisted of 264 reactions and 146 species, is shown in Fig.~\ref{fig:network}A.
+A schematic of the metabolic network, consisting of 264 reactions and 146 species, is shown in Fig.~\ref{fig:network}A.
 The network described the major carbon and energy pathways, as well as amino acid biosynthesis and degradation pathways.
 Using this network, in combination with effective promoter models taken from Moon et al. \cite{Moon:2012ab}, and literature values for cell-free culture parameters (Table ~\ref{tbl:parameters}),
-we simulated the sequence specific production of two model proteins, chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP) using
-different \textit{E.~coli} cell-free extracts.
-In particular, we calculated the transcription rate using effective promoter models in combination with a maximum transcription rate estimated from literature.
-We then maximized the rate of translation, which was bounded from above by an estimate of the maximum translation rate from literature.
-The parameters in the transcription and translation models were largely derived from literature (Table ~\ref{tbl:parameters}).
-Both the transcription and translation rates were subject to resource constraints encoded by the metabolic network.
+we simulated the sequence specific production of two model proteins: chloramphenicol acetyltransferase (CAT) and dual emission green fluorescent protein (deGFP), using different \textit{E.~coli} cell-free extracts.
+%We calculated the transcription rate using effective promoter models in combination with a maximum transcription rate estimated from literature.
+%We then maximized the rate of translation, which was bounded from above by an estimate of the maximum translation rate from literature.
+%The parameters in the transcription and translation models were largely derived from literature (Table ~\ref{tbl:parameters}).
+%Both the transcription and translation rates were subject to resource constraints encoded by the metabolic network.
+We calculated the transcription rate using effective promoter models, then maximized the rate of translation within biologically realistic bounds.
+Transcription and translation rates were subject to resource constraints encoded by the metabolic network, and transcription and translation model parameters were largely derived from literature (Table ~\ref{tbl:parameters}).
 The cell-free metabolic network, along with all model code and parameters, can be downloaded under an MIT software license from the Varnerlab website \cite{varnerlab}.
 
-Cell-free simulations of the time evolution of CAT and deGFP production were consistent with experimental measurements (Fig. ~\ref{fig:network}B and C).
-Chloramphenicol acetyltransferase (CAT), was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress}
-using glucose as a carbon and energy source (Fig.~\ref{fig:network}B).
-With the exception of the first 10-15 min, the cell-free prediction of CAT abundance was consistent with the measured values (coefficient of determination, R\textsuperscript{2} = 0.86).
-On the other hand, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for eight hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
-The cell-free simulation captured the overall trend of deGFP abundance, but failed to capture saturation at the end of the CFPS culture (R\textsuperscript{2} = 0.84).
-Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes,
-transcription and translation elongation rates or the upper bounds for oxygen and glucose consumption rates (modeled as being normally distributed around the parameter values shown in Table ~\ref{tbl:parameters})
-did not qualitatively alter the performance of the model (blue region, 95\% confidence estimate).
-Together, these simulations suggested the description of transcription and translation, and its integration with metabolism encoded in the cell-free metabolic model, were consistent with experimental measurements.
-However, these simulations were conducted at a single plasmid concentration of 5 nM.
+Cell-free simulations of the time evolution of CAT and deGFP production were consistent with experimental measurements (Fig.~\ref{fig:network}B and C).
+CAT was produced under a T7 promoter in a glucose/NMP cell-free system \cite{2005_calhoun_BiotechnologyProgress} using glucose as a source of carbon and energy (Fig.~\ref{fig:network}B).
+Apart from the first 10-15 min, the prediction of CAT abundance was consistent with the measured cell-free values (coefficient of determination, R\textsuperscript{2} = 0.86).
+Meanwhile, deGFP was produced under a P70a promoter in TXTL 2.0 \textit{E.~coli} extract for eight hours using maltose as a carbon and energy source (Fig.~\ref{fig:network}C).
+The cell-free simulation predicted the overall deGFP abundance, but failed to capture saturation at the end of the CFPS culture (R\textsuperscript{2} = 0.84).
+Uncertainty in experimental factors such as the concentration of RNA polymerase, ribosomes, transcription and translation elongation rates, as well as the upper bounds on oxygen and glucose consumption rates (modeled as being normally distributed around the parameter values shown in Table ~\ref{tbl:parameters}), did not qualitatively alter the performance of the model (blue region, 95\% confidence estimate).
+Together, these simulations suggested the description of transcription and translation, and its integration with metabolism encoded in the cell-free model, were consistent with experimental measurements.
+However, these simulations were only conducted at a single plasmid concentration of 5 nM.
 Thus, it was unclear if the model could capture cell-free protein synthesis for a range of plasmid concentrations.
@@
 Cell-free simulations of the deGFP titer for a range of plasmid concentrations were consistent with experimental measurements (Fig.~\ref{fig:network}D).
-The cell-free deGFP titer at each plasmid concentration was calculated by multiplying the flux of deGFP synthesis by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
+The titer at each plasmid concentration was calculated by multiplying the deGFP synthesis flux by the active time of production, approximately 8 hours in TXTL 2.0 \cite{Garamella:2016aa}.
 The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration (R\textsuperscript{2} = 0.97).
-However, while the mean and 95\% confidence estimate of the ensemble were consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating point of 5 nM of plasmid concentration.
-These results, in combination with the time-dependent CAT and deGFP simulations, validated our CFPS modeling approach which required very few adjustable parameters.
-It also showed that the sequence specific template reactions and metabolic network were sufficient to predict the production of different proteins under different promoters, and cell-free systems.
+However, while the mean and 95\% confidence estimate of the ensemble were consistent with measured deGFP levels, the model underpredicted the deGFP titer at the saturating plasmid concentration of 5 nM.
+These results, in combination with the time-dependent CAT and deGFP simulations, validated our modeling approach, which required very few adjustable parameters.
+It also showed that the sequence specific template reactions and metabolic network were sufficient to predict the production of different proteins under different promoters and cell-free systems.
 Next, we used our modeling approach to better understand the theoretical performance limits of CFPS.
@@
 \subsection{Analysis of CFPS performance}
-To better understand the performance of CFPS reactions, we analyzed the productivity, energy efficiency and carbon yield for the cell-free production of eight model proteins with and without amino acid supplementation (Fig.
+To better understand the performance of CFPS reactions, we analyzed the productivity, energy efficiency and carbon yield for the cell-free production of eight proteins with and without amino acid supplementation (Fig.
 \ref{fig:Prof}).
 The expression of each of these proteins was under a P70a promoter, with the exception of CAT which was expressed using a T7 promoter.
-In all cases, the CFPS reaction was supplied with glucose; however, we considered different scenarios for amino acid supplementation.
-First, the CFPS reaction was supplied with amino acids, and simultaneously was also allowed to synthesize amino acids from glucose (AAs supplied and \textit{de~novo} synthesis).
-Next, the CFPS reaction was supplied with both glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AAs supplied w/o \textit{de~novo} synthesis).
-This scenario was consistent with common cell-free extract preparation protocols which often involve amino acid supplementation;
-thus, we expect the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction in these cases.
-Lastly, the cell-free reaction was given glucose but not supplemented with amino acids, and was forced to synthesize them \textit{de~novo} from glucose (\textit{de~novo} synthesis only).
-Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10),
-chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP),
-prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
-An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental metabolic measurements of glucose, organic and amino acid consumption/production rates (see supplemental materials).
-Using these model proteins, we developed effective correlation models that predicted the productivity, energy efficiency and carbon yield given the carbon number of the model protein.
-Finally, we independently validated the effective correlations with a protein not in our original data set, maltose binding protein (MBP).
+%In all cases, the CFPS reaction was supplied with glucose; however, we considered different scenarios for amino acid supplementation.
+In the first case, the CFPS reaction was supplied with glucose and amino acids, and was able to synthesize amino acids from glucose (AAs supplied and \textit{de~novo} synthesis).
+In the second case, the CFPS reaction was supplied with glucose and amino acids, but \textit{de~novo} amino acid biosynthesis was not allowed (AAs supplied w/o \textit{de~novo} synthesis).
+This scenario was consistent with common cell-free extract preparation protocols which often involve amino acid supplementation; thus, we expected the enzymes responsible for amino acid biosynthesis to be largely absent from the CFPS reaction.
+In the final case, the CFPS reaction was supplied with glucose but not amino acids, and was forced to synthesize them \textit{de~novo} from glucose (\textit{de~novo} synthesis only).
+Eight proteins, ranging in size, were selected to evaluate CFPS performance: bone morphogenetic protein 10 (BMP10), chloramphenicol acetyltransferase (CAT), caspase 9 (CASP9), dual emission green fluorescent protein (deGFP), prothrombin (FII), coagulation factor X (FX), fibroblast growth factor 21 (FGF21), and single chain variable fragment R4 (scFvR4).
+An additional case was considered for CAT, where central metabolic fluxes were constrained by experimental measurements of glucose, organic and amino acids (see Supporting Information).
+Using these model proteins, we developed effective correlation models that predicted the productivity, energy efficiency and carbon yield given the carbon number of the protein.
+Finally, we independently validated the correlations with a protein not in our original data set: maltose binding protein (MBP).
 
@@
 The theoretical maximum productivity for proteins expressed using a P70a promoter ($\mu$M/h) was inversely proportional to the carbon number ($C_{POI}$) and varied between 1 and 12 $\mu$M/h for the proteins sampled (Fig.~\ref{fig:Prof}A-B).
-The theoretical maximum productivity was within a standard deviation with and without amino acid supplementation for each protein, but varied significantly between proteins.
-Productivity was a non-linear function of protein length; for instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h whereas the optimal productivity of deGFP (229 aa) was approximately 8.4 $\mu$M/h.
-To examine the influence of protein length, the mean optimal productivity was plotted against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
+The theoretical maximum productivity with and without amino acid supplementation were within a standard deviation of each other for each protein, but varied significantly between proteins.
+Productivity varied non-linearly with protein length; for instance, BMP10 (424 aa) had a optimal productivity of approximately 2.5 $\mu$M/h, whereas the optimal productivity of deGFP (229 aa) was approximately 8.4 $\mu$M/h.
+To examine the influence of protein length, we plotted the mean optimal productivity against the carbon number of each protein (Fig.~\ref{fig:Prof}B).
 The optimal productivity and protein length were related by the power-law relationship $\alpha\times(C_{POI})^{\beta}$, where $\alpha = 6.02\times 10^{6}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.93$ for a P70a promoter.
@@
 The prediction error was less than 8\% for an \emph{a~priori} prediction of CFPS productivity using the effective correlation.
-Thus, the effective productivity correlation could be used as a parameter free method to estimate optimal productivity for cell free protein production using a P70a promoter.
+Thus, the effective productivity correlation could be used as a parameter free method to estimate optimal productivity for cell-free protein production using a P70a promoter.
 For CFPS using other promoters, a similar correlation model could be developed.
-For example, maximal transcription occurs when the promoter model coefficient $u\left(\kappa\right)$ = 1;
-the theoretical maximum productivity correlation for maximum promoter activity also followed a power-law distribution ($\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$) (Fig.~\ref{fig:Prof}B, gray).
-The CAT value under a T7 promoter was similar to the maximal possible productivity as $u_{T7}\left(\kappa\right)\simeq~$0.95 given the T7 promoter model parameters used in this study (Table \ref{tbl:parameters}).
-Taken together, the maximum optimal productivity of a cell-free reaction was found to be inversely proportional to protein size,
-following a power-law relationship for proteins expressed under a P70a promoter.
+For example, maximal transcription occurs when the promoter model coefficient $u\left(\kappa\right)$ = 1; the theoretical maximum productivity correlation for maximum promoter activity also followed a power-law distribution ($\alpha = 1.39\times 10^{7}$ $\mu$M/(h$\cdot$carbon number) and $\beta=-1.99$) (Fig.~\ref{fig:Prof}B, gray).
+The CAT value under a T7 promoter was similar to the maximal productivity as $u_{T7}\left(\kappa\right)\simeq~$0.95 given the T7 promoter model parameters used in this study (Table \ref{tbl:parameters}).
+Taken together, the maximum optimal productivity of a cell-free reaction was found to be inversely proportional to protein size, following a power-law relationship for proteins expressed under a P70a promoter.
 
@@
 %However, the dependence of energy efficiency on carbon number was more pronounced in the absence of amino acid supplementation.
-The optimal energy efficiency of protein synthesis was independent of protein length, with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D).
-The optimal energy efficiency was approximately 86\% for the model proteins sampled.
-The optimal energy efficiency was independent of the protein length; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} =  -4.01\times10^{-4}$ energy efficiency (\%)/carbon number with supplementation,
-and $m_{Y} = 3.03\times10^{-3}$ energy efficiency (\%)/carbon number without supplementation.
-The y-intercept $b_{Y} = 86.20$ energy efficiency (\%) with supplementation, and $b_{Y} = 67.40$ energy efficiency (\%) without supplementation.
-In the presence of amino acids, energy was utilized to power CFPS instead of synthesizing amino acids, thus a constant level of energy efficiency was observed regardless of the protein size.
+The optimal energy efficiency of protein synthesis was independent of protein length, with and without amino acid supplementation (Fig.~\ref{fig:Prof}C-D); it was approximately 86\% for the model proteins sampled.
+The relationship was observed to be linear, but with negligible slopes: $m_{Y}\times(C_{POI})+b_{Y}$, where $m_{Y} = -4.01\times10^{-4}$ energy efficiency (\%)/carbon number for the case with supplementation, and $m_{Y} = 3.03\times10^{-3}$ energy efficiency (\%)/carbon number for the case without supplementation.
+The energy efficiency (y-intercept) was calculated at $b_{Y} = 86.20$ (\%) with supplementation, and $b_{Y} = 67.40$ (\%) without supplementation.
+In the presence of amino acids, energy was utilized to power CFPS instead of synthesizing amino acids; thus, a constant energy efficiency was observed regardless of the protein size.
 In the absence of supplementation, there was a drop in energy efficiency, ranging from 68\% to 76\%.
-In this case, glucose consumption more than doubled (64\% increase for CAT) compared to cases supplemented with amino acids, meanwhile the productivity was similar for each protein (Fig.~\ref{fig:Prof}D).
-Therefore, the energy burden required for synthesizing each amino acid, and powering CFPS, lowered the energy efficiency.
-Surprisingly, without amino acid supplementation, proteins with a higher carbon number had a marginally higher energy efficiency;
-however, this linear trend was mostly independent of protein size (R\textsuperscript{2} = 0.65).
-MBP showed good agreement with the linear efficiency model with and without amino acid supplementation.
-The estimated MBP energy efficiency had a maximum error of 5\% without supplementation, and an error of 1\% - 3\% in the presence of amino acid.
+In this case, glucose consumption more than doubled (64\% increase for CAT) compared to cases supplemented with amino acids; meanwhile, the productivity was similar for each protein (Fig.~\ref{fig:Prof}D).
+Therefore, the energy burden required for synthesizing each amino acid and powering CFPS lowered the energy efficiency.
+Surprisingly, without amino acid supplementation, proteins with a higher carbon number had marginally higher energy efficiency; however, this linear trend was mostly independent of protein size (R\textsuperscript{2} = 0.65).
+MBP was well predicted by the linear efficiency model with and without amino acid supplementation.
+The estimated MBP energy efficiency had a maximum error of 5\% without supplementation, and an error of 1\% - 3\% in the presence of amino acids.
 
@@
 The experimentally constrained energy efficiency was 13.3 $\pm$ 5.0\% compared to the theoretical maximum of approximately 84 $\pm$ 0.1\%.
-Given that the CAT productivity was similar between the simulated and measured systems, differences in
-the glucose consumption rate and the ATP yield per glucose were likely responsible for the difference between the optimal and experimental systems.
+Given that the CAT productivity was similar between the simulated and measured systems, differences in the glucose consumption rate and the ATP yield per glucose were likely responsible for the difference between the optimal and experimental systems.
 The glucose consumption rate was approximately 30 - 40 mM/h in the experimental system (even in the presence of amino acids).
 On the other hand, the constraint based simulation suggested the optimal glucose consumption rate was significantly less than the observed rate, approximately 1 - 7 mM/h (depending upon amino acid supplementation).
-In the constraint based simulation, the CFPS reaction produced only acetate as a by-product, but in the experimental system acetate, lactate, pyruvate, succinate and malate accumulated during the first hour of production.
+In the constraint based simulation, the CFPS reaction produced only acetate as a byproduct, but in the experimental system acetate, lactate, pyruvate, succinate and malate all accumulated during the first hour of production.
 Thus, the constraint based simulation was more carbon efficient.
-Next, the energy produced per unit glucose was different between the optimal and experimental constrained system.
-In the optimal simulation, 12 ATPs were produced per glucose (the theoretical maximum for this network was 21),
-while the experimentally constrained simulation produced only 4 ATPs per glucose.
-Thus, approximately 120 - 160 mM ATP/hr was produced in the experimental system, in contrast to 12 - 84 mM ATP/hr for the optimal simulations.
-We know from measurements that ATP did not accumulate in the experimental system, rather it was consumed by a variety of pathways that were not active
-in the optimal simulation. This suggested that CFPS retained a growth-like operational memory that led to over consumption of glucose, and the over production of ATP.
+The energy produced per unit glucose was different between the optimal and experimentally constrained cases.
+In the optimal simulation, 12 ATPs were produced per unit glucose (the theoretical maximum for this network was 21), while the experimentally constrained simulation produced only 4 ATPs per glucose.
+Thus, approximately 120 - 160 mM ATP/hr was produced in the experimental case, in contrast to 12 - 84 mM ATP/hr for the optimal case.
+We know from measurements that ATP did not accumulate in the experimental system; rather, it was consumed by a variety of pathways that were not active in the optimal simulation.
+This suggested that CFPS retained a growth-like operational memory that led to overconsumption of glucose, and the over production of ATP.
 
@@
 
-\subsubsection{Carbon Yield}
+\subsubsection{Carbon yield}
 The theoretical maximum carbon yield was inversely proportional to protein length and varied between 40\% to 64\% for the proteins sampled (Fig.~\ref{fig:Prof}E-F).
 The relationship between the optimal carbon yield and carbon number was linear; $m_{Y}\times(C_{POI})+b_{Y}$ where $m_{Y} = -8.03\times10^{-3}$ carbon yield (\%)/carbon number with supplementation, and $m_{Y} = -6.98\times10^{-3}$ carbon yield (\%)/carbon number without supplementation.
-The y-intercept $b_{Y} = 63.83$ carbon yield (\%) with supplementation, and $b_{Y} = 58.86$ carbon yield (\%) without supplementation.
-The linear yield models showed good predictability for estimating the optimal carbon yield for a range of carbon numbers ($R^{2} = 0.95$ with amino acids
-and $R^{2} = 0.90$ without supplementation), irrespective of the promoter used to control protein expression.
-The MBP prediction also showed good agreement with the constraint based model, with a prediction error between 3\% to 4\% with and without amino acid supplementation.
+The carbon yield (y-intercept) was $b_{Y} = 63.83$ (\%) with supplementation, and $b_{Y} = 58.86$ (\%) without supplementation.
+The linear models of yield showed good ability to predict the optimal carbon yield for a range of carbon numbers ($R^{2} = 0.95$ with amino acid supplementation and $R^{2} = 0.90$ without supplementation), irrespective of the promoter used to control protein expression.
+MBP also showed good agreement with the constraint based model, with a prediction error between 3\% and 4\% with and without amino acid supplementation.
 The effective yield models were approximately parallel, with a drop in carbon yield of approximately 7\% without amino acids.
-The difference between the cases was associated with glucose consumption.
-In the absence of amino acids, glucose consumption increased, with the change in the carbon yield being
-largely accounted for by increased CO\textsubscript{2} production with supplementation.
+The difference between the cases was associated with a difference in glucose consumption.
+In the absence of amino acids, glucose consumption increased, with the change in the carbon yield being largely accounted for by a change in CO\textsubscript{2} production.
 
 The carbon yield for experimentally constrained CAT production was 6.2\% compared to the theoretical maximum of 58\% (Fig.~\ref{fig:Prof}F, dagger).
-Given that the CAT productivity was similar between the simulated and measured systems, differences in
-the glucose consumption rate and byproduct accumulation were likely responsible for the difference between the optimal and experimental systems.
+Given that the CAT productivity was similar between the simulated and measured systems, differences in the glucose consumption rate and byproduct accumulation were likely responsible for the difference between the optimal and experimental systems.
 The translation rate has been identified as the rate-limiting step for cell-free protein synthesis \cite{2005_underwood_biotech,2014_li_PlosOne}, which was confirmed by the translation rate flux always hitting the upper bound in the simulation, regardless of protein.
-Since CAT productivity could not be improved, the increased carbon consumption accumulated in byproducts such as pyruvate, acetate, lactate, succinate and malate.
+Since CAT productivity could not be improved, the increased carbon consumption accumulated as byproducts: pyruvate, acetate, lactate, succinate and malate.
 In addition, the accumulation of acetate and lactate suggested that oxidative phosphorylation activity was limited.
-Thus to improve carbon yield for CFPS, the translation rate should be improved for increased protein productivity, or energy efficiency should be improved to lower the glucose consumption rate.
+Thus, to improve carbon yield for CFPS, the translation rate should be improved for increased protein productivity, or energy efficiency should be improved to lower the glucose consumption rate.
 %%%%%%%%%%%%%%%%%%%%
@@
 To better understand the effect of substrate utilization and the transcription/translation parameters on CFPS performance, we performed global sensitivity analysis on the productivity, energy efficiency and carbon yield for the CAT protein (Fig.~\ref{fig:SI} and Fig.
-1 in supporting information).
-Surprisingly, RNAP and ribosome abundance had only a modest effect;
-the translation elongation rate had the largest effect on protein productivity.
-On the other hand, oxygen followed by substrate consumption had the largest influence on the energy efficiency and carbon yield, respectively.
+1 in Supporting Information).
+Surprisingly, RNAP and ribosome abundance had only a modest effect; the translation elongation rate had the largest effect on protein productivity.
+On the other hand, oxygen and substrate consumption had the largest and second-largest influences on the energy efficiency and carbon yield, respectively.
 The significance of transcription/translation parameters was robust to amino acid supplementation, with the translation rate being the most sensitive across all cases (Fig.~\ref{fig:SI}A).
@@
 In the model, we assumed that ATP could be produced by both substrate level and oxidative phosphorylation.
-Jewett and coworkers reported that oxidative phosphorylation still operated in cell-free systems, and that the protein
-titer decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
-However, it is unknown how active oxidative phosphorylation is for a glucose powered cell-free system.
-Moreover, the connection between oxidative phosphorylation activity and other performance metrics, such as the carbon yield, is also unclear.
+Jewett and coworkers reported that oxidative phosphorylation still operated in cell-free systems, and that the protein titer decreased from 1.5-fold to 4-fold when oxidative phosphorylation reactions were inhibited in pyruvate-powered CFPS \cite{Jewett:2008aa}.
+However, it is unknown how active oxidative phosphorylation is in a glucose-powered cell-free system.
+Moreover, the connection between oxidative phosphorylation activity and other performance metrics, such as carbon yield, is also unclear.
 
@@
 
-To investigate the connection between carbon yield and oxidative phosphorylation further,
-we calculated the optimal CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
+To investigate the connection between carbon yield and oxidative phosphorylation further, we calculated the optimal CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
 Oxidative phosphorylation had a strong effect on the carbon yield, both with and without amino acid supplementation.
-In the presence of amino acid supplementation, the carbon yield ranged from 20\% to approximately 60\%,
-depending upon the oxidative phosphorylation flux.
+In the presence of amino acid supplementation, the carbon yield ranged from 20\% to approximately 60\%, depending on the oxidative phosphorylation flux.
 However, without amino acid supplementation, the carbon yield dropped to approximately 10\%, and reached a maximum of 50\%.
@@
 In all cases, whenever the carbon yield was below its theoretical maximum, there was an accumulation of both acetate and lactate.
-The experimental dataset exhibited a mixture of acetate and lactate accumulation during CAT synthesis, which suggested the CFPS reaction was not
-operating with optimal oxidative phosphorylation activity.
-Oxidative phosphorylation is a membrane associated process, however CFPS has no cell membrane.
+The experimental dataset exhibited a mixture of acetate and lactate accumulation during CAT synthesis, which suggested the CFPS reaction was not operating with optimal oxidative phosphorylation activity.
+Oxidative phosphorylation is a membrane associated process, while CFPS has no cell membrane.
 Jewett et al. hypothesized that membrane vesicles present in the CFPS reaction carry out oxidative phosphorylation \cite{Jewett:2008aa}.
 In addition, they were able to enhance the production titer of CAT by 33\% when the reaction was augmented with 10mM phosphate.
-However, it is unclear why the additional phosphate increased the CAT yield.
+However, it is unclear why the additional phosphate increased CAT yield.
 Jewett et al. believe it either enhanced oxidative phosphorylation activity or inhibited phosphatase reactions.
-The number, size, protein loading and lifetime of these vesicles remains an open area of study.
-Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity and remains an important area to optimize for improved CFPS performance.
+The number, size, protein loading, and lifetime of these vesicles remains an open area of study.
+Thus, we expect CFPS will have suboptimal oxidative phosphorylation capacity, and it remains an important area for optimization of CFPS performance.
 
@@
 
-\subsection{Optimal Metabolic Flux Distribution}
+\subsection{Optimal metabolic flux distribution}
 Amino acid supplementation altered the optimal metabolic flux distribution predicted for CAT production (Fig.~\ref{fig:flux}).
 To investigate the influence of amino acid supplementation, we compared the simulated metabolic flux distributions for CAT production with and without external amino acids.
-In the presence of amino acid supplementation, and \textit{de~novo} amino acid synthesis, there was an incomplete TCA cycle, where a combination
-of glucose and amino acids powered protein expression (Fig.~\ref{fig:flux}A).
-Glucose was consumed to produce acetyl-coenzyme A, and associated by-products,
-while glutamate was converted to alpha-ketoglutarate which traveled to oxaloacetic acid and pyruvate for additional amino acid biosynthesis.
-On the other hand, in the presence of amino acid supplementation, but without \textit{de~novo} amino acid biosynthesis, there was no TCA cycle flux.
-In this case, ATP was produced by a combination of substrate level and oxidative phosphorylation, where
-ubiquinone was regenerated via \textit{nuo} activity, without relying on succinate dehydrogenase in the TCA cycle (Fig.~\ref{fig:flux}B).
+In the presence of amino acid supplementation, and \textit{de~novo} amino acid synthesis, there was an incomplete TCA cycle, where a combination of glucose and amino acids powered protein expression (Fig.~\ref{fig:flux}A).
+Glucose was consumed to produce acetyl-coenzyme A, and associated byproducts, while glutamate was converted to alpha-ketoglutarate which traveled to oxaloacetic acid and pyruvate for additional amino acid biosynthesis.
+In the presence of amino acid supplementation, but without \textit{de~novo} amino acid biosynthesis, there was no TCA cycle flux.
+In this case, ATP was produced by a combination of substrate level and oxidative phosphorylation, where ubiquinone was regenerated via \textit{nuo} activity, without relying on succinate dehydrogenase in the TCA cycle (Fig.~\ref{fig:flux}B).
 These first two cases where amino acids were available had similar performance, and their respective metabolic flux distributions had a 99\% correlation for all proteins.
-In the absence of amino acid supplementation (where all amino acids were synthesized \textit{de~novo} from glucose),
-the energy efficiency and carbon yield decreased; in this case the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C).
+In the absence of amino acid supplementation (where all amino acids were synthesized \textit{de~novo} from glucose), the energy efficiency and carbon yield decreased; in this case, the TCA cycle was largely complete and there was diversion of metabolic flux into the Entner-Doudoroff pathway to produce NADPH (Fig.~\ref{fig:flux}C).
 However, these simulations represent the theoretical optimum.
-To more accurately describe actual metabolic flux, we constrained the feasible solution space with experimental measurements
-and estimated the optimal CAT flux distribution.
+To more accurately describe actual metabolic flux, we constrained the feasible solution space with experimental measurements and estimated the optimal CAT flux distribution.
 
-The experimentally constrained optimal flux distribution had a 52\% similarity with the theoretically optimal flux distribution (Fig. \ref{fig:flux_exp}).
-Metabolic fluxes were constrained by experimental measurements of glucose, amino and organic acid consumption and production rates for the first hour of the reaction (supporting information) and showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}~=~0.92 (Fig.~\ref{fig:flux_exp}B).
+The experimentally constrained optimal flux distribution had a 52\% correlation with the theoretically optimal flux distribution (Fig. \ref{fig:flux_exp}).
+Metabolic fluxes were constrained by experimental measurements of glucose, amino and organic acid consumption and production rates for the first hour of the reaction (Supporting Information) and showed good agreement with the data with a coefficient of determination of R\textsuperscript{2}~=~0.92 (Fig.~\ref{fig:flux_exp}B).
 The low similarity suggested several differences between the solutions which could be exploited to improve CFPS performance.
-The largest discrepancy was in oxidative phosphorylation,
-where the experimental system relied heavily on \textit{cyd} rather than \textit{cyo} which was less efficient for energy generation.
-In addition, the experimental system had a high flux through the \textit{zwf} yielding an abundance of NADPH used for amino acid biosynthesis.
-The remaining NADPH was interconverted to NADH via the \textit{pnt}1 reaction which was consumed to convert pyruvate to lactate.
+The largest discrepancy was in oxidative phosphorylation, where the experimental system relied heavily on \textit{cyd} rather than \textit{cyo}, which was less efficient for energy generation.
+In addition, the experimental system had a high flux through \textit{zwf}, yielding an abundance of NADPH used for amino acid biosynthesis.
+The remaining NADPH was interconverted to NADH via the \textit{pnt}1 reaction, which was consumed to convert pyruvate to lactate.
 In contrast, the optimal solutions with amino acid supplementation had low activity in \textit{zwf} and \textit{pnt}1 reactions.
-Next, folate, purine, and pyrimidine metabolism along with amino acid biosynthesis were inactive in the optimal system, but active in the experimental system.
-In particular, the experimental system had high alanine and glutamine biosynthetic flux (both accumulated in the media),
-while there was no accumulation of amino acids in the optimal simulations.
+Folate, purine, and pyrimidine metabolism, along with amino acid biosynthesis, were inactive in the optimal system, but active in the experimental system.
+In particular, the experimental system had high alanine and glutamine biosynthetic flux (both accumulated in the media), while there was no accumulation of amino acids in the optimal simulations.
 Even though all amino acids were supplied in the reaction, glucose traveled through the major pathways and was utilized for amino acid biosynthesis.
-Thus, it is unclear which substrate (glucose or amino acids) was used to power CFPS and may in fact be a combination of both.
+Thus, it is unclear which substrate (glucose or amino acids) was used to power CFPS, and it may in fact be a combination of both.
 The optimum solutions only produced the required amount of amino acids necessary; meanwhile, in the experimental system there was an accumulation of alanine, glutamine, pyruvate, lactate, acetate, malate, and succinate which contributed to some of the differences in the flux distributions.
@@
 
-\subsubsection{Potential Alternative Metabolic Optima}
+\subsubsection{Potential alternative metabolic optima}
 Optimal flux distributions predicted using constraint based approaches may not always be unique.
 Alternative optimal solutions have the same objective value, e.g., productivity, but different metabolic flux distributions.
-Several techniques, e.g., flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz119} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
+Several techniques such as flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz119} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
 In this study, we used flux variability analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).
@@
 Ultimately, to determine the metabolic flux distribution occurring in CFPS, we need to add additional constraints to the flux estimation calculation.
-For example, thermodynamic feasibility constraints may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013},
-and $^{13}$C labeling in CFPS could provide significant insight.
-However, while $^{13}$C labeling techniques are well established for \emph{in~vivo} processes \cite{Zamboni:2009},
-application of these techniques to CFPS remains an active area of research.
+For example, thermodynamic feasibility constraints may result in a better depiction of the flux distribution \cite{Henry:2007,Hamilton:2013}, and $^{13}$C labeling in CFPS could provide significant insight.
+However, while $^{13}$C labeling techniques are well established for \emph{in~vivo} processes \cite{Zamboni:2009}, application of these techniques to CFPS remains an active area of research.
 
@@
 In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell-free protein synthesis reactions for a range of proteins.
-We showed first principle predictions for protein production of deGFP and CAT which were in agreement with experimental measurements, under two different promoters and two different cell-free extract systems.
+We showed first principle predictions for protein production of deGFP and CAT, which were in agreement with experimental measurements, under two different promoters and two different cell-free extract systems.
 We are currently working on expanding our library of promoters that can provide good estimates of performance for different proteins.
-Different promoter will help design \emph{de~novo} circuits for optimal functionality and performance.
-This modeling approach suggested correlations for productivity, energy efficiency and carbon yield as a function of the size of the protein.
-Maltose binding protein showed good agreement with these performance metrics with the established model correlations.
+Different promoters will help in the design \emph{de~novo} of circuits for optimal functionality and performance.
+This modeling approach suggested correlations for productivity, energy efficiency and carbon yield as a function of protein size.
+Maltose binding protein showed good agreement with these performance metrics under the established model correlations.
 Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
-The translation rate was identified as the rate limiting step for productivity.
+The translation rate was identified as the rate-limiting step for productivity.
 The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficiency and carbon yield.
 In addition, our modeling approach could be expanded with a more detailed description of gene expression that has been utilized in a \emph{E. coli} genome scale model \cite{Brien693}.
-We would also like to expand our model to include post translation modifications that proteins would require for proper folding and functionality.
+We would also like to include post-translation modifications required for proper protein folding and functionality.
 In conclusion, sequence specific constraint based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
@@
 \subsection*{Formulation and solution of the model equations.}
-The sequence-specific flux balance analysis problem was formulated as a linear program:
+The sequence specific flux balance analysis problem was formulated as a linear program:
 \begin{equation}
@@
 In the third case, amino acid uptake reactions were set to 0 mM/hr.
-In the experimental constrained case, \textit{E.
+In the experimentally constrained case, \textit{E.
 coli} was grown in the presence of 13 amino acids (alanine, arginine, cysteine, serine, aspartate, glutamate, and glutamine were excluded) \cite{Zawada:2003}, thus the synthesis reactions responsible for those 13 amino acid were set to 0 mM/hr.
@@
 $l_{P}$ denotes the number of amino acids in the protein of interest.
-The steady-state mRNA abundance $\rm mRNA^{*}$ was estimated as:
+The steady state mRNA abundance $\rm mRNA^{*}$ was estimated as:
 \begin{equation}
@@
 \begin{equation}\label{eqn:energy-efficiency-definition}
-	\mathcal{E}\equiv\frac{w_{X}\left(2\cdot\rm {ATP}_{X}+\rm {GTP}_{X}\right)+w_{T}\left(2\times\left(\rm {ATP}_{T}+\rm {CTP}_{T}+\rm {GTP}_{T}+\rm {UTP}_{T}\right) \right)}{\texttt{q}_{GLC}\cdot\texttt{ATP}_{GLC}}
+	\mathcal{E}\equiv\frac{w_{X}\left(2\cdot\rm {ATP}_{X}+\rm {GTP}_{X}\right)+w_{T}\left(2\times\left(\rm {ATP}_{T}+\rm {CTP}_{T}+\rm {GTP}_{T}+\rm {UTP}_{T}\right) \right)}{\rm{q}_{GLC}\cdot\rm{ATP}_{GLC}}
 \end{equation}
-where $\rm {ATP}_{T}$, $\rm {CTP}_{T}$, $\rm {GTP}_{T}$, $\rm {UTP}_{T}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{X}$, $\rm {GTP}_{X}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\texttt{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate,
-and $\texttt{ATP}_{GLC}$ denotes the equivalent ATP number produced per glucose.
+where $\rm {ATP}_{T}$, $\rm {CTP}_{T}$, $\rm {GTP}_{T}$, $\rm {UTP}_{T}$ denote the stoichiometric coefficients of each energy species for the transcription of the protein of interest, $\rm {ATP}_{X}$, $\rm {GTP}_{X}$ denote the stoichiometric coefficients of ATP and GTP for the translation of the protein of interest, $\rm{q}_{GLC}=w_{GLC}$ denotes the glucose uptake rate,
+and $\rm{ATP}_{GLC}$ denotes the equivalent ATP number produced per glucose.
 The energy species stoichiometric coefficients are available in the Supporting Information.
@@
 
-We thank Che-Hsiao Shih (Purdue University) for identification of DNA and protein sequences.
 This study was supported by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V. for the support of M.V.

### 2017-05-05 Paper_v1.tex
Small changes

@@
 B. Total order sensitivity of the optimal CAT energy efficiency.
-Metabolic and transcription/translation parameters were varied for amino acid supplementation and synthesis (black), amino acid supplementation without synthesis (dark grey) and amino acid synthesis without supplementation (light gray).
+Metabolic and transcription/translation parameters were varied for amino acid supplementation and synthesis (black),
+amino acid supplementation without synthesis (dark grey) and amino acid synthesis without supplementation (light gray).
 Error bars represent a 95\% CI on the total-oder sensitivity index.}
@@
 \includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation, for three different cases: (A) amino acid uptake and synthesis, (B) amino acid uptake without synthesis, and (C) amino acid synthesis without uptake.
-Mean flux across the ensemble, normalized to glucose uptake flux.
-Thick arrows indicate flux to or from amino acids.}
+\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway,
+TCA cycle, and oxidative phosphorylation, for: (A) amino acid supplementation and \textit{de~novo} synthesis,
+(B) amino acid supplementation without \textit{de~novo} synthesis, and
+(C) \textit{de~novo} amino acid synthesis without supplementation.
+Mean flux across the ensemble, normalized to glucose uptake flux. Thick arrows indicate flux to or from amino acids.}
 \label{fig:flux}
@@
 \includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
-\caption{The deGFP carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 flux balance solutions, for amino acid uptake and synthesis (black), amino acid uptake without synthesis (dark grey), and amino acid synthesis without uptake (light gray).}
+\caption{The CAT carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 flux balance solutions,
+for amino acid supplementation and \textit{de~novo} synthesis (black),
+amino acid supplementation without \textit{de~novo} synthesis (dark grey),
+and \textit{de~novo} amino acid synthesis without supplementation (light gray).}
 \label{fig:oxphos_yield}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribution-Analysis.pdf}
-\caption{The norm effect of pairwise knockouts of subgroups in the cell-free network.
-(A) CAT productivity norm compared to no knockouts in the experimentally constrained case.
-(B) Flux distribution norm compared to no knockouts in the experimentally constrained case.}
+\caption{Pairwise knockouts of reaction subgroup in the cell-free network.
+(A) Difference in the CAT productivity in the presence of reaction knockouts compared to no knockouts for the experimentally constrained case.
+(B) Difference in the optimal flux distribution in the presence of reaction knockouts compared to no knockouts in the experimentally constrained case.
+The difference between perturbed and wild-type productivity and flux distributions was quantified by the $l^{2}$ norm, and then normalized so the maximum change was 1.0.
+Red boxes indicate potential alternative optimal flux distributions.}
 \label{fig:norm}

### 2017-05-18 Paper_v1.tex
Fixes??

@@
 A core \emph{E.~coli} cell-free metabolic model describing glycolysis, pentose phosphate pathway, energy metabolism, amino acid biosynthesis and degradation was developed from literature \cite{Feist:2007aa}; it was then augmented with sequence specific descriptions of promoter function, transcription and translation processes.
-Thus, sequence specific constraint based modeling explicitly coupled transcription and translation with the availability of metabolic resources.
-We tested this approach by simulating the production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield of production for eight different proteins.
-It was observed that higher carbon number proteins typically had lower productivity rates and carbon yields than those of the lower carbon number proteins, whereas energy efficiency was independent of protein size.
-Model correlations were then independently validated with a protein not in the original data set.
-We compared the theoretical optimal solutions with an experimentally constrained case which suggested CFPS retained a growth-like operational memory that led to overconsumption of glucose.
-Further, global sensitivity analysis identified the key metabolic processes that controlled circuit performance, showing oxidative phosphorylation vital to energy efficiency and carbon yield, and translation rate most important for productivity.
-Taken together, sequence specific constraint based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
+Thus, the sequence specific constraint based approach explicitly coupled transcription and translation processes with the availability of metabolic resources in the CFPS reaction.
+We tested this approach by simulating the cell-free production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight additional therapeutically relevant proteins.
+Productivity and carbon yield were found to be inversely proportional to carbon number, while energy efficiency was independent of protein size.
+Based upon these simulations, effective correlations for the productivity, carbon yield and energy efficiency as a function of protein length were developed.
+These correlation models were then independently validated with a protein not in the original data set.
+Further, global sensitivity analysis identified the key metabolic processes that controlled CFPS performance;
+oxidative phosphorylation was vital to energy efficiency and carbon yield, while the translation rate was the most important for productivity.
+Lastly, we compared theoretically optimal flux distributions with an experimentally constrained flux distribution;
+this comparison suggested CFPS retained a growth-like operational memory that led to the overconsumption of glucose which negatively influenced carbon yield and energy efficiency.
+Taken together, sequence specific constraint based modeling of CFPS offered a novel means to \emph{a~priori} estimate the performance of a cell-free system, using only a limited number of
+of adjustable parameters.
 
@@
 In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell-free protein synthesis reactions for a range of proteins.
-We showed first principle predictions for protein production of deGFP and CAT, which were in agreement with experimental measurements, under two different promoters and two different cell-free extract systems.
-We are currently working on expanding our library of promoters that can provide good estimates of performance for different proteins.
-Different promoters will help in the design \emph{de~novo} of circuits for optimal functionality and performance.
-This modeling approach suggested correlations for productivity, energy efficiency and carbon yield as a function of protein size.
-Maltose binding protein showed good agreement with these performance metrics under the established model correlations.
+We showed first principle predictions of the cell-free production of deGFP and CAT, which were in agreement with experimental measurements, under two different promoters.
+Here we considered only the P70a and T7 promoters, but we are currently expanding our library of possible promoters.
+A library of computational promoters, in combination with the cell-free constraints based approach,
+could enable the \emph{de~novo} design of circuits for optimal functionality and performance.
+We also developed effective correlation models for the productivity, energy efficiency and carbon yield as a function of protein size.
+Simulations of maltose binding protein expression showed good agreement with these correlation models, suggesting the effective correlations could be used to quickly prototype CFPS reactions.
 Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
@@
 The sequence specific flux balance linear program was solved using the GNU Linear Programming Kit (GLPK) v4.55 \cite{GLPK}.
-For all cases, amino acid degradation reactions were blocked as these enzymes were knocked out during the cell-free extract preparation \cite{2005_calhoun_BiotechnologyProgress, Garamella:2016aa}.
-In the second case, all amino acid synthesis reactions were set to 0 mM/hr since \textit{E.
-coli} was grown in the presence of amino acids, thus these enzymes would not be present in the cell-free extract media.
-In the third case, amino acid uptake reactions were set to 0 mM/hr.
-In the experimentally constrained case, \textit{E.
-coli} was grown in the presence of 13 amino acids (alanine, arginine, cysteine, serine, aspartate, glutamate, and glutamine were excluded) \cite{Zawada:2003}, thus the synthesis reactions responsible for those 13 amino acid were set to 0 mM/hr.
+For all cases, amino acid degradation reactions were blocked as these enzymes were likely inactivated during the cell-free extract preparation \cite{2005_calhoun_BiotechnologyProgress, Garamella:2016aa}.
+In the absence of \textit{de~novo} amino acid synthesis, all amino acid synthesis reactions were set to 0 mM/hr.
+In the experimentally constrained simulations, \textit{E.
+coli} was grown in the presence of 13 amino acids (alanine, arginine, cysteine, serine, aspartate, glutamate, and glutamine were excluded) \cite{Zawada:2003}, thus the synthesis reactions responsible for those 13 amino acids were set to 0 mM/hr.
 
@@
 Simultaneously, we computed the difference in the flux distribution ($l^{2}$-norm) for each pairwise reaction knockout compared to the flux distribution with no knockouts.
-Those solutions with the same or similar productivity but large changes in the metabolic flux distribution represent alternative optimal soluations.
+Those solutions with the same or similar productivity but large changes in the metabolic flux distribution represent alternative optimal solutions.
 
@@
 This study was supported by an award from the US Army and Systems Biology of Trauma Induced Coagulopathy (W911NF-10-1-0376) to J.V. for the support of M.V.
+The work was also supported by the Center on the Physics of Cancer Metabolism through Award Number 1U54CA210184-01 from the National Cancer Institute.
+The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Cancer Institute or the National Institutes of Health.
 

### 2017-05-18 Paper_v1.tex
Fixes??

@@
 In this study, we developed a sequence specific constraint based modeling approach to predict the performance of cell-free protein synthesis reactions for a range of proteins.
-We showed first principle predictions of the cell-free production of deGFP and CAT, which were in agreement with experimental measurements, under two different promoters.
+First principle predictions of the cell-free production of deGFP and CAT were in agreement with experimental measurements, for two different promoters.
 Here we considered only the P70a and T7 promoters, but we are currently expanding our library of possible promoters.

### 2017-05-18 Paper_v1.tex
changes to the conclusion

@@
 We tested this approach by simulating the cell-free production of two model proteins, and then investigated the productivity, energy efficiency, and carbon yield for eight additional therapeutically relevant proteins.
-Productivity and carbon yield were found to be inversely proportional to carbon number, while energy efficiency was independent of protein size.
+Productivity and carbon yield were inversely proportional to carbon number, while energy efficiency was independent of protein size.
 Based upon these simulations, effective correlations for the productivity, carbon yield and energy efficiency as a function of protein length were developed.
@@
 Several techniques such as flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz119} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
-In this study, we used flux variability analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).
-FVA identified pathways required for CAT production; for example, deletion of the glycolysis/gluconoeogenesis or oxidative phosphorylation pathways resulted in no CAT production.
+In this study, we used group knockout analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).
+In this approach, groups of reactions were removed from the metabolic network, and the translation rate was maximized.
+The difference in the productivity and flux distribution of the nominal and altered system was then calculated.
+Knockout analysis identified pathways required for CAT production; for example, deletion of the glycolysis/gluconoeogenesis or oxidative phosphorylation pathways resulted in no CAT production.
 Likewise, there were pathway knockouts that had no effect on productivity or the metabolic flux distribution, such as removal of isoleucine, leucine, histidine and valine biosynthesis.
@@
 Simulations of maltose binding protein expression showed good agreement with these correlation models, suggesting the effective correlations could be used to quickly prototype CFPS reactions.
-Furthermore, global sensitivity analysis identified oxygen uptake as being instrumental for maintaining a high energy efficiency and carbon yield.
-The translation rate was identified as the rate-limiting step for productivity.
-The model also suggested that cell-free systems can simultaneously operate aerobically and anaerobically, which can lead to inefficient production and should be addressed to optimize energy efficiency and carbon yield.
-In addition, our modeling approach could be expanded with a more detailed description of gene expression that has been utilized in a \emph{E. coli} genome scale model \cite{Brien693}.
-We would also like to include post-translation modifications required for proper protein folding and functionality.
-In conclusion, sequence specific constraint based modeling offers a novel means to \emph{a~priori} estimate the performance of cell-free synthetic circuits.
+Further, global sensitivity analysis identified the key metabolic processes that controlled CFPS performance;
+oxidative phosphorylation was vital to energy efficiency and carbon yield, while the translation rate was the most important for productivity.
+However, ATP was predicted to be produced by both substrate level and oxidative phosphorylation, as both lactate and acetate accumulated in the experimental system.
+While the results of this first study were promising, there are several issues to consider in future work.
+First, a more detailed description of transcription and translation reactions that has been utilized in other genome scale ME models e.g., O’Brien et al \cite{Brien693} could be adapted
+to a cell-free system. This would allow us to consider important facets of protein production, such as the role of chaperones in protein folding.
+We would also like to include post-translation modifications such as glycosylation that are important for the production of therapeutic proteins.
+In conclusion, sequence specific constraint based modeling of cell-free protein production offers a novel means to \emph{a~priori}
+estimate the optimal performance of cell-free synthetic circuits.
+
+% required for proper protein folding and functionality.
+%
 

### 2017-05-19 Paper_v1.tex
Fixed issues w/the refs

@@
 Today, cell-free systems are used in a variety of applications ranging from therapeutic protein production \cite{Lu:2014aa} to synthetic biology \cite{Hodgman:2012aa}.
-However, if CFPS is to become a mainstream technology for advanced applications such as point of care manufacturing \cite{Pardee2016248}, we must first understand the performance limits and costs of these systems \cite{Jewett:2008aa}.
+However, if CFPS is to become a mainstream technology for advanced applications such as point of care manufacturing \cite{Pardee:2016aa}, we must first understand the performance limits and costs of these systems \cite{Jewett:2008aa}.
 One tool to address these questions is constraint based modeling.
@@
 Alternative optimal solutions have the same objective value, e.g., productivity, but different metabolic flux distributions.
-Several techniques such as flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz119} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
+Several techniques such as flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz:2007aa} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
 In this study, we used group knockout analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).

### 2017-05-19 Paper_v1.tex
Changes

@@
 However, these simulations represent the theoretical optimum.
-To more accurately describe actual metabolic flux, we constrained the feasible solution space with experimental measurements and estimated the optimal CAT flux distribution.
+To more accurately describe the metabolic flux, we constrained the feasible solution space with experimental measurements and estimated the optimal CAT flux distribution.
 
@@
 Even though all amino acids were supplied in the reaction, glucose traveled through the major pathways and was utilized for amino acid biosynthesis.
-Thus, it is unclear which substrate (glucose or amino acids) was used to power CFPS, and it may in fact be a combination of both.
-The optimum solutions only produced the required amount of amino acids necessary; meanwhile, in the experimental system there was an accumulation of alanine, glutamine, pyruvate, lactate, acetate, malate, and succinate which contributed to some of the differences in the flux distributions.
-% and may have contributed to the inefficiency of carbon utilization.
+Thus, it is unclear which substrate (glucose or amino acids) was used to power CFPS, and may likely be a combination of both.
+Lastly, the optimal solution produced (or consumed) only the required amount of amino acids;
+meanwhile,  alanine, glutamine, pyruvate, lactate, acetate, malate, and succinate all accumulated in the experimental system.
+This accumulation contributed to the difference in flux distribution.
+
+% the experimental system there was an accumulation of  which contributed to some of the differences in the flux distributions. % and may have contributed to the inefficiency of carbon utilization.
 
@@
 Alternative optimal solutions have the same objective value, e.g., productivity, but different metabolic flux distributions.
-Several techniques such as flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz:2007aa} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
+Techniques such as flux variability analysis (FVA) \cite{Mahadevan2003264,Schuetz:2007aa} or mixed-integer approaches \cite{LEE2000711} can estimate alternative optima.
 In this study, we used group knockout analysis to estimate potential alternative optimal solutions for CAT production constrained by experimental measurements (Fig.~\ref{fig:norm}).
-In this approach, groups of reactions were removed from the metabolic network, and the translation rate was maximized.
-The difference in the productivity and flux distribution of the nominal and altered system was then calculated.
+Groups of reactions were removed from the metabolic network, and the translation rate was maximized.
+The difference between the nominal and altered system was then calculated.
 Knockout analysis identified pathways required for CAT production; for example, deletion of the glycolysis/gluconoeogenesis or oxidative phosphorylation pathways resulted in no CAT production.
@@
 \subsection*{Quantification of uncertainty.}
-Experimental factors taken from literature, for example macromolecular concentrations or elongation rates, have uncertainty associated with their values.
+Experimental factors taken from literature, for example macromolecular concentrations or elongation rates, are uncertain.
 To quantify the influence of this uncertainty on model performance, we randomly sampled the expected physiological ranges for these parameters as determined from literature.
-An ensemble of N = 100 flux distributions was calculated for the three different cases we considered:
+An ensemble of flux distributions was calculated for the three different cases we considered:
 control (with amino acid synthesis and uptake), amino acid uptake without synthesis, and amino acid synthesis without uptake.
@@
 and randomly sampling RNA polymerase levels, ribosome levels, and elongation rates in a physiological range determined from literature.
-RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 12 and 18 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/s
-\cite{2005_underwood_biotech, Garamella:2016aa}.
+RNA polymerase levels were sampled between 60 and 80 nM, ribosome levels between 12 and 18 \textmu M, the RNA polymerase elongation rate between 20 and 30 nt/sec, and the ribosome elongation rate between 1.5 and 3 aa/s \cite{2005_underwood_biotech, Garamella:2016aa}.
 
@@
 \subsection*{Potential alternative optimal metabolic flux solutions.}
-We identified potential alternative optimal metabolic flux distributions by performing single and pairwise reaction group knockout simulations \cite{Mahadevan2003264}.
-Reaction group knockouts were simulated by setting the flux bounds for all the reactions involved in a group to zero.
+We identified potential alternative optimal flux distributions by performing single and pairwise reaction group knockout simulations.
+Reaction group knockouts were simulated by setting the flux bounds for all the reactions involved in a group to zero and then maximizing the translation rate.
 We grouped reactions in the cell-free network into 19 subgroups (available in Supporting Information).
@@
 A.
-Core metabolic network with glycolysis, pentose phoshate pathway, TCA cycle, Entner-Doudoroff pathway.
+Schematic of the core metabolic network describing glycolysis, pentose phosphate pathway, the TCA cycle and the Entner-Doudoroff pathway.
 Thick gray arrows indicate withdrawal of precursors for amino acid synthesis.
-B.
-CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h under glucose consumption.
-Error bars denote the standard deviation of experimental measurements.
-C.
-deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h under maltose consumption.
-Error bars denote a 10\% coefficient of variation.
-D.
-Predicted deGFP concentration at different plasmid concentrations versus measurements of deGFP synthesized in TXTL 2.0.
-95\% CI (blue region) over the ensemble of 100 sets, mean of the ensemble (black line), and experimental measurements (dots).}
+B. CAT production under a T7 promoter in CFPS \textit{E.~coli} extract for 1 h using glucose as a carbon and energy source. Error bars denote the standard deviation of experimental measurements.
+C. deGFP production under a P70 promoter in TXTL 2.0 \textit{E.~coli} extract for 8 h using glucose as a carbon and energy source. Error bars denote a 10\% coefficient of variation.
+D. Predicted versus measured deGFP concentration as a function of plasmid concentration in TXTL 2.0.
+The blue region denotes the 95\% CI over an ensemble of N = 100 sets, the black line denotes the mean of the ensemble, and dots denote experimental measurements.}
 \label{fig:network}
@@
 \includegraphics[width=1.00\textwidth]{./figs/Fig-3-Sensitivity-Analysis.pdf}
-\caption{Total order sensitivity analysis of CFPS.
+\caption{Total order sensitivity analysis of the cell free production of CAT.
 A. Total order sensitivity of the optimal CAT productivity with respect to metabolic and transcription/translation parameters.
@@
 amino acid supplementation without synthesis (dark grey) and amino acid synthesis without supplementation (light gray).
-Error bars represent a 95\% CI on the total-oder sensitivity index.}
+Error bars represent the 95\% CI of the total order sensitivity index.}
 \label{fig:SI}
@@
 \includegraphics[width=1.00\textwidth]{./figs/Fig-5-FluxDistribition-Optimal.pdf}
-\caption{Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway,
-TCA cycle, and oxidative phosphorylation, for: (A) amino acid supplementation and \textit{de~novo} synthesis,
-(B) amino acid supplementation without \textit{de~novo} synthesis, and
-(C) \textit{de~novo} amino acid synthesis without supplementation.
-Mean flux across the ensemble, normalized to glucose uptake flux. Thick arrows indicate flux to or from amino acids.}
+\caption{Optimal metabolic flux distribution for CAT production.
+A. Optimal flux distribution in the presence of amino acid supplementation and \textit{de~novo} synthesis.
+B. Optimal flux distribution in the presence of amino acid supplementation without \textit{de~novo} synthesis.
+C. Optimal flux distribution with \textit{de~novo} amino acid synthesis in the absence of supplementation.
+Mean flux across the ensemble (N = 100), normalized to glucose uptake flux. Thick arrows indicate flux to or from amino acid biosynthesis pathways.}
 \label{fig:flux}
@@
 \includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
-\caption{The CAT carbon yield versus oxidative phosphorylation flux, across an ensemble of 1000 flux balance solutions,
-for amino acid supplementation and \textit{de~novo} synthesis (black),
+\caption{Optimal CAT carbon yield versus oxidative phosphorylation flux calculated across an ensemble (N = 1000) flux balance solutions (points).
+Carbon yield versus oxidative phosphorylation flux for amino acid supplementation and \textit{de~novo} synthesis (black),
 amino acid supplementation without \textit{de~novo} synthesis (dark grey),
@@
 \includegraphics[width=1.00\textwidth]{./figs/Fig-6-FluxDistribition-Experimental.pdf}
-\caption{Constraint based simulation of CAT production for an experimentally constrained case.
-(A) Flux profile for glycolysis, pentose phosphate pathway, Entner-Doudoroff pathway, TCA cycle, and oxidative phosphorylation.
+\caption{Experimentally constraint based simulation of CAT production.
+A. Metabolic flux distribution for CAT production in the presence of experimental constraints for glucose, organic acid and amino acid consumption and production rates.
 Mean flux across the ensemble, normalized to glucose uptake flux. Thick arrows indicate flux to or from amino acids.
-(B) Central carbon metabolite measurements versus simulations over a one hour time course.}
+B. Central carbon metabolite measurements versus simulations over a 1 hour time course.
+The blue region denotes the 95\% CI over an ensemble of N = 100 sets, the black line denotes the mean of the ensemble, and dots denote experimental measurements.}
 \label{fig:flux_exp}
@@
 \includegraphics[width=1.0\textwidth]{./figs/Fig-7-FluxDistribution-Analysis.pdf}
-\caption{Pairwise knockouts of reaction subgroup in the cell-free network.
-(A) Difference in the CAT productivity in the presence of reaction knockouts compared to no knockouts for the experimentally constrained case.
-(B) Difference in the optimal flux distribution in the presence of reaction knockouts compared to no knockouts in the experimentally constrained case.
+\caption{Pairwise knockouts of reaction subgroups in the cell-free network.
+A. Difference in the CAT productivity in the presence of reaction knockouts compared with no knockouts for experimentally constrained CAT production.
+B. Difference in the optimal flux distribution in the presence of reaction knockouts compared with no knockouts for experimentally constrained CAT production.
 The difference between perturbed and wild-type productivity and flux distributions was quantified by the $l^{2}$ norm, and then normalized so the maximum change was 1.0.

### 2017-05-19 Paper_v1.tex
oxphos_yield and edits

@@
 The mean of the ensemble (calculated by sampling the uncertainty in the model parameters) captured the saturation of deGFP production as a function of plasmid concentration (R\textsuperscript{2} = 0.97).
-However, while the mean and 95\% confidence estimate of the ensemble were consistent with measured deGFP levels, the model underpredicted the deGFP titer at the saturating plasmid concentration of 5 nM.
+However, while the mean and 95\% confidence estimate of the ensemble were consistent with measured deGFP levels, the model under predicted the deGFP titer at the saturating plasmid concentration of 5 nM.
 These results, in combination with the time-dependent CAT and deGFP simulations, validated our modeling approach, which required very few adjustable parameters.
@@
 The significance of transcription/translation parameters was robust to amino acid supplementation, with the translation rate being the most sensitive across all cases (Fig.~\ref{fig:SI}A).
-This suggested that the translation elongation rate, and not transcriptional parameters, controlled productivity.
+This suggested that the translation elongation rate, and not transcription parameters, controlled productivity.
 Underwood and coworkers showed that an increase in ribosome levels did not significantly increase protein yields or rates; however, adding elongation factors increased protein synthesis rates by 27\% \cite{2005_underwood_biotech}.
@@
 To investigate the connection between carbon yield and oxidative phosphorylation further, we calculated the optimal CAT carbon yield as a function of the oxidative phosphorylation flux (Fig.~\ref{fig:oxphos_yield}).
+We calculated yield across an ensemble of 1000 flux balance solutions by varying the oxygen uptake rate with transcription and translation parameters.
 Oxidative phosphorylation had a strong effect on the carbon yield, both with and without amino acid supplementation.
@@
 \includegraphics[width=0.7\textwidth]{./figs/Fig-4-OxyPhox-versus-Carbon-Yield.pdf}
-\caption{Optimal CAT carbon yield versus oxidative phosphorylation flux calculated across an ensemble (N = 1000) flux balance solutions (points).
+\caption{Optimal CAT carbon yield versus oxidative phosphorylation flux calculated across an ensemble (N = 1000) flux balance solutions (points) by varying the oxygen uptake rate with transcription and translation parameters.
 Carbon yield versus oxidative phosphorylation flux for amino acid supplementation and \textit{de~novo} synthesis (black),
